### 2018-04-04 paper.tex
first pass on the abstract

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[10pt, sigplan]{acmart}
%%\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[sigplan,10pt,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,10pt,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[sigplan,10pt]{acmart}\settopmatter{}

%% Conference information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
%\acmConference[PL'17]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
%\acmYear{2017}
%\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
%\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
%\startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2017}           %% If different from \acmYear

%% Bibliography style
%%\bibliographystyle{ACM-Reference-Format}
%% Citation style
%\citestyle{acmauthoryear}  %% For author/year citations
%\citestyle{acmnumeric}     %% For numeric citations
%\setcitestyle{nosort}      %% With 'acmnumeric', to disable automatic
                            %% sorting of references within a single citation;
                            %% e.g., \cite{Smith99,Carpenter05,Baker12}
                            %% rendered as [14,5,2] rather than [2,5,14].
%\setcitesyle{nocompress}   %% With 'acmnumeric', to disable automatic
                            %% compression of sequential references within a
                            %% single citation;
                            %% e.g., \cite{Baker12,Baker14,Baker16}
                            %% rendered as [2,3,4] rather than [2-4].

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Note: Authors migrating a paper from traditional SIGPLAN
%% proceedings format to PACMPL format must update the
%% '\documentclass' and topmatter commands above; see
%% 'acmart-pacmpl-template.tex'.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption
\usepackage{xspace}
\usepackage{graphicx}
\usepackage{ifthen}
%\input{macros.tex}

\usepackage{xcolor}
\newcommand{\sd}[1]{\color{red}\fbox{\bfseries\sffamily\scriptsize Stef:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\sk}[1]{\color{red}\fbox{\bfseries\sffamily\scriptsize Sophie:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\cb}[1]{\color{red}\fbox{\bfseries\sffamily\scriptsize Clement:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}

\begin{document}

%% Title information
\title[Short Title]{Assessing primitives performance\\ on multi-stage execution}

%% Author with single affiliation.
\author{First1 Last1}
                                        %% can be repeated if necessary
\affiliation{
  \position{Position1}
  \department{Department1}              %% \department is recommended
  \institution{Institution1}            %% \institution is required
  \streetaddress{Street1 Address1}
  \city{City1}
  \state{State1}
  \postcode{Post-Code1}
  \country{Country1}                    %% \country is recommended
}
\email{first1.last1@inst1.edu}          %% \email is recommended

%% Author with two affiliations and emails.
\author{First2 Last2}
                                        %% can be repeated if necessary
\affiliation{
  \position{Position2a}
  \department{Department2a}             %% \department is recommended
  \institution{Institution2a}           %% \institution is required
  \streetaddress{Street2a Address2a}
  \city{City2a}
  \state{State2a}
  \postcode{Post-Code2a}
  \country{Country2a}                   %% \country is recommended
}
\email{first2.last2@inst2a.com}         %% \email is recommended

%% Author with two affiliations and emails.
\author{First3 Last3}
\affiliation{
  \position{Position2b}
  \department{Department2b}             %% \department is recommended
  \institution{Institution2b}           %% \institution is required
  \streetaddress{Street3b Address2b}
  \city{City2b}
  \state{State2b}
  \postcode{Post-Code2b}
  \country{Country2b}                   %% \country is recommended
}
\email{first2.last2@inst2b.org}         %% \email is recommended

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Most of programming managed languages use primitive functions to speed up their execution time; these primitives usually implement core operations (such as addition, subtraction,...) or frequently used operations.
They are also massively used in the context of dynamic or just-in-time compilation: the compiler then executes this alternative version instead of the standard one to gain performance.
%%Problem

One of the challenge is that dynamic compilation is slower then static compilation: in this context, one has to set up strategies to reach higher performance. In particular since execution may require crossing borders of different paradigms (assembly, C calls,...) with their associated costs. Once such strategies are understood and accessed, primitives can thus be implemented in the context of just in time compilation.

%%\item Aim \\
This paper describes and compares the recent implementation of optimised versions of the String comparison primitive at different stages of the code execution and their impact on execution time. This analysis is performed on the opensmalltalk virtual machine (known as Cog) that dynamically compiles oriented-object languages such as Newspeak, Pharo and Squeak.
%%\item Perspectives \\
[describes briefly the 3 implementations and compares with V8 and Java?]\\
\end{abstract}

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
%\begin{CCSXML}
%<ccs2012>
%<concept>
%<concept_id>10011007.10011006.10011008</concept_id>
%<concept_desc>Software and its engineering~General programming languages</concept_desc>
%<concept_significance>500</concept_significance>
%</concept>
%<concept>
%<concept_id>10003456.10003457.10003521.10003525</concept_id>
%<concept_desc>Social and professional topics~History of programming languages</concept_desc>
%<concept_significance>300</concept_significance>
%</concept>
%</ccs2012>
%\end{CCSXML}

%\ccsdesc[500]{Software and its engineering~General programming languages}
%\ccsdesc[300]{Social and professional topics~History of programming languages}
%% End of generated code

%% Keywords
%% comma separated list
%%\keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission

%% \maketitle
%% Note: \maketitle command must come after title commands, author
%% commands, abstract environment, Computing Classification System
%% environment and commands, and keywords command.
\maketitle

\section{Introduction}

Short description of the String class (String, ByteString, WideString)?

\sd{this is a test}

\section{Terminology}

\subsection{definitions}
\begin{itemize}
	\item primitive (but already defined in abstract?)
	\item 
\end{itemize}

Understanding primitive implementation and execution

\subsection{MiscPlugin}
Code compiled into bytecode and interpreted\\
trade-off 
\begin{itemize}
	\item +: easy to code, easy to read, implementation in-image, no need to recompile th vm sources in case of modification
	\item -: slow
\end{itemize}

\subsection{primitive in Slang}
Slang is compiled to C (by usual c compilers gcc/llvm), and the resulting machine code is executed\\
trade-off
\begin{itemize}
	\item +: faster execution
	\item -: costly jumps between Smalltalk and C runtime (trampolines), need to recompile vm sources, little control over outcome (c compilers optimizations)
\end{itemize}

\subsection{primitive in baseline JIT}
generate functions as machine code
partial implementation (frequent cases are implemented)
Sometimes, fallback to less-optimised version\\

trade-off
\begin{itemize}
	\item +: faster (faster) execution (no c - Smalltalk runtime jumps), control over the outcome
	\item -: need to recompile vm sources (trampolines), code readability 
\end{itemize}

\subsection{primitive in optimizing JIT}

add appendix with code

\section{Assessment}
All these versions of the primitives have been implemented. Compare performances

\subsection{Micro-benchmarks}
Results on strings of different lengths (3, 10, 1000)

\subsection{Macro-benchmarks}
JSON
PetitParser
XML

\section{Related work}

\begin{itemize}
	\item Java: platform dependent
	\item V8: interpreter compiled by JIT
	optimised backend
	as if we were compiling slang using the JIT compiler (ahead of time)
	no more virtual call
	\item VisualWorks: baseline JIT but no interpreter
\end{itemize}

\section{Future work and conclusion}
1. number of representations\\
2. common representation

%% Acknowledgments
%%\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
%  This material is based upon work supported by the
%  \grantsponsor{GS100000001}{National Science
%    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
%  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
%  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
%  conclusions or recommendations expressed in this material are those
%  of the author and do not necessarily reflect the views of the
%  National Science Foundation.
%\end{acks}

%% Bibliography
%\bibliography{bibfile}

%% Appendix
\appendix
\section{Appendix}

Text of appendix \ldots

\end{document}

### 2018-04-27 paper.tex
Added comments everywhere

@@
 \newcommand{\sd}[1]{\color{red}\fbox{\bfseries\sffamily\scriptsize Stef:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
-\newcommand{\sk}[1]{\color{red}\fbox{\bfseries\sffamily\scriptsize Sophie:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
-\newcommand{\cb}[1]{\color{red}\fbox{\bfseries\sffamily\scriptsize Clement:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
+\newcommand{\sk}[1]{\color{blue}\fbox{\bfseries\sffamily\scriptsize Sophie:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
+\newcommand{\cba}[1]{\color{purple}\fbox{\bfseries\sffamily\scriptsize Clement:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
 
@@
 %% Author with single affiliation.
-\author{First1 Last1}
+\author{Sophie Kaleba}
                                         %% can be repeated if necessary
@@
 %% Author with two affiliations and emails.
-\author{First2 Last2}
-                                        %% can be repeated if necessary
+\author{Cl\'ement B\'era}
 \affiliation{
-  \position{Position2a}
-  \department{Department2a}             %% \department is recommended
-  \institution{Institution2a}           %% \institution is required
-  \streetaddress{Street2a Address2a}
-  \city{City2a}
-  \state{State2a}
-  \postcode{Post-Code2a}
-  \country{Country2a}                   %% \country is recommended
+  % \position{}
+	\department{Software Languages Lab}              %% \department is recommended
+	\institution{Vrije Universiteit Brussel}            %% \institution is required
+	\city{Brussel}
+  % \state{}
+  % \postcode{}
+	\country{Belgium}                    %% \country is recommended
 }
-\email{first2.last2@inst2a.com}         %% \email is recommended
+\email{clement.bera@vub.ac.be}          %% \email is recommended
 
 %% Author with two affiliations and emails.
-\author{First3 Last3}
+\author{St\'ephane Ducasse}
 \affiliation{
-  \position{Position2b}
-  \department{Department2b}             %% \department is recommended
-  \institution{Institution2b}           %% \institution is required
-  \streetaddress{Street3b Address2b}
-  \city{City2b}
-  \state{State2b}
-  \postcode{Post-Code2b}
-  \country{Country2b}                   %% \country is recommended
+ % \position{Position2b}
+  \department{RMoD}             %% \department is recommended
+  \institution{Inria}           %% \institution is required
+  %\streetaddress{Street3b Address2b}
+  \city{Lille}
+  %\state{State2b}
+  %\postcode{Post-Code2b}
+  \country{France}                   %% \country is recommended
 }
-\email{first2.last2@inst2b.org}         %% \email is recommended
+\email{stephane.ducasse@inria.fr}         %% \email is recommended
 
@@
 \begin{abstract}
+
 Most of programming managed languages use primitive functions to speed up their execution time; these primitives usually implement core operations (such as addition, subtraction,...) or frequently used operations.
+\cba{Le papier est sur les primitives d'optimisations, pas sur les primitives essentielles. Je parlerais que des prims d'optimisations dans l'abstract et expliquerais la difference dans l'intro.}
 They are also massively used in the context of dynamic or just-in-time compilation: the compiler then executes this alternative version instead of the standard one to gain performance.
+\cba{Il y a des problemes de terminologie.
+Le JIT compile, n'exécute rien.
+La VM exécute le code.
+A l'oral, par abus de langage, on peut dire le JIT execute, mais dans un papier il faut etre rigoureux}
 %%Problem
 
-One of the challenge is that dynamic compilation is slower then static compilation: in this context, one has to set up strategies to reach higher performance.
+\textbf{One of the challenge is that dynamic compilation is slower then static compilation}:
+\cba{Ce que j'ai mis en gras, je le vois en tant que reviewer de manière aussi directe, je met strong reject et je quote la phrase dans la review.
+Je suis un des reviewers les plus sympas d'ICOOOLPS, les autres seront plus aggressifs que ca.}
+ in this context, one has to set up strategies to reach higher performance.
 In particular since execution may require crossing borders of different paradigms (assembly, C calls,...) with their associated costs.
 Once such strategies are understood and accessed, primitives can thus be implemented in the context of just in time compilation.
+ \cba{Je comprend pas ce que tu veux dire}
+
+ \cba{Je me demande si c'est pas trop verbeux jusque là.
+Ecrire des phrases introduisant: (1) Managed runtime, (2) primitive d'optimisation, (3) la possibilité de l'implem dans le language bas niveau où est implémenté la VM comme C/C++ ou dans une representation intermédiaire du JIT.
+(4) les avantages / problèmes des 2 approches me parait mieux}
 
 %%\item Aim \\
-This paper describes and compares the recent implementation of optimised versions of the String comparison primitive at different stages of the code execution and their impact on execution time.
+This paper describes and compares the recent implementation of optimised versions of the String comparison primitive \textit{at different stages of the code execution} and their impact on execution time.
+\cba{peut etre enleve ça, l'essentiel c'est ce qui est écrit avant}
+\cba{Introduit clairement a quoi corresponde chaque implem, C external lib, C internal code, back-end de JIT, etc.}
  This analysis is performed on the opensmalltalk virtual machine (known as Cog) that dynamically compiles oriented-object languages such as Newspeak, Pharo and Squeak.
+ \cba{Verbeux, a fusionner avec une phrase sur les resultats. Peut etre mettre une phrase comme quoi ca pourrait s'appliquer a d'autres MV avec runtime LLevel C/C++ typiquement + JIT vers code natif}
 %%\item Perspectives \\
 [describes briefly the 3 implementations and compares with V8 and Java?]\\
+\cba{Personnellement je ne parlerais pas des related work dans l'abstract dans ce papier car la direction principale n'est pas une comparaison avec un related work.
+A la place je finirais l'abstract par une phrase sur les résultats en parlant du trade-off, genre dans la back-end du JIT la vitesse d'execution est plus fiable (pas de pb avec les differences de compil LLVM/GCC par ex), est rapide, mais par contre le cout d'ingenierie est supérieur.
+Il faut donc mesure l'utilisation de chaque primitive de performance et mettre un place un ratio complexité/performance }
+
 \end{abstract}
@@
 %% comma separated list
-%%\keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission
+\keywords{Just-in-Time compiler, Primitive, Virtual machine, Managed runtime}  %% \keywords are mandatory in final camera-ready submission
 
@@
 
+\cba{Ici il faut répéter l'abstract en detaillant plus chaque chose.
+Comme le coeur du papier est sur String prim, tu introduis la primitive sur laquelle tu vas travailler, dont la representation des strings, ses cas d'utilisation
+Il faut que tu précises bien le coût de changement entre la pile St et la pile C et que tu expliques le probleme des primitives C courtes appelées depuis le code généré par le JIT.
+Je ferais une figure.
+C'est important pour la suite.
+Il faut definir primitive, primitive essentielle, primitive d'optimisation.
+Il faut expliquer brievement les primitives Smalltalk, comment elles sont activées, dans quelles cas le fall back code est appelé, que si elles fail c'est side-effect free.
+=> Dans mon papier ISMM'15 il y a une explication section 4.2 le début de Primitive operations.
+dont tu peux t'inspirer, voir copier coller comme je suis auteur des 2.}
+
 Short description of the String class (String, ByteString, WideString)?
-
-\sd{this is a test}
+\cba{oui, explique la primitive, dont la representation des strings, ses cas d'utilisation}
 
 \section{Terminology}
+\cba{Je mettrais section 2 "Implementations evaluated", tu ne definis par vraiment de terminology a part la premiere subsection que je bougerais en intro}
 
+\cba{Les definitions de primitive d'optimisation et essentielles je les mettrais dans l'intro}
 \subsection{definitions}
 \begin{itemize}
-	\item primitive (but already defined in abstract?)
+	\item primitive (but already defined in abstract?) \cba{Dans le papier tu dois faire comme si l'abstract n'existait pas et tout redéfinir. Par contre tu peux définir dans l'intro}
 	\item 
@@
 
+\cba{Je pense il faut comparer 
+(1) Smalltalk pur (ta baseline, pas grand chose à discuter, tu peux exprimer les autres en relative speed-up par rapport a la baseline dans les graphes de resultats.)
+(2) SmalltalkSmartSyntax + Misc
+(3) SlangSmartSyntax + Misc, 
+(4) Slang, 
+(5) Cog's RTL
+(6) eventuellement Cog's x86 back-end si tu veux faire une implem spécifique a x86 en plus avec SSE4.}
+\cba{Dans les parties suivante tu définis chaque cas, tu évalues les perfs sur micro et macro benchmarks, tu discutes le code machine généré par les compilo C et le JIT, tu montres les boucles de comparaison ASM générées par les compilo C et le JIT et u mets l'intégralité des fonctions avec tous les codes sources et générés en appendices.}
+
 \subsection{MiscPlugin}
@@
 \begin{itemize}
-	\item +: easy to code, easy to read, implementation in-image, no need to recompile th vm sources in case of modification
+	\item +: easy to code, easy to read, implementation in-image, no need to recompile th vm sources in case of modification \cba{Ah bon, Y'a pas besoin de recompiler les sources en cas de modif VM?
+Ca sort d'ou ca?
+Le problème est que l'image VMMaker qui génère le code C de la VM utilise sa version de la primitive justement}
 	\item -: slow
@@
 \begin{itemize}
-	\item +: faster execution
+	\item +: faster execution \cba{Pourquoi c'est plus rapide? Access to VM internal API, etc.}
 	\item -: costly jumps between Smalltalk and C runtime (trampolines), need to recompile vm sources, little control over outcome (c compilers optimizations)
@@
 add appendix with code
+
+\cba{Future work -> primitive processor dependent (i.e. SSE4 instrs), pros and cons and why you did not do it (ingeniering cost per back-end)}
 
@@
 
+\cba{Ajoute le related work que tu as trouvé de ICOOOLPS avec le travail de Tim.}
+
 \section{Future work and conclusion}
@@
 2. common representation
+
+\cba{Ici tu mets 
+(6) eventuellement Cog's x86 back-end si tu veux faire une implem spécifique a x86 en plus avec SSE4.
+si tu l'as pas mis avant}
 
@@
 
+\cba{Source code et generated C and ASM code for all versions}
+
 Text of appendix \ldots

### 2018-05-02 paper.tex
pass on abstract

--- 2018-04-04 paper.tex
@@
 %% Author with single affiliation.
-\author{First1 Last1}
+\author{Sophie Kaleba}
                                         %% can be repeated if necessary
 \affiliation{
-  \position{Position1}
-  \department{Department1}              %% \department is recommended
-  \institution{Institution1}            %% \institution is required
-  \streetaddress{Street1 Address1}
-  \city{City1}
-  \state{State1}
-  \postcode{Post-Code1}
-  \country{Country1}                    %% \country is recommended
+  %\position{Position1}
+  \department{RMoD}              %% \department is recommended
+  \institution{Inria}            %% \institution is required
+ % \streetaddress{Street1 Address1}
+  \city{Lille}
+  %\state{France}
+  %\postcode{Post-Code1}
+  \country{France}                    %% \country is recommended
 }
-\email{first1.last1@inst1.edu}          %% \email is recommended
+\email{sophie.kaleba@etudiant.univ-lille1.fr}          %% \email is recommended
 
 %% Author with two affiliations and emails.
-\author{First2 Last2}
-                                        %% can be repeated if necessary
+\author{Cl\'ement B\'era}
 \affiliation{
-  \position{Position2a}
-  \department{Department2a}             %% \department is recommended
-  \institution{Institution2a}           %% \institution is required
-  \streetaddress{Street2a Address2a}
-  \city{City2a}
-  \state{State2a}
-  \postcode{Post-Code2a}
-  \country{Country2a}                   %% \country is recommended
+  % \position{}
+	\department{Software Languages Lab}              %% \department is recommended
+	\institution{Vrije Universiteit Brussel}            %% \institution is required
+	\city{Brussel}
+  % \state{}
+  % \postcode{}
+	\country{Belgium}                    %% \country is recommended
 }
-\email{first2.last2@inst2a.com}         %% \email is recommended
+\email{clement.bera@vub.ac.be}          %% \email is recommended
 
 %% Author with two affiliations and emails.
-\author{First3 Last3}
+\author{St\'ephane Ducasse}
 \affiliation{
-  \position{Position2b}
-  \department{Department2b}             %% \department is recommended
-  \institution{Institution2b}           %% \institution is required
-  \streetaddress{Street3b Address2b}
-  \city{City2b}
-  \state{State2b}
-  \postcode{Post-Code2b}
-  \country{Country2b}                   %% \country is recommended
+ % \position{Position2b}
+  \department{RMoD}             %% \department is recommended
+  \institution{Inria}           %% \institution is required
+  %\streetaddress{Street3b Address2b}
+  \city{Lille}
+  %\state{State2b}
+  %\postcode{Post-Code2b}
+  \country{France}                   %% \country is recommended
 }
-\email{first2.last2@inst2b.org}         %% \email is recommended
+\email{stephane.ducasse@inria.fr}         %% \email is recommended
 
@@
 \begin{abstract}
-Most of programming managed languages use primitive functions to speed up their execution time; these primitives usually implement core operations (such as addition, subtraction,...) or frequently used operations.
-They are also massively used in the context of dynamic or just-in-time compilation: the compiler then executes this alternative version instead of the standard one to gain performance.
-%%Problem
 
-One of the challenge is that dynamic compilation is slower then static compilation: in this context, one has to set up strategies to reach higher performance.
-In particular since execution may require crossing borders of different paradigms (assembly, C calls,...) with their associated costs.
-Once such strategies are understood and accessed, primitives can thus be implemented in the context of just in time compilation.
+Virtual machines, besides the interpreter and just-in-time compiler optimization facilities, also include a set of primitive operations that the client language can use.
+Some of these are essential and cannot be performed in any other way.
+Others are optional, they can be expressed in the client language but are often implemented in the virtual machine to improve performance when the just-in-time compiler is unable to do so (start-up performance, speculative optimizations not implemented or not mature enough, etc.).
 
-%%\item Aim \\
-This paper describes and compares the recent implementation of optimised versions of the String comparison primitive at different stages of the code execution and their impact on execution time.
-This analysis is performed on the opensmalltalk virtual machine (known as Cog) that dynamically compiles oriented-object languages such as Newspeak, Pharo and Squeak.
-%%\item Perspectives \\
-[describes briefly the 3 implementations and compares with V8 and Java?]\\
+In a hybrid runtime, where code is executed by an interpreter and a just-in-time compiler, the implementor can choose to implement optional primitives in the client language, in the virtual machine implementation language (typically C or C++) or on top of the just-in-time compiler back-end.
+We implemented the String comparison optional primitive in each case.
+This paper describes the different implementations and compares the execution time in Cog, a Smalltalk virtual machine.
+
+The paper shows that the implementation with the most reliable and high performance is also the most complex and difficult to implement.
+Since hundreds of optional primitives can be available, the virtual machine implementor has to carefully choose how to implement each of them to balance between maintenance cost and high performance.
+
 \end{abstract}
@@
 
+%Managed lang
+
+Dynamically typed programming languages are traditionally
+implemented as virtual machines (VMs). As the popularity
+of VM-based languages has increased, VM implementation
+techniques have evolved to the point where most mature
+VMs utilize some form of just-in-time (JIT) compilation, with
+two main technologies contending at the moment: method
+JITs [12], which generally compile methods at a time into
+machine code, and utilize techniques such as inlining and
+branch pruning to simplify the code or merge multiple methods;
+and tracing JITs [1], which record the actions of the
+program including branches taken and methods dispatched
+into a specific trace and compile it into machine code
+
+Define JIT VM acronym
+
+\subsection{Primitive execution}
+
+ prim, optional and essential MENTION built-in
+ Il faut definir primitive, primitive essentielle, primitive d'optimisation.
+Il faut expliquer brievement les primitives Smalltalk, comment elles sont activées, dans quelles cas le fall back code est appelé, que si elles fail c'est side-effect free.
+=> Dans mon papier ISMM'15 il y a une explication section 4.2 le début de Primitive operations.
+dont tu peux t'inspirer, voir copier coller comme je suis auteur des 2.
+
+primitive performance: VM without opt JIT, before reaching peak perf, or jit incorrectly optimizing specific part of code.
+
+\subsection{VM execution model}
+
+Il faut expliquer le process de compilation de la VM. (Slang -> C, C-> ASM, au runtime JIT ST->ASM)
+ execution with C and St stack
+
+\begin{figure}[tb]
+		\centering
+		\includegraphics[width=0.6\linewidth]{figures/VMCompilation}
+		\caption{VM compilation}
+		\label{fig:VMCompilation}
+\end{figure}
+
+ Il faut que tu précises bien le coût de changement entre la pile St et la pile C et que tu expliques le probleme des primitives C courtes appelées depuis le code généré par le JIT.
+Je ferais une figure.
+C'est important pour la suite.
+
+\subsection{String comparison primitive}
+
+ Comme le coeur du papier est sur String prim, tu introduis la primitive sur laquelle tu vas travailler, dont la representation des strings, ses cas d'utilisation
  Short description of the String class (String, ByteString, WideString)?
 
-\sd{this is a test}
+\section{Different primitive implementations}
 
-\section{Terminology}
+ \subsection{Context}
 
-\subsection{definitions}
-\begin{itemize}
-	\item primitive (but already defined in abstract?)
-	\item 
-\end{itemize}
+ [Max 1/2 page, justifie 1,2,3 et non neg,0,pos]
 
-Understanding primitive implementation and execution
+We had SmartSyntax primitive.
+Explain 1,2,3 results
+Problem compatibility other languages
+expect neg,0,pos
 
-\subsection{MiscPlugin}
-Code compiled into bytecode and interpreted\\
-trade-off 
-\begin{itemize}
-	\item +: easy to code, easy to read, implementation in-image, no need to recompile th vm sources in case of modification
-	\item -: slow
-\end{itemize}
+Wanting to improve performance
 
-\subsection{primitive in Slang}
-Slang is compiled to C (by usual c compilers gcc/llvm), and the resulting machine code is executed\\
-trade-off
-\begin{itemize}
-	\item +: faster execution
-	\item -: costly jumps between Smalltalk and C runtime (trampolines), need to recompile vm sources, little control over outcome (c compilers optimizations)
-\end{itemize}
+2 critical problems:
+(1) changing the prim => changing all clients (VM compil from Pharo or Squeak different results)
+(2) no support in SmartSyntax for read-only objects. [plugins but not String comparison]
 
-\subsection{primitive in baseline JIT}
+Premiere tentative Eliot -> Slang plugin. Keep 1,2,3. Resoud Read-only / changing the prim problems.
+
+Deuxieme tentative Slang internal, then Cog's RTL.
+=> pour simplifier le RTL neg,0,pos etait mieux => minus instead of branches.
+=> compat execution model, Slang internal neg,0, pos too.
+
+ \subsection{Different implementation and execution}
+
+- Pure Smalltalk: baseline
+Smalltalk code on top of VM. Used as our baseline for benchmarks.
+Write down the Smalltalk code here.
+Expliquer to:do: over do: (to explain expresiveness) => core lib often use to:do: and not do:
+
+- Smart Syntax Plugin
+%= aString 	
+%	self == aString ifTrue: [ ^true ].
+%	aString isString ifFalse: [ ^false ].
+%	self size = aString size ifFalse: [ ^false ].
+%	1 to: self size do: [:i | (self at: i) = (aString at: i) ifFalse: [^false] ].
+%	^ true
+Restricted Smalltalk
+
+written in Smalltalk with restrictions.
+Compiled to C AOT into a plugin.
+
+- Slang plugin 
+Slang, but no access au fonctions de la VM drectement.
+ interpreterProxy. (C compiler inlining, linking time ? cmacro ?)
+
+- Slang pure
+
+- Slang pure + Cog RTL
+
 generate functions as machine code
 partial implementation (frequent cases are implemented)
-Sometimes, fallback to less-optimised version\\
+Sometimes, fallback to less-optimised version
 
-trade-off
-\begin{itemize}
-	\item +: faster (faster) execution (no c - Smalltalk runtime jumps), control over the outcome
-	\item -: need to recompile vm sources (trampolines), code readability 
-\end{itemize}
+ \subsection{Pros and cons}
 
-\subsection{primitive in optimizing JIT}
+  Can be changed like Smalltalk code (No C compilation whatsoever). Readability. Only impl without VM recompilation.
 
-add appendix with code
+Baseline performance is not that great, jtted performance is good only with mature optimizing JIT we don't have.
 
-\section{Assessment}
+Number of implementations: 1, 2 or 3: same implem for primitive in C and fall back code in Smalltalk, relatively easy to read/code.
+recompilation. 
+changing the prim => changing all clients (VM compil from Pharo or Squeak different results
+Performance: cross file inlining not possible by default. (InterpreterProxy indirection)
+Control on generated code => difference LLVM/GCC
+ Pure C performance (no interpreter proxy).
+ no C-St jump. Control over the outcome (reliable performance)
+ costly jumps between Smalltalk and C runtime (trampolines), need to recompile vm sources, little control over outcome (c compilers 
+ need to recompile vm sources (trampolines), code readability
+
+ \subsection{Comparison}
+
+\cba{expliquer tableau comparatif}
+
+Trade-off engineering time and performance.
+- recompiler la VM ou pas (Plugin externes, code Smalltalk)
+- complexite d'implem (comprehension Smalltalk, C, ASM)
+
+%%%%%
+% END OLD
+%%%%%
+
+\section{Evaluation}
 All these versions of the primitives have been implemented. Compare performances
+
+\subsection{Set-up}
+Describe set-up:
+GCC eval Ubuntu + machine spec
+LLVM eval Mac OS X + machine spec.
 
@@
 
+4 string length. (1er <=3, dernier > 1000) 
+line 1: GCC
+line 2: LLVM
+String evaluation here.
+
+\cba{Show ASM loops and discuss the differences
+add appendix with full code}
+
 \subsection{Macro-benchmarks}
 JSON
-PetitParser
-XML
 
@@
 \begin{itemize}
-	\item Java: platform dependent
+	%\item Java: platform dependent
 	\item V8: interpreter compiled by JIT
@@
 	no more virtual call
-	\item VisualWorks: baseline JIT but no interpreter
+	\item VisualWorks: baseline JIT but no interpreter => only 1 implement to write and maintain.
+	\item Self - Strongtalk -> tout en Smalltalk avec mature optimizing JIT. pb baseline perf, haut cout de maintenance implem pour l'adaptive optimizer
 \end{itemize}
+
+\cba{Ajoute le related work que tu as trouvé de ICOOOLPS avec le travail de Tim.}
 
@@
 2. common representation
+\cba{Future work -> primitive processor dependent (i.e. SSE4 instrs), pros and cons and why you did not do it (ingeniering cost per back-end)}
+\cba{Inlining in Sista + ref}
+
+Conclusion: trade off expressiveness performance
 

### 2018-05-02 paper.tex
wrote an introduciton I like and added some references

@@
 %% Bibliography style
-%%\bibliographystyle{ACM-Reference-Format}
+\bibliographystyle{ACM-Reference-Format}
 %% Citation style
 %\citestyle{acmauthoryear}  %% For author/year citations
-%\citestyle{acmnumeric}     %% For numeric citations
+\citestyle{acmnumeric}     %% For numeric citations
 %\setcitestyle{nosort}      %% With 'acmnumeric', to disable automatic
@@
 \section{Introduction}
+\label{sec:intro}
 
-%Managed lang
+High level object oriented programming languages are often implemented on top of a virtual machine (VM).
+As VMs has become more popular, different techniques have been set-up using Just-In-Time (JIT) compilation to improve the overall performance of the runtime such as method JITs \cite{ursPHD} which usually compile methods to native code and tracing JITs \cite{Dynamo, PyPyTracing} which usually compile linear traces of execution into native code.
 
-Dynamically typed programming languages are traditionally
-implemented as virtual machines (VMs). As the popularity
-of VM-based languages has increased, VM implementation
-techniques have evolved to the point where most mature
-VMs utilize some form of just-in-time (JIT) compilation, with
-two main technologies contending at the moment: method
-JITs [12], which generally compile methods at a time into
-machine code, and utilize techniques such as inlining and
-branch pruning to simplify the code or merge multiple methods;
-and tracing JITs [1], which record the actions of the
-program including branches taken and methods dispatched
-into a specific trace and compile it into machine code
+On top of its virtual machine, the client language can use a set of \emph{primitives}~\footnote{We use in the paper the Smalltalk terminology (primitive) as this is the programming language used for our evaluation.
+Some other programming languages, such as Javascript, prefer to use the term built-in instead of primitive.}, which are performed directly by the interpreter rather than by evaluating expressions in a method.
+We distinguish two kinds of primitives:
+\begin{itemize}
+	\item \emph{Essential primitives} cannot be performed in any other way.
+A high-level object oriented language without primitives can move values from one variable to another, but cannot add two integers together.
+Many arithmetic and comparison operations between numbers are primitives.
+Some primitives allow to communicate with I/O devices such as the disk, the display and the keyboard.
+	\item \emph{Optional primitives} exist only to make the system run faster.
+They can be implemented in the client language directly, making its implementation as a primitive optional, but they are implemented in the VM to improve performance.
+\end{itemize}
 
-Define JIT VM acronym
+The implementation of an optional primitive in the VM rather than the client language improves performance but also often increases the implementation engineering cost.
+For example, accessing objects in the VM usually requires some understanding of the memory layout or implementation details, which are not needed in the client language.
 
-\subsection{Primitive execution}
+For this reason, some virtual machine implementors attempted to remove most or all optional primitives.
+Their goal was to improve the execution of the client language with a JIT performing speculative optimizations \cite{ursPHD} for the execution of the optional primitives implemented in the client language to be as fast as the performance of the same primitives written in the VM.
+Self \cite{ursPHD} and Strongtalk \cite{Strongtalk} were to first to try this approach.
+Early versions of the Javascript engine V8 \cite{V8} also implemented most of the primitives, including for example Array operations, in Javascript itself.
+However, without optional primitives implemented in the VM, several problems rise:
+\begin{itemize}
+	\item High performance relies on a JIT with speculative optimizations, which is hard to implement, maintain and evolve,
+	\item The performance of the code not yet optimized is much worse than the peak performance code,
+	\item Even mature JIT fail to optimize some narrow cases, where the performance is drastically slower.
+\end{itemize}
+Overall, this approach requires high engineering time to get a mature optimizing JIT with speculative optimizations and even then, it leads in practice to unreliable performance.
 
- prim, optional and essential MENTION built-in
- Il faut definir primitive, primitive essentielle, primitive d'optimisation.
-Il faut expliquer brievement les primitives Smalltalk, comment elles sont activées, dans quelles cas le fall back code est appelé, que si elles fail c'est side-effect free.
-=> Dans mon papier ISMM'15 il y a une explication section 4.2 le début de Primitive operations.
-dont tu peux t'inspirer, voir copier coller comme je suis auteur des 2.
+VMs are traditionally implemented in a low-level language such as C or C++.
+To balance between memory footprint, start-up performance and peak performance, the execution of code is usually done through multiple execution tiers: the first few executions of a code snippets are done through an interpreter and the JIT compiles at runtime the frequently used code snippets for further executions to be performed quicker.
+This leads to annoying concerns.
+For example, let's say we implement a primitive in C.
+Frequently used portion of code calling that primitive are going to be compiled to native code by the JIT.
+This is problematic since the native code generated by the JIT is not directly compatible with native code generated by the C compiler: a call from one to another may require to edit the stack pointer and the frame pointer from the client stack to the C stack and to spill or move multiple registers.
+Switching between both can waste up to around a dozen native instructions.
+If the primitive is going to execute many instructions, the switch overhead might be negligible, but if the primitive executes only a few instructions, the overhead can be noticeable.
 
-primitive performance: VM without opt JIT, before reaching peak perf, or jit incorrectly optimizing specific part of code.
+In the paper, we took the example of a string comparison optional primitive.
+The primitive compares two strings and answers if one string is greater, equal or lower than the other string.
+The exact specification are detailed in Section \ref{subsec:primSpec}.
+We implemented a first version in Smalltalk.
+We then implemented multiple version compiling through C compilers to native code, using different part of our VM infrastructure.
+Lastly, we implemented a version in the back-end of the JIT.
+Each time, we tried to write the code in the most efficient way possible.
+Then, we compared the execution time of all versions, showing that overall the most complex implementation is the fastest to execute.
 
-\subsection{VM execution model}
+Section \ref{sec:implContext} describes our implementation context, \emph{i.e.,} the execution model of the VM we used for our evaluation, how it executes primitives and the specification of our string comparison primitive.
+Section \ref{sec:implem} details the various implementations of the primitive we used for the evaluation.
+In Section \ref{sec:eval}, we evaluate the performance of the different implementations with different C compilers and string sizes.
+We also show the native code of the performance critical part of the primitive generated by each version and discuss it.
+Further sections compare our work to related works, discuss future work and conclude.
+
+\section{Implementation context}
+\label{sec:implContext}
+
+Our evaluation is based on the Cog
+
+\subsection{VM execution.}
+
+\cite{blueBook}
+\cite{PharoByExample,SqueakByExample, NewspeakOopsla}
+\cite{BackToTheFuture, CogJIT}
+exec hybrid interp JIT.
+
+Impl langage.
 
@@
 Il faut que tu précises bien le coût de changement entre la pile St et la pile C et que tu expliques le probleme des primitives C courtes appelées depuis le code généré par le JIT.
-Je ferais une figure.
 C'est important pour la suite.
 
-\subsection{String comparison primitive}
+\subsection{Primitive execution.}
+
+ Il faut definir primitive, primitive essentielle, primitive d'optimisation.
+Il faut expliquer brievement les primitives Smalltalk, comment elles sont activées, dans quelles cas le fall back code est appelé, que si elles fail c'est side-effect free.
+=> Dans mon papier ISMM'15 il y a une explication section 4.2 le début de Primitive operations.
+dont tu peux t'inspirer, voir copier coller comme je suis auteur des 2.
+
+primitive performance: VM without opt JIT, before reaching peak perf, or jit incorrectly optimizing specific part of code.
+
+\subsection{String comparison primitive.}
+\label{subsec:primSpec}
+
+Section \ref{subsec:context} explain why we changed it.
 
@@
 \section{Different primitive implementations}
+\label{sec:implem}
 
  \subsection{Context}
+ \label{subsec:context}
+
+ \cba{Move that to past section?}
 
@@
 \section{Evaluation}
+\label{sec:eval}
+
 All these versions of the primitives have been implemented. Compare performances
@@
 
-\begin{itemize}
+Work on reducing the number of prim
+
+RPython \cite{NoPrimTracing}
+
+Self - Strongtalk -> tout en Smalltalk avec mature optimizing JIT. pb baseline perf, haut cout de maintenance implem pour l'adaptive optimizer \cite{Strongtalk, ursPHD}
+
 	%\item Java: platform dependent
-	\item V8: interpreter compiled by JIT
+	 V8: interpreter compiled by JIT \cite{V8} Mention Ignition.
 	optimised backend
@@
 	no more virtual call
-	\item VisualWorks: baseline JIT but no interpreter => only 1 implement to write and maintain.
-	\item Self - Strongtalk -> tout en Smalltalk avec mature optimizing JIT. pb baseline perf, haut cout de maintenance implem pour l'adaptive optimizer
-\end{itemize}
+
+ VisualWorks: baseline JIT but no interpreter => only 1 implement to write and maintain.
 
@@
 %% Bibliography
-%\bibliography{bibfile}
+\bibliography{sista}
 

### 2018-05-02 paper.tex
started implem ctxt

@@
 %% Title information
-\title[Short Title]{Assessing primitives performance\\ on multi-stage execution}
+\title[Assessing primitives performance on multi-stage execution]{Assessing primitives performance\\ on multi-stage execution}
 
@@
 
-Our evaluation is based on the Cog
+Our evaluation is based on the Cog VM a Smalltalk virtual machine.
+The Cog VM is based on Dan Ingalls' Smalltalk VM \cite{BackToTheFuture} and have been enhanced with a JIT \cite{CogJIT} compiling one method or closure at a time to native code.
+Historically, the VM implemented a 32-bits adaptation of the Smalltalk-80 specifications \cite{blueBook}, but the Cog VM is now the default VM for multiple programming languages such as Pharo \cite{PharoByExample}, Squeak \cite{SqueakByExample} and Newspeak \cite{NewspeakOopsla}.
 
-\subsection{VM execution.}
+Section \ref{sec:VMCompil} describes briefly the implementation languages and compilation process of the Cog VM.
+Section \ref{sec:VMexec} explains briefly the hybrid execution model with an interpreter and a JIT.
+In Section \ref{sec:primExec}, we detail how primitives are executed.
+Section \ref{subsec:primSpec} discusses the specifications of the string comparison primitive used for our evaluation.
 
-\cite{blueBook}
-\cite{PharoByExample,SqueakByExample, NewspeakOopsla}
-\cite{BackToTheFuture, CogJIT}
-exec hybrid interp JIT.
+\subsection{VM compilation}
+\label{sec:VMCompil}
 
-Impl langage.
+Similar to \cite{RPythonToolchain}
 
-Il faut expliquer le process de compilation de la VM. (Slang -> C, C-> ASM, au runtime JIT ST->ASM)
- execution with C and St stack
+Impl langage. JIT.
+
+Il faut expliquer le process de compilation de la VM. (Slang -> C, C-> ASM,
 
@@
 
+\subsection{VM execution}
+\label{sec:VMexec}
+
+exec hybrid interp 
+ au runtime JIT ST->ASM)
+ execution with C and St stack
+
 Il faut que tu précises bien le coût de changement entre la pile St et la pile C et que tu expliques le probleme des primitives C courtes appelées depuis le code généré par le JIT.
@@
 
-\subsection{Primitive execution.}
+\subsection{Primitive execution}
+\label{sec:primExec}
 
@@
 
-\subsection{String comparison primitive.}
+\subsection{String comparison primitive}
 \label{subsec:primSpec}

### 2018-05-02 paper.tex
wrote prim specification subsection in the train station

@@
 
-Similar to \cite{RPythonToolchain}
-
-Impl langage. JIT.
-
-Il faut expliquer le process de compilation de la VM. (Slang -> C, C-> ASM,
-
 \begin{figure}[tb]
@@
 		\includegraphics[width=0.6\linewidth]{figures/VMCompilation}
-		\caption{VM compilation}
+		\caption{Cog VM compilation}
 		\label{fig:VMCompilation}
@@
 
+Most of the Cog VM code base is written in Slang, a subset of Smalltalk.
+Slang is compiled to C and then to native code through standard C compilers.
+The execution engine (the memory manager, the interpreter and the baseline JIT) are entirely written in Slang.
+The two main purposes of using slang code over plain C are:
+\begin{itemize}
+	\item to specify with annotations what function needs to be inlined or duplicated with constant parameters, and
+	\item to be able to simulate VM execution, executing the Slang code as Smalltalk code on top of a compiled VM for debugging purposes.
+\end{itemize}
+
+The executable is generated in two steps as shown on Figure \ref{fig:VMCompilation}, similarly to the RPython toolchain \cite{RPythonToolchain}.
+The first step is to generate the two C files representing the whole execution engine written in Slang using the Slang-to-C compiler.
+During the second step, a C compiler is called to compile the execution engine and the platform-specific code written directly in C to the executable VM.
+
 \subsection{VM execution}
 \label{sec:VMexec}
+
+%TODO
+
+primitive implemented in C or Cog's back-end.
 
@@
 
-Il faut que tu précises bien le coût de changement entre la pile St et la pile C et que tu expliques le probleme des primitives C courtes appelées depuis le code généré par le JIT.
-C'est important pour la suite.
+coût de changement entre la pile St et la pile C et que tu expliques le probleme des primitives C courtes appelées depuis le code généré par le JIT. C'est important pour la suite.
 
@@
 
- Il faut definir primitive, primitive essentielle, primitive d'optimisation.
-Il faut expliquer brievement les primitives Smalltalk, comment elles sont activées, dans quelles cas le fall back code est appelé, que si elles fail c'est side-effect free.
-=> Dans mon papier ISMM'15 il y a une explication section 4.2 le début de Primitive operations.
-dont tu peux t'inspirer, voir copier coller comme je suis auteur des 2.
+In the Cog VM, primitives are always associated with compiled methods.
+A compiled method has information in its header to inform the virtual machine if it has a primitive operation or not.
+When activating a method with a primitive, the primitive function is executed before the method's bytecode.
+If the primitive succeeds, the primitive returns a result, as if it were the result of the virtual call.
+If the primitive fails, it does so without side-effects and execution continues executing the method's bytecode, as if the primitive had not been present.
 
-primitive performance: VM without opt JIT, before reaching peak perf, or jit incorrectly optimizing specific part of code.
+To fail without side-effects a primitive must validate any parameters and any state fetched from them, before changing execution state by performing its operations.
+Validation involves any of testing for a specific class, testing for bit vs pointer objects, bounds checks, and recursively applying these tests to substructure of the parameters.
+For example, the primitive that installs a cursor examines the first parameter to check that it represents a valid cursor object, comprised of two bitmaps, one for the image and one for the shape, plus a point to specify the cursor's hotspot.
 
@@
 
-Section \ref{subsec:context} explain why we changed it.
+\cba{Cette partie je faisais la queue dans la gare en mm temps, a relire}
 
- Comme le coeur du papier est sur String prim, tu introduis la primitive sur laquelle tu vas travailler, dont la representation des strings, ses cas d'utilisation
- Short description of the String class (String, ByteString, WideString)?
+ The primitive specification is a bit tedious to explain, since as we were trying to improve its performance, we had to change the specifications.
+In this section we describe first the string implementation, then the original primitive specification and why they were problematic and lastly the new specifications.
+
+\paragraph{String representations.} Strings are represented in all programming languages on top of the Cog VM in two possible forms, ByteStrings and WideStrings.
+ByteStrings encode each character of the string in a single byte.
+The encoding usually follows the extended ASCII standard.
+WideStrings encode each character of the string in 32 bits.
+The encoding usually follows Unicode specifications.
+In this paper, we will discuss only the implementation of the string comparison primitive for ByteStrings.
+WideStrings use an implementation entirely implemented in the client language, and so far, it has not been reported as a performance bottleneck for any program of any users.
+To sort arrays of strings, strings can be compared.
+They can be compared in the default order (ASCII or Unicode) or using different orders, for example, the case insensitive order.
+
+\paragraph{Original specification.}  The original primitive took three operands.
+The two first operands were the two ByteStrings to compare and the third operand is the order table, a ByteArray encoding the order of characters.
+The default order is the ASCII order, hence each entry corresponds to its index in the byte array (The ByteArray has a size of 256, and accessing the element i, with 0 <= i <= 255, answered i).
+However, in some case, the program may do case insensitive ByteStrings comparison.
+In this case, the order ByteArray is different, since some entries answer the same value (for example, the entry for the 41 and 61, respectively characters A and a, both answer the same value).
+To compare the two ByteStrings, the primitive conceptually iterates byte by byte over the two objects until it reaches the end of one them, comparing for each byte that the value present in the order ByteArray at the index is greater, equal or smaller than the value computed from the other ByteString.
+If a difference is met, the value returns based on the result, if not, at the end of the iteration the primitive compares the ByteString sizes and returns.
+Based on the three parameters, the primitive answered 3,2 or 1 depending if the first ByteString was greater, equal or smaller than the second one.
+
+\paragraph{Specification issues.} As we were trying to optimize string comparison but specifically string equality, we encountered two main issues:
+\begin{itemize}
+	\item \emph{Returned value:} Other programming languages implement the primitive to return a negative integer, 0 or a positive integer instead of 1,2 or 3.
+It looks like a minor difference, but it is quite important for performance.
+With a negative, 0 or positive value, the primitive can just answer the difference between two different characters or the difference in size if all first characters are equal.
+With 1,2 or 3, the primitive has to end with multiple branches to know what number to return.
+In native code, a subtraction is on common architecture quicker to perform than one or multiple branches.
+In addition, the primitive code is easier to write with the subtraction when writing directly the assembly instructions and we did exactly that when writing the primitive on top of the JIT's back-end.
+	\item \emph{Order:} Having the order as the third operands is convenient in some cases (case insensitive comparison), but is just overhead in the common case (checking for string equality).
+\end{itemize}
+
+\paragraph{Final specification.} The final specification we productized last month takes two or three parameters.
+The third parameter, the order, is optional.
+If the third parameter is absent, the VM assumes the byte ordering (ASCII order) is the order to compare against.
+Therefore, the VM does not need to go through the indirection order array in the common case.
+The primitive answers a negative value if the first parameter is smaller than the second, 0 if they are equal, and a positive value if the first parameter is greater than the second one.
 
@@
 
- \subsection{Context}
- \label{subsec:context}
-
- \cba{Move that to past section?}
-
- [Max 1/2 page, justifie 1,2,3 et non neg,0,pos]
-
-We had SmartSyntax primitive.
-Explain 1,2,3 results
-Problem compatibility other languages
-expect neg,0,pos
-
-Wanting to improve performance
-
-2 critical problems:
-(1) changing the prim => changing all clients (VM compil from Pharo or Squeak different results)
-(2) no support in SmartSyntax for read-only objects. [plugins but not String comparison]
-
-Premiere tentative Eliot -> Slang plugin. Keep 1,2,3. Resoud Read-only / changing the prim problems.
-
-Deuxieme tentative Slang internal, then Cog's RTL.
-=> pour simplifier le RTL neg,0,pos etait mieux => minus instead of branches.
-=> compat execution model, Slang internal neg,0, pos too.
+ \cba{petite intro avec plan de section}
 
@@
 - Smart Syntax Plugin
+%CHECK how to import code env in latex.
+%\begin{code}
 %= aString 	
@@
 %	self size = aString size ifFalse: [ ^false ].
-%	1 to: self size do: [:i | (self at: i) = (aString at: i) ifFalse: [^false] ].
+%	1 to: self size do: [:i | (self at: i) = (aString at: i) ifFalse: [ ^ false ] ].
 %	^ true
+%\end{code}
 Restricted Smalltalk
@@
 
+\cba{Je pense string de taille 0-1 c'est cool pour montrer que le code C va moins vite que pur smalltalk dans ce cas (si c'est vraiment le cas)}
 4 string length. (1er <=3, dernier > 1000) 

### 2018-05-03 paper.tex
comments

@@
 
-\begin{figure}[tb]
-		\centering
-		\includegraphics[width=0.6\linewidth]{figures/VMCompilation}
-		\caption{Cog VM compilation}
-		\label{fig:VMCompilation}
-\end{figure}
-
 Most of the Cog VM code base is written in Slang, a subset of Smalltalk.
@@
 
+\begin{figure}[bth]
+		\centering
+		\includegraphics[width=0.6\linewidth]{figures/VMCompilation}
+		\caption{Cog VM compilation}
+		\label{fig:VMCompilation}
+\end{figure}
+
 The executable is generated in two steps as shown on Figure \ref{fig:VMCompilation}, similarly to the RPython toolchain \cite{RPythonToolchain}.
@@
 
-%TODO
+The Cog VM executes the client's code using an hybrid runtime with an interpreter and a JIT.
+The interpreter uses a global look-up cache to speed-up virtual calls.
+On cache hits, it requests the JIT to compile that method and the native code version is used instead.
+In most case, the JIT is used at the second activation of the method.
+Closures have similar heuristics.
 
-primitive implemented in C or Cog's back-end.
+The VM code, including the interpreter code, is compiled by a C compiler and uses the C stack at runtime.
+During its execution, the interpreter modifies the execution stack of the client language, which is different from its own code stack as shown on Figure \ref{fig:TwoStacks}.
+This means for example that the frame pointer register in the processor refers to a stack frame on the C stack and that a processor push instruction pushes a value on the C stack.
+When executing the native code generated by JIT, only the client language stack is used.
+This means for example that the frame pointer register in the processor refers to a stack frame on the client stack and that a processor push instruction pushes a value on the client stack.
 
-exec hybrid interp 
- au runtime JIT ST->ASM)
- execution with C and St stack
+\begin{figure}[tbh]
+		\centering
+		\includegraphics[width=0.99\linewidth]{figures/TwoStacks}
+		\caption{Client and C stacks}
+		\label{fig:TwoStacks}
+\end{figure}
 
-coût de changement entre la pile St et la pile C et que tu expliques le probleme des primitives C courtes appelées depuis le code généré par le JIT. C'est important pour la suite.
+Because of this design, switching the execution between the interpreter and the native code generated by the JIT has a cost.
+Effectively, the VM needs to switch the frame pointer, the stack pointer and depending on architecture the link register from one stack to the other one.
+It also needs to save those pointers to be able to resume execution from one stack to the other one.
+In addition, caller-saved registers on the C conventions or the JIT convention need to be saved since the VM cannot tell ahead what code is going to be used in the other execution model and what registers are required.
+We estimate the overhead to up to around a dozen of processor instruction, the exact number varies a lot so it's difficult to precise it (the number of register to edit depends on the processor and on the code executed).
+
+Most of the primitives are implemented either in Slang or directly in C.
+They benefit from the C compiler optimizations to be efficient.
+When called from the interpreter, activating a primitive is just a C call (the primitive function pointer is cached in the look-up cache next to the method to activate).
+When called from the native code generated by the JIT, activating a primitive requires to switch the execution form the client stack to the C stack, perform the primitive, and switch back to resume execution.
+If the primitive takes a significant amount of instructions to execute, for example, the primitive is copying 100kb from one array to another one, the stack switch dance overhead is negligible.
+If the primitive is very quick to perform, for example, the primitive is adding two integers which do not overflow, the overhead is massive: the execution cost of the addition goes from a few processor instructions to a few dozen instructions.
+
+\cba{TODO!}
+To solve this performance bottleneck, the Cog VM allows to redefine primitives on top of its back-end. 
+integrated as part of the native code generated.
+No switch cost.
+No need to reimpl long running code -> direct in Abs assembly, hard to write and maintain, have to be smart to compete with optimized C code.
+=> often partial impl for quick to perform prim, for ex, add of 2 int, if overflow just go back to C prim.
 
@@
 The primitive answers a negative value if the first parameter is smaller than the second, 0 if they are equal, and a positive value if the first parameter is greater than the second one.
+\cba{Maybe rewrite the last paragraph to make it clearer / more formal.}
 
@@
 
- \cba{petite intro avec plan de section}
+ \cba{petite intro avec plan de section, context was moved before.}
 
@@
 
-- Pure Smalltalk: baseline
+ \cba{Maybe wrte version A, B, C , D , E or a word to represent what they are (Smalltalk) (SmartSyntax) (Plugin) (Slang) (Slang+JIT)
+ so you can refer to that to talk about any of these later and not wirte each time : the implem with ths wirtten that wya and exec that way.
+ Myabe those could be paragraph titles instead of the following}
+
+\paragraph{Baseline: pure Smalltalk.}
+: baseline
 Smalltalk code on top of VM. Used as our baseline for benchmarks.
@@
 
-- Smart Syntax Plugin
+New spec
+
+\paragraph{Existing: Smart Syntax Plugin.}
 %CHECK how to import code env in latex.
+%Note we're talking about comparison, not =, so not sure we should show = or bench =
 %\begin{code}
@@
 Compiled to C AOT into a plugin.
+Write once, fall-back + compiled.
 
-- Slang plugin 
+old spec
+
+\paragraph{Existing: Slang Plugin.}
 Slang, but no access au fonctions de la VM drectement.
@@
 
-- Slang pure
+ old spec
 
-- Slang pure + Cog RTL
+\paragraph{Pure Slang.}
+
+new spec
+
+\paragraph{Pure Slang with JIT support.}
+Cog RTL
 
@@
 
+new spec
+
  \subsection{Pros and cons}
@@
 
-Baseline performance is not that great, jtted performance is good only with mature optimizing JIT we don't have.
+Baseline performance is not that great.
 
-Number of implementations: 1, 2 or 3: same implem for primitive in C and fall back code in Smalltalk, relatively easy to read/code.
+Number of implementations to maintain: 1, 2 or 3: same implem for primitive in C and fall back code in Smalltalk, relatively easy to read/code.
 recompilation. 
@@
 
+Il faut expliquer en une phrase: 
+petit string = cout important C-St stack switch.
+Gde string = cout de la boucle.
+
 \cba{Show ASM loops and discuss the differences
@@
 \subsection{Macro-benchmarks}
+Intro sentence with ref to squeak speed center.
 JSON
 
-\section{Related work}
+\section{Discussion and related work}
+
+\cba{C'est un peu de la discussion dans le sens que le related work c'est que celui de Tim, le reste c'est plus des solutions pour le pb du stack switch, qui n'est pas le problème résolu dans ce papier}
 

### 2018-05-07 paper.tex
related work + section 2finish

@@
 \usepackage{ifthen}
+\usepackage{pgfplots}
+\usepgfplotslibrary{statistics}
 %\input{macros.tex}
@@
 
-\cba{TODO!}
-To solve this performance bottleneck, the Cog VM allows to redefine primitives on top of its back-end. 
-integrated as part of the native code generated.
-No switch cost.
-No need to reimpl long running code -> direct in Abs assembly, hard to write and maintain, have to be smart to compete with optimized C code.
-=> often partial impl for quick to perform prim, for ex, add of 2 int, if overflow just go back to C prim.
+To solve this performance bottleneck, the Cog VM allows to redefine primitives in Cog's Register Transfer Language (RTL), an abstract assembly which is compiled to native code through Cog's JIT back-end.
+Such primitives are then generated to native code at runtime when a method annotated with the primitive is compiled to native code.
+This allows the primitive to be performed in the client stack with the client calling convention, saving register moves and the stack switch dance.
+Primitives redefined in such way can be entirely or partially redefined.
+In the later case, only the common cases are re-implemented on top of the JIT back-end, and an unimplemented case happens at runtime, the generated native code calls the C primitive instead, adding overhead only in uncommon cases.
 
@@
 Write down the Smalltalk code here.
-Expliquer to:do: over do: (to explain expresiveness) => core lib often use to:do: and not do:
-
-New spec
-
-\paragraph{Existing: Smart Syntax Plugin.}
 %CHECK how to import code env in latex.
@@
 %\end{code}
+Expliquer to:do: over do: (to explain expresiveness) => core lib often use to:do: and not do:
+
+New spec
+
+\paragraph{Existing: Smart Syntax Plugin.}
 Restricted Smalltalk
@@
 
+taille 0
+
+\begin{tikzpicture}
+  \begin{axis}
+  [
+boxplot/draw direction
+=y,
+    xtick={1,2,3},
+    xticklabels={Plugin, Slang, Slang+JIT},
+    ]
+    \addplot+[
+    boxplot prepared={
+      median=2.942,
+      upper quartile=2.977,
+      lower quartile=2.911,
+      upper whisker=3.098,
+      lower whisker=2.797
+    },
+    ] coordinates {};
+    \addplot+[
+    boxplot prepared={
+      median=0.9594,
+      upper quartile=0.9791,
+      lower quartile=0.9230,
+      upper whisker=1.1059,
+      lower whisker=0.8705
+    },
+    ] coordinates {};
+    \addplot+[
+    boxplot prepared={
+      median=0.4258,
+      upper quartile=0.4310,
+      lower quartile=0.4227,
+      upper whisker=0.5181,
+      lower whisker=0.4074
+    },
+    ] coordinates {};
+  \end{axis}
+\end{tikzpicture}
+
+taille 5
+
+\begin{tikzpicture}
+  \begin{axis}
+  [
+boxplot/draw direction=y,
+    xtick={1,2,3},
+    xticklabels={Plugin, Slang, Slang+JIT},
+    ]
+    \addplot+[
+    boxplot prepared={
+      median=0.6904,
+      upper quartile=0.6998,
+      lower quartile=0.6835,
+      upper whisker=0.7353,
+      lower whisker=0.6407
+    },
+    ] coordinates {};
+    \addplot+[
+    boxplot prepared={
+      median=0.2530,
+      upper quartile=0.2594,
+      lower quartile=0.2449,
+      upper whisker=0.2836,
+      lower whisker=0.2323
+    },
+    ] coordinates {};
+    \addplot+[
+    boxplot prepared={
+      median=0.1455,
+      upper quartile=0.1475,
+      lower quartile=0.1425,
+      upper whisker=0.1636,
+      lower whisker=0.1266
+    },
+    ] coordinates {};
+  \end{axis}
+\end{tikzpicture}
+
+taille 100
+\begin{tikzpicture}
+  \begin{axis}
+  [
+boxplot/draw direction=y,
+    xtick={1,2,3},
+    xticklabels={Plugin, Slang, Slang+JIT},
+    ]
+    \addplot+[
+    boxplot prepared={
+      median=0.10378,
+      upper quartile=0.10435,
+      lower quartile=0.10355,
+      upper whisker=0.10586,
+      lower whisker=0.09964
+    },
+    ] coordinates {};
+    \addplot+[
+    boxplot prepared={
+      median=0.05689,
+      upper quartile=0.05706,
+      lower quartile=0.05648,
+      upper whisker=0.05985,
+      lower whisker=0.05527
+    },
+    ] coordinates {};
+    \addplot+[
+    boxplot prepared={
+      median=0.05564,
+      upper quartile=0.05560,
+      lower quartile=0.05552,
+      upper whisker=0.05985,
+      lower whisker=0.05382
+    },
+    ] coordinates {};
+  \end{axis}
+\end{tikzpicture}
+
+taille 10000
+\begin{tikzpicture}
+  \begin{axis}
+  [
+boxplot/draw direction
+=y,
+    xtick={1,2,3, 4},
+    xticklabels={Baseline, Plugin, Slang, Slang+JIT},
+    ]
+    \addplot+[
+    boxplot prepared={
+      median=10474,
+      upper quartile=10506,
+      lower quartile=10455,
+      upper whisker=11905,
+      lower whisker=10099
+    },
+    ] coordinates {};
+    \addplot+[
+    boxplot prepared={
+      median=761,
+      upper quartile=761,
+      lower quartile=761,
+      upper whisker=765,
+      lower whisker=760
+    },
+    ] coordinates {};
+    \addplot+[
+    boxplot prepared={
+      median=476.7,
+      upper quartile=477,
+      lower quartile=476,
+      upper whisker=482,
+      lower whisker=475
+    },
+    ] coordinates {};
+    \addplot+[
+    boxplot prepared={
+      median=500,
+      upper quartile=501.8,
+      lower quartile=500,
+      upper whisker=512,
+      lower whisker=499
+    },
+    ] coordinates {};
+  \end{axis}
+\end{tikzpicture}
+
 \cba{Je pense string de taille 0-1 c'est cool pour montrer que le code C va moins vite que pur smalltalk dans ce cas (si c'est vraiment le cas)}
@@
 
-\cba{C'est un peu de la discussion dans le sens que le related work c'est que celui de Tim, le reste c'est plus des solutions pour le pb du stack switch, qui n'est pas le problème résolu dans ce papier}
+\cba{Maybe talk in those cases the number of representation of the primitives}
 
-Work on reducing the number of prim
+We identified two directly related works.
+In some languages (Strongtalk, Self, RSqueak), the implementors tried to decrease the overall number of optional primitives by improving the client language performance, reducing the need for such primitives \cite{Strongtalk, ursPHD, NoPrimTracing}.
+In addition, in the V8 project  \cite{V8}, the interpreter was entirely built on top of the V8 JIT's back-end using a specific DSL similar to Cog's RTL.
 
-RPython \cite{NoPrimTracing}
+\subsection{Decreasing the number of primitives}
 
-Self - Strongtalk -> tout en Smalltalk avec mature optimizing JIT. pb baseline perf, haut cout de maintenance implem pour l'adaptive optimizer \cite{Strongtalk, ursPHD}
+The philosophy of Self and Strongtalk, both high performance VMs, were to improve the performance of the client language with a mature optimizing JIT featuring advanced optimizations such as speculative optimizations \cite{Strongtalk, ursPHD}.
+In this context, the need of primitive is reduced, since the performance gain from using a primitive over the client language is low or inexistant.
+More recently, Felgentreff \emph{et al.} \cite{NoPrimTracing} analysed the performance difference between primitives implemented in C, RPython, and Smalltalk.
+They showed that with the RPython toolchain framework, the performance difference between all three implementation was not that big, hence they could choose to implement some optional primitives in the client language over C or RPython without performance drop to reduce the number of primitives to maintain.
 
-	%\item Java: platform dependent
-	 V8: interpreter compiled by JIT \cite{V8} Mention Ignition.
-	optimised backend
-	as if we were compiling slang using the JIT compiler (ahead of time)
-	no more virtual call
+These approaches have the advantage of decreasing the primitive maintenance cost since fewer primitives have to be maintained.
+The performance problem is however partially solved: only the peak performance is similar with and without the primitives.
+Since those system rely on JITs, start-up performance is worse than peak performance, increasing the performance difference at start-up between the client code and the primitive.
+In addition, the performance can be  unreliable with optimizing JITs.
+The implementors of Morphic \cite{MorphicSelf} on top of the Self VM were complaining that changing a single line of code could lead to important performance drop and they would not understand why.
+Those performance drops were due to the optimizing JIT taking different optimization decisions.
+This problem was so significant that recently the V8 team \cite{V8} rewrote their primitives written in Javascript (named built-ins in their project) in a DSL compiled ahead of time through their JIT back-end to improve the performance of the primitives and therefore decrease the performance drop between baseline and peak performance.
+Aside from the start-up performance and the performance unreliability problems described, these approaches also require a mature optimizing JIT, hard to implement and maintain.
 
- VisualWorks: baseline JIT but no interpreter => only 1 implement to write and maintain.
+\subsection{No stack switch overhead}
 
-\cba{Ajoute le related work que tu as trouvé de ICOOOLPS avec le travail de Tim.}
+To avoid paying the stack switch cost between the client Stack and the C stack, the V8 team recently wrote an interpreter Ignition in a DSL compiled ahead of time through their JIT back-end to native code.
+With their primtives and their interpreter written in the DSL, the V8 runtime rarely needs to switch between the client and the C stack (no need to switch to interpret code, no need to switch for primitives).
+In this context, the stack switch overhead is not a problem anymore.
+
+In our context, implementing a similar approach would require us to compile Slang to native code through our JIT compiler back-end or to rewrite the interpreter on top of Cog JIT's RTL.
+Slang has some abstractions over native code, similarly to C.
+Compiling Slang ahead of time to native code efficiently requires us to re-implement optimizations implemented in C compilers and we don't want that complexity or to lower the abstractions level of Slang to express native code optimizations, which would increase the complexity of Slang, which we don't want either.
+Rewriting the interpreter on top of Cog JIT's RTL is possible but requires a significant amount of work.
 

### 2018-05-07 paper.tex
add missing DOI
added figure * for bench
add gcc boxplots - all versions. the display will be improved

@@
 
-%%%%%
-% END OLD
-%%%%%
-
 \section{Evaluation}
@@
 
-taille 0
+\cba{Personnellement je ferais une figure, ou tu places tes graphes dedans, et pares tu referes a la figure dans le texte.}
+
+\begin{figure*}
+  \centering
+
+    \begin{subfigure}[b]{.45\textwidth}
 
@@
     =y,
-    xtick={1,2,3},
-    xticklabels={Plugin, Slang, Slang+JIT},
+        xtick={1,2,3,4,5},
+        xticklabels={Baseline, oldPlugin, SlangPlugin, PureSlang, Slang+JIT},
+        font={\tiny}
         ]
@@
         boxplot prepared={
-      median=2.942,
-      upper quartile=2.977,
-      lower quartile=2.911,
-      upper whisker=3.098,
-      lower whisker=2.797
+          median=1,
+          upper quartile=1,
+          lower quartile=1,
+          upper whisker=1,
+          lower whisker=1
+        },        ] coordinates {};
+        \addplot+[
+        boxplot prepared={
+          median=0.3413,
+          upper quartile=0.3421,
+          lower quartile=0.3398,
+          upper whisker=0.3559,
+          lower whisker=0.3351
+        },        ] coordinates {};
+        \addplot+[
+        boxplot prepared={
+          median=0.3399,
+          upper quartile=0.3409,
+          lower quartile=0.3393,
+          upper whisker=0.3569,
+          lower whisker=0.3140
         },
@@
         boxplot prepared={
-      median=0.9594,
-      upper quartile=0.9791,
-      lower quartile=0.9230,
-      upper whisker=1.1059,
-      lower whisker=0.8705
+          median=1.089,
+          upper quartile=1.090,
+          lower quartile=1.084,
+          upper whisker=1.143,
+          lower whisker=1.016
         },
@@
         boxplot prepared={
-      median=0.4258,
-      upper quartile=0.4310,
-      lower quartile=0.4227,
-      upper whisker=0.5181,
-      lower whisker=0.4074
+          median=2.486,
+          upper quartile=2.488,
+          lower quartile=2.483,
+          upper whisker=2.612,
+          lower whisker=2.477
         },
@@
 
-taille 5
+    \subcaption{Taille 0}
+    \end{subfigure}
+    \begin{subfigure}[b]{.45\textwidth}
 
@@
     boxplot/draw direction=y,
-    xtick={1,2,3},
-    xticklabels={Plugin, Slang, Slang+JIT},
+        xtick={1,2,3,4,5},
+        xticklabels={Baseline, oldPlugin, SlangPlugin, PureSlang, Slang+JIT},
+        font={\tiny}
         ]
@@
         boxplot prepared={
-      median=0.6904,
-      upper quartile=0.6998,
-      lower quartile=0.6835,
-      upper whisker=0.7353,
-      lower whisker=0.6407
+          median=1,
+          upper quartile=1,
+          lower quartile=1,
+          upper whisker=1,
+          lower whisker=1
         },
@@
         boxplot prepared={
-      median=0.2530,
-      upper quartile=0.2594,
-      lower quartile=0.2449,
-      upper whisker=0.2836,
-      lower whisker=0.2323
+          median=1.394,
+          upper quartile=1.396,
+          lower quartile=1.388,
+          upper whisker=1.416,
+          lower whisker=1.332
         },
@@
         boxplot prepared={
-      median=0.1455,
-      upper quartile=0.1475,
-      lower quartile=0.1425,
-      upper whisker=0.1636,
-      lower whisker=0.1266
+          median=1.361,
+          upper quartile=1.366,
+          lower quartile=1.358,
+          upper whisker=1.380,
+          lower whisker=1.007
+        },
+        ] coordinates {};
+                \addplot+[
+        boxplot prepared={
+          median=3.887,
+          upper quartile=3.887,
+          lower quartile=3.884,
+          upper whisker=3.942,
+          lower whisker=2.294
+        },
+        ] coordinates {};
+                \addplot+[
+        boxplot prepared={
+          median=6.716,
+          upper quartile=6.744,
+          lower quartile=6.712,
+          upper whisker=6.970,
+          lower whisker=6.676
         },
@@
 
-taille 100
+    \subcaption{Taille 5}
+    \end{subfigure}
+    \begin{subfigure}[b]{.45\textwidth}
+
     \begin{tikzpicture}
@@
         boxplot prepared={
-      median=0.10378,
-      upper quartile=0.10435,
-      lower quartile=0.10355,
-      upper whisker=0.10586,
-      lower whisker=0.09964
+          median=1,
+          upper quartile=1,
+          lower quartile=1,
+          upper whisker=1,
+          lower whisker=1
         },
@@
         boxplot prepared={
-      median=0.05689,
-      upper quartile=0.05706,
-      lower quartile=0.05648,
-      upper whisker=0.05985,
-      lower whisker=0.05527
+          median=8.569,
+          upper quartile=8.602,
+          lower quartile=8.562,
+          upper whisker=9.671,
+          lower whisker=8.510
         },
@@
         boxplot prepared={
-      median=0.05564,
-      upper quartile=0.05560,
-      lower quartile=0.05552,
-      upper whisker=0.05985,
-      lower whisker=0.05382
+          median=8.938,
+          upper quartile=8.967,
+          lower quartile=8.935,
+          upper whisker=10.022,
+          lower whisker=6.044
+        },
+        ] coordinates {};
+        \addplot+[
+        boxplot prepared={
+          median=16.22,
+          upper quartile=16.24,
+          lower quartile=16.21,
+          upper whisker=18.20,
+          lower whisker=10.40
+        },
+        ] coordinates {};
+        \addplot+[
+        boxplot prepared={
+          median=16.66,
+          upper quartile=16.68,
+          lower quartile=16.44,
+          upper whisker=18.44,
+          lower whisker=16.43
         },
@@
 
-taille 10000
+    \subcaption{Taille 100}
+    \end{subfigure}
+    \begin{subfigure}[b]{.45\textwidth}
+
     \begin{tikzpicture}
@@
     =y,
-    xtick={1,2,3, 4},
-    xticklabels={Baseline, Plugin, Slang, Slang+JIT},
+        xtick={1,2,3,4,5},
+        xticklabels={Baseline, oldPlugin, SlangPlugin, PureSlang, Slang+JIT},
+        font={\tiny}
         ]
@@
         boxplot prepared={
-      median=10474,
-      upper quartile=10506,
-      lower quartile=10455,
-      upper whisker=11905,
-      lower whisker=10099
+          median=1,
+          upper quartile=1,
+          lower quartile=1,
+          upper whisker=1,
+          lower whisker=1
         },
@@
         boxplot prepared={
-      median=761,
-      upper quartile=761,
-      lower quartile=761,
-      upper whisker=765,
-      lower whisker=760
+          median=13.50,
+          upper quartile=13.55,
+          lower quartile=13.46,
+          upper whisker=14.42,
+          lower whisker=12.90
         },
@@
         boxplot prepared={
-      median=476.7,
-      upper quartile=477,
-      lower quartile=476,
-      upper whisker=482,
-      lower whisker=475
+          median=14.76,
+          upper quartile=14.82,
+          lower quartile=14.76,
+          upper whisker=15.71,
+          lower whisker=10.83
         },
@@
         boxplot prepared={
-      median=500,
-      upper quartile=501.8,
-      lower quartile=500,
-      upper whisker=512,
-      lower whisker=499
+          median=24.39,
+          upper quartile=24.41,
+          lower quartile=24.26,
+          upper whisker=25.98,
+          lower whisker=19.77
+        },
+        ] coordinates {};
+        \addplot+[
+        boxplot prepared={
+          median=24.25,
+          upper quartile=24.30,
+          lower quartile=24.25,
+          upper whisker=25.82,
+          lower whisker=24.08
         },
@@
     \end{tikzpicture}
+
+   \subcaption{Taille 1000}
+   \end{subfigure}
+   \caption{GCC-Linux results}
+\end{figure*}
 
@@
 \cba{Inlining in Sista + ref}
+\cite{SistaFastStartUp}
 

### 2018-05-07 paper.tex
llvm boxplot. display might change as well

@@
 
-    \subcaption{Taille 0}
+    \subcaption{Size 0}
     \end{subfigure}
@@
 
-    \subcaption{Taille 5}
+    \subcaption{Size 5}
     \end{subfigure}
@@
 
-    \subcaption{Taille 100}
+    \subcaption{Size 100}
     \end{subfigure}
@@
 
-   \subcaption{Taille 1000}
+   \subcaption{Size 1000}
    \end{subfigure}
    \caption{GCC-Linux results}
+\end{figure*}
+
+\begin{figure*}
+  \centering
+
+    \begin{subfigure}[b]{.45\textwidth}
+
+    \begin{tikzpicture}
+      \begin{axis}
+      [
+    boxplot/draw direction
+    =y,
+        xtick={1,2,3,4,5},
+        xticklabels={Baseline, oldPlugin, SlangPlugin, PureSlang, Slang+JIT},
+        font={\tiny}
+        ]
+        \addplot+[
+        boxplot prepared={
+          median=1,
+          upper quartile=1,
+          lower quartile=1,
+          upper whisker=1,
+          lower whisker=1
+        },        ] coordinates {};
+        \addplot+[
+        boxplot prepared={
+          median=0.3162,
+          upper quartile=0.3179,
+          lower quartile=0.3072,
+          upper whisker=0.3720,
+          lower whisker=0.3022
+        },        ] coordinates {};
+        \addplot+[
+        boxplot prepared={
+          median=0.3023,
+          upper quartile=0.3029,
+          lower quartile=0.3006,
+          upper whisker=0.3547,
+          lower whisker=0.2984
+        },
+        ] coordinates {};
+        \addplot+[
+        boxplot prepared={
+          median=0.8538,
+          upper quartile=0.8591,
+          lower quartile=0.8482,
+          upper whisker=0.9920,
+          lower whisker=0.7516
+        },
+        ] coordinates {};
+        \addplot+[
+        boxplot prepared={
+          median=2.142,
+          upper quartile=2.429,
+          lower quartile=2.120,
+          upper whisker=2.990,
+          lower whisker=2.038
+        },
+        ] coordinates {};
+      \end{axis}
+    \end{tikzpicture}
+
+    \subcaption{Size 0}
+    \end{subfigure}
+    \begin{subfigure}[b]{.45\textwidth}
+
+    \begin{tikzpicture}
+      \begin{axis}
+      [
+    boxplot/draw direction=y,
+        xtick={1,2,3,4,5},
+        xticklabels={Baseline, oldPlugin, SlangPlugin, PureSlang, Slang+JIT},
+        font={\tiny}
+        ]
+        \addplot+[
+        boxplot prepared={
+          median=1,
+          upper quartile=1,
+          lower quartile=1,
+          upper whisker=1,
+          lower whisker=1
+        },
+        ] coordinates {};
+        \addplot+[
+        boxplot prepared={
+          median=1.261,
+          upper quartile=1.270,
+          lower quartile=1.227,
+          upper whisker=1.282,
+          lower whisker=1.205
+        },
+        ] coordinates {};
+        \addplot+[
+        boxplot prepared={
+          median=1.229,
+          upper quartile=1.235,
+          lower quartile=1.227,
+          upper whisker=1.248,
+          lower whisker=1.203
+        },
+        ] coordinates {};
+                \addplot+[
+        boxplot prepared={
+          median=2.832,
+          upper quartile=2.847,
+          lower quartile=2.814,
+          upper whisker=2.875,
+          lower whisker=2.706
+        },
+        ] coordinates {};
+                \addplot+[
+        boxplot prepared={
+          median=7.094,
+          upper quartile=7.138,
+          lower quartile=7.057,
+          upper whisker=7.212,
+          lower whisker=6.854
+        },
+        ] coordinates {};
+      \end{axis}
+    \end{tikzpicture}
+
+    \subcaption{Size 5}
+    \end{subfigure}
+    \begin{subfigure}[b]{.45\textwidth}
+
+    \begin{tikzpicture}
+      \begin{axis}
+      [
+    boxplot/draw direction=y,
+        xtick={1,2,3,4,5},
+        xticklabels={Baseline, oldPlugin, SlangPlugin, PureSlang, Slang+JIT},
+        font={\tiny}
+        ]
+        \addplot+[
+        boxplot prepared={
+          median=1,
+          upper quartile=1,
+          lower quartile=1,
+          upper whisker=1,
+          lower whisker=1
+        },
+        ] coordinates {};
+         \addplot+[
+        boxplot prepared={
+          median=6.746,
+          upper quartile=6.826,
+          lower quartile=6.622,
+          upper whisker=7.384,
+          lower whisker=5.307
+        },
+        ] coordinates {};
+        \addplot+[
+        boxplot prepared={
+          median=6.581,
+          upper quartile=6.626,
+          lower quartile=6.553,
+          upper whisker=7.087,
+          lower whisker=6.362
+        },
+        ] coordinates {};
+        \addplot+[
+        boxplot prepared={
+          median=10.330,
+          upper quartile=10.122,
+          lower quartile=10.461,
+          upper whisker=11.110,
+          lower whisker=9.246
+        },
+        ] coordinates {};
+        \addplot+[
+        boxplot prepared={
+          median=18.08,
+          upper quartile=18.16,
+          lower quartile=17.79,
+          upper whisker=19.53,
+          lower whisker=16.36
+        },
+        ] coordinates {};
+      \end{axis}
+    \end{tikzpicture}
+
+    \subcaption{Size 100}
+    \end{subfigure}
+    \begin{subfigure}[b]{.45\textwidth}
+
+    \begin{tikzpicture}
+      \begin{axis}
+      [
+    boxplot/draw direction
+    =y,
+        xtick={1,2,3,4,5},
+        xticklabels={Baseline, oldPlugin, SlangPlugin, PureSlang, Slang+JIT},
+        font={\tiny}
+        ]
+        \addplot+[
+        boxplot prepared={
+          median=1,
+          upper quartile=1,
+          lower quartile=1,
+          upper whisker=1,
+          lower whisker=1
+        },
+        ] coordinates {};
+        \addplot+[
+        boxplot prepared={
+          median=9.485,
+          upper quartile=9.521,
+          lower quartile=9.314,
+          upper whisker=9.720,
+          lower whisker=8.831
+        },
+        ] coordinates {};
+        \addplot+[
+        boxplot prepared={
+          median=9.212,
+          upper quartile=9.271,
+          lower quartile=9.173,
+          upper whisker=9.364,
+          lower whisker=8.979
+        },
+        ] coordinates {};
+        \addplot+[
+        boxplot prepared={
+          median=13.48,
+          upper quartile=13.61,
+          lower quartile=13.33,
+          upper whisker=13.70,
+          lower whisker=12.81
+        },
+        ] coordinates {};
+        \addplot+[
+        boxplot prepared={
+          median=26.54,
+          upper quartile=26.80,
+          lower quartile=26.07,
+          upper whisker=27.12,
+          lower whisker=25.53
+        },
+        ] coordinates {};
+      \end{axis}
+    \end{tikzpicture}
+
+   \subcaption{Size 1000}
+   \end{subfigure}
+   \caption{LLVM-Mac results}
 \end{figure*}

### 2018-05-08 paper.tex
proper plots for bench results

--- 2018-05-07 paper.tex
@@
       [
-boxplot/draw direction
-=y,
-    xtick={1,2,3},
-    xticklabels={Plugin, Slang, Slang+JIT},
+        xtick={1,...,5},
+        xticklabels={Baseline, oldPlugin, SlangPlugin, PureSlang, Slang+JIT},
+        %x tick label as interval,
+        font={\tiny},
+		ybar=10pt,
+		xbar width=0.30,
+		no markers,Baseline/.style={red},2/.style={green},3/.style={blue}
+		%cycle list={red, orange, green, purple, black}
         ]
-    \addplot+[
-    boxplot prepared={
-      median=2.942,
-      upper quartile=2.977,
-      lower quartile=2.911,
-      upper whisker=3.098,
-      lower whisker=2.797
-    },
-    ] coordinates {};
-    \addplot+[
-    boxplot prepared={
-      median=0.9594,
-      upper quartile=0.9791,
-      lower quartile=0.9230,
-      upper whisker=1.1059,
-      lower whisker=0.8705
-    },
-    ] coordinates {};
-    \addplot+[
-    boxplot prepared={
-      median=0.4258,
-      upper quartile=0.4310,
-      lower quartile=0.4227,
-      upper whisker=0.5181,
-      lower whisker=0.4074
-    },
-    ] coordinates {};
+\addplot[
+    fill=black!25,
+    %draw=none,
+    %fill,
+    draw=black,
+    fill,
+    point meta=y,
+    every node near coord/.style={inner ysep=5pt},
+    error bars/.cd,
+        y dir=both,
+        y explicit
+]  
+table [y error=error] {
+x   y           error    
+1   1           0 
+2   0.3147064   0.002288488 
+3   0.3035965   0.001786185
+4   0.8545431   0.006132950 
+5   2.250732    0.04133638 
+};            
+      \end{axis}
+    \end{tikzpicture}
+    \subcaption{Size 0}
+    \end{subfigure}
+    \begin{subfigure}[b]{.45\textwidth}
+    \begin{tikzpicture}
+      \begin{axis}
+      [
+        xtick={1,...,5},
+        xticklabels={Baseline, oldPlugin, SlangPlugin, PureSlang, Slang+JIT},
+        %x tick label as interval,
+        font={\tiny},
+		ybar=10pt,
+		xbar width=0.30
+        ]
+\addplot[
+    fill=black!25,
+    draw=black,
+    point meta=y,
+    every node near coord/.style={inner ysep=5pt},
+    error bars/.cd,
+        y dir=both,
+        y explicit
+] 
+table [y error=error] {
+x   y           error    
+1   1           0 
+2   1.2505193   0.004611354 
+3   1.2295801   0.001810171
+4   2.8205415   0.008035024 
+5   7.089806    0.01403785 
+};            
       \end{axis}
@@
 
-taille 5
+    \subcaption{Size 5}
+    \end{subfigure}
+    \begin{subfigure}[b]{.45\textwidth}
+ \begin{tikzpicture}
+      \begin{axis}
+      [
+        xtick={1,...,5},
+        xticklabels={Baseline, oldPlugin, SlangPlugin, PureSlang, Slang+JIT},
+        %x tick label as interval,
+        font={\tiny},
+		ybar=10pt,
+		ymin=0,
+		xbar width=0.30
+        ]
+\addplot[
+    fill=black!25,
+    draw=black,
+    point meta=y,
+    every node near coord/.style={inner ysep=5pt},
+    error bars/.cd,
+        y dir=both,
+        y explicit
+] 
+table [y error=error] {
+x   y           error    
+1   1           0 
+2   6.6951738   0.062391122 
+3   6.6032572   0.024774416
+4   10.3029394   0.061481532 
+5   17.979611    0.10127936 
+};            
+      \end{axis}
+    \end{tikzpicture}
+
+    \subcaption{Size 100}
+    \end{subfigure}
+    \begin{subfigure}[b]{.45\textwidth}
 
@@
       [
-boxplot/draw direction=y,
-    xtick={1,2,3},
-    xticklabels={Plugin, Slang, Slang+JIT},
+        xtick={1,...,5},
+        xticklabels={Baseline, oldPlugin, SlangPlugin, PureSlang, Slang+JIT},
+        %x tick label as interval,
+        font={\tiny},
+		ybar=10pt,
+		ymin=0,
+		xbar width=0.30
         ]
-    \addplot+[
-    boxplot prepared={
-      median=0.6904,
-      upper quartile=0.6998,
-      lower quartile=0.6835,
-      upper whisker=0.7353,
-      lower whisker=0.6407
-    },
-    ] coordinates {};
-    \addplot+[
-    boxplot prepared={
-      median=0.2530,
-      upper quartile=0.2594,
-      lower quartile=0.2449,
-      upper whisker=0.2836,
-      lower whisker=0.2323
-    },
-    ] coordinates {};
-    \addplot+[
-    boxplot prepared={
-      median=0.1455,
-      upper quartile=0.1475,
-      lower quartile=0.1425,
-      upper whisker=0.1636,
-      lower whisker=0.1266
-    },
-    ] coordinates {};
+\addplot[
+    fill=black!25,
+    draw=black,
+    point meta=y,
+    every node near coord/.style={inner ysep=5pt},
+    error bars/.cd,
+        y dir=both,
+        y explicit
+] 
+table [y error=error] {
+x   y           error    
+1   1           0 
+2   9.4064534   0.036698187 
+3   9.2138415   0.013982843
+4   13.4384830   0.040202717 
+5   26.404792    0.09298324 
+};            
       \end{axis}
@@
 
-taille 100
+   \subcaption{Size 1000}
+   \end{subfigure}
+   \caption{GCC-Linux results}
+\end{figure*}
+
+\begin{figure*}
+  \centering
+
+    \begin{subfigure}[b]{.45\textwidth}
+
     \begin{tikzpicture}
@@
       [
-boxplot/draw direction=y,
-    xtick={1,2,3},
-    xticklabels={Plugin, Slang, Slang+JIT},
+        xtick={1,...,5},
+        xticklabels={Baseline, oldPlugin, SlangPlugin, PureSlang, Slang+JIT},
+        %x tick label as interval,
+        font={\tiny},
+		ybar=10pt,
+		ymin=0,
+		xbar width=0.30
         ]
-    \addplot+[
-    boxplot prepared={
-      median=0.10378,
-      upper quartile=0.10435,
-      lower quartile=0.10355,
-      upper whisker=0.10586,
-      lower whisker=0.09964
-    },
-    ] coordinates {};
-    \addplot+[
-    boxplot prepared={
-      median=0.05689,
-      upper quartile=0.05706,
-      lower quartile=0.05648,
-      upper whisker=0.05985,
-      lower whisker=0.05527
-    },
-    ] coordinates {};
-    \addplot+[
-    boxplot prepared={
-      median=0.05564,
-      upper quartile=0.05560,
-      lower quartile=0.05552,
-      upper whisker=0.05985,
-      lower whisker=0.05382
-    },
-    ] coordinates {};
+\addplot[
+    fill=black!25,
+    draw=black,
+    point meta=y,
+    every node near coord/.style={inner ysep=5pt},
+    error bars/.cd,
+        y dir=both,
+        y explicit
+] 
+table [y error=error] {
+x   y           error    
+1   1           0 
+2   0.3147064   0.002288488 
+3   0.3035965   0.001786185
+4   0.8545431   0.006132950 
+5   2.250732    0.04133638 
+};            
       \end{axis}
@@
 
-taille 10000
+    \subcaption{Size 0}
+    \end{subfigure}
+    \begin{subfigure}[b]{.45\textwidth}
+
     \begin{tikzpicture}
@@
       [
-boxplot/draw direction
-=y,
-    xtick={1,2,3, 4},
-    xticklabels={Baseline, Plugin, Slang, Slang+JIT},
+        xtick={1,...,5},
+        xticklabels={Baseline, oldPlugin, SlangPlugin, PureSlang, Slang+JIT},
+        %x tick label as interval,
+        font={\tiny},
+		ybar=10pt,
+		xbar width=0.30
         ]
-    \addplot+[
-    boxplot prepared={
-      median=10474,
-      upper quartile=10506,
-      lower quartile=10455,
-      upper whisker=11905,
-      lower whisker=10099
-    },
-    ] coordinates {};
-    \addplot+[
-    boxplot prepared={
-      median=761,
-      upper quartile=761,
-      lower quartile=761,
-      upper whisker=765,
-      lower whisker=760
-    },
-    ] coordinates {};
-    \addplot+[
-    boxplot prepared={
-      median=476.7,
-      upper quartile=477,
-      lower quartile=476,
-      upper whisker=482,
-      lower whisker=475
-    },
-    ] coordinates {};
-    \addplot+[
-    boxplot prepared={
-      median=500,
-      upper quartile=501.8,
-      lower quartile=500,
-      upper whisker=512,
-      lower whisker=499
-    },
-    ] coordinates {};
+\addplot[
+    fill=black!25,
+    draw=black,
+    point meta=y,
+    every node near coord/.style={inner ysep=5pt},
+    error bars/.cd,
+        y dir=both,
+        y explicit
+] 
+table [y error=error] {
+x   y           error    
+1   1           0 
+2   1.2505193   0.004611354 
+3   1.2295801   0.001810171
+4   2.8205415   0.008035024 
+5   7.089806    0.01403785 
+};            
       \end{axis}
     \end{tikzpicture}
+    \subcaption{Size 5}
+    \end{subfigure}
+    \begin{subfigure}[b]{.45\textwidth}
+ \begin{tikzpicture}
+      \begin{axis}
+      [
+        xtick={1,...,5},
+        xticklabels={Baseline, oldPlugin, SlangPlugin, PureSlang, Slang+JIT},
+        %x tick label as interval,
+        font={\tiny},
+		ybar=10pt,
+		ymin=0,
+		xbar width=0.30
+        ]
+\addplot[
+    fill=black!25,
+    draw=black,
+    point meta=y,
+    every node near coord/.style={inner ysep=5pt},
+    error bars/.cd,
+        y dir=both,
+        y explicit
+] 
+table [y error=error] {
+x   y           error    
+1   1           0 
+2   6.6951738   0.062391122 
+3   6.6032572   0.024774416
+4   10.3029394  0.061481532 
+5   17.979611   0.10127936 
+};            
+      \end{axis}
+    \end{tikzpicture}
+    \subcaption{Size 100}
+    \end{subfigure}
+    \begin{subfigure}[b]{.45\textwidth}
+
+    \begin{tikzpicture}
+      \begin{axis}
+      [
+        xtick={1,...,5},
+        xticklabels={Baseline, oldPlugin, SlangPlugin, PureSlang, Slang+JIT},
+        %x tick label as interval,
+        font={\tiny},
+		ybar=10pt,
+		ymin=0,
+		xbar width=0.30
+        ]
+\addplot[
+    fill=black!25,
+    draw=black,
+    point meta=y,
+    every node near coord/.style={inner ysep=5pt},
+    error bars/.cd,
+        y dir=both,
+        y explicit
+] 
+table [y error=error] {
+x   y           error    
+1   1           0 
+2   9.4064534   0.036698187 
+3   9.2138415   0.013982843
+4   13.4384830  0.040202717 
+5   26.404792   0.09298324 
+};            
+      \end{axis}
+    \end{tikzpicture}
+
+   \subcaption{Size 1000}
+   \end{subfigure}
+   \caption{LLVM-Mac results}
+\end{figure*}
 

### 2018-05-08 paper.tex
update plots, micro bench parts + start asm explanation

@@
 \usepackage{pgfplots}
+\usepackage{listings}
 \usepgfplotslibrary{statistics}
@@
 \newcommand{\cba}[1]{\color{purple}\fbox{\bfseries\sffamily\scriptsize Clement:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
+\newcommand*\rot{\rotatebox{90}}
+\input{macros}
 
@@
 Their goal was to improve the execution of the client language with a JIT performing speculative optimizations \cite{ursPHD} for the execution of the optional primitives implemented in the client language to be as fast as the performance of the same primitives written in the VM.
-Self \cite{ursPHD} and Strongtalk \cite{Strongtalk} were to first to try this approach.
-Early versions of the Javascript engine V8 \cite{V8} also implemented most of the primitives, including for example Array operations, in Javascript itself.
+Self \cite{ursPHD} and Strongtalk \cite{Strongtalk} were \sk{the?} to first to try this approach.
+Early versions of the Javascript engine V8 \cite{V8} also implemented most of the primitives, including for example Array operations in Javascript itself.
 However, without optional primitives implemented in the VM, several problems rise:
@@
 VMs are traditionally implemented in a low-level language such as C or C++.
-To balance between memory footprint, start-up performance and peak performance, the execution of code is usually done through multiple execution tiers: the first few executions of a code snippets are done through an interpreter and the JIT compiles at runtime the frequently used code snippets for further executions to be performed quicker.
-This leads to annoying concerns.
-For example, let's say we implement a primitive in C.
+To balance between memory footprint, start-up performance and peak performance, the execution of code is usually done through multiple execution tiers: the first few executions of a code snippet are done through an interpreter and the JIT compiler at runtime the frequently used code snippets for further executions to be performed quicker.
+This leads to annoying concerns: for example, let's say we implement a primitive in C.
 Frequently used portion of code calling that primitive are going to be compiled to native code by the JIT.
@@
 
-In the paper, we took the example of a string comparison optional primitive.
+In this paper, we take the example of a string comparison optional primitive.
 The primitive compares two strings and answers if one string is greater, equal or lower than the other string.
-The exact specification are detailed in Section \ref{subsec:primSpec}.
+The exact specifications are detailed in Section \ref{subsec:primSpec}.
 We implemented a first version in Smalltalk.
@@
 In addition, caller-saved registers on the C conventions or the JIT convention need to be saved since the VM cannot tell ahead what code is going to be used in the other execution model and what registers are required.
-We estimate the overhead to up to around a dozen of processor instruction, the exact number varies a lot so it's difficult to precise it (the number of register to edit depends on the processor and on the code executed).
+We estimate the overhead to up to around a dozen of processor instruction, the exact number varies a lot so it is difficult to precise it (the number of register to edit depends on the processor and on the code executed).
 
@@
 When called from the interpreter, activating a primitive is just a C call (the primitive function pointer is cached in the look-up cache next to the method to activate).
-When called from the native code generated by the JIT, activating a primitive requires to switch the execution form the client stack to the C stack, perform the primitive, and switch back to resume execution.
+When called from the native code generated by the JIT, activating a primitive requires to switch the execution from the client stack to the C stack, perform the primitive, and switch back to resume execution.
 If the primitive takes a significant amount of instructions to execute, for example, the primitive is copying 100kb from one array to another one, the stack switch dance overhead is negligible.
@@
 
- \cba{petite intro avec plan de section, context was moved before.}
+ This section first describes the different ways of implementing a primitive. We then explain how each of these version are executed by the VM and the pros and cons of each and every.
 
- \subsection{Different implementation and execution}
+ \subsection{Different implementations and executions}
 
@@
 Smalltalk code on top of VM. Used as our baseline for benchmarks.
-Write down the Smalltalk code here.
-%CHECK how to import code env in latex.
-%Note we're talking about comparison, not =, so not sure we should show = or bench =
-%\begin{code}
-%= aString 	
-%	self == aString ifTrue: [ ^true ].
-%	aString isString ifFalse: [ ^false ].
-%	self size = aString size ifFalse: [ ^false ].
-%	1 to: self size do: [:i | (self at: i) = (aString at: i) ifFalse: [ ^ false ] ].
-%	^ true
-%\end{code}
+
+\begin{code}
+compareWithPureST: aString
+
+	| c1 c2 length1 length2 |
+	length1 := self size.
+	length2 := aString size.
+		1 to: (length1 min: length2) do: [ :i |
+			(c1 := self basicAt: i) = (c2 := aString basicAt: i) 
+			ifFalse: [
+				c1 < c2
+					ifTrue: [ ^1 ]
+					ifFalse: [ ^3 ] ] ].
+		length1 = length2 ifTrue: [ ^2 ].
+		length1 < length2
+			ifTrue: [ ^1 ]
+			ifFalse: [ ^3 ] 
+\end{code}
+
 Expliquer to:do: over do: (to explain expresiveness) => core lib often use to:do: and not do:
-
 New spec
@@
 
+\begin{table}
+\centering
+\begin{tabular}{|c||c|c|c|c|}
+  \hline
+    & \rot{Engineering time} & \rot{Execution time} & \rot{Readability}  \\
+  \hline
+  Baseline & + & +++ & +++ \\
+  SmartSyntax & ++ & ++ & ++ \\
+  SlangPlugin & ++ & ++ & ++ \\
+  PureSlang & +++ & + & ++ \\
+  Slang+JIT & +++ & + & + \\
+  \hline
+\end{tabular}
+\caption{Trade-offs between implementations: synthetized}
+\label{tab:comparisontable}
+\end{table}
+
+Each of these implementations has its pros and cons considering the context. In the end, there is always a trade-off between engineering time and performance. In
+
 Trade-off engineering time and performance.
@@
 \subsection{Micro-benchmarks}
-Results on strings of different lengths (3, 10, 1000)
 
@@
         xtick={1,...,5},
-        xticklabels={Baseline, oldPlugin, SlangPlugin, PureSlang, Slang+JIT},
+        xticklabels={Baseline, SmartSyntax, Plugin, PureSlang, Slang+JIT},
         %x tick label as interval,
@@
 1   1           0 
-2   0.3147064   0.002288488 
-3   0.3035965   0.001786185
-4   0.8545431   0.006132950 
-5   2.250732    0.04133638 
+2   0.3411212   0.0006746299 
+3   0.3375418   0.001577878
+4   1.077762   0.005363974 
+5   2.490423    0.004304882 
 };            
@@
 1   1           0 
-2   1.2505193   0.004611354 
-3   1.2295801   0.001810171
-4   2.8205415   0.008035024 
-5   7.089806    0.01403785 
+2   1.3891212   0.0029489731 
+3   1.3149840   0.020471694
+4   3.818961   0.054523074 
+5   6.740474    0.010805042 
 };            
@@
 1   1           0 
-2   6.6951738   0.062391122 
-3   6.6032572   0.024774416
-4   10.3029394   0.061481532 
-5   17.979611    0.10127936 
+2   8.6795487   0.0586762222 
+3   8.9585208   0.116958804
+4   15.809515   0.300447885 
+5   16.752667    0.106255682 
 };            
@@
 1   1           0 
-2   9.4064534   0.036698187 
-3   9.2138415   0.013982843
-4   13.4384830   0.040202717 
-5   26.404792    0.09298324 
+2   13.6051267   0.0589164968 
+3   14.7789145   0.147140189
+4   24.417751   0.186326223 
+5   24.480071    0.095038374 
 };            
@@
    \caption{GCC-Linux results}
+   \label{fig:gcc}
 \end{figure*}
@@
    \caption{LLVM-Mac results}
+   \label{fig:llvm}
 \end{figure*}
 
-\cba{Je pense string de taille 0-1 c'est cool pour montrer que le code C va moins vite que pur smalltalk dans ce cas (si c'est vraiment le cas)}
-4 string length. (1er <=3, dernier > 1000) 
-line 1: GCC
-line 2: LLVM
-String evaluation here.
+%\cba{Je pense string de taille 0-1 c'est cool pour montrer que le code C va moins vite que pur smalltalk dans ce cas (si c'est vraiment le cas)}
+We performed the benchmarks upon 4 different string sizes: 0, 5, 100 and 1000 characters, using 2 different compilers: GCC on one hand and LLVM on the other hand.
 
-Il faut expliquer en une phrase: 
-petit string = cout important C-St stack switch.
-Gde string = cout de la boucle.
+The results are shown in terms of relative speed-up towards the baseline implementation, i.e. the pure smalltalk version. 
+We can notice some patterns over the results in figures \ref{fig:gcc} and \ref{fig:llvm}:
+\begin{itemize}
+\item The execution time of the SmartSyntax and Plugin implementations is either very close or greater than the baseline's for the small strings.
+This can be explained by the high cost of the switching between the C and Smalltalk stacks, as mentionned in \sk{faire ref à la partie}.
+\item The Slang+JIT implementation is more performant in every case. Nevertheless, 
+the performance gap between this implementation and the PureSlang one tends to decrease as the string size increase. The comparison loop \sk{ref to appendix}
+\end{itemize} proves to be more and more time-consuming because of the higher number of iterations.
+
+Howerver, one could notice a rather striking performance difference between the 2 compilers: when compiling with GCC, we notive in Figure \ref{gcc} that the performances between the PureSlang and Slang+JIT implementations are really close for long strings (100 characters and longer), with respectively 24,42 and 24,48 relative speed-up factor for a 1000-characters string.
+They still greatly differ when LLVM is used, with respectively 13,44 and 26,40 relative speed-up factor.
+
+%\begin{lstlisting}[language={[x64]Assembler}]
+%movzbl 0x8(%eax,%edx,1),%ebp
+%movzbl 0x8(%eax,%ebx,1),%ecx
+%cmp    %ecx,%ebp
+%jne    6e8 <primitiveCompareWith+0x1f8>
+%add    $0x1,%eax
+%cmp    %eax,%edi
+%jne    610 <primitiveCompareWith+0x120> 
+%\end{lstlisting}
+
+\begin{table}
+\centering
+\begin{tabular}{|r|l|}
+  \hline
+   GCC & LLVM  \\
+  \hline
+  `movzbl 0x8(\%eax,\%edx,1),\%ebp` & `movl -0x14(\%ebp), \%esi`  \\
+  movzbl 0x8(\%eax,\%ebx,1),\%ecx & movzbl 0x8(\%edx,\%esi), \%esi \\
+  cmp    \%ecx,\%ebp & movl -0x18(\%ebp), \%edi \\
+  jne    0x8558 &  movzbl 0x8(\%edx,\%edi), \%edi \\
+  add    \$0x1,\%eax & subl \%edi, \%esi \\
+  cmp    \%eax,\%edi &  jne 0x1f660 \\
+  jne    0x8480 & incl \%edx \\
+   & cmpl \%ebx, \%edx \\
+   & jl 0x1f644 \\
+  \hline
+\end{tabular}
+\caption{Different compilers optimizations: GCC and LLVM}
+\label{tab:cmp_asm}
+\end{table}
 

### 2018-05-08 paper.tex
change tikz figs
adds the asm loops and comparison

@@
 \item The execution time of the SmartSyntax and Plugin implementations is either very close or greater than the baseline's for the small strings.
-This can be explained by the high cost of the switching between the C and Smalltalk stacks, as mentionned in \sk{faire ref à la partie}.
-\item The Slang+JIT implementation is more performant in every case. Nevertheless, 
+This can be explained by the high cost of the switching between the C and Smalltalk stacks, as mentionned in Section \ref{sec:VMexec}.
+\item The Slang+JIT implementation is more efficient in every case. Nevertheless, 
 the performance gap between this implementation and the PureSlang one tends to decrease as the string size increase. The comparison loop \sk{ref to appendix}
@@
 
-Howerver, one could notice a rather striking performance difference between the 2 compilers: when compiling with GCC, we notive in Figure \ref{gcc} that the performances between the PureSlang and Slang+JIT implementations are really close for long strings (100 characters and longer), with respectively 24,42 and 24,48 relative speed-up factor for a 1000-characters string.
-They still greatly differ when LLVM is used, with respectively 13,44 and 26,40 relative speed-up factor.
+However, one could notice a rather striking performance difference between the 2 compilers: when compiling with GCC, we notive in Figure \ref{gcc} that the performances between the PureSlang and Slang+JIT implementations are really close for long strings (100 characters and longer), with respectively 24,42 and 24,48 relative speed-up factors for a 1000-characters string.
+Yet they greatly differ when LLVM is used, with respectively 13,44 and 26,40 relative speed-up factors.
 
-%\begin{lstlisting}[language={[x64]Assembler}]
-%movzbl 0x8(%eax,%edx,1),%ebp
-%movzbl 0x8(%eax,%ebx,1),%ecx
-%cmp    %ecx,%ebp
-%jne    6e8 <primitiveCompareWith+0x1f8>
-%add    $0x1,%eax
-%cmp    %eax,%edi
-%jne    610 <primitiveCompareWith+0x120> 
-%\end{lstlisting}
+This difference is mainly due to the optimizations performed by both the compilers: as depicted in Figure \ref{Asm} GCC generates instructions that are very close to the ones generated by the JIT, hence the similar speed-up factors.
+On the contrary, LLVM generates more instructions: it especially fails at perserving the registries holding the strings (edi and esi), leading to 2 extra instructions (bolded in the figure) that will load the strings at each iteration.
 
-\begin{table}
+\begin{figure}[htp!]
 		\centering
-\begin{tabular}{|r|l|}
-  \hline
-   GCC & LLVM  \\
-  \hline
-  `movzbl 0x8(\%eax,\%edx,1),\%ebp` & `movl -0x14(\%ebp), \%esi`  \\
-  movzbl 0x8(\%eax,\%ebx,1),\%ecx & movzbl 0x8(\%edx,\%esi), \%esi \\
-  cmp    \%ecx,\%ebp & movl -0x18(\%ebp), \%edi \\
-  jne    0x8558 &  movzbl 0x8(\%edx,\%edi), \%edi \\
-  add    \$0x1,\%eax & subl \%edi, \%esi \\
-  cmp    \%eax,\%edi &  jne 0x1f660 \\
-  jne    0x8480 & incl \%edx \\
-   & cmpl \%ebx, \%edx \\
-   & jl 0x1f644 \\
-  \hline
-\end{tabular}
-\caption{Different compilers optimizations: GCC and LLVM}
-\label{tab:cmp_asm}
-\end{table}
+		\includegraphics[width=0.99\linewidth]{figures/cmpasm.png}
+		\caption{GCC and LLVM: different optimizations on the comparison loop}
+		\label{fig:Asm}
+\end{figure}
 

### 2018-05-08 paper.tex
changed tikz pic bench result in 1 pic

--- 2018-05-08 paper.tex
@@
 The primitive answers a negative value if the first parameter is smaller than the second, 0 if they are equal, and a positive value if the first parameter is greater than the second one.
-\cba{Maybe rewrite the last paragraph to make it clearer / more formal.}
+\cba{Maybe rewrite the last paragraph to make it clearer / more formal. - CF Java spec ?}
 
@@
 
-\cba{Personnellement je ferais une figure, ou tu places tes graphes dedans, et pares tu referes a la figure dans le texte.}
+%%%
+% Legend definition (so I can change it once for all)
+%%%
+
+\def\legBase{Baseline}
+\def\legSmartSyntaxPlug{Smart Syntax}
+\def\legSlangPlug{Plugin}
+\def\legSlang{Pure Slang}
+\def\legRTL{Slang+JIT}
+
+%%%
+% End Legend
+%%%
 
@@
   \centering
-
-    \begin{subfigure}[b]{.45\textwidth}
-
+    \begin{subfigure}[b]{\textwidth}
+    \begin{subfigure}[b]{.19\textwidth}
     \begin{tikzpicture}
@@
       [
+        x=0.5cm,
         xtick={1,...,5},
-        xticklabels={Baseline, SmartSyntax, Plugin, PureSlang, Slang+JIT},
-        %x tick label as interval,
-        font={\tiny},
+        xticklabel style={rotate=90},
+        bar width=0.3cm,
+        enlarge x limits={abs=0.45cm},
+        xticklabels={\legBase, \legSmartSyntaxPlug, \legSlangPlug, \legSlang, \legRTL},
+        font={\footnotesize},
 	ybar=10pt,
-		xbar width=0.30,
-		no markers,Baseline/.style={red},2/.style={green},3/.style={blue}
-		%cycle list={red, orange, green, purple, black}
+	ymin=0
         ]
@@
     \end{subfigure}
-    \begin{subfigure}[b]{.45\textwidth}
+     \begin{subfigure}[b]{.19\textwidth}
     \begin{tikzpicture}
@@
       [
+        x=0.5cm,
         xtick={1,...,5},
-        xticklabels={Baseline, SmartSyntax, Plugin, PureSlang, Slang+JIT},
-        %x tick label as interval,
-        font={\tiny},
+        xticklabel style={rotate=90},
+        bar width=0.3cm,
+        enlarge x limits={abs=0.45cm},
+        xticklabels={\legBase, \legSmartSyntaxPlug, \legSlangPlug, \legSlang, \legRTL},
+        font={\footnotesize},
 	ybar=10pt,
-		xbar width=0.30
+	ymin=0
+        ]
+\addplot[
+    fill=black!25,
+    %draw=none,
+    %fill,
+    draw=black,
+    fill,
+    point meta=y,
+    every node near coord/.style={inner ysep=5pt},
+    error bars/.cd,
+        y dir=both,
+        y explicit
+]  
+table [y error=error] {
+x   y           error    
+1   1           0 
+2   0.3411212   0.0006746299 
+3   0.3375418   0.001577878
+4   1.077762   0.005363974 
+5   2.490423    0.004304882 
+};            
+      \end{axis}
+    \end{tikzpicture}
+    \subcaption{Size 1 TO UPDATE}
+    \end{subfigure}
+    \begin{subfigure}[b]{.19\textwidth}
+    \begin{tikzpicture}
+      \begin{axis}
+      [
+        x=0.5cm,
+        xtick={1,...,5},
+        xticklabel style={rotate=90},
+        bar width=0.3cm,
+        enlarge x limits={abs=0.45cm},
+        xticklabels={\legBase, \legSmartSyntaxPlug, \legSlangPlug, \legSlang, \legRTL},
+        font={\footnotesize},
+	ybar=10pt,
+	ymin=0
         ]
@@
     \end{tikzpicture}
-
     \subcaption{Size 5}
     \end{subfigure}
-    \begin{subfigure}[b]{.45\textwidth}
+    \begin{subfigure}[b]{.19\textwidth}
  \begin{tikzpicture}
@@
       [
+        x=0.5cm,
         xtick={1,...,5},
-        xticklabels={Baseline, SmartSyntax, Plugin, PureSlang, Slang+JIT},
-        %x tick label as interval,
-        font={\tiny},
+        xticklabel style={rotate=90},
+        bar width=0.3cm,
+        enlarge x limits={abs=0.45cm},
+        xticklabels={\legBase, \legSmartSyntaxPlug, \legSlangPlug, \legSlang, \legRTL},
+        font={\footnotesize},
 	ybar=10pt,
-		ymin=0,
-		xbar width=0.30
+	ymin=0
         ]
@@
     \end{tikzpicture}
-
     \subcaption{Size 100}
     \end{subfigure}
-    \begin{subfigure}[b]{.45\textwidth}
-
+    \begin{subfigure}[b]{.19\textwidth}
      \begin{tikzpicture}
@@
       [
+        x=0.5cm,
         xtick={1,...,5},
-        xticklabels={Baseline, SmartSyntax, Plugin, PureSlang, Slang+JIT},
-        %x tick label as interval,
-        font={\tiny},
+        xticklabel style={rotate=90},
+        bar width=0.3cm,
+        enlarge x limits={abs=0.45cm},
+        xticklabels={\legBase, \legSmartSyntaxPlug, \legSlangPlug, \legSlang, \legRTL},
+        font={\footnotesize},
 	ybar=10pt,
-		ymin=0,
-		xbar width=0.30
+	ymin = 0
         ]
@@
     \end{tikzpicture}
-
    \subcaption{Size 1000}
    \end{subfigure}
-   \caption{GCC-Linux results}
-   \label{fig:gcc}
-\end{figure*}
-
-\begin{figure*}
-  \centering
-
-    \begin{subfigure}[b]{.45\textwidth}
+  % \caption{GCC-Linux results}
+   % \label{fig:gcc}
+   \subcaption*{\textbf{GCC-Linux results}\vspace{0.3cm}}     
+   \end{subfigure}
+ \begin{subfigure}[b]{\textwidth}
+    \begin{subfigure}[b]{.19\textwidth}
 
@@
     \end{subfigure}
-    \begin{subfigure}[b]{.45\textwidth}
+     \begin{subfigure}[b]{.19\textwidth}
 
@@
       [
+        x=0.5cm,
         xtick={1,...,5},
-        xticklabels={Baseline, SmartSyntax, Plugin, PureSlang, Slang+JIT},
-        %x tick label as interval,
-        font={\tiny},
+        xticklabel style={rotate=90},
+        bar width=0.3cm,
+        enlarge x limits={abs=0.45cm},
+        xticklabels={\legBase, \legSmartSyntaxPlug, \legSlangPlug, \legSlang, \legRTL},
+        font={\footnotesize},
 	ybar=10pt,
-		xbar width=0.30
+	ymin=0
+        ]
+\addplot[
+    fill=black!25,
+    draw=black,
+    point meta=y,
+    every node near coord/.style={inner ysep=5pt},
+    error bars/.cd,
+        y dir=both,
+        y explicit
+] 
+table [y error=error] {
+x   y           error    
+1   1           0 
+2   0.3147064   0.002288488 
+3   0.3035965   0.001786185
+4   0.8545431   0.006132950 
+5   2.250732    0.04133638 
+};            
+      \end{axis}
+    \end{tikzpicture}
+
+    \subcaption{Size 1 TO UPDATE}
+    \end{subfigure}
+    \begin{subfigure}[b]{.19\textwidth}
+
+    \begin{tikzpicture}
+      \begin{axis}
+      [
+        x=0.5cm,
+        xtick={1,...,5},
+        xticklabel style={rotate=90},
+        bar width=0.3cm,
+        enlarge x limits={abs=0.45cm},
+        xticklabels={\legBase, \legSmartSyntaxPlug, \legSlangPlug, \legSlang, \legRTL},
+        font={\footnotesize},
+	ybar=10pt,
+	ymin=0
         ]
@@
     \end{subfigure}
-    \begin{subfigure}[b]{.45\textwidth}
+    \begin{subfigure}[b]{.19\textwidth}
  \begin{tikzpicture}
@@
     \end{subfigure}
-    \begin{subfigure}[b]{.45\textwidth}
+    \begin{subfigure}[b]{.19\textwidth}
 
@@
    \end{subfigure}
-   \caption{LLVM-Mac results}
+   \subcaption*{\textbf{LLVM-Mac results}}
+    \end{subfigure}
    \label{fig:llvm}

### 2018-05-08 paper.tex
implem comparison + asm comparison figure

--- 2018-05-08 paper.tex
@@
 The two first operands were the two ByteStrings to compare and the third operand is the order table, a ByteArray encoding the order of characters.
-The default order is the ASCII order, hence each entry corresponds to its index in the byte array (The ByteArray has a size of 256, and accessing the element i, with 0 <= i <= 255, answered i).
+The default order is the ASCII order, hence each entry corresponds to its index in the byte array (The ByteArray has a size of 256, and accessing the element i, with 0 $\leq$ i $\leq$ 255, answered i).
 However, in some case, the program may do case insensitive ByteStrings comparison.
@@
  \subsection{Different implementations and executions}
-
- \cba{Maybe wrte version A, B, C , D , E or a word to represent what they are (Smalltalk) (SmartSyntax) (Plugin) (Slang) (Slang+JIT)
- so you can refer to that to talk about any of these later and not wirte each time : the implem with ths wirtten that wya and exec that way.
- Myabe those could be paragraph titles instead of the following}
 
@@
 \centering
-\begin{tabular}{|c||c|c|c|c|}
+\begin{tabular}{|c|c|c||c||c|c|c||c|}
   \hline
-    & \rot{Engineering time} & \rot{Execution time} & \rot{Readability}  \\
+    & \rot{readability} & \rot{maintenance} & \rot{Engineering time} & \rot{re-compilation} & \rot{plugin} & \rot{stack switching} & \rot{Execution time} \\
   \hline
-  Baseline & + & +++ & +++ \\
-  SmartSyntax & ++ & ++ & ++ \\
-  SlangPlugin & ++ & ++ & ++ \\
-  PureSlang & +++ & + & ++ \\
-  Slang+JIT & +++ & + & + \\
+  Baseline    & X &   & +   &    &   &   & +++ \\
+  SmartSyntax & X &   & ++  &    & X & X & ++\\
+  SlangPlugin &   &   & ++  &    & X & X & ++ \\
+  PureSlang   &   & X & +++ &  X &   & X & ++\\
+  Slang+JIT   &   & X & +++ &  X &   &   & + \\
   \hline
 \end{tabular}
-\caption{Trade-offs between implementations: synthetized}
-\label{tab:comparisontable}
+\caption{Trade-offs between implementations: summarized}
+\label{tab:cmp_implem}
 \end{table}
 
-Each of these implementations has its pros and cons considering the context. In the end, there is always a trade-off between engineering time and performance. In
+One or the other implementation could be chosen over another one, depending on what one's aiming at: the main trade-off will be between the engineering time and the performance.
+The table \ref{cmp_implem} summarizes the main pros and cons related to each implementation
 
@@
 \subsection{Micro-benchmarks}
-
-\cba{Personnellement je ferais une figure, ou tu places tes graphes dedans, et pares tu referes a la figure dans le texte.}
 
@@
 We performed the benchmarks upon 4 different string sizes: 0, 5, 100 and 1000 characters, using 2 different compilers: GCC on one hand and LLVM on the other hand.
+Figure \ref{fig:gcc} and \ref{fig:llvm} show the results in terms of relative speed-up towards the baseline implementation\footnote{i.e.
+the pure Smalltalk version}, respectively for GCC and LLVM.
 
-The results are shown in terms of relative speed-up towards the baseline implementation, i.e. the pure smalltalk version. 
-We can notice some patterns over the results in figures \ref{fig:gcc} and \ref{fig:llvm}:
+We can draw the following patterns from these results:
 \begin{itemize}
@@
 
+\sk{parler des autres optis qui n'ont pas d'impact négatif sur le temps d'exec ? ou le conserver pour les future work pour des potentielles améliorations?}
+
 \begin{figure}[htp!]
@@
 
-\cba{Show ASM loops and discuss the differences
-add appendix with full code}
+\cba{appendix with full code}
 

### 2018-05-08 paper.tex

--- 2018-05-08 paper.tex
@@
 \subsection{Micro-benchmarks}
-
 %%%

### 2018-05-08 paper.tex
update plot value for size 1
comments

@@
 The primitive answers a negative value if the first parameter is smaller than the second, 0 if they are equal, and a positive value if the first parameter is greater than the second one.
-\cba{Maybe rewrite the last paragraph to make it clearer / more formal. - CF Java spec ?}
+\cba{Maybe rewrite the last paragraph to make it clearer / more formal. - CF Java spec ? => This is critical, maybe in bold or in a box}
 
@@
     fill=black!25,
-    %draw=none,
-    %fill,
     draw=black,
@@
 1   1           0 
-2   0.3411212   0.0006746299 
-3   0.3375418   0.001577878
-4   1.077762   0.005363974 
-5   2.490423    0.004304882 
+2   0.6013716   0.0023206431 
+3   0.5653467   0.007829291
+4   1.816869    0.019114984 
+5   3.537800    0.012357500 
 };            
@@
     \end{tikzpicture}
-    \subcaption{Size 1 TO UPDATE}
+    \subcaption{Size 1}
     \end{subfigure}
@@
 1   1           0 
-2   0.3147064   0.002288488 
-3   0.3035965   0.001786185
-4   0.8545431   0.006132950 
-5   2.250732    0.04133638 
+2   0.5605682   0.002198180 
+3   0.5566675   0.001023225
+4   1.5605275   0.006372279 
+5   3.754532    0.01109086 
 };            
@@
 
-    \subcaption{Size 1 TO UPDATE}
+    \subcaption{Size 1}
     \end{subfigure}
@@
 
-\cba{appendix with full code}
+\cba{I think in the fig we should also see the ASM cog generate}
 

### 2018-05-08 paper.tex
fixed abstract and first intro paragraph https://www.merriam-webster.com/dictionary/high-level

@@
 Some of these are essential and cannot be performed in any other way.
-Others are optional, they can be expressed in the client language but are often implemented in the virtual machine to improve performance when the just-in-time compiler is unable to do so (start-up performance, speculative optimizations not implemented or not mature enough, etc.).
+Others are optional: they can be expressed in the client language but are often implemented in the virtual machine to improve performance when the just-in-time compiler is unable to do so (start-up performance, speculative optimizations not implemented or not mature enough, etc.).
 
-In a hybrid runtime, where code is executed by an interpreter and a just-in-time compiler, the implementor can choose to implement optional primitives in the client language, in the virtual machine implementation language (typically C or C++) or on top of the just-in-time compiler back-end.
+In a hybrid runtime, where code is executed by an interpreter and a just-in-time compiler, the implementor can choose to implement optional primitives in the client language, in the virtual machine implementation language (typically C or C++), or on top of the just-in-time compiler back-end.
+This raises the questions of the tradeoffs of the different alternatives.
 We implemented the String comparison optional primitive in each case.
@@
 
-The paper shows that the implementation with the most reliable and high performance is also the most complex and difficult to implement.
-Since hundreds of optional primitives can be available, the virtual machine implementor has to carefully choose how to implement each of them to balance between maintenance cost and high performance.
-
+Without surprise, the paper shows that the implementation with the most reliable and high performance is also the most complex and difficult to implement.
+Since hundreds of optional primitives could be wished, the virtual machine implementor has to carefully choose how to implement each of them to balance between maintenance cost and high performance.
 \end{abstract}
@@
 
-High level object oriented programming languages are often implemented on top of a virtual machine (VM).
+High-level object-oriented programming languages are often implemented on top of a virtual machine (VM).
 As VMs has become more popular, different techniques have been set-up using Just-In-Time (JIT) compilation to improve the overall performance of the runtime such as method JITs \cite{ursPHD} which usually compile methods to native code and tracing JITs \cite{Dynamo, PyPyTracing} which usually compile linear traces of execution into native code.

### 2018-05-08 paper.tex
f

--- 2018-05-08 paper.tex
@@
 Therefore, the VM does not need to go through the indirection order array in the common case.
-The primitive answers a negative value if the first parameter is smaller than the second, 0 if they are equal, and a positive value if the first parameter is greater than the second one.
-\cba{Maybe rewrite the last paragraph to make it clearer / more formal. - CF Java spec ? => This is critical, maybe in bold or in a box}
+The primitive answers a negative integer, zero, or a positive integer as the first parameter is less than, equal to, or greater than the second parameter.
 
@@
 \paragraph{Baseline: pure Smalltalk.}
+\cba{Hum I believed that one should follow new specification and not old one ? With neg, 0 or pos value ?}
 : baseline

### 2018-05-08 paper.tex

--- 2018-05-08 paper.tex
@@
 
+\cba{We need to introduce here the concept of VM plugin and interpreterproxy indirection/overhead. This is very important since later we just say this is a plugin and we don't redefine it}
+
 \subsection{VM execution}
@@
 Therefore, the VM does not need to go through the indirection order array in the common case.
-The primitive answers a negative value if the first parameter is smaller than the second, 0 if they are equal, and a positive value if the first parameter is greater than the second one.
-\cba{Maybe rewrite the last paragraph to make it clearer / more formal. - CF Java spec ? => This is critical, maybe in bold or in a box}
+The primitive answers a negative integer, zero, or a positive integer as the first parameter is less than, equal to, or greater than the second parameter.
+
+In the paper we will focus on the comparison of two strings with the default ASCII ordering and measure the performance only of this case.
+The goal of this work was to improve the performance of common string operations, typically string equality which internally use the primitive, and the cases where the order is non ASCII are uncommon.
 
@@
 \paragraph{Baseline: pure Smalltalk.}
-: baseline
-Smalltalk code on top of VM. Used as our baseline for benchmarks.
+To evaluate the primitive performance, we use as baseline an implementation in pure Smalltalk.
+There are multiple ways of writing the comparison of two strings in Smalltalk, following Smalltalk coding conventions, a standard developer would likely use high-level constructs which apply a closure on each element of the strings.
+We did not use high-level construct and chose to implement the Smalltalk version the most optimized way Smalltalk allows, \emph{i.e.,} we used the contructs to:do: and ifFalse:, which are both recognized by the Smalltalk to bytecode compiler and compiled respectively to loops and branches at the bytecode level.
+The implementation follows the new specification, hence two methods are available, baselineCompareWith: and baselineCompareWith:order:.
+Since the focus is only on ASCII ordered comparison, we show the code only of baselineCompareWith:.
 
 \begin{code}
-compareWithPureST: aString
-
+baselineCompareWith: aString 
 	| c1 c2 length1 length2 |
@@
 	1 to: (length1 min: length2) do: [ :i |
-			(c1 := self basicAt: i) = (c2 := aString basicAt: i) 
-			ifFalse: [
-				c1 < c2
-					ifTrue: [ ^1 ]
-					ifFalse: [ ^3 ] ] ].
-		length1 = length2 ifTrue: [ ^2 ].
-		length1 < length2
-			ifTrue: [ ^1 ]
-			ifFalse: [ ^3 ] 
+		(c1 := self basicAt: i) = (c2 := aString basicAt: i) ifFalse: [ ^ c1 - c2 ] ].
+	^ length1 - length2
 \end{code}
 
-Expliquer to:do: over do: (to explain expresiveness) => core lib often use to:do: and not do:
-New spec
+\paragraph{Existing: Smart Syntax Plugin.}
 
-\paragraph{Existing: Smart Syntax Plugin.}
-Restricted Smalltalk
-
-written in Smalltalk with restrictions.
-Compiled to C AOT into a plugin.
-Write once, fall-back + compiled.
-
-old spec
+The Smart Syntax Plugin is one of the version existing until we started this work.
+The idea behind Smart Syntax was the following: the programmer can write code in Smalltalk using exclusively constructs that are recognized by the Smart Syntax compiler and that can be directly compiled to C (computing the size of a byte array, accessing its fields and doing integer arithmetic for example).
+With the addition of type annotations in the form of pragmas \cite{Duca16a} \footnote{Pragma are a form of method annotation.}, the Smart Syntax plugin was able to generate C code from the Smalltalk code.
+With this approach, the programmer has to write the optional primitive code a single time in Smalltalk.
+This code is used both as the fall-back code if the VM does not implement the primitive and at Slang to C compilation time to generate a C plugin from the code and the type annotations.
+This primitive follows the old specification.
 
 \paragraph{Existing: Slang Plugin.}
-Slang, but no access au fonctions de la VM drectement.
- interpreterProxy. (C compiler inlining, linking time ? cmacro ?)
 
- old spec
+The Slang plugin is a recent complete rewrite of the Smart Syntax plugin.
+The main difference is that the Smalltalk version is used only as fall-back code, while at Slang to C compilation time the VM uses code written in Slang to generate the C code.
+One of the side-effect is that the plugin's code is now part of the VM code base and not the client code bases, which is convenient for maintenance since the Cog VM has multiple client languages.
+However, the implementor now needs to maintain and evolve two version of the primitive, the fall-back code in Smalltalk and the code in Slang.
+This primitive follows the old specification.
 
@@
 One or the other implementation could be chosen over another one, depending on what one's aiming at: the main trade-off will be between the engineering time and the performance.
-The table \ref{cmp_implem} summarizes the main pros and cons related to each implementation
+The table \ref{tab:cmp_implem} summarizes the main pros and cons related to each implementation
 
@@
 
-This difference is mainly due to the optimizations performed by both the compilers: as depicted in Figure \ref{Asm} GCC generates instructions that are very close to the ones generated by the JIT, hence the similar speed-up factors.
+This difference is mainly due to the optimizations performed by both the compilers: as depicted in Figure \ref{fig:Asm} GCC generates instructions that are very close to the ones generated by the JIT, hence the similar speed-up factors.
 On the contrary, LLVM generates more instructions: it especially fails at perserving the registries holding the strings (edi and esi), leading to 2 extra instructions (bolded in the figure) that will load the strings at each iteration.
@@
 %% Bibliography
-\bibliography{sista}
+\bibliography{sista,rmod}
 

### 2018-05-08 paper.tex
reading up where it makes sense

@@
 High-level object-oriented programming languages are often implemented on top of a virtual machine (VM).
-As VMs has become more popular, different techniques have been set-up using Just-In-Time (JIT) compilation to improve the overall performance of the runtime such as method JITs \cite{ursPHD} which usually compile methods to native code and tracing JITs \cite{Dynamo, PyPyTracing} which usually compile linear traces of execution into native code.
+As VMs has become more popular, different techniques have been set up using Just-In-Time (JIT) compilation to improve the runtime overall performance: method JITs \cite{ursPHD} which usually compile methods to native code and tracing JITs \cite{Dynamo, PyPyTracing} which usually compile linear traces of execution into native code.
 
-On top of its virtual machine, the client language can use a set of \emph{primitives}~\footnote{We use in the paper the Smalltalk terminology (primitive) as this is the programming language used for our evaluation.
+On top of its virtual machine, the client language can use a set of \emph{primitives}\footnote{We use in the paper the Smalltalk terminology (primitive) as this is the programming language used for our evaluation.
 Some other programming languages, such as Javascript, prefer to use the term built-in instead of primitive.}, which are performed directly by the interpreter rather than by evaluating expressions in a method.
@@
 	\item \emph{Essential primitives} cannot be performed in any other way.
-A high-level object oriented language without primitives can move values from one variable to another, but cannot add two integers together.
+A high-level object-oriented language without primitives can move values from one variable to another, but cannot add two integers together.
 Many arithmetic and comparison operations between numbers are primitives.
-Some primitives allow to communicate with I/O devices such as the disk, the display and the keyboard.
-	\item \emph{Optional primitives} exist only to make the system run faster.
-They can be implemented in the client language directly, making its implementation as a primitive optional, but they are implemented in the VM to improve performance.
+Some primitives allow one to communicate with I/O devices such as the disk, the display and the keyboard.
+	\item \emph{Optional primitives} exist only to make the system run faster \cite{blueBook}.
+They can be implemented in the client language directly, making its implementation as a primitive optional, but they are often implemented directly in the VM to improve performance.
 \end{itemize}
@@
 The implementation of an optional primitive in the VM rather than the client language improves performance but also often increases the implementation engineering cost.
-For example, accessing objects in the VM usually requires some understanding of the memory layout or implementation details, which are not needed in the client language.
+For example, accessing objects in the VM usually requires understanding of the memory layout or implementation details, which are not needed in the client language.
 
 For this reason, some virtual machine implementors attempted to remove most or all optional primitives.
-Their goal was to improve the execution of the client language with a JIT performing speculative optimizations \cite{ursPHD} for the execution of the optional primitives implemented in the client language to be as fast as the performance of the same primitives written in the VM.
+\sd{sentence was too long: please reread my version:}Their goal was to use a JIT performing speculative optimizations of optional primitives implemented in the client language.
+Ideally we wanted to get   the performance of the same primitives written in the VM.
 Self \cite{ursPHD} and Strongtalk \cite{Strongtalk} were \sk{the?} to first to try this approach.
@@
 VMs are traditionally implemented in a low-level language such as C or C++.
-To balance between memory footprint, start-up performance and peak performance, the execution of code is usually done through multiple execution tiers: the first few executions of a code snippet are done through an interpreter and the JIT compiler at runtime the frequently used code snippets for further executions to be performed quicker.
+To balance between memory footprint, start-up performance and peak performance, the execution of code is usually done through multiple execution tiers: the first few executions of a code snippet are done through an interpreter and the JIT compiler \sd{optimizes} at runtime the frequently used code snippets \sd{rm: for further executions to be performed quicker}.
+
 This leads to annoying concerns: for example, let's say we implement a primitive in C.
@@
 This is problematic since the native code generated by the JIT is not directly compatible with native code generated by the C compiler: a call from one to another may require to edit the stack pointer and the frame pointer from the client stack to the C stack and to spill or move multiple registers.
-Switching between both can waste up to around a dozen native instructions.
+Switching between both usually wastes up to around a dozen native instructions.
 If the primitive is going to execute many instructions, the switch overhead might be negligible, but if the primitive executes only a few instructions, the overhead can be noticeable.
+
+\sd{the link is missing: In this article we want to measure the complexities and gain of this different implementations}
 
@@
 The exact specifications are detailed in Section \ref{subsec:primSpec}.
-We implemented a first version in Smalltalk.
+We implemented a first version in Pharo/Smalltalk.
 We then implemented multiple version compiling through C compilers to native code, using different part of our VM infrastructure.
@@
 Our evaluation is based on the Cog VM a Smalltalk virtual machine.
-The Cog VM is based on Dan Ingalls' Smalltalk VM \cite{BackToTheFuture} and have been enhanced with a JIT \cite{CogJIT} compiling one method or closure at a time to native code.
+The Cog VM is based on Dan Ingalls' Smalltalk VM \cite{BackToTheFuture} and \sd{has} have been enhanced with a JIT \cite{CogJIT} compiling one method or closure at a time to native code.
 Historically, the VM implemented a 32-bits adaptation of the Smalltalk-80 specifications \cite{blueBook}, but the Cog VM is now the default VM for multiple programming languages such as Pharo \cite{PharoByExample}, Squeak \cite{SqueakByExample} and Newspeak \cite{NewspeakOopsla}.
@@
 In Section \ref{sec:primExec}, we detail how primitives are executed.
-Section \ref{subsec:primSpec} discusses the specifications of the string comparison primitive used for our evaluation.
+Section \ref{subsec:primSpec} discusses the string comparison primitive specifications  used for our evaluation.
 
@@
 The execution engine (the memory manager, the interpreter and the baseline JIT) are entirely written in Slang.
-The two main purposes of using slang code over plain C are:
+The two main purposes of using Slang code over plain C are:
 \begin{itemize}
@@
 
-The executable is generated in two steps as shown on Figure \ref{fig:VMCompilation}, similarly to the RPython toolchain \cite{RPythonToolchain}.
+The executable is generated in two steps as shown on Figure~\ref{fig:VMCompilation}, similarly to the RPython toolchain \cite{RPythonToolchain}.
 The first step is to generate the two C files representing the whole execution engine written in Slang using the Slang-to-C compiler.
@@
 The VM code, including the interpreter code, is compiled by a C compiler and uses the C stack at runtime.
-During its execution, the interpreter modifies the execution stack of the client language, which is different from its own code stack as shown on Figure \ref{fig:TwoStacks}.
+During its execution, the interpreter modifies the execution stack of the client language, which is different from its own code stack as shown on Figure~\ref{fig:TwoStacks}.
 This means for example that the frame pointer register in the processor refers to a stack frame on the C stack and that a processor push instruction pushes a value on the C stack.
@@
 
-To solve this performance bottleneck, the Cog VM allows to redefine primitives in Cog's Register Transfer Language (RTL), an abstract assembly which is compiled to native code through Cog's JIT back-end.
+To solve this performance bottleneck, the Cog VM allows one to redefine primitives in Cog's Register Transfer Language (RTL), an abstract assembly which is compiled to native code through Cog's JIT back-end.
 Such primitives are then generated to native code at runtime when a method annotated with the primitive is compiled to native code.
@@
 
-\paragraph{String representations.} Strings are represented in all programming languages on top of the Cog VM in two possible forms, ByteStrings and WideStrings.
+\paragraph{String representations.} Strings are represented on top of the Cog VM in two possible forms: ByteStrings and WideStrings.
 ByteStrings encode each character of the string in a single byte.
 The encoding usually follows the extended ASCII standard.
-WideStrings encode each character of the string in 32 bits.
+WideStrings encode each character in 32 bits.
 The encoding usually follows Unicode specifications.
@@
 \paragraph{Original specification.}  The original primitive took three operands.
-The two first operands were the two ByteStrings to compare and the third operand is the order table, a ByteArray encoding the order of characters.
+The two first operands were the two ByteStrings to compare and the third operand was the order table, a ByteArray encoding the order of characters.
 The default order is the ASCII order, hence each entry corresponds to its index in the byte array (The ByteArray has a size of 256, and accessing the element i, with 0 $\leq$ i $\leq$ 255, answered i).
@@
 In this case, the order ByteArray is different, since some entries answer the same value (for example, the entry for the 41 and 61, respectively characters A and a, both answer the same value).
-To compare the two ByteStrings, the primitive conceptually iterates byte by byte over the two objects until it reaches the end of one them, comparing for each byte that the value present in the order ByteArray at the index is greater, equal or smaller than the value computed from the other ByteString.
+To compare two ByteStrings, the primitive conceptually iterates byte by byte over the two objects until it reaches the end of one them, comparing for each byte that the value present in the order ByteArray at the index is greater, equal or smaller than the value computed from the other ByteString.
 If a difference is met, the value returns based on the result, if not, at the end of the iteration the primitive compares the ByteString sizes and returns.
-Based on the three parameters, the primitive answered 3,2 or 1 depending if the first ByteString was greater, equal or smaller than the second one.
+Based on the three parameters, the primitive answered 3, 2, or 1 depending if the first ByteString was greater, equal or smaller than the second one.
 
@@
 \begin{itemize}
-	\item \emph{Returned value:} Other programming languages implement the primitive to return a negative integer, 0 or a positive integer instead of 1,2 or 3.
-It looks like a minor difference, but it is quite important for performance.
+	\item \emph{Returned value:} Other programming languages implement the primitive to return a negative integer, 0 or a positive integer instead of 1, 2 or 3.
+It looks like a minor difference, but it is important for performance.
 With a negative, 0 or positive value, the primitive can just answer the difference between two different characters or the difference in size if all first characters are equal.
-With 1,2 or 3, the primitive has to end with multiple branches to know what number to return.
+With 1, 2, or 3, the primitive has to end with multiple branches to know what number to return.
 In native code, a subtraction is on common architecture quicker to perform than one or multiple branches.
@@
 
- This section first describes the different ways of implementing a primitive. We then explain how each of these version are executed by the VM and the pros and cons of each and every.
+ This section first describes the different ways of implementing a primitive. We then explain how each of these versions are executed by the VM and the pros and cons of each and every.
 

### 2018-05-08 paper.tex
update

--- 2018-05-08 paper.tex
@@
 \subsection{Set-up}
-Describe set-up:
+
+Evaluaiton on x86 and not x64.
+
 GCC eval Ubuntu + machine spec
-LLVM eval Mac OS X + machine spec.
+
+The LLVM-Mac evaluaiton was performed on a MacBook pro with Mac OS 10.11.6, a 2.9Ghz processor Intel Core 5, 8 Gb 1867MHz DDR3 of RAM.
+The Mac VM was compiled using Apple LLVM version 8.0.0 (clang-800.0.38).
 

### 2018-05-08 paper.tex
ff
added appendix

@@
    \subcaption*{\textbf{GCC-Linux results}\vspace{0.3cm}}     
+   \label{fig:gcc}
    \end{subfigure}
@@
     \end{subfigure}
-   \label{fig:llvm}
+   \label{fig:bench}
 \end{figure*}
@@
 We performed the benchmarks upon 4 different string sizes: 0, 5, 100 and 1000 characters, using 2 different compilers: GCC on one hand and LLVM on the other hand.
-Figure \ref{fig:gcc} and \ref{fig:llvm} show the results in terms of relative speed-up towards the baseline implementation\footnote{i.e.
+Figure \ref{fig:bench} show the results in terms of relative speed-up towards the baseline implementation\footnote{i.e.
 the pure Smalltalk version}, respectively for GCC and LLVM.
@@
 \appendix
-\section{Appendix}
+\section{Smart Syntax code}
+\label{SmartSyntaxAppendix}
 
-\cba{Source code et generated C and ASM code for all versions}
+\cba{Source code et generated C code et generated ASM code}
 
-Text of appendix \ldots
+\section{Slang Plugin code}
+\label{SmartSyntaxAppendix}
+
+\cba{Source code et generated C code et generated ASM code}
+
+\section{Slang code}
+\label{SlangAppendix}
+
+\cba{Source code et generated C code et generated ASM code}
+
+\section{Cog's RTL code}
+\label{RTLAppendix}
+
+\cba{Source code et generated ASM code}
 

### 2018-05-08 paper.tex
added appendix and section 3.1 done

@@
 This code is used both as the fall-back code if the VM does not implement the primitive and at Slang to C compilation time to generate a C plugin from the code and the type annotations.
-This primitive follows the old specification.
+This primitive follows the old specification, the full code is available in Appendix \ref{SmartSyntaxAppendix}.
 
@@
 However, the implementor now needs to maintain and evolve two version of the primitive, the fall-back code in Smalltalk and the code in Slang.
-This primitive follows the old specification.
+This primitive follows the old specification, the full code is available in Appendix \ref{PluginAppendix}.
 
@@
 
-new spec
+The Slang implementation is written inside the interpreter and has access to all interpreter APIs.
+Since the same primitive index \cba{Concept to introduce before or rephrase} is used for the primitive with 2 or 3 arguments, the primitive needs to deal with both cases in a way that the common case is efficient.
+The primitive is now part of the core of the VM, it cannot be easily added or removed as plugin primitives, but it is simpler to read and write.
+This primitive follows the new specification, the full code is available in Appendix \ref{SlangAppendix}.
 
 \paragraph{Pure Slang with JIT support.}
-Cog RTL
-
-generate functions as machine code
-partial implementation (frequent cases are implemented)
-Sometimes, fallback to less-optimised version
-
-new spec
+
+In addition to the previous implementation, we've re-implemented here the primitive on top of Cog's JIT back-end.
+The primitive is written in Cog's RTL, an abstract assembly which instructions compile almost one-to-one to native instructions.
+This primitive requires a more important amount of work to be written: the number of line of code is greater, the implementor has to be careful to generate efficient machine code (there is no C compiler to optimize the code) and debugging such code requires the use of a processor simulator.
+Since only performance critical implementations matter in this context, we've re-implemented only the case where there is 2 parameters.
+For the three parameter primitive, the VM falls-back to the Slang version.
+This primitive follows the new specification, the full code is available in Appendix \ref{RTLAppendix}.
 
@@
 
+\begin{code}
+compare: string1 with: string2 collated: order
+	"Return 1, 2 or 3, if string1 is <, =, or > string2, with the collating order of characters given by the order array."
+
+	| len1 len2 c1 c2 |
+	<primitive: 'primitiveCompareString' module: 'MiscPrimitivePlugin'>
+	<var: #string1 declareC: 'unsigned char *string1'>
+	<var: #string2 declareC: 'unsigned char *string2'>
+	<var: #order declareC: 'unsigned char *order'>
+
+	len1 := string1 size.
+	len2 := string2 size.
+	1 to: (len1 min: len2) do:
+		[:i |
+		c1 := order at: (string1 basicAt: i) + 1.
+		c2 := order at: (string2 basicAt: i) + 1.
+		c1 = c2 ifFalse: 
+			[c1 < c2 ifTrue: [^ 1] ifFalse: [^ 3]]].
+	len1 = len2 ifTrue: [^ 2].
+	len1 < len2 ifTrue: [^ 1] ifFalse: [^ 3].
+\end{code}
+
 \cba{Source code et generated C code et generated ASM code}
@@
 \section{Slang Plugin code}
-\label{SmartSyntaxAppendix}
+\label{PluginAppendix}
+
+\begin{code}
+primitiveCompareString
+	"ByteString (class) compare: string1 with: string2 collated: order"
+	<export: true>
+	| len1 len2 order string1 string2 |
+
+	<var: 'order' type: #'unsigned char *'>
+	<var: 'string1' type: #'unsigned char *'>
+	<var: 'string2' type: #'unsigned char *'>
+	((interpreterProxy isBytes: (interpreterProxy stackValue: 0))
+	and: [(interpreterProxy isBytes: (interpreterProxy stackValue: 1))
+	and: [interpreterProxy isBytes: (interpreterProxy stackValue: 2)]]) ifFalse:
+		[^interpreterProxy primitiveFailFor: PrimErrBadArgument].
+	string1 := interpreterProxy firstIndexableField: (interpreterProxy stackValue: 2).
+	string2 := interpreterProxy firstIndexableField: (interpreterProxy stackValue: 1).
+	order := interpreterProxy firstIndexableField: (interpreterProxy stackValue: 0).
+	len1 := interpreterProxy sizeOfSTArrayFromCPrimitive: string1.
+	len2 := interpreterProxy sizeOfSTArrayFromCPrimitive: string2.
+	(interpreterProxy sizeOfSTArrayFromCPrimitive: order) < 256 ifTrue:
+		[^interpreterProxy primitiveFailFor: PrimErrBadArgument].
+	0 to: (len1 min: len2) - 1 do: 
+		[ :i | | c1 c2 |
+		c1 := order at: (string1 at: i).
+		c2 := order at: (string2 at: i).
+		c1 = c2 ifFalse:
+			[^interpreterProxy methodReturnInteger: (c1 < c2 ifTrue: [1] ifFalse: [3])]].
+	interpreterProxy methodReturnInteger:
+		(len1 = len2 ifTrue: [2] ifFalse: [len1 < len2 ifTrue: [1] ifFalse: [3]])
+\end{code}
 
@@
 
+\begin{code}
+primitiveCompareWith
+	"<string1> primitiveCompareWith: string2 [collated: order] "
+	<export: true>	
+	| string1 string2 order strLength1 strLength2 result |
+	"1 - fetch the parameters from the stack"	
+	argumentCount = 1 ifFalse:
+		[argumentCount ~= 2 ifTrue:
+			[^self primitiveFailFor: PrimErrBadNumArgs].
+			 order := self stackTop.
+			 ((objectMemory isBytes: order)
+			 and: [(objectMemory numBytesOfBytes: order) = 256]) ifFalse:
+				[^self primitiveFailFor: PrimErrBadArgument]].
+	string1 := self stackValue: argumentCount.
+	string2 := self stackValue: argumentCount - 1. 		
+	"2 - check their types - all parameters are ByteObject"
+	((objectMemory isBytes: string1)
+	 and: [objectMemory isBytes: string2]) ifFalse: 
+		[^self primitiveFailFor: PrimErrBadArgument].
+	"3 - compare the strings"	
+	strLength1 := objectMemory numBytesOfBytes: string1.
+	strLength2 := objectMemory numBytesOfBytes: string2.
+	result := order 
+		ifNil: [self rawCompare: string1 length: strLength1 with: string2 length: strLength2 accessBlock:
+				[:str :index | objectMemory fetchByte: index ofObject: str ]]
+		ifNotNil: 
+			[self rawCompare: string1 length: strLength1 with: string2 length: strLength2 accessBlock:
+				[:str :index | objectMemory fetchByte: (objectMemory fetchByte: index ofObject: str) + 1 ofObject: order ]].
+
+	self methodReturnInteger: result
+\end{code}
+\begin{code}
+rawCompare: string1 length: strLength1 with: string2 length: strLength2 accessBlock: accessBlock
+	| c1 c2 min |
+	<inline: true> "needs to be forced else slang does not inline it by default"
+	min := strLength1 min: strLength2.
+	0 to: min-1 do: 
+		[:i | c1 := accessBlock value: string1 value: i.
+			c2 := accessBlock value: string2 value: i.
+			c1 = c2 ifFalse: [^c1 - c2]].
+	^strLength1 - strLength2
+\end{code}
+
 \cba{Source code et generated C code et generated ASM code}
@@
 
+\begin{code}
+genPrimitiveStringCompareWith
+	| instr jump jumpAbove jumpIncorrectFormat1 jumpIncorrectFormat2 jumpIncorrectFormat3 jumpIncorrectFormat4 jumpMidFailure jumpSuccess minSizeReg string1CharOrByteSizeReg string2CharOrByteSizeReg string1Reg string2Reg |
+	<var: #jumpIncorrectFormat1 type: #'AbstractInstruction *'>
+	<var: #jumpIncorrectFormat2 type: #'AbstractInstruction *'>
+	<var: #jumpIncorrectFormat3 type: #'AbstractInstruction *'>
+	<var: #jumpIncorrectFormat4 type: #'AbstractInstruction *'>
+	<var: #jumpAbove type: #'AbstractInstruction *'>
+	<var: #jumpSuccess type: #'AbstractInstruction *'>
+	<var: #jump type: #'AbstractInstruction *'>
+	<var: #jumpMidFailure type: #'AbstractInstruction *'>
+	"I redefine those name to ease program comprehension"
+	string1Reg := ReceiverResultReg.
+	string2Reg := Arg0Reg.
+	string1CharOrByteSizeReg := Arg1Reg.
+	string2CharOrByteSizeReg := ClassReg.
+	minSizeReg := SendNumArgsReg.
+	"Load arguments in reg"
+	cogit genLoadArgAtDepth: 0 into: string2Reg.
+	"checks if string1 is a byteobject and get its size in bytes"
+	self genGetFormatOf: string1Reg into: TempReg.
+	cogit CmpCq: objectMemory firstByteFormat R: TempReg.
+	jumpIncorrectFormat1 := cogit JumpLess: 0.
+	cogit CmpCq: objectMemory firstCompiledMethodFormat R: TempReg.
+	jumpIncorrectFormat2 := cogit JumpAboveOrEqual: 0.	
+	self genGetNumSlotsOf: string1Reg into: string1CharOrByteSizeReg.
+	(cogit LogicalShiftLeftCq: objectMemory shiftForWord R: string1CharOrByteSizeReg). 
+	cogit AndCq: objectMemory wordSize - 1 R: TempReg R: TempReg. 
+	cogit SubR: TempReg R: string1CharOrByteSizeReg. 
+	"checks if string2 is a byteobject and get its size in bytes"
+	self genGetFormatOf: string2Reg into: TempReg.
+	cogit CmpCq: objectMemory firstByteFormat R: TempReg.
+	jumpIncorrectFormat3 := cogit JumpLess: 0.
+	cogit CmpCq: objectMemory firstCompiledMethodFormat R: TempReg.
+	jumpIncorrectFormat4 := cogit JumpAboveOrEqual: 0.	
+	self genGetNumSlotsOf: string2Reg into: string2CharOrByteSizeReg.
+	(cogit LogicalShiftLeftCq: objectMemory shiftForWord R: string2CharOrByteSizeReg).
+	cogit AndCq: objectMemory wordSize - 1 R: TempReg R: TempReg.
+	cogit SubR: TempReg R: string2CharOrByteSizeReg.
+	"Type and number of arguments are correct"
+	"Compute the min"	 
+	cogit CmpR: string1CharOrByteSizeReg R: string2CharOrByteSizeReg.
+	jumpAbove := cogit JumpBelow: 0. 
+	cogit MoveR: string1CharOrByteSizeReg R: minSizeReg. 
+	jump := cogit Jump: 0. 
+	jumpAbove jmpTarget: (cogit MoveR: string2CharOrByteSizeReg R: minSizeReg). 
+	jump jmpTarget: (cogit CmpCq: 0 R: minSizeReg). 
+	jumpSuccess := cogit JumpZero: 0. "if one of the string is empty, no need to go through the comparing loop"
+	"Compare the bytes"
+	cogit MoveCq: objectMemory baseHeaderSize  R: TempReg.
+	cogit AddCq: objectMemory baseHeaderSize R: minSizeReg.
+	instr := cogit MoveXbr: TempReg R: string1Reg R: string1CharOrByteSizeReg.
+	cogit MoveXbr: TempReg R: string2Reg R: string2CharOrByteSizeReg.
+	cogit SubR: string2CharOrByteSizeReg R: string1CharOrByteSizeReg. 
+	jumpMidFailure := cogit JumpNonZero: 0. "the 2 compared characters are different, exit the loop"
+	cogit AddCq: 1 R: TempReg.
+	cogit CmpR: TempReg R: minSizeReg. 
+	cogit JumpNonZero: instr.
+	"all bytes from 1 to minSize are equal"
+	self genGetNumBytesOf: string1Reg into: string1CharOrByteSizeReg.
+	self genGetNumBytesOf: string2Reg into: string2CharOrByteSizeReg.
+	jumpSuccess jmpTarget: (cogit SubR: string2CharOrByteSizeReg R: string1CharOrByteSizeReg).
+	jumpMidFailure  jmpTarget: (cogit MoveR: string1CharOrByteSizeReg R: ReceiverResultReg).	
+	self genConvertIntegerToSmallIntegerInReg: ReceiverResultReg.
+	cogit genPrimReturn.
+	jumpIncorrectFormat4 
+		jmpTarget: (jumpIncorrectFormat3 
+			jmpTarget: (jumpIncorrectFormat2 
+				jmpTarget: (jumpIncorrectFormat1 jmpTarget: cogit Label))).
+	^ CompletePrimitive
+\end{code}
+
 \cba{Source code et generated ASM code}

### 2018-05-08 paper.tex
adds section about comparison

--- 2018-05-08 paper.tex
@@
 Some of these are essential and cannot be performed in any other way.
-Others are optional: they can be expressed in the client language but are often implemented in the virtual machine to improve performance when the just-in-time compiler is unable to do so (start-up performance, speculative optimizations not implemented or not mature enough, etc.).
+Others are optional, they can be expressed in the client language but are often implemented in the virtual machine to improve performance when the just-in-time compiler is unable to do so (start-up performance, speculative optimizations not implemented or not mature enough, etc.).
 
-In a hybrid runtime, where code is executed by an interpreter and a just-in-time compiler, the implementor can choose to implement optional primitives in the client language, in the virtual machine implementation language (typically C or C++), or on top of the just-in-time compiler back-end.
-This raises the questions of the tradeoffs of the different alternatives.
+In a hybrid runtime, where code is executed by an interpreter and a just-in-time compiler, the implementor can choose to implement optional primitives in the client language, in the virtual machine implementation language (typically C or C++) or on top of the just-in-time compiler back-end.
 We implemented the String comparison optional primitive in each case.
@@
 
-Without surprise, the paper shows that the implementation with the most reliable and high performance is also the most complex and difficult to implement.
-Since hundreds of optional primitives could be wished, the virtual machine implementor has to carefully choose how to implement each of them to balance between maintenance cost and high performance.
+The paper shows that the implementation with the most reliable and high performance is also the most complex and difficult to implement.
+Since hundreds of optional primitives can be available, the virtual machine implementor has to carefully choose how to implement each of them to balance between maintenance cost and high performance.
+
 \end{abstract}
@@
 
-High-level object-oriented programming languages are often implemented on top of a virtual machine (VM).
-As VMs has become more popular, different techniques have been set up using Just-In-Time (JIT) compilation to improve the runtime overall performance: method JITs \cite{ursPHD} which usually compile methods to native code and tracing JITs \cite{Dynamo, PyPyTracing} which usually compile linear traces of execution into native code.
+High level object oriented programming languages are often implemented on top of a virtual machine (VM).
+As VMs has become more popular, different techniques have been set-up using Just-In-Time (JIT) compilation to improve the overall performance of the runtime such as method JITs \cite{ursPHD} which usually compile methods to native code and tracing JITs \cite{Dynamo, PyPyTracing} which usually compile linear traces of execution into native code.
 
-On top of its virtual machine, the client language can use a set of \emph{primitives}\footnote{We use in the paper the Smalltalk terminology (primitive) as this is the programming language used for our evaluation.
+On top of its virtual machine, the client language can use a set of \emph{primitives}~\footnote{We use in the paper the Smalltalk terminology (primitive) as this is the programming language used for our evaluation.
 Some other programming languages, such as Javascript, prefer to use the term built-in instead of primitive.}, which are performed directly by the interpreter rather than by evaluating expressions in a method.
@@
 	\item \emph{Essential primitives} cannot be performed in any other way.
-A high-level object-oriented language without primitives can move values from one variable to another, but cannot add two integers together.
+A high-level object oriented language without primitives can move values from one variable to another, but cannot add two integers together.
 Many arithmetic and comparison operations between numbers are primitives.
-Some primitives allow one to communicate with I/O devices such as the disk, the display and the keyboard.
-	\item \emph{Optional primitives} exist only to make the system run faster \cite{blueBook}.
-They can be implemented in the client language directly, making its implementation as a primitive optional, but they are often implemented directly in the VM to improve performance.
+Some primitives allow to communicate with I/O devices such as the disk, the display and the keyboard.
+	\item \emph{Optional primitives} exist only to make the system run faster.
+They can be implemented in the client language directly, making its implementation as a primitive optional, but they are implemented in the VM to improve performance.
 \end{itemize}
@@
 The implementation of an optional primitive in the VM rather than the client language improves performance but also often increases the implementation engineering cost.
-For example, accessing objects in the VM usually requires understanding of the memory layout or implementation details, which are not needed in the client language.
+For example, accessing objects in the VM usually requires some understanding of the memory layout or implementation details, which are not needed in the client language.
 
 For this reason, some virtual machine implementors attempted to remove most or all optional primitives.
-\sd{sentence was too long: please reread my version:}Their goal was to use a JIT performing speculative optimizations of optional primitives implemented in the client language.
-Ideally we wanted to get   the performance of the same primitives written in the VM.
+Their goal was to improve the execution of the client language with a JIT performing speculative optimizations \cite{ursPHD} for the execution of the optional primitives implemented in the client language to be as fast as the performance of the same primitives written in the VM.
 Self \cite{ursPHD} and Strongtalk \cite{Strongtalk} were \sk{the?} to first to try this approach.
@@
 VMs are traditionally implemented in a low-level language such as C or C++.
-To balance between memory footprint, start-up performance and peak performance, the execution of code is usually done through multiple execution tiers: the first few executions of a code snippet are done through an interpreter and the JIT compiler \sd{optimizes} at runtime the frequently used code snippets \sd{rm: for further executions to be performed quicker}.
-
+To balance between memory footprint, start-up performance and peak performance, the execution of code is usually done through multiple execution tiers: the first few executions of a code snippet are done through an interpreter and the JIT compiler at runtime the frequently used code snippets for further executions to be performed quicker.
 This leads to annoying concerns: for example, let's say we implement a primitive in C.
@@
 This is problematic since the native code generated by the JIT is not directly compatible with native code generated by the C compiler: a call from one to another may require to edit the stack pointer and the frame pointer from the client stack to the C stack and to spill or move multiple registers.
-Switching between both usually wastes up to around a dozen native instructions.
+Switching between both can waste up to around a dozen native instructions.
 If the primitive is going to execute many instructions, the switch overhead might be negligible, but if the primitive executes only a few instructions, the overhead can be noticeable.
-
-\sd{the link is missing: In this article we want to measure the complexities and gain of this different implementations}
 
@@
 The exact specifications are detailed in Section \ref{subsec:primSpec}.
-We implemented a first version in Pharo/Smalltalk.
+We implemented a first version in Smalltalk.
 We then implemented multiple version compiling through C compilers to native code, using different part of our VM infrastructure.
@@
 Our evaluation is based on the Cog VM a Smalltalk virtual machine.
-The Cog VM is based on Dan Ingalls' Smalltalk VM \cite{BackToTheFuture} and \sd{has} have been enhanced with a JIT \cite{CogJIT} compiling one method or closure at a time to native code.
+The Cog VM is based on Dan Ingalls' Smalltalk VM \cite{BackToTheFuture} and have been enhanced with a JIT \cite{CogJIT} compiling one method or closure at a time to native code.
 Historically, the VM implemented a 32-bits adaptation of the Smalltalk-80 specifications \cite{blueBook}, but the Cog VM is now the default VM for multiple programming languages such as Pharo \cite{PharoByExample}, Squeak \cite{SqueakByExample} and Newspeak \cite{NewspeakOopsla}.
@@
 In Section \ref{sec:primExec}, we detail how primitives are executed.
-Section \ref{subsec:primSpec} discusses the string comparison primitive specifications  used for our evaluation.
+Section \ref{subsec:primSpec} discusses the specifications of the string comparison primitive used for our evaluation.
 
@@
 The execution engine (the memory manager, the interpreter and the baseline JIT) are entirely written in Slang.
-The two main purposes of using Slang code over plain C are:
+The two main purposes of using slang code over plain C are:
 \begin{itemize}
@@
 
-The executable is generated in two steps as shown on Figure~\ref{fig:VMCompilation}, similarly to the RPython toolchain \cite{RPythonToolchain}.
+The executable is generated in two steps as shown on Figure \ref{fig:VMCompilation}, similarly to the RPython toolchain \cite{RPythonToolchain}.
 The first step is to generate the two C files representing the whole execution engine written in Slang using the Slang-to-C compiler.
 During the second step, a C compiler is called to compile the execution engine and the platform-specific code written directly in C to the executable VM.
-
-\cba{We need to introduce here the concept of VM plugin and interpreterproxy indirection/overhead. This is very important since later we just say this is a plugin and we don't redefine it}
 
@@
 The VM code, including the interpreter code, is compiled by a C compiler and uses the C stack at runtime.
-During its execution, the interpreter modifies the execution stack of the client language, which is different from its own code stack as shown on Figure~\ref{fig:TwoStacks}.
+During its execution, the interpreter modifies the execution stack of the client language, which is different from its own code stack as shown on Figure \ref{fig:TwoStacks}.
 This means for example that the frame pointer register in the processor refers to a stack frame on the C stack and that a processor push instruction pushes a value on the C stack.
@@
 
-To solve this performance bottleneck, the Cog VM allows one to redefine primitives in Cog's Register Transfer Language (RTL), an abstract assembly which is compiled to native code through Cog's JIT back-end.
+To solve this performance bottleneck, the Cog VM allows to redefine primitives in Cog's Register Transfer Language (RTL), an abstract assembly which is compiled to native code through Cog's JIT back-end.
 Such primitives are then generated to native code at runtime when a method annotated with the primitive is compiled to native code.
@@
 
-\paragraph{String representations.} Strings are represented on top of the Cog VM in two possible forms: ByteStrings and WideStrings.
+\paragraph{String representations.} Strings are represented in all programming languages on top of the Cog VM in two possible forms, ByteStrings and WideStrings.
 ByteStrings encode each character of the string in a single byte.
 The encoding usually follows the extended ASCII standard.
-WideStrings encode each character in 32 bits.
+WideStrings encode each character of the string in 32 bits.
 The encoding usually follows Unicode specifications.
@@
 \paragraph{Original specification.}  The original primitive took three operands.
-The two first operands were the two ByteStrings to compare and the third operand was the order table, a ByteArray encoding the order of characters.
+The two first operands were the two ByteStrings to compare and the third operand is the order table, a ByteArray encoding the order of characters.
 The default order is the ASCII order, hence each entry corresponds to its index in the byte array (The ByteArray has a size of 256, and accessing the element i, with 0 $\leq$ i $\leq$ 255, answered i).
@@
 In this case, the order ByteArray is different, since some entries answer the same value (for example, the entry for the 41 and 61, respectively characters A and a, both answer the same value).
-To compare two ByteStrings, the primitive conceptually iterates byte by byte over the two objects until it reaches the end of one them, comparing for each byte that the value present in the order ByteArray at the index is greater, equal or smaller than the value computed from the other ByteString.
+To compare the two ByteStrings, the primitive conceptually iterates byte by byte over the two objects until it reaches the end of one them, comparing for each byte that the value present in the order ByteArray at the index is greater, equal or smaller than the value computed from the other ByteString.
 If a difference is met, the value returns based on the result, if not, at the end of the iteration the primitive compares the ByteString sizes and returns.
-Based on the three parameters, the primitive answered 3, 2, or 1 depending if the first ByteString was greater, equal or smaller than the second one.
+Based on the three parameters, the primitive answered 3,2 or 1 depending if the first ByteString was greater, equal or smaller than the second one.
 
@@
 \begin{itemize}
-	\item \emph{Returned value:} Other programming languages implement the primitive to return a negative integer, 0 or a positive integer instead of 1, 2 or 3.
-It looks like a minor difference, but it is important for performance.
+	\item \emph{Returned value:} Other programming languages implement the primitive to return a negative integer, 0 or a positive integer instead of 1,2 or 3.
+It looks like a minor difference, but it is quite important for performance.
 With a negative, 0 or positive value, the primitive can just answer the difference between two different characters or the difference in size if all first characters are equal.
-With 1, 2, or 3, the primitive has to end with multiple branches to know what number to return.
+With 1,2 or 3, the primitive has to end with multiple branches to know what number to return.
 In native code, a subtraction is on common architecture quicker to perform than one or multiple branches.
@@
 Therefore, the VM does not need to go through the indirection order array in the common case.
-The primitive answers a negative integer, zero, or a positive integer as the first parameter is less than, equal to, or greater than the second parameter.
-
-In the paper we will focus on the comparison of two strings with the default ASCII ordering and measure the performance only of this case.
-The goal of this work was to improve the performance of common string operations, typically string equality which internally use the primitive, and the cases where the order is non ASCII are uncommon.
+The primitive answers a negative value if the first parameter is smaller than the second, 0 if they are equal, and a positive value if the first parameter is greater than the second one.
+\cba{Maybe rewrite the last paragraph to make it clearer / more formal. - CF Java spec ?}
 
@@
 
- This section first describes the different ways of implementing a primitive. We then explain how each of these versions are executed by the VM and the pros and cons of each and every.
+ This section first describes the different ways of implementing a primitive. We then explain how each of these version are executed by the VM and the pros and cons of each and every.
 
@@
 \paragraph{Baseline: pure Smalltalk.}
-To evaluate the primitive performance, we use as baseline an implementation in pure Smalltalk.
-There are multiple ways of writing the comparison of two strings in Smalltalk, following Smalltalk coding conventions, a standard developer would likely use high-level constructs which apply a closure on each element of the strings.
-We did not use high-level construct and chose to implement the Smalltalk version the most optimized way Smalltalk allows, \emph{i.e.,} we used the contructs to:do: and ifFalse:, which are both recognized by the Smalltalk to bytecode compiler and compiled respectively to loops and branches at the bytecode level.
-The implementation follows the new specification, hence two methods are available, baselineCompareWith: and baselineCompareWith:order:.
-Since the focus is only on ASCII ordered comparison, we show the code only of baselineCompareWith:.
+: baseline
+Smalltalk code on top of VM. Used as our baseline for benchmarks.
 
 \begin{code}
-baselineCompareWith: aString 
+compareWithPureST: aString
+
 	| c1 c2 length1 length2 |
@@
 		1 to: (length1 min: length2) do: [ :i |
-		(c1 := self basicAt: i) = (c2 := aString basicAt: i) ifFalse: [ ^ c1 - c2 ] ].
-	^ length1 - length2
+			(c1 := self basicAt: i) = (c2 := aString basicAt: i) 
+			ifFalse: [
+				c1 < c2
+					ifTrue: [ ^1 ]
+					ifFalse: [ ^3 ] ] ].
+		length1 = length2 ifTrue: [ ^2 ].
+		length1 < length2
+			ifTrue: [ ^1 ]
+			ifFalse: [ ^3 ] 
 \end{code}
 
+Expliquer to:do: over do: (to explain expresiveness) => core lib often use to:do: and not do:
+New spec
+
 \paragraph{Existing: Smart Syntax Plugin.}
+Restricted Smalltalk
 
-The Smart Syntax Plugin is one of the version existing until we started this work.
-The idea behind Smart Syntax was the following: the programmer can write code in Smalltalk using exclusively constructs that are recognized by the Smart Syntax compiler and that can be directly compiled to C (computing the size of a byte array, accessing its fields and doing integer arithmetic for example).
-With the addition of type annotations in the form of pragmas \cite{Duca16a} \footnote{Pragma are a form of method annotation.}, the Smart Syntax plugin was able to generate C code from the Smalltalk code.
-With this approach, the programmer has to write the optional primitive code a single time in Smalltalk.
-This code is used both as the fall-back code if the VM does not implement the primitive and at Slang to C compilation time to generate a C plugin from the code and the type annotations.
-This primitive follows the old specification.
+written in Smalltalk with restrictions.
+Compiled to C AOT into a plugin.
+Write once, fall-back + compiled.
+
+old spec
 
 \paragraph{Existing: Slang Plugin.}
+Slang, but no access au fonctions de la VM drectement.
+ interpreterProxy. (C compiler inlining, linking time ? cmacro ?)
 
-The Slang plugin is a recent complete rewrite of the Smart Syntax plugin.
-The main difference is that the Smalltalk version is used only as fall-back code, while at Slang to C compilation time the VM uses code written in Slang to generate the C code.
-One of the side-effect is that the plugin's code is now part of the VM code base and not the client code bases, which is convenient for maintenance since the Cog VM has multiple client languages.
-However, the implementor now needs to maintain and evolve two version of the primitive, the fall-back code in Smalltalk and the code in Slang.
-This primitive follows the old specification.
+ old spec
 
@@
 
-\cba{expliquer tableau comparatif}
-
 \begin{table}
@@
   \hline
-    & \rot{readability} & \rot{maintenance} & \rot{Engineering time} & \rot{re-compilation} & \rot{plugin} & \rot{stack switching} & \rot{Execution time} \\
+    & \rot{readability} & \rot{maintenance} & \rot{Engineering time} & \rot{re-compilation} & \rot{plugin} & \rot{stack switching} & \rot{Performance} \\
   \hline
@@
 
-One or the other implementation could be chosen over another one, depending on what one's aiming at: the main trade-off will be between the engineering time and the performance.
-The table \ref{tab:cmp_implem} summarizes the main pros and cons related to each implementation
-
-Trade-off engineering time and performance.
-- recompiler la VM ou pas (Plugin externes, code Smalltalk)
-- complexite d'implem (comprehension Smalltalk, C, ASM)
+One or another implementation could be preferred depending on the intended aim: the main trade-off will be found between the engineering time and the performance.
+The table \ref{cmp_implem} summarizes the main pros and cons related to each implementation: the engineering time may tend to increase when more low-level languages are used, such as Cog's RLT for instance.
+Also, a VM recompilation affects the overall complexity of the development process, but the resulting code might be faster to run.
+Likewise, the use of plugins may bring modularity, but as you will see in Section \ref{sec:eval}, the performance, although better than baseline bersion, could still be improved.
 
@@
 \subsection{Set-up}
-
-Evaluaiton on x86 and not x64.
-
+Describe set-up:
 GCC eval Ubuntu + machine spec
-
-The LLVM-Mac evaluaiton was performed on a MacBook pro with Mac OS 10.11.6, a 2.9Ghz processor Intel Core 5, 8 Gb 1867MHz DDR3 of RAM.
-The Mac VM was compiled using Apple LLVM version 8.0.0 (clang-800.0.38).
+LLVM eval Mac OS X + machine spec.
 
@@
 
-This difference is mainly due to the optimizations performed by both the compilers: as depicted in Figure \ref{fig:Asm} GCC generates instructions that are very close to the ones generated by the JIT, hence the similar speed-up factors.
+This difference is mainly due to the optimizations performed by both the compilers: as depicted in Figure \ref{Asm} GCC generates instructions that are very close to the ones generated by the JIT, hence the similar speed-up factors.
 On the contrary, LLVM generates more instructions: it especially fails at perserving the registries holding the strings (edi and esi), leading to 2 extra instructions (bolded in the figure) that will load the strings at each iteration.
@@
 
-\cba{I think in the fig we should also see the ASM cog generate}
+\cba{appendix with full code}
 
@@
 %% Bibliography
-\bibliography{sista,rmod}
+\bibliography{sista}
 

### 2018-05-08 paper.tex
new criterion for comparison table

--- 2018-05-08 paper.tex
@@
 
-  Can be changed like Smalltalk code (No C compilation whatsoever). Readability. Only impl without VM recompilation.
-
-Baseline performance is not that great.
-
-Number of implementations to maintain: 1, 2 or 3: same implem for primitive in C and fall back code in Smalltalk, relatively easy to read/code.
-recompilation. 
-changing the prim => changing all clients (VM compil from Pharo or Squeak different results
-Performance: cross file inlining not possible by default. (InterpreterProxy indirection)
-Control on generated code => difference LLVM/GCC
- Pure C performance (no interpreter proxy).
- no C-St jump. Control over the outcome (reliable performance)
- costly jumps between Smalltalk and C runtime (trampolines), need to recompile vm sources, little control over outcome (c compilers 
- need to recompile vm sources (trampolines), code readability
+ - Requires recompilation of VM or plugin
+ - Implementation language familiar to client dev
+ - Implementation language familiar to vm dev
+ - readability: number of lines of code ? 
+ - Number of impl to maintain 1,2,3 or more if multiple clients.
+ - Performance: stack switch, C performacne, cross file inlining, (InterpreterProxy indirection)
+ - Control on generated code, diff llvm gcc
 
  \subsection{Comparison}
-
-\cba{expliquer tableau comparatif}
 
@@
 \centering
-\begin{tabular}{|c|c|c||c||c|c|c||c|}
+\begin{tabular}{cccccccc}
   \hline
-    & \rot{readability} & \rot{maintenance} & \rot{Engineering time} & \rot{re-compilation} & \rot{plugin} & \rot{stack switching} & \rot{Execution time} \\
+    & \rot{recompilation needed} & \rot{familiar to client dev} & \rot{familiar to vm dev}} & \rot{number of LoC} & \rot{Implementations to maintain} & \rot{Performance} & \rot{Control over generated code} \\
   \hline
-  Baseline    & X &   & +   &    &   &   & +++ \\
-  SmartSyntax & X &   & ++  &    & X & X & ++\\
-  SlangPlugin &   &   & ++  &    & X & X & ++ \\
-  PureSlang   &   & X & +++ &  X &   & X & ++\\
-  Slang+JIT   &   & X & +++ &  X &   &   & + \\
+  Baseline    &             & \checkmark  &            & 10  & 3+ &  +  &  \\
+   \hline
+  SmartSyntax &             &  \checkmark &            & 20  & 3+ & ++  & -\\
+   \hline
+  SlangPlugin &             &  \checkmark &            & 35  & 2  & ++  & - \\
+   \hline
+  PureSlang   & \checkmark  &             & \checkmark & 45  & 1  & ++  & -\\
+   \hline
+  Slang+JIT   & \checkmark  &             & \checkmark & 85  & 1  & ++++ & + \\
   \hline
@@
 
-One or the other implementation could be chosen over another one, depending on what one's aiming at: the main trade-off will be between the engineering time and the performance.
-The table \ref{tab:cmp_implem} summarizes the main pros and cons related to each implementation
-
-Trade-off engineering time and performance.
-- recompiler la VM ou pas (Plugin externes, code Smalltalk)
-- complexite d'implem (comprehension Smalltalk, C, ASM)
+One or another implementation could be preferred depending on the intended aim: the main trade-off will be found between the engineering time and the performance.
+The table \ref{cmp_implem} summarizes the main pros and cons related to each implementation: the engineering time may tend to increase when more low-level languages are used, such as Cog's RLT for instance.
+Also, a VM recompilation affects the overall complexity of the development process, but the resulting code might be faster to run.
+Likewise, the use of plugins may bring modularity, but as you will see in Section \ref{sec:eval}, the performance, although better than baseline version, could still be improved.
 
@@
 
-All these versions of the primitives have been implemented. Compare performances
+The SlangPlugin, PureSlang and Slang+JIT implementations of this string comparison primitive have been recently (re)implemented in the Cog VM.
+In this section, we will compare the performances of these primitives against the other existing implementations performances.
+We have measures their respective execution times for different string sizes.
 
@@
 
-Evaluaiton on x86 and not x64.
-
-GCC eval Ubuntu + machine spec
-
-The LLVM-Mac evaluaiton was performed on a MacBook pro with Mac OS 10.11.6, a 2.9Ghz processor Intel Core 5, 8 Gb 1867MHz DDR3 of RAM.
+The GCC-Linux evaluation was performed on an Asus Zenbook with Ubuntu 16.04 LTS, a 2,2 Ghz processor Intel Core 5, 8 Gb 1600MHz DDR3 of RAM. The Linux VM was compiled using GCC version 5.4.0.
+The LLVM-Mac evaluation was performed on a MacBook pro with Mac OS 10.11.6, a 2.9Ghz processor Intel Core 5, 8 Gb 1867MHz DDR3 of RAM.
 The Mac VM was compiled using Apple LLVM version 8.0.0 (clang-800.0.38).
+The evaluation was performed on 32 bits VMs.
 
@@
 \item The Slang+JIT implementation is more efficient in every case. Nevertheless,
-the performance gap between this implementation and the PureSlang one tends to decrease as the string size increase. The comparison loop \sk{ref to appendix}
-\end{itemize} proves to be more and more time-consuming because of the higher number of iterations.
+the performance gap between this implementation and the PureSlang one tends to decrease as the string size increase.
+The comparison loop (\ref{fig:Asm}) proves to be more and more time-consuming because of the higher number of iterations.
+\end{itemize}
 
@@
 
-\begin{figure}[htp!]
+\begin{figure*}[htp!]
 		\centering
-		\includegraphics[width=0.99\linewidth]{figures/cmpasm.png}
+		\includegraphics[width=1\linewidth]{figures/asm_cmp.pdf}
 		\caption{GCC and LLVM: different optimizations on the comparison loop}
 		\label{fig:Asm}
-\end{figure}
-
-\cba{I think in the fig we should also see the ASM cog generate}
-
-\subsection{Macro-benchmarks}
-Intro sentence with ref to squeak speed center.
-JSON
+\end{figure*}
 

### 2018-05-08 paper.tex
add mention of numbered and named primitives

@@
 
+The primitives can be associated either with a name, either with an index in the Cog VM. The named primitives can be added and modified without the need to recompile the VM.
+
 \subsection{String comparison primitive}
@@
 The Slang implementation is written inside the interpreter and has access to all interpreter APIs.
-Since the same primitive index \cba{Concept to introduce before or rephrase} is used for the primitive with 2 or 3 arguments, the primitive needs to deal with both cases in a way that the common case is efficient.
+Since the same primitive index is used for the primitive with 2 or 3 arguments, the primitive needs to deal with both cases in a way that the common case is efficient.
 The primitive is now part of the core of the VM, it cannot be easily added or removed as plugin primitives, but it is simpler to read and write.

### 2018-05-08 paper.tex
add small section about plugins and interpreter proxy

@@
 \cba{We need to introduce here the concept of VM plugin and interpreterproxy indirection/overhead. This is very important since later we just say this is a plugin and we don't redefine it}
+\sk{done but someone better check what I wrote :-)}
+
+VM plugins enable the addition of new features to the VM. There are 2 types of plugins: the built-ins and the externals. 
+\begin{itemize}
+\item External plugins. Plugins that can be added or removed from the VM without the need to recompile the VM sources.
+\item Built-in plugins.
+Plugins that cannot be installed without recompiling the VM sources.
+However, these plugins can reach some parts of the VM internals\footnote{as data structures, routines...} that remains unreachable from an external plugin.
+These operations are performed via an InterpreterProxy.
+Although, this solution results in a substential overhead if called often.
+\end{itemize}
 

### 2018-05-09 paper.tex
various fixes here and there, fixed refs

@@
 
- \subsection{Pros and cons}
-
- - Requires recompilation of VM or plugin
- - Implementation language familiar to client dev
- - Implementation language familiar to vm dev
- - readability: number of lines of code ? 
- - Number of impl to maintain 1,2,3 or more if multiple clients.
- - Performance: stack switch, C performacne, cross file inlining, (InterpreterProxy indirection)
- - Control on generated code, diff llvm gcc
-
  \subsection{Comparison}
+
+ In this section we compare the five different implementations according to the following criteria:
+ \begin{itemize} 
+ 	\item \emph{Recompilation needed:} Diff no, plugin, full
+	\item \emph{Implementation language familiar to developers:} Merged the 2 latter, explain ST dev ST, VM dev ST-Slang-C-ASM, maybe result is + ou - ? or + or 0
+	\item \emph{Readability:} explain why number of lines of code. 
+	\item \emph{Number of implementations to maintain:} explain fall-back code which is in all languages, Slang/C, JIT => maybe we ignore the fall-back code ?
+	\item \emph{Stack switch from generated native code:} 
+	\item \emph{Optimization potential:} discuss inlining C comp cross file and InterpreterProxy indirection
+	\item \emph{Control over the generated code:} diff llvm gcc while JIT always the same
+\end{itemize}
 
@@
 \centering
-\begin{tabular}{cccccccc}
+\begin{tabular}{lccccccc}
   \hline
-    & \rot{recompilation needed} & \rot{familiar to client dev} & \rot{familiar to vm dev}} & \rot{number of LoC} & \rot{Implementations to maintain} & \rot{Performance} & \rot{Control over generated code} \\
+    & \rot{recompilation needed} & \rot{familiar to client dev} & \rot{familiar to vm dev} & \rot{number of LoC} & \rot{Implementations to maintain} & \rot{Performance} & \rot{Control over generated code} \\
   \hline
@@
 One or another implementation could be preferred depending on the intended aim: the main trade-off will be found between the engineering time and the performance.
-The table \ref{cmp_implem} summarizes the main pros and cons related to each implementation: the engineering time may tend to increase when more low-level languages are used, such as Cog's RLT for instance.
+Table \ref{tab:cmp_implem} summarizes the main pros and cons related to each implementation: the engineering time may tend to increase when more low-level languages are used, such as Cog's RTL for instance.
 Also, a VM recompilation affects the overall complexity of the development process, but the resulting code might be faster to run.
@@
 
-The SlangPlugin, PureSlang and Slang+JIT implementations of this string comparison primitive have been recently (re)implemented in the Cog VM.
+The SlangPlugin, PureSlang and Slang+JIT implementations of this string comparison primitive have been recently implemented in the Cog VM.
 In this section, we will compare the performances of these primitives against the other existing implementations performances.
@@
 
-The GCC-Linux evaluation was performed on an Asus Zenbook with Ubuntu 16.04 LTS, a 2,2 Ghz processor Intel Core 5, 8 Gb 1600MHz DDR3 of RAM. The Linux VM was compiled using GCC version 5.4.0.
-The LLVM-Mac evaluation was performed on a MacBook pro with Mac OS 10.11.6, a 2.9Ghz processor Intel Core 5, 8 Gb 1867MHz DDR3 of RAM.
+\paragraph{GCC-Linux.}The GCC-Linux evaluation was performed on an Asus Zenbook with Ubuntu 16.04 LTS, a 2,2 Ghz processor Intel Core 5, 8 Gb 1600MHz DDR3 of RAM.
+The Linux VM was compiled using GCC version 5.4.0.
+
+\paragraph{LLVM-Mac.}The LLVM-Mac evaluation was performed on a MacBook pro with Mac OS 10.11.6, a 2.9Ghz processor Intel Core 5, 8 Gb 1867MHz DDR3 of RAM.
 The Mac VM was compiled using Apple LLVM version 8.0.0 (clang-800.0.38).
-The evaluation was performed on 32 bits VMs.
+
+\paragraph{Processor.}The following evaluation was performed on 32 bits Intel VMs (x86).
 
@@
    \end{subfigure}
-   \subcaption*{\textbf{LLVM-Mac results}}
+   \subcaption*{\textbf{LLVM-Mac results}\vspace{-0.5cm}}
     \end{subfigure}
+   \caption{String comparison benchmarks for strings of different sizes on different architectures}
    \label{fig:bench}
@@
 
-However, one could notice a rather striking performance difference between the 2 compilers: when compiling with GCC, we notive in Figure \ref{gcc} that the performances between the PureSlang and Slang+JIT implementations are really close for long strings (100 characters and longer), with respectively 24,42 and 24,48 relative speed-up factors for a 1000-characters string.
+However, one could notice a rather striking performance difference between the 2 compilers: when compiling with GCC, we notive in Figure \ref{fig:bench} that the performances between the PureSlang and Slang+JIT implementations are really close for long strings (100 characters and longer), with respectively 24,42 and 24,48 relative speed-up factors for a 1000-characters string.
 Yet they greatly differ when LLVM is used, with respectively 13,44 and 26,40 relative speed-up factors.
@@
 		\includegraphics[width=1\linewidth]{figures/asm_cmp.pdf}
-		\caption{GCC and LLVM: different optimizations on the comparison loop}
+		\caption{GCC and LLVM: different optimizations on the comparison loop \cba{I want another column with comments, what does each line do.
+I need some different representation, no addresses, labels for jumps, something standard would be nice.}}
 		\label{fig:Asm}
@@
 				jmpTarget: (jumpIncorrectFormat1 jmpTarget: cogit Label))).
-	^ CompletePrimitive
+	^ CompletePrimitive}
 \end{code}

### 2018-05-09 paper.tex
add extra collumn to the asm comparison table

@@
 \usepackage{xcolor}
+\newcommand{\todo}[1]{\color{orange}\fbox{\bfseries\sffamily\scriptsize TODO:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
 \newcommand{\sd}[1]{\color{red}\fbox{\bfseries\sffamily\scriptsize Stef:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
@@
 Our evaluation is based on the Cog VM a Smalltalk virtual machine.
-The Cog VM is based on Dan Ingalls' Smalltalk VM \cite{BackToTheFuture} and \sd{has} have been enhanced with a JIT \cite{CogJIT} compiling one method or closure at a time to native code.
+The Cog VM is based on Dan Ingalls' Smalltalk VM \cite{BackToTheFuture} and has been enhanced with a JIT \cite{CogJIT} compiling one method or closure at a time to native code.
 Historically, the VM implemented a 32-bits adaptation of the Smalltalk-80 specifications \cite{blueBook}, but the Cog VM is now the default VM for multiple programming languages such as Pharo \cite{PharoByExample}, Squeak \cite{SqueakByExample} and Newspeak \cite{NewspeakOopsla}.
@@
  \subsection{Comparison}
+
+ \todo{Finish section}
 
@@
 
+\subsection{Methodology}
+
+\todo{fill methodology section}
+\cba{missing here: what are the benchmark actually run ?
+Can be short code sample or explanation of both.
+It has to include: each bench was run x times, each time being y comparison.
+We removed the loop overhead in the measurements.}
+\cba{+ we bench only = strings, which could make a difference on the plugin version (one less branch) - maybe we should change to number of character compared instead of string size so it does not imply the string are equals.}
+\cba{SM methodology: what is done? which benchmarks are used? how are they executed? what is measured? how do you address warmup etc? what kind of machine is used for the measurements?}
+
+\todo{Add exact results}
+\cba{We are also missing exact results, aside from the graphs I usually expect a table with exact results and error margins...
+That's why the plot is often generated from a file, the table can reuse the same file and you can re-do your bench without changing the latex code.
+Cf https://hal.inria.fr/hal-01596321/document page 9, it would be nice here to have a similar table below the benchs}
+
 \subsection{Micro-benchmarks}
@@
 
-\begin{figure*}
+\begin{figure*}[bht]
   \centering
@@
    \end{subfigure}
-   \subcaption*{\textbf{LLVM-Mac results}\vspace{-0.5cm}}
+   \subcaption*{\textbf{LLVM-Mac results}\vspace{-0.2cm}} %change vspace based on fig position
     \end{subfigure}
@@
 %\cba{Je pense string de taille 0-1 c'est cool pour montrer que le code C va moins vite que pur smalltalk dans ce cas (si c'est vraiment le cas)}
-We performed the benchmarks upon 4 different string sizes: 0, 5, 100 and 1000 characters, using 2 different compilers: GCC on one hand and LLVM on the other hand.
+We performed the benchmarks upon five different string sizes: 0, 1, 5, 100 and 1000 characters, using two different compilers: GCC on one hand and LLVM on the other hand.
 Figure \ref{fig:bench} show the results in terms of relative speed-up towards the baseline implementation\footnote{i.e.
@@
 		\includegraphics[width=1\linewidth]{figures/asm_cmp.pdf}
-		\caption{GCC and LLVM: different optimizations on the comparison loop \cba{I want another column with comments, what does each line do.
-I need some different representation, no addresses, labels for jumps, something standard would be nice.}}
+		\caption{GCC and LLVM: different optimizations on the comparison loop}
 		\label{fig:Asm}
@@
 
-\cba{Maybe talk in those cases the number of representation of the primitives}
+\cba{Maybe talk in those cases the number of representation of the primitives, this is a bit disconnected with the rest of the paper I feel}
 
@@
 \section{Future work and conclusion}
-1. number of representations\\
-2. common representation
-\cba{Future work -> primitive processor dependent (i.e. SSE4 instrs), pros and cons and why you did not do it (ingeniering cost per back-end)}
-\cba{Inlining in Sista + ref}
-\cite{SistaFastStartUp}
-
-Conclusion: trade off expressiveness performance
+
+In this section we discuss two future work directions and conclude.
+
+\paragraph{Future work: back-end specific.}
+
+All the primitive we compare are written in a processor independent way, due to the C compiler abstraction for the Slang versions, the abstract assembly abstraction for Cog's RTL version or the VM abstraction for the baseline implementation in Smalltalk.
+We could rewrite the primitive differently for each back-end currently supported (x86, x64, MIPSEL, ARMv6) and yet to support (ARMv8, ...).
+It could be relevant in some cases performance-wise, for example, on intel processors, the JVM implements string comparison using the SSE4.2 string comparison instructions when available \cite{JVMSSE4StringComp}.
+We did not go in this direction because the maintenance cost in this case is definitely too high since we would have to maintain a different implementation per back-end  (already 4 implementations today).
+
+\paragraph{Future work: adaptive optimizer integration.}
+
+For the past few years, an experimental adaptive optimizer (a JIT featuring advanced optimizations such as speculative inlining) has been developed on top of the Cog VM \cite{SistaFastStartUp}.
+The optimizer has specific rules to be able to inline primitive operations without performance loss.
+We need to carefully measure, in different benchmarks, the performance of the inlined string comparison primitive.
+We also need to discuss the maintenance cost, for example, do we need a new representation of the primitive for the optimizing JIT to inline it?
+The micro-benchmarks used in the paper were relevant to show the client stack to C stack switch overhead as well as baseline performance.
+In the optimizing JIT context, such micro-benchmarks are not really relevant anymore since the optimizing JIT usually entirely optimizes away such benchmarks and real-application performance depends on how well the primitive is inlined and specialized using information from the optimized method (typically, if one operand of the string comparison is a constant string, can the generated code benefit from this knowledge?).
+
+\paragraph{Conclusion.}
+
+In this paper we compare five different implementations of the string primitives on top of the Cog VM.
+Choosing the right implementation for one's VM is a trade-off between maintenance cost and execution time, the quickest to execute the primitive is, the harder it is to maintain.
 
@@
 
+\todo{Add some intro sentence here and there + maybe ASM in appendix}
 %% Appendix

### 2018-05-09 paper.tex
fixed appendixes

@@
 
-\todo{Add some intro sentence here and there + maybe ASM in appendix}
 %% Appendix
+\vfill\null %switch column
 \appendix
@@
 \label{SmartSyntaxAppendix}
+
+Source code of the smart syntax method, which is used to generate the C code for the plugin but also as Smalltalk fall-back code.
 
@@
 
-\cba{Source code et generated C code et generated ASM code}
-
+\vfill\null %switch column
 \section{Slang Plugin code}
 \label{PluginAppendix}
+
+Source code of the plugin method generating the C code for the Slang plugin code. Most operations needs to go through the interpreter proxy.
 
@@
 
-\cba{Source code et generated C code et generated ASM code}
-
+\vfill\null %switch column
 \section{Slang code}
 \label{SlangAppendix}
+
+Source code of the slang method generating the C code for the Slang string comparison primitive.
+We split it in two methods, and forced inlining at Slang-to-C compilation time to remove the closures while avoiding code duplication.
 
@@
 	self methodReturnInteger: result
-\end{code}
-\begin{code}
+
 rawCompare: string1 length: strLength1 with: string2 length: strLength2 accessBlock: accessBlock
@@
 
-\cba{Source code et generated C code et generated ASM code}
-
+\vfill\null %switch column
 \section{Cog's RTL code}
 \label{RTLAppendix}
+
+Source code of the method generating the native code of the string comparison primitive on top of Cog's RTL.
+Method names starting with uppercase (in blue) are directly generating an abstract assembly instruction.
+Other methods called, starting with gen, generate a sequence of abstract assembly instructions.
+We duplicated the code checking if each operand is a ByteString and extract its size to generate better native code than doing the two operations separatedly.
 
@@
 	cogit SubR: TempReg R: string2CharOrByteSizeReg.
+	\end{code}
+	\vfill\null %switch column
+	\begin{code}
 	"Type and number of arguments are correct"
@@
 
-\cba{Source code et generated ASM code}
-
 \end{document}

### 2018-05-09 paper.tex
add methodology. detailed table incoming

--- 2018-05-09 paper.tex
@@
 
-\todo{fill methodology section}
-\cba{missing here: what are the benchmark actually run ?
-Can be short code sample or explanation of both.
-It has to include: each bench was run x times, each time being y comparison.
-We removed the loop overhead in the measurements.}
+We assessed the execution time of the different implementations by running the following benchmark:
+
+\begin{code}
+Smalltalk garbageCollect.
+#(0 1 5 100 1000) collect: [ :size |
+	| iterations time overhead |
+	collection := ByteString new: size.
+	collection2 := ByteString new: size.
+	size < 100 ifTrue: [iterations := (100000000 // (size*10 max: 1) sqrt) floor.]
+			   ifFalse: [iterations := (10000000 // size sqrt) floor.].
+	overhead := [ 1 to: iterations do: [ :i | ] ] timeToRun.
+	time := [ 1 to: iterations do: [ :i | 
+		collection compareWithSlang: collection2 ] ] timeToRun.
+	stream nextPutAll: ((time - overhead) asString)].
+\end{code}\\
+
+Each benchmark was measured 30 times.
+
+In each benchmark, we run a string comparison in a loop and measure the total execution time.
+The number of iterations of the loop depends on the string size: the comparison is repeated between 1414 and 10000 times for a short string (5 characters or less), and between 100 and 316 times for a long string (100 characters or more).
+This addresses the warm-up of the JIT and allows the benchmarks to run at least 1 second in the slower implementations.
+The loop overhead is measured and removed from the measurements.
+
 \cba{+ we bench only = strings, which could make a difference on the plugin version (one less branch) - maybe we should change to number of character compared instead of string size so it does not imply the string are equals.}
-\cba{SM methodology: what is done? which benchmarks are used? how are they executed? what is measured? how do you address warmup etc? what kind of machine is used for the measurements?}
 

### 2018-05-09 paper.tex
comparison, appendix

--- 2018-05-09 paper.tex
@@
 \newcommand{\cba}[1]{\color{purple}\fbox{\bfseries\sffamily\scriptsize Clement:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
-\newcommand*\rot{\rotatebox{90}}
+%\newcommand*{rotatebox{75}}
 \input{macros}
@@
 
- \todo{Finish section}
-
- In this section we compare the five different implementations according to the following criteria:
- \begin{itemize} 
- 	\item \emph{Recompilation needed:} Diff no, plugin, full
-	\item \emph{Implementation language familiar to developers:} Merged the 2 latter, explain ST dev ST, VM dev ST-Slang-C-ASM, maybe result is + ou - ? or + or 0
-	\item \emph{Readability:} explain why number of lines of code. 
-	\item \emph{Number of implementations to maintain:} explain fall-back code which is in all languages, Slang/C, JIT => maybe we ignore the fall-back code ?
-	\item \emph{Stack switch from generated native code:} 
-	\item \emph{Optimization potential:} discuss inlining C comp cross file and InterpreterProxy indirection
-	\item \emph{Control over the generated code:} diff llvm gcc while JIT always the same
-\end{itemize}
-
-\begin{table}
+\begin{table*} [bth]
 \centering
@@
   \hline
-    & \rot{recompilation needed} & \rot{familiar to client dev} & \rot{familiar to vm dev} & \rot{number of LoC} & \rot{Implementations to maintain} & \rot{Performance} & \rot{Control over generated code} \\
+    & Recompilation & Familiar to & Lignes & Number of & Stack switch & Optimization & Code generation \\
+    & required & developers & of code & implementations & overhead & potential & control \\
   \hline
-  Baseline    &             & \checkmark  &            & 10  & 3+ &  +  &  \\
+  Baseline    & No & All  & 10  & 0 & No & - & - \\
    \hline
-  SmartSyntax &             &  \checkmark &            & 20  & 3+ & ++  & -\\
+  SmartSyntax & Plugin & VM & 20  & 3+ & Yes  & Limited & Low \\
    \hline
-  SlangPlugin &             &  \checkmark &            & 35  & 2  & ++  & - \\
+  SlangPlugin & Plugin & VM & 35  & 1  & Yes & Limited & Low \\
    \hline
-  PureSlang   & \checkmark  &             & \checkmark & 45  & 1  & ++  & -\\
+  PureSlang   & Yes  & VM & 45  & 1  & Yes & High & Low \\
    \hline
-  Slang+JIT   & \checkmark  &             & \checkmark & 85  & 1  & ++++ & + \\
+  Slang+JIT   & Yes  & VM-JIT & 85  & 2  & No & High & Full \\
   \hline
 \end{tabular}
-\caption{Trade-offs between implementations: summarized}
+\caption{Comparison between implementations\vspace{-0.5cm}}
 \label{tab:cmp_implem}
-\end{table}
+\end{table*}
+
+ In this section we compare the five different implementations according to the following criteria:
+ \begin{itemize} 
+ 	\item \emph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (-) or requires to recompile the full VM (Yes).
+	\item \emph{Implementation language familiar to developers:} Smalltalk application developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code.
+Only Cog's JIT implementors are familiar with its RTL, which is a subset of the VM developers.
+We use \emph{All} if all communities are familiar with the language, \emph{VM} if all VM developers are and \emph{VM-JIT} if only a subset of VM developers are.
+	\item \emph{Readability:} To evaluate the readability of the approach, we use one of the most used metric, the number of lines of code.
+	\item \emph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to execute if the primitive fails.
+This code is maintain by the client language implementors, and not by the VM implementors.
+Hence we do not count the fall-back code in the number of implementation to maintain, nor the pure Smalltalk implementation.
+However, as a VM implementor, we need to maintain what generates the C code of the VM, hence, in the case of Smart Syntax, we need to maintain a different method in each client language.
+We also count all the Slang implementations as well as the ones on Cog's RTL.
+	\item \emph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
+	\item \emph{Optimization potential:} \todo{}
+	discuss inlining C comp cross file and InterpreterProxy indirection
+	Smalltalk we control indirectly.
+	\item \emph{Control over the generated code:} \todo{}diff llvm gcc while JIT always the same
+\end{itemize}
 
@@
 
-However, one could notice a rather striking performance difference between the 2 compilers: when compiling with GCC, we notive in Figure \ref{fig:bench} that the performances between the PureSlang and Slang+JIT implementations are really close for long strings (100 characters and longer), with respectively 24,42 and 24,48 relative speed-up factors for a 1000-characters string.
-Yet they greatly differ when LLVM is used, with respectively 13,44 and 26,40 relative speed-up factors.
-
-This difference is mainly due to the optimizations performed by both the compilers: as depicted in Figure \ref{fig:Asm} GCC generates instructions that are very close to the ones generated by the JIT, hence the similar speed-up factors.
-On the contrary, LLVM generates more instructions: it especially fails at perserving the registries holding the strings (edi and esi), leading to 2 extra instructions (bolded in the figure) that will load the strings at each iteration.
-
-\sk{parler des autres optis qui n'ont pas d'impact négatif sur le temps d'exec ? ou le conserver pour les future work pour des potentielles améliorations?}
-
-\begin{figure*}[htp!]
+\begin{figure*}[bth]
 		\centering
@@
 \end{figure*}
+
+However, one could notice a rather striking performance difference between the 2 compilers: when compiling with GCC, we notive in Figure \ref{fig:bench} that the performances between the PureSlang and Slang+JIT implementations are really close for long strings (100 characters and longer), with respectively 24,42 and 24,48 relative speed-up factors for a 1000-characters string.
+Yet they greatly differ when LLVM is used, with respectively 13,44 and 26,40 relative speed-up factors.
+
+This difference is mainly due to the optimizations performed by both the compilers: as depicted in Figure \ref{fig:Asm} GCC generates instructions that are very close to the ones generated by the JIT, hence the similar speed-up factors.
+On the contrary, LLVM generates more instructions: it especially fails at perserving the registries holding the strings (edi and esi), leading to 2 extra instructions (bolded in the figure) that will load the strings at each iteration.
+
+\sk{parler des autres optis qui n'ont pas d'impact négatif sur le temps d'exec ? ou le conserver pour les future work pour des potentielles améliorations?}
 
@@
 
+\vspace{10cm} %Adapt to make vfill work
+
 %% Appendix
-\vfill\null %switch column
+\null \vfill\null %switch column
+
 \appendix

### 2018-05-09 paper.tex
rm hack
added comment
detailed bench results

--- 2018-05-09 paper.tex
@@
 \usepackage{listings}
+\usepackage{multirow}
 \usepgfplotslibrary{statistics}
@@
 
-\todo{Add exact results}
-\cba{We are also missing exact results, aside from the graphs I usually expect a table with exact results and error margins...
-That's why the plot is often generated from a file, the table can reuse the same file and you can re-do your bench without changing the latex code.
-Cf https://hal.inria.fr/hal-01596321/document page 9, it would be nice here to have a similar table below the benchs}
+\begin{table*}[bht]
+\centering
+\begin{tabular}{|c|c|c|c|c|c|c|}
+  \hline
+    & & 0 & 1 & 5 & 100 & 1000 \\
+   \hline
+  \multirow{2}{*}{SmartSyntax} &GCC & 0.3411 +- 0.0007 & 0.6014 +- 0.0023 & 1.3891 +- 0.0023 &  8.6796 +- 0.0587 & 13.6051 +- 0.0589\\
+ 							   &LLVM & 0.3147 +- 0.0023 & 0.5606 +- 0.0022 & 1.2505 +- 0.0046 & 10.3029 +- 0.0624 & 9.4065 +- 0.0367\\
+   \hline    
+  \multirow{2}{*}{SlangPlugin} &GCC& 0.3375 +- 0.0016 & 0.5654 +- 0.0078 & 1.3150 +- 0.0205 & 8.9585 +- 0.1170  & 14.7789 +- 0.1471 \\
+  							   &LLVM& 0.3036 +- 0.0018 & 0.5567 +- 0.0010 & 1.2296 +- 0.0018 & 6.6033 +- 0.0248  & 9.2138 +- 0.0140 \\
+   \hline
+  \multirow{2}{*}{PureSlang} &GCC& 1.0778 +- 0.0054 & 1.8169 +- 0.0191 & 3.8190 +- 0.0545 & 15.8095 +- 0.3005 & 24.4178 +- 0.1863\\
+ 							 &LLVM& 0.8545 +- 0.0061 & 1.5605 +- 0.0064 & 2.8205 +- 0.0080 & 10.3029 +- 0.0615  & 13.4385 +- 0.0402 \\
+   \hline
+  \multirow{2}{*}{Slang+JIT} &GCC& 2.4904 +- 0.0043 & 3.5378 +- 0.0124  & 6.7405 +- 0.0108 & 16.7527 +- 0.1063 & 24.4801 +- 0.0950\\
+ 							&LLVM & 2.2507 +- 0.0413 & 3.7545 +- 0.0111 & 7.0898 +- 0.0140 & 17.9796 +- 0.1013  & 26.4048 +- 0.0930 \\
+  \hline
+\end{tabular}
+\caption{Benchmark results with standard errors. Average relative speed-up regarding baseline}
+\label{tab:detail_bench}
+\end{table*}
 
@@
 
+\todo{Add some intro sentence here and there + maybe ASM in appendix}
 %% Appendix
-\vfill\null %switch column
 \appendix
@@
 \label{SmartSyntaxAppendix}
-
-Source code of the smart syntax method, which is used to generate the C code for the plugin but also as Smalltalk fall-back code.
 
@@
 
-\vfill\null %switch column
+\cba{Source code et generated C code et generated ASM code}
+
 \section{Slang Plugin code}
 \label{PluginAppendix}
-
-Source code of the plugin method generating the C code for the Slang plugin code. Most operations needs to go through the interpreter proxy.
 
@@
 
-\vfill\null %switch column
+\cba{Source code et generated C code et generated ASM code}
+
 \section{Slang code}
 \label{SlangAppendix}
-
-Source code of the slang method generating the C code for the Slang string comparison primitive.
-We split it in two methods, and forced inlining at Slang-to-C compilation time to remove the closures while avoiding code duplication.
 
@@
 	self methodReturnInteger: result
-
+\end{code}
+\begin{code}
 rawCompare: string1 length: strLength1 with: string2 length: strLength2 accessBlock: accessBlock
@@
 
-\vfill\null %switch column
+\cba{Source code et generated C code et generated ASM code}
+
 \section{Cog's RTL code}
 \label{RTLAppendix}
-
-Source code of the method generating the native code of the string comparison primitive on top of Cog's RTL.
-Method names starting with uppercase (in blue) are directly generating an abstract assembly instruction.
-Other methods called, starting with gen, generate a sequence of abstract assembly instructions.
-We duplicated the code checking if each operand is a ByteString and extract its size to generate better native code than doing the two operations separatedly.
 
@@
 	cogit SubR: TempReg R: string2CharOrByteSizeReg.
-	\end{code}
-	\vfill\null %switch column
-	\begin{code}
 	"Type and number of arguments are correct"
@@
 
+\cba{Source code et generated ASM code}
+
 \end{document}

### 2018-05-09 paper.tex

--- 2018-05-09 paper.tex
@@
 
-\todo{Add exact results}
-\cba{We are also missing exact results, aside from the graphs I usually expect a table with exact results and error margins...
-That's why the plot is often generated from a file, the table can reuse the same file and you can re-do your bench without changing the latex code.
-Cf https://hal.inria.fr/hal-01596321/document page 9, it would be nice here to have a similar table below the benchs}
+\begin{table*}[bht]
+\centering
+\begin{tabular}{|c|c|c|c|c|c|c|}
+  \hline
+    & & 0 & 1 & 5 & 100 & 1000 \\
+   \hline
+  \multirow{2}{*}{SmartSyntax} &GCC & 0.3411 +- 0.0007 & 0.6014 +- 0.0023 & 1.3891 +- 0.0023 &  8.6796 +- 0.0587 & 13.6051 +- 0.0589\\
+ 							   &LLVM & 0.3147 +- 0.0023 & 0.5606 +- 0.0022 & 1.2505 +- 0.0046 & 10.3029 +- 0.0624 & 9.4065 +- 0.0367\\
+   \hline
+  \multirow{2}{*}{SlangPlugin} &GCC& 0.3375 +- 0.0016 & 0.5654 +- 0.0078 & 1.3150 +- 0.0205 & 8.9585 +- 0.1170  & 14.7789 +- 0.1471 \\
+  							   &LLVM& 0.3036 +- 0.0018 & 0.5567 +- 0.0010 & 1.2296 +- 0.0018 & 6.6033 +- 0.0248  & 9.2138 +- 0.0140 \\
+   \hline
+  \multirow{2}{*}{PureSlang} &GCC& 1.0778 +- 0.0054 & 1.8169 +- 0.0191 & 3.8190 +- 0.0545 & 15.8095 +- 0.3005 & 24.4178 +- 0.1863\\
+ 							 &LLVM& 0.8545 +- 0.0061 & 1.5605 +- 0.0064 & 2.8205 +- 0.0080 & 10.3029 +- 0.0615  & 13.4385 +- 0.0402 \\
+   \hline
+  \multirow{2}{*}{Slang+JIT} &GCC& 2.4904 +- 0.0043 & 3.5378 +- 0.0124  & 6.7405 +- 0.0108 & 16.7527 +- 0.1063 & 24.4801 +- 0.0950\\
+ 							&LLVM & 2.2507 +- 0.0413 & 3.7545 +- 0.0111 & 7.0898 +- 0.0140 & 17.9796 +- 0.1013  & 26.4048 +- 0.0930 \\
+  \hline
+\end{tabular}
+\caption{Benchmark results with standard errors. Average relative speed-up regarding baseline}
+\label{tab:detail_bench}
+\end{table*}
 
@@
 		\includegraphics[width=1\linewidth]{figures/asm_cmp.pdf}
-		\caption{GCC and LLVM: different optimizations on the comparison loop}
+		\caption{GCC and LLVM: different optimizations on the comparison loop \cba{on peut enlever les adresses je pense ca sert a rien en vrai et comme ca les fonts pourront etre plus grosses}}
 		\label{fig:Asm}
@@
 
-\vspace{10cm} %Adapt to make vfill work
+%\vspace{10cm} %Adapt to make vfill work
 

### 2018-05-09 paper.tex
bench section: add a sentence about the equal strings

@@
 \begin{code}
-Smalltalk garbageCollect.
 #(0 1 5 100 1000) collect: [ :size |
@@
 
-\cba{+ we bench only = strings, which could make a difference on the plugin version (one less branch) - maybe we should change to number of character compared instead of string size so it does not imply the string are equals.}
+The 2 compared strings are equal as it corresponds to the worst scenario for most of our benched implementations \footnote{There is one branch less in the SlangPlugin implementation in this case}.
 
@@
 		\includegraphics[width=1\linewidth]{figures/asm_cmp.pdf}
-		\caption{GCC and LLVM: different optimizations on the comparison loop \cba{on peut enlever les adresses je pense ca sert a rien en vrai et comme ca les fonts pourront etre plus grosses}}
+		\caption{GCC and LLVM: different optimizations on the comparison loop}
 		\label{fig:Asm}

### 2018-05-09 paper.tex
comparison section: control over generated code

@@
 	Smalltalk we control indirectly.
-	\item \emph{Control over the generated code:} \todo{}diff llvm gcc while JIT always the same
+	\item \emph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code.
+The generated code can therefore differ according the compilers and the optimizations they performed.
+On the contrary, the JIT will always generate the same native code.
 \end{itemize}
@@
 	collection2 := ByteString new: size.
-	size < 100 ifTrue: [iterations := (100000000 // (size*10 max: 1) sqrt) floor.]
+	size < 100 
+		ifTrue: [iterations := (100000000 // (size*10 max: 1) sqrt) floor.]
 		ifFalse: [iterations := (10000000 // size sqrt) floor.].
@@
 On the contrary, LLVM generates more instructions: it especially fails at perserving the registries holding the strings (edi and esi), leading to 2 extra instructions (bolded in the figure) that will load the strings at each iteration.
-
-\sk{parler des autres optis qui n'ont pas d'impact négatif sur le temps d'exec ? ou le conserver pour les future work pour des potentielles améliorations?}
 

### 2018-05-09 paper.tex
fix my bad merge on appendix

@@
 
+Source code of the smart syntax method, which is used to generate the C code for the plugin but also as Smalltalk fall-back code.
+
 \begin{code}

### 2018-05-09 paper.tex
section comparison: about optimization potential

@@
 	\item \emph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
-	\item \emph{Optimization potential:} \todo{}
-	discuss inlining C comp cross file and InterpreterProxy indirection
-	Smalltalk we control indirectly.
+	\item \emph{Optimization potential:} 
+	C cross-file inlining is not performed by default for internal plugins.
+External plugins don't have access to the whole VM API and have to go through InterpreterProxy to do so, leading to an overhead.
+
+	\todo{}
+	Smalltalk we control indirectly.\sk{?}
 	\item \emph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code.

### 2018-05-10 paper.tex
moved the bench and table figures to the same page, add package to have bottom figures and put the ASM loop comp at the bottom

@@
 \usepgfplotslibrary{statistics}
+\usepackage{dblfloatfix} %enable fig at bottom of page
 %\input{macros.tex}
@@
 External plugins don't have access to the whole VM API and have to go through InterpreterProxy to do so, leading to an overhead.
-
-	\todo{}
-	Smalltalk we control indirectly.\sk{?}
 	\item \emph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code.
@@
 
-In each benchmark, we run a string comparison in a loop and measure the total execution time.
-The number of iterations of the loop depends on the string size: the comparison is repeated between 1414 and 10000 times for a short string (5 characters or less), and between 100 and 316 times for a long string (100 characters or more).
-This addresses the warm-up of the JIT and allows the benchmarks to run at least 1 second in the slower implementations.
-The loop overhead is measured and removed from the measurements.
-
-The 2 compared strings are equal as it corresponds to the worst scenario for most of our benched implementations \footnote{There is one branch less in the SlangPlugin implementation in this case}.
-
-\begin{table*}[bht]
+%%%
+% Legend definition (so I can change it once for all)
+%%%
+
+\def\legBase{Baseline}
+\def\legSmartSyntaxPlug{Smart syntax}
+\def\legSlangPlug{Plugin}
+\def\legSlang{Pure slang}
+\def\legRTL{Slang+JIT}
+\def\plotHeight{6.4cm}
+
+%%%
+% End Legend
+%%%
+
+\begin{figure*}[htb!]
+  \centering
+    \begin{subfigure}[b]{\textwidth}
+    \begin{subfigure}[b]{.19\textwidth}
+    \subcaption*{Size 0}
+    \begin{tikzpicture}
+      \begin{axis}
+      [
+        height=\plotHeight,
+        x=0.5cm,
+        xtick={1,...,5},
+        xticklabel style={rotate=90},
+        bar width=0.3cm,
+        enlarge x limits={abs=0.45cm},
+        xticklabels={\legBase, \legSmartSyntaxPlug, \legSlangPlug, \legSlang, \legRTL},
+        font={\footnotesize},
+	ybar=10pt,
+	ymin=0
+        ]
+\addplot[
+    fill=black!25,
+    draw=black,
+    fill,
+    point meta=y,
+    every node near coord/.style={inner ysep=5pt},
+    error bars/.cd,
+        y dir=both,
+        y explicit
+]
+table [y error=error] {
+x   y           error
+1   1           0
+2   0.3411212   0.0006746299
+3   0.3375418   0.001577878
+4   1.077762   0.005363974
+5   2.490423    0.004304882
+};
+      \end{axis}
+    \end{tikzpicture}
+    \end{subfigure}
+     \begin{subfigure}[b]{.19\textwidth}
+    \subcaption*{Size 1}
+    \begin{tikzpicture}
+      \begin{axis}
+      [
+        height=\plotHeight,
+        x=0.5cm,
+        xtick={1,...,5},
+        xticklabel style={rotate=90},
+        bar width=0.3cm,
+        enlarge x limits={abs=0.45cm},
+        xticklabels={\legBase, \legSmartSyntaxPlug, \legSlangPlug, \legSlang, \legRTL},
+        font={\footnotesize},
+	ybar=10pt,
+	ymin=0
+        ]
+\addplot[
+    fill=black!25,
+    %draw=none,
+    %fill,
+    draw=black,
+    fill,
+    point meta=y,
+    every node near coord/.style={inner ysep=5pt},
+    error bars/.cd,
+        y dir=both,
+        y explicit
+]
+table [y error=error] {
+x   y           error
+1   1           0
+2   0.6013716   0.0023206431
+3   0.5653467   0.007829291
+4   1.816869    0.019114984
+5   3.537800    0.012357500
+};
+      \end{axis}
+    \end{tikzpicture}
+    \end{subfigure}
+    \begin{subfigure}[b]{.19\textwidth}
+    \subcaption*{Size 5}
+    \begin{tikzpicture}
+      \begin{axis}
+      [
+        height=\plotHeight,
+        x=0.5cm,
+        xtick={1,...,5},
+        xticklabel style={rotate=90},
+        bar width=0.3cm,
+        enlarge x limits={abs=0.45cm},
+        xticklabels={\legBase, \legSmartSyntaxPlug, \legSlangPlug, \legSlang, \legRTL},
+        font={\footnotesize},
+	ybar=10pt,
+	ymin=0
+        ]
+\addplot[
+    fill=black!25,
+    draw=black,
+    point meta=y,
+    every node near coord/.style={inner ysep=5pt},
+    error bars/.cd,
+        y dir=both,
+        y explicit
+]
+table [y error=error] {
+x   y           error
+1   1           0
+2   1.3891212   0.0029489731
+3   1.3149840   0.020471694
+4   3.818961   0.054523074
+5   6.740474    0.010805042
+};
+      \end{axis}
+    \end{tikzpicture}
+    \end{subfigure}
+    \begin{subfigure}[b]{.19\textwidth}
+    \subcaption*{\hspace{0.4cm}Size 100}
+ \begin{tikzpicture}
+      \begin{axis}
+      [
+        height=\plotHeight,
+        x=0.5cm,
+        xtick={1,...,5},
+        xticklabel style={rotate=90},
+        bar width=0.3cm,
+        enlarge x limits={abs=0.45cm},
+        xticklabels={\legBase, \legSmartSyntaxPlug, \legSlangPlug, \legSlang, \legRTL},
+        font={\footnotesize},
+	ybar=10pt,
+	ymin=0
+        ]
+\addplot[
+    fill=black!25,
+    draw=black,
+    point meta=y,
+    every node near coord/.style={inner ysep=5pt},
+    error bars/.cd,
+        y dir=both,
+        y explicit
+]
+table [y error=error] {
+x   y           error
+1   1           0
+2   8.6795487   0.0586762222
+3   8.9585208   0.116958804
+4   15.809515   0.300447885
+5   16.752667    0.106255682
+};
+      \end{axis}
+    \end{tikzpicture}
+    \end{subfigure}
+    \begin{subfigure}[b]{.19\textwidth}
+    \subcaption*{\hspace{0.5cm}Size 1000}
+     \begin{tikzpicture}
+      \begin{axis}
+      [
+        height=\plotHeight,
+        x=0.5cm,
+        xtick={1,...,5},
+        xticklabel style={rotate=90},
+        bar width=0.3cm,
+        enlarge x limits={abs=0.45cm},
+        xticklabels={\legBase, \legSmartSyntaxPlug, \legSlangPlug, \legSlang, \legRTL},
+        font={\footnotesize},
+	ybar=10pt,
+	ymin = 0
+        ]
+\addplot[
+    fill=black!25,
+    draw=black,
+    point meta=y,
+    every node near coord/.style={inner ysep=5pt},
+    error bars/.cd,
+        y dir=both,
+        y explicit
+]
+table [y error=error] {
+x   y           error
+1   1           0
+2   13.6051267   0.0589164968
+3   14.7789145   0.147140189
+4   24.417751   0.186326223
+5   24.480071    0.095038374
+};
+      \end{axis}
+    \end{tikzpicture}
+   \end{subfigure}
+  % \caption{GCC-Linux results}
+   % \label{fig:gcc}
+   \subcaption*{\textbf{GCC-Linux}\vspace{0.25cm}}
+   \label{fig:gcc}
+   \end{subfigure}
+ \begin{subfigure}[b]{\textwidth}
+    \begin{subfigure}[b]{.19\textwidth}
+
+    \begin{tikzpicture}
+      \begin{axis}
+      [
+        height=\plotHeight,
+        x=0.5cm,
+        xtick={1,...,5},
+        xticklabel style={rotate=90},
+        bar width=0.3cm,
+        enlarge x limits={abs=0.45cm},
+        xticklabels={\legBase, \legSmartSyntaxPlug, \legSlangPlug, \legSlang, \legRTL},
+        font={\footnotesize},
+	ybar=10pt,
+	ymin=0,
+	]
+\addplot[
+    fill=black!25,
+    draw=black,
+    point meta=y,
+    every node near coord/.style={inner ysep=5pt},
+    error bars/.cd,
+        y dir=both,
+        y explicit
+]
+table [y error=error] {
+x   y           error
+1   1           0
+2   0.3147064   0.002288488
+3   0.3035965   0.001786185
+4   0.8545431   0.006132950
+5   2.250732    0.04133638
+};
+      \end{axis}
+    \end{tikzpicture}
+
+    \end{subfigure}
+     \begin{subfigure}[b]{.19\textwidth}
+
+    \begin{tikzpicture}
+      \begin{axis}
+      [
+        height=\plotHeight,
+        x=0.5cm,
+        xtick={1,...,5},
+        xticklabel style={rotate=90},
+        bar width=0.3cm,
+        enlarge x limits={abs=0.45cm},
+        xticklabels={\legBase, \legSmartSyntaxPlug, \legSlangPlug, \legSlang, \legRTL},
+        font={\footnotesize},
+	ybar=10pt,
+	ymin=0
+        ]
+\addplot[
+    fill=black!25,
+    draw=black,
+    point meta=y,
+    every node near coord/.style={inner ysep=5pt},
+    error bars/.cd,
+        y dir=both,
+        y explicit
+]
+table [y error=error] {
+x   y           error
+1   1           0
+2   0.5605682   0.002198180
+3   0.5566675   0.001023225
+4   1.5605275   0.006372279
+5   3.754532    0.01109086
+};
+      \end{axis}
+    \end{tikzpicture}
+
+    \end{subfigure}
+    \begin{subfigure}[b]{.19\textwidth}
+
+    \begin{tikzpicture}
+      \begin{axis}
+      [
+        height=\plotHeight,
+        x=0.5cm,
+        xtick={1,...,5},
+        xticklabel style={rotate=90},
+        bar width=0.3cm,
+        enlarge x limits={abs=0.45cm},
+        xticklabels={\legBase, \legSmartSyntaxPlug, \legSlangPlug, \legSlang, \legRTL},
+        font={\footnotesize},
+	ybar=10pt,
+	ymin=0
+        ]
+\addplot[
+    fill=black!25,
+    draw=black,
+    point meta=y,
+    every node near coord/.style={inner ysep=5pt},
+    error bars/.cd,
+        y dir=both,
+        y explicit
+]
+table [y error=error] {
+x   y           error
+1   1           0
+2   1.2505193   0.004611354
+3   1.2295801   0.001810171
+4   2.8205415   0.008035024
+5   7.089806    0.01403785
+};
+      \end{axis}
+    \end{tikzpicture}
+    \end{subfigure}
+    \begin{subfigure}[b]{.19\textwidth}
+ \begin{tikzpicture}
+      \begin{axis}
+      [
+        height=\plotHeight,
+        x=0.5cm,
+        xtick={1,...,5},
+        xticklabel style={rotate=90},
+        bar width=0.3cm,
+        enlarge x limits={abs=0.45cm},
+        xticklabels={\legBase, \legSmartSyntaxPlug, \legSlangPlug, \legSlang, \legRTL},
+        font={\footnotesize},
+	ybar=10pt,
+	ymin=0
+        ]
+\addplot[
+    fill=black!25,
+    draw=black,
+    point meta=y,
+    every node near coord/.style={inner ysep=5pt},
+    error bars/.cd,
+        y dir=both,
+        y explicit
+]
+table [y error=error] {
+x   y           error
+1   1           0
+2   6.6951738   0.062391122
+3   6.6032572   0.024774416
+4   10.3029394  0.061481532
+5   17.979611   0.10127936
+};
+      \end{axis}
+    \end{tikzpicture}
+    \end{subfigure}
+    \begin{subfigure}[b]{.19\textwidth}
+
+    \begin{tikzpicture}
+      \begin{axis}
+      [
+        height=\plotHeight,
+        x=0.5cm,
+        xtick={1,...,5},
+        xticklabel style={rotate=90},
+        bar width=0.3cm,
+        enlarge x limits={abs=0.45cm},
+        xticklabels={\legBase, \legSmartSyntaxPlug, \legSlangPlug, \legSlang, \legRTL},
+        font={\footnotesize},
+	ybar=10pt,
+	ymin=0
+        ]
+\addplot[
+    fill=black!25,
+    draw=black,
+    point meta=y,
+    every node near coord/.style={inner ysep=5pt},
+    error bars/.cd,
+        y dir=both,
+        y explicit
+]
+table [y error=error] {
+x   y           error
+1   1           0
+2   9.4064534   0.036698187
+3   9.2138415   0.013982843
+4   13.4384830  0.040202717
+5   26.404792   0.09298324
+};
+      \end{axis}
+    \end{tikzpicture}
+
+   \end{subfigure}
+   \subcaption*{\textbf{LLVM-Mac}\vspace{-0.2cm}} %change vspace based on fig position
+    \end{subfigure}
+   \caption{String comparison benchmarks for strings of different sizes on different architectures}
+   \label{fig:bench}
+\end{figure*}
+\begin{table*}
 \centering
@@
 \end{tabular} 
+\vspace{0.2cm}
 \caption{Benchmark results with standard errors. Average relative speed-up regarding baseline}
@@
 
+\begin{figure*}[b!]
+		\centering
+		\includegraphics[width=1\linewidth]{figures/asm_cmp.pdf}
+		\caption{GCC and LLVM: different optimizations on the comparison loop}
+		\label{fig:Asm}
+\end{figure*}
+
+In each benchmark, we run a string comparison in a loop and measure the total execution time.
+The number of iterations of the loop depends on the string size: the comparison is repeated between 1414 and 10000 times for a short string (5 characters or less), and between 100 and 316 times for a long string (100 characters or more).
+This addresses the warm-up of the JIT and allows the benchmarks to run at least 1 second in the slower implementations.
+The loop overhead is measured and removed from the measurements.
+
+The 2 compared strings are equal as it corresponds to the worst scenario for most of our benched implementations \footnote{There is one branch less in the SlangPlugin implementation in this case}.
+
 \subsection{Micro-benchmarks}
-%%%
-% Legend definition (so I can change it once for all)
-%%%
-
-\def\legBase{Baseline}
-\def\legSmartSyntaxPlug{Smart Syntax}
-\def\legSlangPlug{Plugin}
-\def\legSlang{Pure Slang}
-\def\legRTL{Slang+JIT}
-
-%%%
-% End Legend
-%%%
-
-\begin{figure*}[bht]
-  \centering
-    \begin{subfigure}[b]{\textwidth}
-    \begin{subfigure}[b]{.19\textwidth}
-    \begin{tikzpicture}
-      \begin{axis}
-      [
-        x=0.5cm,
-        xtick={1,...,5},
-        xticklabel style={rotate=90},
-        bar width=0.3cm,
-        enlarge x limits={abs=0.45cm},
-        xticklabels={\legBase, \legSmartSyntaxPlug, \legSlangPlug, \legSlang, \legRTL},
-        font={\footnotesize},
-	ybar=10pt,
-	ymin=0
-        ]
-\addplot[
-    fill=black!25,
-    draw=black,
-    fill,
-    point meta=y,
-    every node near coord/.style={inner ysep=5pt},
-    error bars/.cd,
-        y dir=both,
-        y explicit
-]
-table [y error=error] {
-x   y           error
-1   1           0
-2   0.3411212   0.0006746299
-3   0.3375418   0.001577878
-4   1.077762   0.005363974
-5   2.490423    0.004304882
-};
-      \end{axis}
-    \end{tikzpicture}
-    \subcaption{Size 0}
-    \end{subfigure}
-     \begin{subfigure}[b]{.19\textwidth}
-    \begin{tikzpicture}
-      \begin{axis}
-      [
-        x=0.5cm,
-        xtick={1,...,5},
-        xticklabel style={rotate=90},
-        bar width=0.3cm,
-        enlarge x limits={abs=0.45cm},
-        xticklabels={\legBase, \legSmartSyntaxPlug, \legSlangPlug, \legSlang, \legRTL},
-        font={\footnotesize},
-	ybar=10pt,
-	ymin=0
-        ]
-\addplot[
-    fill=black!25,
-    %draw=none,
-    %fill,
-    draw=black,
-    fill,
-    point meta=y,
-    every node near coord/.style={inner ysep=5pt},
-    error bars/.cd,
-        y dir=both,
-        y explicit
-]
-table [y error=error] {
-x   y           error
-1   1           0
-2   0.6013716   0.0023206431
-3   0.5653467   0.007829291
-4   1.816869    0.019114984
-5   3.537800    0.012357500
-};
-      \end{axis}
-    \end{tikzpicture}
-    \subcaption{Size 1}
-    \end{subfigure}
-    \begin{subfigure}[b]{.19\textwidth}
-    \begin{tikzpicture}
-      \begin{axis}
-      [
-        x=0.5cm,
-        xtick={1,...,5},
-        xticklabel style={rotate=90},
-        bar width=0.3cm,
-        enlarge x limits={abs=0.45cm},
-        xticklabels={\legBase, \legSmartSyntaxPlug, \legSlangPlug, \legSlang, \legRTL},
-        font={\footnotesize},
-	ybar=10pt,
-	ymin=0
-        ]
-\addplot[
-    fill=black!25,
-    draw=black,
-    point meta=y,
-    every node near coord/.style={inner ysep=5pt},
-    error bars/.cd,
-        y dir=both,
-        y explicit
-]
-table [y error=error] {
-x   y           error
-1   1           0
-2   1.3891212   0.0029489731
-3   1.3149840   0.020471694
-4   3.818961   0.054523074
-5   6.740474    0.010805042
-};
-      \end{axis}
-    \end{tikzpicture}
-    \subcaption{Size 5}
-    \end{subfigure}
-    \begin{subfigure}[b]{.19\textwidth}
- \begin{tikzpicture}
-      \begin{axis}
-      [
-        x=0.5cm,
-        xtick={1,...,5},
-        xticklabel style={rotate=90},
-        bar width=0.3cm,
-        enlarge x limits={abs=0.45cm},
-        xticklabels={\legBase, \legSmartSyntaxPlug, \legSlangPlug, \legSlang, \legRTL},
-        font={\footnotesize},
-	ybar=10pt,
-	ymin=0
-        ]
-\addplot[
-    fill=black!25,
-    draw=black,
-    point meta=y,
-    every node near coord/.style={inner ysep=5pt},
-    error bars/.cd,
-        y dir=both,
-        y explicit
-]
-table [y error=error] {
-x   y           error
-1   1           0
-2   8.6795487   0.0586762222
-3   8.9585208   0.116958804
-4   15.809515   0.300447885
-5   16.752667    0.106255682
-};
-      \end{axis}
-    \end{tikzpicture}
-    \subcaption{Size 100}
-    \end{subfigure}
-    \begin{subfigure}[b]{.19\textwidth}
-     \begin{tikzpicture}
-      \begin{axis}
-      [
-        x=0.5cm,
-        xtick={1,...,5},
-        xticklabel style={rotate=90},
-        bar width=0.3cm,
-        enlarge x limits={abs=0.45cm},
-        xticklabels={\legBase, \legSmartSyntaxPlug, \legSlangPlug, \legSlang, \legRTL},
-        font={\footnotesize},
-	ybar=10pt,
-	ymin = 0
-        ]
-\addplot[
-    fill=black!25,
-    draw=black,
-    point meta=y,
-    every node near coord/.style={inner ysep=5pt},
-    error bars/.cd,
-        y dir=both,
-        y explicit
-]
-table [y error=error] {
-x   y           error
-1   1           0
-2   13.6051267   0.0589164968
-3   14.7789145   0.147140189
-4   24.417751   0.186326223
-5   24.480071    0.095038374
-};
-      \end{axis}
-    \end{tikzpicture}
-   \subcaption{Size 1000}
-   \end{subfigure}
-  % \caption{GCC-Linux results}
-   % \label{fig:gcc}
-   \subcaption*{\textbf{GCC-Linux results}\vspace{0.3cm}}
-   \label{fig:gcc}
-   \end{subfigure}
- \begin{subfigure}[b]{\textwidth}
-    \begin{subfigure}[b]{.19\textwidth}
-
-    \begin{tikzpicture}
-      \begin{axis}
-      [
-        x=0.5cm,
-        xtick={1,...,5},
-        xticklabel style={rotate=90},
-        bar width=0.3cm,
-        enlarge x limits={abs=0.45cm},
-        xticklabels={\legBase, \legSmartSyntaxPlug, \legSlangPlug, \legSlang, \legRTL},
-        font={\footnotesize},
-	ybar=10pt,
-	ymin=0
-	]
-\addplot[
-    fill=black!25,
-    draw=black,
-    point meta=y,
-    every node near coord/.style={inner ysep=5pt},
-    error bars/.cd,
-        y dir=both,
-        y explicit
-]
-table [y error=error] {
-x   y           error
-1   1           0
-2   0.3147064   0.002288488
-3   0.3035965   0.001786185
-4   0.8545431   0.006132950
-5   2.250732    0.04133638
-};
-      \end{axis}
-    \end{tikzpicture}
-
-    \subcaption{Size 0}
-    \end{subfigure}
-     \begin{subfigure}[b]{.19\textwidth}
-
-    \begin{tikzpicture}
-      \begin{axis}
-      [
-        x=0.5cm,
-        xtick={1,...,5},
-        xticklabel style={rotate=90},
-        bar width=0.3cm,
-        enlarge x limits={abs=0.45cm},
-        xticklabels={\legBase, \legSmartSyntaxPlug, \legSlangPlug, \legSlang, \legRTL},
-        font={\footnotesize},
-	ybar=10pt,
-	ymin=0
-        ]
-\addplot[
-    fill=black!25,
-    draw=black,
-    point meta=y,
-    every node near coord/.style={inner ysep=5pt},
-    error bars/.cd,
-        y dir=both,
-        y explicit
-]
-table [y error=error] {
-x   y           error
-1   1           0
-2   0.5605682   0.002198180
-3   0.5566675   0.001023225
-4   1.5605275   0.006372279
-5   3.754532    0.01109086
-};
-      \end{axis}
-    \end{tikzpicture}
-
-    \subcaption{Size 1}
-    \end{subfigure}
-    \begin{subfigure}[b]{.19\textwidth}
-
-    \begin{tikzpicture}
-      \begin{axis}
-      [
-        x=0.5cm,
-        xtick={1,...,5},
-        xticklabel style={rotate=90},
-        bar width=0.3cm,
-        enlarge x limits={abs=0.45cm},
-        xticklabels={\legBase, \legSmartSyntaxPlug, \legSlangPlug, \legSlang, \legRTL},
-        font={\footnotesize},
-	ybar=10pt,
-	ymin=0
-        ]
-\addplot[
-    fill=black!25,
-    draw=black,
-    point meta=y,
-    every node near coord/.style={inner ysep=5pt},
-    error bars/.cd,
-        y dir=both,
-        y explicit
-]
-table [y error=error] {
-x   y           error
-1   1           0
-2   1.2505193   0.004611354
-3   1.2295801   0.001810171
-4   2.8205415   0.008035024
-5   7.089806    0.01403785
-};
-      \end{axis}
-    \end{tikzpicture}
-    \subcaption{Size 5}
-    \end{subfigure}
-    \begin{subfigure}[b]{.19\textwidth}
- \begin{tikzpicture}
-      \begin{axis}
-      [
-        x=0.5cm,
-        xtick={1,...,5},
-        xticklabel style={rotate=90},
-        bar width=0.3cm,
-        enlarge x limits={abs=0.45cm},
-        xticklabels={\legBase, \legSmartSyntaxPlug, \legSlangPlug, \legSlang, \legRTL},
-        font={\footnotesize},
-	ybar=10pt,
-	ymin=0
-        ]
-\addplot[
-    fill=black!25,
-    draw=black,
-    point meta=y,
-    every node near coord/.style={inner ysep=5pt},
-    error bars/.cd,
-        y dir=both,
-        y explicit
-]
-table [y error=error] {
-x   y           error
-1   1           0
-2   6.6951738   0.062391122
-3   6.6032572   0.024774416
-4   10.3029394  0.061481532
-5   17.979611   0.10127936
-};
-      \end{axis}
-    \end{tikzpicture}
-    \subcaption{Size 100}
-    \end{subfigure}
-    \begin{subfigure}[b]{.19\textwidth}
-
-    \begin{tikzpicture}
-      \begin{axis}
-      [
-        x=0.5cm,
-        xtick={1,...,5},
-        xticklabel style={rotate=90},
-        bar width=0.3cm,
-        enlarge x limits={abs=0.45cm},
-        xticklabels={\legBase, \legSmartSyntaxPlug, \legSlangPlug, \legSlang, \legRTL},
-        font={\footnotesize},
-	ybar=10pt,
-	ymin=0
-        ]
-\addplot[
-    fill=black!25,
-    draw=black,
-    point meta=y,
-    every node near coord/.style={inner ysep=5pt},
-    error bars/.cd,
-        y dir=both,
-        y explicit
-]
-table [y error=error] {
-x   y           error
-1   1           0
-2   9.4064534   0.036698187
-3   9.2138415   0.013982843
-4   13.4384830  0.040202717
-5   26.404792   0.09298324
-};
-      \end{axis}
-    \end{tikzpicture}
-
-   \subcaption{Size 1000}
-   \end{subfigure}
-   \subcaption*{\textbf{LLVM-Mac results}\vspace{-0.2cm}} %change vspace based on fig position
-    \end{subfigure}
-   \caption{String comparison benchmarks for strings of different sizes on different architectures}
-   \label{fig:bench}
-\end{figure*}
 
@@
 We performed the benchmarks upon five different string sizes: 0, 1, 5, 100 and 1000 characters, using two different compilers: GCC on one hand and LLVM on the other hand.
-Figure \ref{fig:bench} show the results in terms of relative speed-up towards the baseline implementation\footnote{i.e.
-the pure Smalltalk version}, respectively for GCC and LLVM.
+Figure \ref{fig:bench} show the results in terms of relative speed-up towards the baseline implementation, respectively for GCC and LLVM.
 
@@
 \end{itemize}
-
-\begin{figure*}[bth]
-		\centering
-		\includegraphics[width=1\linewidth]{figures/asm_cmp.pdf}
-		\caption{GCC and LLVM: different optimizations on the comparison loop}
-		\label{fig:Asm}
-\end{figure*}
 

### 2018-05-10 paper.tex
added comments

@@
 
+\cba{We may use paragraph instead of itemize here to compact}
  In this section we compare the five different implementations according to the following criteria:
@@
 The number of iterations of the loop depends on the string size: the comparison is repeated between 1414 and 10000 times for a short string (5 characters or less), and between 100 and 316 times for a long string (100 characters or more).
-This addresses the warm-up of the JIT and allows the benchmarks to run at least 1 second in the slower implementations.
+This addresses the warm-up of the JIT \cba{???
+This needs to be explained differently.
+We get into jitted code after a small number of loop iteration (I think 20 currently) or second call.
+The JIT overhead is negligible here for 1 second because we could not see any time spent in the JIT in the profiler + ref to your paper, i.e.
+less than 0.5ms in the JIT} and allows the benchmarks to run at least 1 second in the slower implementations \cba{Ok, 1 second to avoid processor noises}.
 The loop overhead is measured and removed from the measurements.
 
-The 2 compared strings are equal as it corresponds to the worst scenario for most of our benched implementations \footnote{There is one branch less in the SlangPlugin implementation in this case}.
-
+The 2 compared strings are equal as it corresponds to the worst scenario for most of our benched implementations. There is one less branch in the two plugin implementations in this case.
+I
 \subsection{Micro-benchmarks}
@@
 \end{itemize}
+
+\cba{We need to explain: small string, overhead stack switch high, baseline and JIT version fastest. Large string, loop is where the time is spent, slang and JIT fastest.
+We need size 1000 so that the stack switch overhead cannot be seen anymore on gcc (size 100 we still C it)
+Size 0 is only '' = '', not really interesting in real app, but show well the stack switch overhead.
+Example use cases are comparing tokens in parsers (Size ~4-10 [self, super, thisContext, this, etc.])
+Other example are long urls, hundreds of characters, which are represented with large string comp bench}
 

### 2018-05-10 paper.tex
added comments and todo

@@
 \centering
-\begin{tabular}{lccccccc}
+\begin{tabular}{c|cccc|ccc}
+   &  \multicolumn{4}{c|}{Software maintenance \& evolution} & \multicolumn{3}{c}{Primitive execution time}\\
   \hline
@@
    \hline
-  Slang+JIT   & Yes  & VM-JIT & 85  & 2  & No & High & Full \\
+  Slang+JIT   & Yes  & VM-JIT & 85  & 2  & No & High & High \\
   \hline
@@
 	\item \emph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
-	\item \emph{Optimization potential:} 
+	\item \emph{Optimization potential:} \cba{need to mention that this is for large strings}
 	C cross-file inlining is not performed by default for internal plugins.
@@
 Likewise, the use of plugins may bring modularity, but as you will see in Section \ref{sec:eval}, the performance, although better than baseline version, could still be improved.
+
+\cba{In this section somewhere we need to say that criteria are split in maintenance and performance, the evaluation in next section is on last 3 criteria: Stack switch overhead, Optimization potential and control over the generated code.}
 
@@
 The JIT overhead is negligible here for 1 second because we could not see any time spent in the JIT in the profiler + ref to your paper, i.e.
-less than 0.5ms in the JIT} and allows the benchmarks to run at least 1 second in the slower implementations \cba{Ok, 1 second to avoid processor noises}.
+less than 0.5ms in the JIT} and allows the benchmarks to run at least 1 second in the slower implementations \cba{Ok, 1 second to avoid processor / OS noises, not just for fun}.
 The loop overhead is measured and removed from the measurements.
@@
 
-\cba{We need to explain: small string, overhead stack switch high, baseline and JIT version fastest. Large string, loop is where the time is spent, slang and JIT fastest.
-We need size 1000 so that the stack switch overhead cannot be seen anymore on gcc (size 100 we still C it)
+\todo{Following comments are the most important for paper}
+
+\cba{we had 3 criterias:
+- 1 C ST stack switch overhead. 
+- 2 optimization potential, that we may change to large string cmp perf
+- 3 control over generated code
+We need to explain the results based on each criteria, ideally in 3 paragraphs entitled with the 3 criteria.
+The 2 paragraph below explain only point 3 and lacks a conclusion sentence "blabla we have control and if C compiler is not good such as uncommon back-end or x86 which is still quite common in this case..."
+}
+\cba{This is the kind of missing things: C ST stack switch overhead: small string, overhead stack switch high, baseline and JIT version fastest. 
+Large string perf: Large string, loop is where the time is spent, slang and JIT fastest.
+We need size 1000 so that the stack switch overhead cannot be seen anymore on gcc (size 100 we still C it)}
+
+\cba{Maybe small discussion on relevance
 Size 0 is only '' = '', not really interesting in real app, but show well the stack switch overhead.
 Example use cases are comparing tokens in parsers (Size ~4-10 [self, super, thisContext, this, etc.])
-Other example are long urls, hundreds of characters, which are represented with large string comp bench}
+Other example are long urls, hundreds of characters, which are represented with large string comp bench
+or something like that.}
 

### 2018-05-10 paper.tex
bench comparison: more detailed explanation and references to previous sections

@@
 \subsection{Comparison}
+\label{sec:implem_cmp}
 
@@
 The 2 compared strings are equal as it corresponds to the worst scenario for most of our benched implementations. There is one less branch in the two plugin implementations in this case.
-I
+
 \subsection{Micro-benchmarks}
@@
 \item The execution time of the SmartSyntax and Plugin implementations is either very close or greater than the baseline's for the small strings.
+\item The Slang+JIT implementation is more efficient than the baseline in every case.
+Nevertheless,the performance gap between this version and the PureSlang one tends to decrease as the string size increases.
+\item The compiler used (either GCC or LLVM) does not lead to significant differences in terms of performance, except for the 100+-characters strings. 
+\end{itemize}
+
+We can explain these results based on the last 3 criteria detailed in Section \ref{sec:implem_cmp}:
+
+\paragraph{Stack switch overhead}
+As mentionned in Section \ref{sec:VMexec}, all implementations written in Slang come with an overhead due to the switch between the C and Smalltalk stacks.
+This overhead is clearly visible in the bench results in Figure \ref{fig:bench} for small strings (size 1 or less): the SmartSyntax and Plugin implementations are around 30 to 60\% slower than the baseline's, and the PureSlang performances are close to the baseline's.
+As the string size goes larger, the overhead is less significant because it is **absorbed** by other performance gains.
+Regarding this criteria, the Slang+JIT version is always the fastest: it is not really impacted by the switch overhead, because the Slang part is used for a very short time.
+
+\paragraph{Optimization potential OR Large strings comparison performance}
+The plugins come with an overhead due to the cross-file inlining and the passage through an InterpreterProxy, like said in Section \ref{sec:implem_cmp}.
+This is one of the factor explaining the performance gap between these versions and the Slang and Slang+JIT ones.
+When it comes to comparing large strings (100 characters and more), the execution time is mainly spent in the comparison loop (Figure \ref{fig:Asm}).
+However, the switch overhead is still noticeable in the PureSlang implementation when strings of 100 characters are compared: PureSlang is 15.8 times faster while Slang+JIT is 16.8 times faster.
+This is not the case anymore on 1000-characters strings.
+This last observation only applies to the GCC results. The LLVM version
+
+\paragraph{Control over generated code}
+One could notice a rather striking performance difference between the 2 compilers: when compiling with GCC, the performances between the PureSlang and Slang+JIT implementations are really close for long strings (100 characters and longer), with respectively 24,42 and 24,48 relative speed-up factors for a 1000-characters string.
+Yet they greatly differ when LLVM is used, with respectively 13,44 and 26,40 relative speed-up factors.
+
+This difference is mainly due to the optimizations performed by both the compilers: as depicted in Figure \ref{fig:Asm}, GCC generates instructions that are very close to the ones generated by the JIT, hence the similar speed-up factors.
+On the contrary, LLVM generates more instructions: it especially fails at preserving the registries holding the strings (edi and esi), leading to 2 extra instructions (bolded in the figure) that will load the strings at each iteration.
+
+%We can draw the following patterns from these results:
+%\begin{itemize}
+%\item The execution time of the SmartSyntax and Plugin implementations is either very close or greater than the baseline's for the small strings.
 This can be explained by the high cost of the switching between the C and Smalltalk stacks, as mentionned in Section \ref{sec:VMexec}.
-\item The Slang+JIT implementation is more efficient in every case. Nevertheless,
-the performance gap between this implementation and the PureSlang one tends to decrease as the string size increase.
+%\item The Slang+JIT implementation is more efficient in every case. Nevertheless,
+%the performance gap between this implementation and the PureSlang one tends to decrease as the string size increase.
 The comparison loop (\ref{fig:Asm}) proves to be more and more time-consuming because of the higher number of iterations.
-\end{itemize}
-
-\todo{Following comments are the most important for paper}
-
-\cba{we had 3 criterias:
-- 1 C ST stack switch overhead. 
-- 2 optimization potential, that we may change to large string cmp perf
-- 3 control over generated code
-We need to explain the results based on each criteria, ideally in 3 paragraphs entitled with the 3 criteria.
-The 2 paragraph below explain only point 3 and lacks a conclusion sentence "blabla we have control and if C compiler is not good such as uncommon back-end or x86 which is still quite common in this case..."
-}
-\cba{This is the kind of missing things: C ST stack switch overhead: small string, overhead stack switch high, baseline and JIT version fastest. 
-Large string perf: Large string, loop is where the time is spent, slang and JIT fastest.
-We need size 1000 so that the stack switch overhead cannot be seen anymore on gcc (size 100 we still C it)}
-
-\cba{Maybe small discussion on relevance
-Size 0 is only '' = '', not really interesting in real app, but show well the stack switch overhead.
-Example use cases are comparing tokens in parsers (Size ~4-10 [self, super, thisContext, this, etc.])
-Other example are long urls, hundreds of characters, which are represented with large string comp bench
-or something like that.}
-
-However, one could notice a rather striking performance difference between the 2 compilers: when compiling with GCC, we notive in Figure \ref{fig:bench} that the performances between the PureSlang and Slang+JIT implementations are really close for long strings (100 characters and longer), with respectively 24,42 and 24,48 relative speed-up factors for a 1000-characters string.
+%\end{itemize}
+%
+%\todo{Following comments are the most important for paper}
+%
+%\cba{we had 3 criterias:
+%- 1 C ST stack switch overhead. 
+%- 2 optimization potential, that we may change to large string cmp perf
+%- 3 control over generated code
+%We need to explain the results based on each criteria, ideally in 3 paragraphs entitled with the 3 criteria.
+%The 2 paragraph below explain only point 3 and lacks a conclusion sentence "blabla we have control and if C compiler is not good such as uncommon back-end or x86 which is still quite common in this case..."
+%}
+%\cba{This is the kind of missing things: C ST stack switch overhead: small string, overhead stack switch high, baseline and JIT version fastest. 
+%Large string perf: Large string, loop is where the time is spent, slang and JIT fastest.
+%We need size 1000 so that the stack switch overhead cannot be seen anymore on gcc (size 100 we still C it)}
+%
+%\cba{Maybe small discussion on relevance
+%Size 0 is only '' = '', not really interesting in real app, but show well the stack switch overhead.
+%Example use cases are comparing tokens in parsers (Size ~4-10 [self, super, thisContext, this, etc.])
+%Other example are long urls, hundreds of characters, which are represented with large string comp bench
+%or something like that.}
+%
+%However, one could notice a rather striking performance difference between the 2 compilers: when compiling with GCC, we notive in Figure \ref{fig:bench} that the performances between the PureSlang and Slang+JIT implementations are really close for long strings (100 characters and longer), with respectively 24,42 and 24,48 relative speed-up factors for a 1000-characters string.
 Yet they greatly differ when LLVM is used, with respectively 13,44 and 26,40 relative speed-up factors.
-
-This difference is mainly due to the optimizations performed by both the compilers: as depicted in Figure \ref{fig:Asm} GCC generates instructions that are very close to the ones generated by the JIT, hence the similar speed-up factors.
+%
+%This difference is mainly due to the optimizations performed by both the compilers: as depicted in Figure \ref{fig:Asm} GCC generates instructions that are very close to the ones generated by the JIT, hence the similar speed-up factors.
 On the contrary, LLVM generates more instructions: it especially fails at perserving the registries holding the strings (edi and esi), leading to 2 extra instructions (bolded in the figure) that will load the strings at each iteration.

### 2018-05-10 paper.tex
fix one title

@@
 
-\paragraph{Optimization potential OR Large strings comparison performance}
+\paragraph{Optimization potential on large strings comparison}
 The plugins come with an overhead due to the cross-file inlining and the passage through an InterpreterProxy, like said in Section \ref{sec:implem_cmp}.

### 2018-05-10 paper.tex
more on bench details

@@
 \paragraph{Optimization potential on large strings comparison}
-The plugins come with an overhead due to the cross-file inlining and the passage through an InterpreterProxy, like said in Section \ref{sec:implem_cmp}.
+The plugins come with an overhead \sk{not an overhead when it comes to inlining.
+Just not optimized.
+need rephrasing} due to the cross-file inlining and the passage through an InterpreterProxy, like said in Section \ref{sec:implem_cmp}.
 This is one of the factor explaining the performance gap between these versions and the Slang and Slang+JIT ones.
@@
 This is not the case anymore on 1000-characters strings.
-This last observation only applies to the GCC results. The LLVM version
-
+This last observation only applies to the GCC results: the code generated by the LLVM compiler induces a longer execution time, as explained in the next paragraph. \sk{is it legal to ref that way?}
 \paragraph{Control over generated code}
@@
 On the contrary, LLVM generates more instructions: it especially fails at preserving the registries holding the strings (edi and esi), leading to 2 extra instructions (bolded in the figure) that will load the strings at each iteration.
+
+\todo{add conclusion}
 

### 2018-05-10 paper.tex
typos here and there

@@
 VMs are traditionally implemented in a low-level language such as C or C++.
-To balance between memory footprint, start-up performance and peak performance, the execution of code is usually done through multiple execution tiers: the first few executions of a code snippet are done through an interpreter and the JIT compiler \sd{optimizes} at runtime the frequently used code snippets \sd{rm: for further executions to be performed quicker}.
+To balance between memory footprint, start-up performance and peak performance, the execution of code is usually done through multiple execution tiers: the first few executions of a code snippet are done through an interpreter and the JIT compiler optimizes at runtime the frequently used code snippets.
 
@@
  \begin{itemize}
- 	\item \emph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (-) or requires to recompile the full VM (Yes).
+ 	\item \emph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (Plugin) or requires to recompile the full VM (Yes).
 	\item \emph{Implementation language familiar to developers:} Smalltalk application developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code.
@@
 
-The 2 compared strings are equal as it corresponds to the worst scenario for most of our benched implementations. There is one less branch in the two plugin implementations in this case.
+The two compared strings are equal as it corresponds to the worst scenario for most of our benched implementations. There is one less branch in the two plugin implementations in this case.
 
@@
 
-\paragraph{Stack switch overhead}
+\paragraph{Stack switch overhead.}
 As mentionned in Section \ref{sec:VMexec}, all implementations written in Slang come with an overhead due to the switch between the C and Smalltalk stacks.
@@
 
-\paragraph{Optimization potential on large strings comparison}
+\paragraph{Optimization potential on large strings comparison.}
 The plugins come with an overhead \sk{not an overhead when it comes to inlining.
@@
 This is not the case anymore on 1000-characters strings.
-This last observation only applies to the GCC results: the code generated by the LLVM compiler induces a longer execution time, as explained in the next paragraph. \sk{is it legal to ref that way?}
-\paragraph{Control over generated code}
+This last observation only applies to the GCC results: the code generated by the LLVM compiler induces a longer execution time, as explained in the next paragraph.
+\sk{is it legal to ref that way?} \cba{it's ok, but not recommended since sometimes we move things around without checking other sections}
+\paragraph{Control over generated code.}
 One could notice a rather striking performance difference between the 2 compilers: when compiling with GCC, the performances between the PureSlang and Slang+JIT implementations are really close for long strings (100 characters and longer), with respectively 24,42 and 24,48 relative speed-up factors for a 1000-characters string.

### 2018-05-10 paper.tex
minor fixes with Sophie

@@
 For this reason, some virtual machine implementors attempted to remove most or all optional primitives.
-\sd{sentence was too long: please reread my version:}Their goal was to use a JIT performing speculative optimizations of optional primitives implemented in the client language.
+Their goal was to use a JIT performing speculative optimizations to be able to implement the optional primitives in the client language without performance loss.
 Ideally we wanted to get   the performance of the same primitives written in the VM.
-Self \cite{ursPHD} and Strongtalk \cite{Strongtalk} were \sk{the?} to first to try this approach.
+Self \cite{ursPHD} and Strongtalk \cite{Strongtalk} were the first to try this approach.
 Early versions of the Javascript engine V8 \cite{V8} also implemented most of the primitives, including for example Array operations in Javascript itself.
@@
 \begin{itemize}
-\item External plugins. Plugins that can be added or removed from the VM without the need to recompile the VM sources.
-\item Built-in plugins.
-Plugins that cannot be installed without recompiling the VM sources.
+\item \emph{External plugins:} Plugins that can be added or removed from the VM without the need to recompile the VM sources.
+\item \emph{Internal plugins:} Plugins that cannot be installed without recompiling the VM sources.
 However, these plugins can reach some parts of the VM internals\footnote{as data structures, routines...} that remains unreachable from an external plugin.
@@
 Although, this solution results in a substential overhead if called often.
+\cba{Remove from paper InterpreterProxy}
 \end{itemize}
@@
 
-\cba{We may use paragraph instead of itemize here to compact}
  In this section we compare the five different implementations according to the following criteria:
@@
 External plugins don't have access to the whole VM API and have to go through InterpreterProxy to do so, leading to an overhead.
+\cba{InterpreterProxy not defined before}
 	\item \emph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code.
@@
 
-\paragraph{GCC-Linux.}The GCC-Linux evaluation was performed on an Asus Zenbook with Ubuntu 16.04 LTS, a 2,2 Ghz processor Intel Core 5, 8 Gb 1600MHz DDR3 of RAM.
+\paragraph{GCC-Linux.}The GCC-Linux evaluation was performed on an Asus Zenbook with Ubuntu 16.04 LTS, a 2.2 Ghz processor Intel Core 5, 8 Gb 1600MHz DDR3 of RAM.
 The Linux VM was compiled using GCC version 5.4.0.
 
-\paragraph{LLVM-Mac.}The LLVM-Mac evaluation was performed on a MacBook pro with Mac OS 10.11.6, a 2.9Ghz processor Intel Core 5, 8 Gb 1867MHz DDR3 of RAM.
+\paragraph{LLVM-Mac.}The LLVM-Mac evaluation was performed on a MacBook pro with Mac OS 10.11.6, a 2.9 Ghz processor Intel Core 5, 8 Gb 1867MHz DDR3 of RAM.
 The Mac VM was compiled using Apple LLVM version 8.0.0 (clang-800.0.38).
@@
 The JIT overhead is negligible here for 1 second because we could not see any time spent in the JIT in the profiler + ref to your paper, i.e.
-less than 0.5ms in the JIT} and allows the benchmarks to run at least 1 second in the slower implementations \cba{Ok, 1 second to avoid processor / OS noises, not just for fun}.
+less than 0.5ms in the JIT} \cite{CogProfiler} and allows the benchmarks to run at least 1 second in the slower implementations \cba{Ok, 1 second to avoid processor / OS noises, not just for fun}.
 The loop overhead is measured and removed from the measurements.
@@
 This last observation only applies to the GCC results: the code generated by the LLVM compiler induces a longer execution time, as explained in the next paragraph.
-\sk{is it legal to ref that way?} \cba{it's ok, but not recommended since sometimes we move things around without checking other sections}
-\paragraph{Control over generated code.}
+
 One could notice a rather striking performance difference between the 2 compilers: when compiling with GCC, the performances between the PureSlang and Slang+JIT implementations are really close for long strings (100 characters and longer), with respectively 24,42 and 24,48 relative speed-up factors for a 1000-characters string.

### 2018-05-12 paper.tex
using [DT90] instead of [5]
up to sectio 3

@@
 %% Bibliography style
-\bibliographystyle{ACM-Reference-Format}
+
+%\bibliographystyle{ACM-Reference-Format}
+
 %% Citation style
@@
 In a hybrid runtime, where code is executed by an interpreter and a just-in-time compiler, the implementor can choose to implement optional primitives in the client language, in the virtual machine implementation language (typically C or C++), or on top of the just-in-time compiler back-end.
-This raises the questions of the tradeoffs of the different alternatives.
-We implemented the String comparison optional primitive in each case.
+This raises the question of the tradeoffs of the different alternatives.
+As a case study, we implemented the String comparison optional primitive in each case.
 This paper describes the different implementations and compares the execution time in Cog, a Smalltalk virtual machine.
 
-Without surprise, the paper shows that the implementation with the most reliable and high performance is also the most complex and difficult to implement.
-Since hundreds of optional primitives could be wished, the virtual machine implementor has to carefully choose how to implement each of them to balance between maintenance cost and high performance.
+Without surprise, the paper shows that the implementation with the most reliable and highest performance is also the most complex and difficult to implement.
+Nevertheless, since hundreds of optional primitives could be wished, the virtual machine implementor has to carefully choose how to implement each of them to balance between maintenance cost and high performance.
 \end{abstract}
@@
 
+\paragraph{Optional primitives as client code.}
 The implementation of an optional primitive in the VM rather than the client language improves performance but also often increases the implementation engineering cost.
@@
 Their goal was to use a JIT performing speculative optimizations to be able to implement the optional primitives in the client language without performance loss.
-Ideally we wanted to get   the performance of the same primitives written in the VM.
+Ideally they wanted to get   the performance of the same primitives written in the VM.
 Self \cite{ursPHD} and Strongtalk \cite{Strongtalk} were the first to try this approach.
@@
 \begin{itemize}
-	\item High performance relies on a JIT with speculative optimizations, which is hard to implement, maintain and evolve,
-	\item The performance of the code not yet optimized is much worse than the peak performance code,
-	\item Even mature JIT fail to optimize some narrow cases, where the performance is drastically slower.
+	\item High performance relies on a JIT with speculative optimizations, which is hard to implement, maintain and evolve.
+	\item The performance of the code not yet optimized is much worse than the peak performance code \sd{I did not really get it. indeed non optimized code is slower than optimized code. So what?}.
+	\item Even mature JITs fail to optimize some narrow cases, where the performance is drastically slower. \sd{do you have a ref?}
 \end{itemize}
@@
 
+\paragraph{World border cost.}
 VMs are traditionally implemented in a low-level language such as C or C++.
@@
 
-Section \ref{sec:implContext} describes our implementation context, \emph{i.e.,} the execution model of the VM we used for our evaluation, how it executes primitives and the specification of our string comparison primitive.
-Section \ref{sec:implem} details the various implementations of the primitive we used for the evaluation.
-In Section \ref{sec:eval}, we evaluate the performance of the different implementations with different C compilers and string sizes.
+Section~\ref{sec:implContext} describes our implementation context, \emph{i.e.,} the execution model of the VM we used for our evaluation, how it executes primitives and the specification of our string comparison primitive.
+Section~\ref{sec:implem} details the various implementations of the primitive we used for the evaluation.
+In Section~\ref{sec:eval}, we evaluate the performance of the different implementations with different C compilers and string sizes.
 We also show the native code of the performance critical part of the primitive generated by each version and discuss it.
@@
 
-Our evaluation is based on the Cog VM a Smalltalk virtual machine.
+Our evaluation is based on the Cog VM: a Smalltalk virtual machine.
 The Cog VM is based on Dan Ingalls' Smalltalk VM \cite{BackToTheFuture} and has been enhanced with a JIT \cite{CogJIT} compiling one method or closure at a time to native code.
@@
 
-Section \ref{sec:VMCompil} describes briefly the implementation languages and compilation process of the Cog VM.
-Section \ref{sec:VMexec} explains briefly the hybrid execution model with an interpreter and a JIT.
-In Section \ref{sec:primExec}, we detail how primitives are executed.
-Section \ref{subsec:primSpec} discusses the string comparison primitive specifications  used for our evaluation.
+Section~\ref{sec:VMCompil} describes briefly %\sd{no it does not describe that: the implementation languages and}
+the compilation process of the Cog VM.
+Section~\ref{sec:VMexec} explains briefly the hybrid execution model with an interpreter and a JIT.
+In Section~ \ref{sec:primExec}, we detail how primitives are executed.
+Section~\ref{subsec:primSpec} discusses the string comparison primitive specifications  used for our evaluation.
 
@@
 
-VM plugins enable the addition of new features to the VM. There are 2 types of plugins: the built-ins and the externals.
+Cog VM can be extended using plugins \cite{Guzd01b}.
+VM plugins enable the addition of new features to the VM. There are 2 types of plugins:  the external and the  built-ins ones.
 \begin{itemize}
@@
 Because of this design, switching the execution between the interpreter and the native code generated by the JIT has a cost.
-Effectively, the VM needs to switch the frame pointer, the stack pointer and depending on architecture the link register from one stack to the other one.
+Effectively, the VM needs to switch the frame pointer, the stack pointer and depending on architecture the link register \sd{I do not get "the link register"} from one stack to the other one.
 It also needs to save those pointers to be able to resume execution from one stack to the other one.
@@
 \cba{Cette partie je faisais la queue dans la gare en mm temps, a relire}
-
+\sd{je ne suis pas sur que je garderais la specification originale car cela n'apporte pas grand-chose.}
  The primitive specification is a bit tedious to explain, since as we were trying to improve its performance, we had to change the specifications.
@@
 
- This section first describes the different ways of implementing a primitive. We then explain how each of these versions are executed by the VM and the pros and cons of each and every.
+ This section first describes the different ways of implementing a primitive. We then explain how each of these versions is executed by the VM and the pros and cons of each and every.
 
@@
 %% Bibliography
-\bibliography{sista,rmod}
+\bibliographystyle{alpha}
+\bibliography{sista,rmod,others}
 

### 2018-05-13 paper.tex
typos

@@
-%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
+%%%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
 \documentclass[10pt, sigplan]{acmart}
@@
 There are multiple ways of writing the comparison of two strings in Smalltalk, following Smalltalk coding conventions, a standard developer would likely use high-level constructs which apply a closure on each element of the strings.
-We did not use high-level construct and chose to implement the Smalltalk version the most optimized way Smalltalk allows, \emph{i.e.,} we used the contructs to:do: and ifFalse:, which are both recognized by the Smalltalk to bytecode compiler and compiled respectively to loops and branches at the bytecode level.
+We did not use high-level construct and chose to implement the Smalltalk version the most optimized way Smalltalk allows, \emph{i.e.,} we used the constructs to:do: and ifFalse:, which are both recognized by the Smalltalk to bytecode compiler and compiled respectively to loops and branches at the bytecode level.
 The implementation follows the new specification, hence two methods are available, baselineCompareWith: and baselineCompareWith:order:.
@@
   \hline
-    & Recompilation & Familiar to & Lignes & Number of & Stack switch & Optimization & Code generation \\
+    & Recompilation & Familiar to & Lines & Number of & Stack switch & Optimization & Code generation \\
     & required & developers & of code & implementations & overhead & potential & control \\
@@
 
-\cba{In this section somewhere we need to say that criteria are split in maintenance and performance, the evaluation in next section is on last 3 criteria: Stack switch overhead, Optimization potential and control over the generated code.}
+\cba{In this section somewhere we need to say that criteria are split in maintenance and performance, the evaluation in next section is on last three criteria: Stack switch overhead, Optimization potential and control over the generated code.}
 
@@
 \paragraph{Stack switch overhead.}
-As mentionned in Section \ref{sec:VMexec}, all implementations written in Slang come with an overhead due to the switch between the C and Smalltalk stacks.
+As mentioned in Section~\ref{sec:VMexec}, all implementations written in Slang come with an overhead due to the switch between the C and Smalltalk stacks.
 This overhead is clearly visible in the bench results in Figure \ref{fig:bench} for small strings (size 1 or less): the SmartSyntax and Plugin implementations are around 30 to 60\% slower than the baseline's, and the PureSlang performances are close to the baseline's.
-As the string size goes larger, the overhead is less significant because it is **absorbed** by other performance gains.
+As the string size goes larger, the overhead is less significant because it is absorbed by other performance gains.
 Regarding this criteria, the Slang+JIT version is always the fastest: it is not really impacted by the switch overhead, because the Slang part is used for a very short time.
@@
 Just not optimized.
-need rephrasing} due to the cross-file inlining and the passage through an InterpreterProxy, like said in Section \ref{sec:implem_cmp}.
+need rephrasing} due to the cross-file inlining and the passage through an InterpreterProxy, like said in Section~\ref{sec:implem_cmp}.
 This is one of the factor explaining the performance gap between these versions and the Slang and Slang+JIT ones.
@@
 In some languages (Strongtalk, Self, RSqueak), the implementors tried to decrease the overall number of optional primitives by improving the client language performance, reducing the need for such primitives \cite{Strongtalk, ursPHD, NoPrimTracing}.
-In addition, in the V8 project  \cite{V8}, the interpreter was entirely built on top of the V8 JIT's back-end using a specific DSL similar to Cog's RTL.
+In addition, in the V8 project~\cite{V8}, the interpreter was entirely built on top of the V8 JIT's back-end using a specific DSL similar to Cog's RTL.
 
@@
 To avoid paying the stack switch cost between the client Stack and the C stack, the V8 team recently wrote an interpreter Ignition in a DSL compiled ahead of time through their JIT back-end to native code.
-With their primtives and their interpreter written in the DSL, the V8 runtime rarely needs to switch between the client and the C stack (no need to switch to interpret code, no need to switch for primitives).
+With their primitives and their interpreter written in the DSL, the V8 runtime rarely needs to switch between the client and the C stack (no need to switch to interpret code, no need to switch for primitives).
 In this context, the stack switch overhead is not a problem anymore.

### 2018-05-13 paper.tex
paper.tex

@@
 
-The Smart Syntax Plugin is one of the version existing until we started this work.
-The idea behind Smart Syntax was the following: the programmer can write code in Smalltalk using exclusively constructs that are recognized by the Smart Syntax compiler and that can be directly compiled to C (computing the size of a byte array, accessing its fields and doing integer arithmetic for example).
-With the addition of type annotations in the form of pragmas \cite{Duca16a} \footnote{Pragma are a form of method annotation.}, the Smart Syntax plugin was able to generate C code from the Smalltalk code.
+The Smart Syntax Plugin is one of the version \sd{of what} existing until we started this work.
+The idea behind Smart Syntax is the following: the programmer writes code in Smalltalk using exclusively constructs that are recognized by the Smart Syntax compiler and that can be directly compiled to C (\emph{e.g.}, computing the size of a byte array, accessing its fields and doing integer arithmetic).
+With the addition of type annotations in the form of pragmas \cite{Duca16a} \footnote{Pragma are method annotation.}, the Smart Syntax plugin is able to generate C code from the Smalltalk code.
+\sd{A word on QuickTalk \cite{Ball86a}}
 With this approach, the programmer has to write the optional primitive code a single time in Smalltalk.
 This code is used both as the fall-back code if the VM does not implement the primitive and at Slang to C compilation time to generate a C plugin from the code and the type annotations.
-This primitive follows the old specification, the full code is available in Appendix \ref{SmartSyntaxAppendix}.
+This primitive follows the old specification, the full code is available in Appendix~\ref{SmartSyntaxAppendix}.
+\sd{do we have a ref to the quickTalk approach? Because it was like that too if I remember.}
 
 \paragraph{Existing: Slang Plugin.}
-
 The Slang plugin is a recent complete rewrite of the Smart Syntax plugin.
-The main difference is that the Smalltalk version is used only as fall-back code, while at Slang to C compilation time the VM uses code written in Slang to generate the C code.
+The main difference is that the Smalltalk version is used only as fall-back code, while at Slang-to-C compilation time the VM uses code written in Slang to generate the C code.
 One of the side-effect is that the plugin's code is now part of the VM code base and not the client code bases, which is convenient for maintenance since the Cog VM has multiple client languages.
 However, the implementor now needs to maintain and evolve two version of the primitive, the fall-back code in Smalltalk and the code in Slang.
-This primitive follows the old specification, the full code is available in Appendix \ref{PluginAppendix}.
+\sd{do we need the next sentence?}This primitive follows the old specification, the full code is available in Appendix~\ref{PluginAppendix}.
 

### 2018-05-13 paper.tex
oing over comparison

@@
 
-The Slang implementation is written inside the interpreter and has access to all interpreter APIs.
-Since the same primitive index is used for the primitive with 2 or 3 arguments, the primitive needs to deal with both cases in a way that the common case is efficient.
+The Slang implementation is written inside the interpreter and has access to \sd{sd?full instead all} full interpreter APIs.
+Since the same primitive index \sd{no idea what is it} is used for the primitive with 2 or 3 arguments, the primitive needs to deal with both cases in a way that the common case is efficient.
 The primitive is now part of the core of the VM, it cannot be easily added or removed as plugin primitives, but it is simpler to read and write.
@@
 In addition to the previous implementation, we've re-implemented here the primitive on top of Cog's JIT back-end.
-The primitive is written in Cog's RTL, an abstract assembly which instructions compile almost one-to-one to native instructions.
-This primitive requires a more important amount of work to be written: the number of line of code is greater, the implementor has to be careful to generate efficient machine code (there is no C compiler to optimize the code) and debugging such code requires the use of a processor simulator.
-Since only performance critical implementations matter in this context, we've re-implemented only the case where there is 2 parameters.
-For the three parameter primitive, the VM falls-back to the Slang version.
+The primitive is written in Cog's RTL, an abstract assembly which \sd{whose?} instructions compile almost one-to-one to native instructions.
+This primitive requires a more important amount of work to be written: the number of line of code is larger, the implementor has to be careful to generate efficient machine code (there is no C compiler to optimize the code) and debugging such code requires the use of a processor simulator \sd{such as Bosch}.
+Since only performance critical implementations matter in this context, we've only re-implemented the case where there are two parameters.
+For the three parameter primitive, the VM falls back to the Slang version.
 This primitive follows the new specification, the full code is available in Appendix \ref{RTLAppendix}.
@@
  	\item \emph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (Plugin) or requires to recompile the full VM (Yes).
+\sd{why not ST, Plugin, VM?}
 	\item \emph{Implementation language familiar to developers:} Smalltalk application developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code.
@@
 We use \emph{All} if all communities are familiar with the language, \emph{VM} if all VM developers are and \emph{VM-JIT} if only a subset of VM developers are.
-	\item \emph{Readability:} To evaluate the readability of the approach, we use one of the most used metric, the number of lines of code.
+	\item \emph{Readability:} To evaluate the approach readability, we use the number of lines of code even if there is no direct link between readability and mere lines of code.
+\sd{well this is kind of weak.
+but I do not know what do propose.
+Cyclomatic complexity counts the number of different exits.
+And it would be a nice complement}
 	\item \emph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to execute if the primitive fails.

### 2018-05-13 paper.tex
final pass

@@
 
-\paragraph{Pure Slang with JIT support.}
+\paragraph{Pure Slang with JIT support.}\sd{Why not calling it Slang+RTL}
 
@@
 Only Cog's JIT implementors are familiar with its RTL, which is a subset of the VM developers.
-We use \emph{All} if all communities are familiar with the language, \emph{VM} if all VM developers are and \emph{VM-JIT} if only a subset of VM developers are.
+We use All if all communities are familiar with the language, VM if all VM developers are and VM-JIT if only a subset of VM developers are.
 	\item \emph{Readability:} To evaluate the approach readability, we use the number of lines of code even if there is no direct link between readability and mere lines of code.
@@
 And it would be a nice complement}
-	\item \emph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to execute if the primitive fails.
-This code is maintain by the client language implementors, and not by the VM implementors.
-Hence we do not count the fall-back code in the number of implementation to maintain, nor the pure Smalltalk implementation.
+	\item \emph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to be executed if the primitive fails.
+This code is maintained by the client language implementors, and not by the VM implementors.
+Hence we do not count the fall-back code in the number of implementation to maintain, nor the pure Smalltalk implementation\sd{why?}.
 However, as a VM implementor, we need to maintain what generates the C code of the VM, hence, in the case of Smart Syntax, we need to maintain a different method in each client language.
@@
 
-One or another implementation could be preferred depending on the intended aim: the main trade-off will be found between the engineering time and the performance.
-Table \ref{tab:cmp_implem} summarizes the main pros and cons related to each implementation: the engineering time may tend to increase when more low-level languages are used, such as Cog's RTL for instance.
+One or another implementation could be preferred depending on the intended aim: the main trade-off is between  engineering time and  performance.
+Table \ref{tab:cmp_implem} summarizes the main pros and cons related to each implementation: the engineering time tends to increase when more low-level languages are used, such as Cog's RTL for instance.
 Also, a VM recompilation affects the overall complexity of the development process, but the resulting code might be faster to run.
-Likewise, the use of plugins may bring modularity, but as you will see in Section \ref{sec:eval}, the performance, although better than baseline version, could still be improved.
+Likewise, the use of plugins brings modularity, but as you will see in Section~\ref{sec:eval}, the performance, although better than baseline version, could still be improved.
 
@@
 The SlangPlugin, PureSlang and Slang+JIT implementations of this string comparison primitive have been recently implemented in the Cog VM.
-In this section, we will compare the performances of these primitives against the other existing implementations performances.
-We have measures their respective execution times for different string sizes.
+In this section, we compare execution speed of these primitives against the other existing implementations.
+We measure their respective execution times for different string sizes.
+\sd{mention microbenchmarks here.}\sd{why there is only 4.3 and not something else?
+Because this is strange from a shape structure.}
 
@@
     \end{subfigure}
-   \caption{String comparison benchmarks for strings of different sizes on different architectures}
+   \caption{String comparison benchmarks for strings of different sizes on different architectures.}
    \label{fig:bench}
@@
 
-The two compared strings are equal as it corresponds to the worst scenario for most of our benched implementations. There is one less branch in the two plugin implementations in this case.
+The two compared strings are equal as this situation corresponds to the worst scenario for most implementations. There is one less branch in the two plugin implementations in this case.
 
@@
 We performed the benchmarks upon five different string sizes: 0, 1, 5, 100 and 1000 characters, using two different compilers: GCC on one hand and LLVM on the other hand.
-Figure \ref{fig:bench} show the results in terms of relative speed-up towards the baseline implementation, respectively for GCC and LLVM.
+Figure~\ref{fig:bench} shows the results in terms of relative speed-up towards the baseline implementation, respectively for GCC and LLVM.
 
@@
 \begin{itemize}
-\item The execution time of the SmartSyntax and Plugin implementations is either very close or greater than the baseline's for the small strings.
+\item The execution time of the SmartSyntax and Plugin implementations is either very close or greater than \sd{within which ranges? 10\% 200\%} the baseline's for the small strings.
 \item The Slang+JIT implementation is more efficient than the baseline in every case.
-Nevertheless,the performance gap between this version and the PureSlang one tends to decrease as the string size increases.
+Nevertheless, the performance gap between this version and the PureSlang one tends to decrease as the string size increases.
 \item The compiler used (either GCC or LLVM) does not lead to significant differences in terms of performance, except for the 100+-characters strings. 
@@
 
-We can explain these results based on the last 3 criteria detailed in Section \ref{sec:implem_cmp}:
+We can explain these results based on the last 3 criteria detailed in Section~\ref{sec:implem_cmp}:
 
@@
 As mentioned in Section~\ref{sec:VMexec}, all implementations written in Slang come with an overhead due to the switch between the C and Smalltalk stacks.
-This overhead is clearly visible in the bench results in Figure \ref{fig:bench} for small strings (size 1 or less): the SmartSyntax and Plugin implementations are around 30 to 60\% slower than the baseline's, and the PureSlang performances are close to the baseline's.
+This overhead is clearly visible in the bench results in Figure~\ref{fig:bench} for small strings (size 1 or less): the SmartSyntax and Plugin implementations are around 30 to 60\% slower than the baseline's, and the PureSlang performance is close to the baseline's.
 As the string size goes larger, the overhead is less significant because it is absorbed by other performance gains.
-Regarding this criteria, the Slang+JIT version is always the fastest: it is not really impacted by the switch overhead, because the Slang part is used for a very short time.
+Regarding this criteria, the Slang+JIT version is always the fastest: it is not really impacted by the switch overhead, because the Slang part is used for a very short time \sd{Why that?
+just during the time the JIT does not kick in?}.
 
@@
 This is one of the factor explaining the performance gap between these versions and the Slang and Slang+JIT ones.
-When it comes to comparing large strings (100 characters and more), the execution time is mainly spent in the comparison loop (Figure \ref{fig:Asm}).
-However, the switch overhead is still noticeable in the PureSlang implementation when strings of 100 characters are compared: PureSlang is 15.8 times faster while Slang+JIT is 16.8 times faster.
+When it comes to comparing large strings (100 characters and more), the execution time is mainly spent in the comparison loop (Figure~\ref{fig:Asm}).
+However, the switch overhead is still noticeable in the PureSlang implementation when strings of 100 characters are compared: PureSlang is 15.8 times faster \sd{than what} while Slang+JIT is 16.8 times faster \sd{than what}.
 This is not the case anymore on 1000-characters strings.
@@
 
-One could notice a rather striking performance difference between the 2 compilers: when compiling with GCC, the performances between the PureSlang and Slang+JIT implementations are really close for long strings (100 characters and longer), with respectively 24,42 and 24,48 relative speed-up factors for a 1000-characters string.
-Yet they greatly differ when LLVM is used, with respectively 13,44 and 26,40 relative speed-up factors.
-
-This difference is mainly due to the optimizations performed by both the compilers: as depicted in Figure \ref{fig:Asm}, GCC generates instructions that are very close to the ones generated by the JIT, hence the similar speed-up factors.
-On the contrary, LLVM generates more instructions: it especially fails at preserving the registries holding the strings (edi and esi), leading to 2 extra instructions (bolded in the figure) that will load the strings at each iteration.
+One could notice a rather striking performance difference between the 2 compilers: when compiling with GCC, the performance between the PureSlang and Slang+JIT implementations is really close for long strings (100 characters and longer), with respectively 24,42 and 24,48 relative speed-up factors for a 1000-characters string.
+Yet performance greatly differs when LLVM is used, with respectively 13,44 and 26,40 relative speed-up factors.
+
+This difference is mainly due to the optimizations performed by both the compilers: as depicted in Figure~\ref{fig:Asm}, GCC generates instructions that are very close to the ones generated by the JIT, hence the similar speed-up factors.
+On the contrary, LLVM generates more instructions: it especially fails at preserving the registries \sd{registers?} holding the strings (edi and esi), leading to two extra instructions (bolded in the figure) that  load the strings at each iteration.
 
@@
 In this context, the need of primitive is reduced, since the performance gain from using a primitive over the client language is low or inexistant.
-More recently, Felgentreff \emph{et al.} \cite{NoPrimTracing} analysed the performance difference between primitives implemented in C, RPython, and Smalltalk.
-They showed that with the RPython toolchain framework, the performance difference between all three implementation was not that big, hence they could choose to implement some optional primitives in the client language over C or RPython without performance drop to reduce the number of primitives to maintain.
+More recently, Felgentreff \emph{et al.} \cite{NoPrimTracing} analysed the speed difference between primitives implemented in C, RPython, and Smalltalk.
+They showed that with the RPython toolchain framework, the speed difference between all three implementations is not that big, hence they could choose to implement some optional primitives in the client language over C or RPython without performance drop to reduce the number of primitives to maintain.
 
@@
 The performance problem is however partially solved: only the peak performance is similar with and without the primitives.
-Since those system rely on JITs, start-up performance is worse than peak performance, increasing the performance difference at start-up between the client code and the primitive.
+Since those systems rely on JITs, start-up performance is worse than peak performance, increasing the performance difference at start-up between the client code and the primitive.
 In addition, the performance can be  unreliable with optimizing JITs.
@@
 
-To avoid paying the stack switch cost between the client Stack and the C stack, the V8 team recently wrote an interpreter Ignition in a DSL compiled ahead of time through their JIT back-end to native code.
+To avoid paying the stack switch cost between the client Stack and the C stack, the V8 team recently wrote an interpreter Ignition \sd{what it is?} in a DSL compiled ahead of time through their JIT back-end to native code.
 With their primitives and their interpreter written in the DSL, the V8 runtime rarely needs to switch between the client and the C stack (no need to switch to interpret code, no need to switch for primitives).
@@
 
-All the primitive we compare are written in a processor independent way, due to the C compiler abstraction for the Slang versions, the abstract assembly abstraction for Cog's RTL version or the VM abstraction for the baseline implementation in Smalltalk.
+All the primitives we compared are written in a processor independent way, due to the C compiler abstraction for the Slang versions, the abstract assembly abstraction for Cog's RTL version or the VM abstraction for the baseline implementation in Smalltalk.
 We could rewrite the primitive differently for each back-end currently supported (x86, x64, MIPSEL, ARMv6) and yet to support (ARMv8, ...).
-It could be relevant in some cases performance-wise, for example, on intel processors, the JVM implements string comparison using the SSE4.2 string comparison instructions when available \cite{JVMSSE4StringComp}.
-We did not go in this direction because the maintenance cost in this case is definitely too high since we would have to maintain a different implementation per back-end  (already 4 implementations today).
+It could be relevant in some cases performance-wise, for example, on Intel processors, the JVM implements string comparison using the SSE4.2 string comparison instructions when available~\cite{JVMSSE4StringComp}.
+We did not go in this direction because the maintenance cost in this case is definitely too high since we would have to maintain a different implementation per back-end (already 4 implementations today \sd{that we have or that we would have}).
 

### 2018-05-13 paper.tex
adding clement's phd

@@
 
-For the past few years, an experimental adaptive optimizer (a JIT featuring advanced optimizations such as speculative inlining) has been developed on top of the Cog VM \cite{SistaFastStartUp}.
+For the past few years, an experimental adaptive optimizer (a JIT featuring advanced optimizations such as speculative inlining) has been developed on top of the Cog VM \cite{SistaFastStartUp,Bera17a}.
 The optimizer has specific rules to be able to inline primitive operations without performance loss.

### 2018-05-13 paper.tex
with clement and sophie

@@
 	\item The performance of the code not yet optimized is much worse than the peak performance code \sd{I did not really get it. indeed non optimized code is slower than optimized code. So what?}.
-	\item Even mature JITs fail to optimize some narrow cases, where the performance is drastically slower. \sd{do you have a ref?}
+	\item Even mature JITs fail to optimize some narrow cases, where the performance is drastically slower. \sd{do you have a ref?}\sd{clement said: Hotspot paper}
 \end{itemize}
@@
 Cog VM can be extended using plugins \cite{Guzd01b}.
-VM plugins enable the addition of new features to the VM. There are 2 types of plugins:  the external and the  built-ins ones.
+VM plugins enable the addition of new features to the VM. There are two types of plugins:  the external and the  built-ins ones.
 \begin{itemize}
@@
 		\includegraphics[width=0.99\linewidth]{figures/TwoStacks}
-		\caption{Client and C stacks}
+		\caption{Client and C stacks.}
 		\label{fig:TwoStacks}
@@
  \subsection{Different implementations and executions}
+We distinguish several implementations: the first one
 
@@
 This primitive follows the old specification, the full code is available in Appendix~\ref{SmartSyntaxAppendix}.
-\sd{do we have a ref to the quickTalk approach? Because it was like that too if I remember.}
 
@@
 
-The Slang implementation is written inside the interpreter and has access to \sd{sd?full instead all} full interpreter APIs.
-Since the same primitive index \sd{no idea what is it} is used for the primitive with 2 or 3 arguments, the primitive needs to deal with both cases in a way that the common case is efficient.
+The Slang implementation is written inside the interpreter and has access to the full interpreter APIs.
+Since the same primitive index \sd{no idea what is it} is used for the primitive with two or three arguments, the primitive needs to deal with both cases in a way that the common case is efficient.
 The primitive is now part of the core of the VM, it cannot be easily added or removed as plugin primitives, but it is simpler to read and write.
@@
 In addition to the previous implementation, we've re-implemented here the primitive on top of Cog's JIT back-end.
-The primitive is written in Cog's RTL, an abstract assembly which \sd{whose?} instructions compile almost one-to-one to native instructions.
-This primitive requires a more important amount of work to be written: the number of line of code is larger, the implementor has to be careful to generate efficient machine code (there is no C compiler to optimize the code) and debugging such code requires the use of a processor simulator \sd{such as Bosch}.
+The primitive is written in Cog's RTL, an abstract assembly whose instructions compile almost one-to-one to native instructions.
+This primitive requires a more important amount of work to be written: the number of line of code is larger, the implementor has to be careful to generate efficient machine code (there is no C compiler to optimize the code) and debugging such code requires the use of a processor simulator \sd{such as Boch}.
 Since only performance critical implementations matter in this context, we've only re-implemented the case where there are two parameters.
@@
 \end{tabular}
-\caption{Comparison between implementations\vspace{-0.5cm}}
+\caption{Comparison between implementations.\vspace{-0.5cm}}
 \label{tab:cmp_implem}
@@
 We use All if all communities are familiar with the language, VM if all VM developers are and VM-JIT if only a subset of VM developers are.
-	\item \emph{Readability:} To evaluate the approach readability, we use the number of lines of code even if there is no direct link between readability and mere lines of code.
-\sd{well this is kind of weak.
-but I do not know what do propose.
-Cyclomatic complexity counts the number of different exits.
-And it would be a nice complement}
+	\item \emph{Readability:} To evaluate the approach readability, we use the number of lines of code even if this is arguable but used regularly in other papers \cite{xxx}.
+A better metric would be to use the cyclomatic complexity, however since the  Cog' s RTL generates branches at runtime, it is difficult to use it.
 	\item \emph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to be executed if the primitive fails.
@@
 Likewise, the use of plugins brings modularity, but as you will see in Section~\ref{sec:eval}, the performance, although better than baseline version, could still be improved.
-
-\cba{In this section somewhere we need to say that criteria are split in maintenance and performance, the evaluation in next section is on last three criteria: Stack switch overhead, Optimization potential and control over the generated code.}
 
@@
 \vspace{0.2cm}
-\caption{Benchmark results with standard errors. Average relative speed-up regarding baseline}
+\caption{Benchmark results with standard errors. Average relative speed-up regarding baseline.}
 \label{tab:detail_bench}
@@
 		\includegraphics[width=1\linewidth]{figures/asm_cmp.pdf}
-		\caption{GCC and LLVM: different optimizations on the comparison loop}
+		\caption{GCC and LLVM: different optimizations on the comparison loop.}
 		\label{fig:Asm}
@@
 When it comes to comparing large strings (100 characters and more), the execution time is mainly spent in the comparison loop (Figure~\ref{fig:Asm}).
-However, the switch overhead is still noticeable in the PureSlang implementation when strings of 100 characters are compared: PureSlang is 15.8 times faster \sd{than what} while Slang+JIT is 16.8 times faster \sd{than what}.
+However, the switch overhead is still noticeable in the PureSlang implementation when strings of 100 characters are compared: PureSlang is 15.8 times faster than Baseline while Slang+JIT is 16.8 times faster than Baseline.
 This is not the case anymore on 1000-characters strings.
@@
 This difference is mainly due to the optimizations performed by both the compilers: as depicted in Figure~\ref{fig:Asm}, GCC generates instructions that are very close to the ones generated by the JIT, hence the similar speed-up factors.
-On the contrary, LLVM generates more instructions: it especially fails at preserving the registries \sd{registers?} holding the strings (edi and esi), leading to two extra instructions (bolded in the figure) that  load the strings at each iteration.
+On the contrary, LLVM generates more instructions: it especially fails at preserving registers holding the strings (edi and esi), leading to two extra instructions (bolded in the figure) that  load the strings at each iteration.
 
@@
 
-To avoid paying the stack switch cost between the client Stack and the C stack, the V8 team recently wrote an interpreter Ignition \sd{what it is?} in a DSL compiled ahead of time through their JIT back-end to native code.
+To avoid paying the stack switch cost between the client Stack and the C stack, the V8 team recently wrote an interpreter named \emph{Ignition} in a DSL.
+This interpreter is compiled ahead of time through their JIT back-end to native code.
 With their primitives and their interpreter written in the DSL, the V8 runtime rarely needs to switch between the client and the C stack (no need to switch to interpret code, no need to switch for primitives).
@@
 It could be relevant in some cases performance-wise, for example, on Intel processors, the JVM implements string comparison using the SSE4.2 string comparison instructions when available~\cite{JVMSSE4StringComp}.
-We did not go in this direction because the maintenance cost in this case is definitely too high since we would have to maintain a different implementation per back-end (already 4 implementations today \sd{that we have or that we would have}).
+We did not go in this direction because the maintenance cost is too high since we would have to maintain a different implementation per back-end.
+We already have four back-ends (x86, x64, MIPSEL, ARMv6) in production.
 

### 2018-05-16 paper.tex
replace interProxy by interface, take some other comments into account

@@
 
-\cba{We need to introduce here the concept of VM plugin and interpreterproxy indirection/overhead. This is very important since later we just say this is a plugin and we don't redefine it}
-\sk{done but someone better check what I wrote :-)}
-
 Cog VM can be extended using plugins \cite{Guzd01b}.
-VM plugins enable the addition of new features to the VM. There are two types of plugins:  the external and the  built-ins ones.
+VM plugins enable the addition of new features to the VM. There are two types of plugins:  the external and the internal ones.
 \begin{itemize}
@@
 \item \emph{Internal plugins:} Plugins that cannot be installed without recompiling the VM sources.
-However, these plugins can reach some parts of the VM internals\footnote{as data structures, routines...} that remains unreachable from an external plugin.
-These operations are performed via an InterpreterProxy.
-Although, this solution results in a substential overhead if called often.
-\cba{Remove from paper InterpreterProxy}
+However, these plugins can reach via an interface some parts of the VM internals such as data structures, routines,...
+that remains unreachable from an external plugin.
+This solution results in a substential overhead if called often.
 \end{itemize}
@@
 The primitive is written in Cog's RTL, an abstract assembly whose instructions compile almost one-to-one to native instructions.
-This primitive requires a more important amount of work to be written: the number of line of code is larger, the implementor has to be careful to generate efficient machine code (there is no C compiler to optimize the code) and debugging such code requires the use of a processor simulator \sd{such as Boch}.
+This primitive requires a more important amount of work to be written: the number of line of code is larger, the implementor has to be careful to generate efficient machine code (there is no C compiler to optimize the code) and debugging such code requires the use of a processor simulator such as Bochs.
 Since only performance critical implementations matter in this context, we've only re-implemented the case where there are two parameters.
@@
 	\item \emph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
-	\item \emph{Optimization potential:} \cba{need to mention that this is for large strings}
-	C cross-file inlining is not performed by default for internal plugins.
-External plugins don't have access to the whole VM API and have to go through InterpreterProxy to do so, leading to an overhead.
-\cba{InterpreterProxy not defined before}
+	\item \emph{Optimization potential on large string comparison:}
+	C cross-file inlining is not performed by default for internal plugins. External plugins don't have access to the whole VM API and have to go through an interface to do so, leading to an overhead.
 	\item \emph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code.
@@
 The JIT overhead is negligible here for 1 second because we could not see any time spent in the JIT in the profiler + ref to your paper, i.e.
-less than 0.5ms in the JIT} \cite{CogProfiler} and allows the benchmarks to run at least 1 second in the slower implementations \cba{Ok, 1 second to avoid processor / OS noises, not just for fun}.
+less than 0.5ms in the JIT} \cite{CogProfiler} and allows the benchmarks to run at least 1 second in the slower implementations to avoid processor / OS noises.
 The loop overhead is measured and removed from the measurements.
@@
 As the string size goes larger, the overhead is less significant because it is absorbed by other performance gains.
-Regarding this criteria, the Slang+JIT version is always the fastest: it is not really impacted by the switch overhead, because the Slang part is used for a very short time \sd{Why that?
-just during the time the JIT does not kick in?}.
+Regarding this criteria, the Slang+JIT version is always the fastest as it is not impacted by the switch overhead.
 

### 2018-05-16 paper.tex
add a bunch of comments

@@
  The primitive specification is a bit tedious to explain, since as we were trying to improve its performance, we had to change the specifications.
-In this section we describe first the string implementation, then the original primitive specification and why they were problematic and lastly the new specifications.
+\sk{rephrase bc subsections deleted} In this section we describe first the string implementation, then the original primitive specification and why they were problematic and lastly the new specifications.
 
@@
 
-\paragraph{Original specification.}  The original primitive took three operands.
+\paragraph{Original specification.}  \sk{delete}The original primitive took three operands.
 The two first operands were the two ByteStrings to compare and the third operand was the order table, a ByteArray encoding the order of characters.
@@
 
-\paragraph{Specification issues.} As we were trying to optimize string comparison but specifically string equality, we encountered two main issues:
+\paragraph{Specification issues.} \sk{delete} As we were trying to optimize string comparison but specifically string equality, we encountered two main issues:
 \begin{itemize}
@@
 
-\paragraph{Final specification.} The final specification we productized last month takes two or three parameters.
+\paragraph{Final specification.} \sk{rename and possibly rephrase}The final specification we productized last month takes two or three parameters.
 The third parameter, the order, is optional.
@@
 
+\sk{So, do we agree that: 1- we bench all implem with new spec (i am rewriting the plugins right now) 2- we drop the sections "original spec", "spec issues", rename "final spec" and rephrase the implementations details on the sections below?}
  This section first describes the different ways of implementing a primitive. We then explain how each of these versions is executed by the VM and the pros and cons of each and every.
@@
  \subsection{Different implementations and executions}
-We distinguish several implementations: the first one
+We distinguish several implementations: the first one \sk{delete "the first one"? sounds odd}
 
@@
 
-\paragraph{Existing: Smart Syntax Plugin.}
+\paragraph{Existing: Smart Syntax.}
 
@@
 
-\paragraph{Existing: Slang Plugin.}
+\paragraph{Existing: Plugin.}
 The Slang plugin is a recent complete rewrite of the Smart Syntax plugin.
@@
 
-\paragraph{Pure Slang with JIT support.}\sd{Why not calling it Slang+RTL}
+\paragraph{Slang+JIT.} \sk{Id like to keep slang+JIT and not slang+RLT as JIT may ring a bell for the non-smalltalkers (while RLT may not).
+but as we are already using a lot of ST words (primitives, Slang...), does that actually make any sense ?}
 
@@
  	\item \emph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (Plugin) or requires to recompile the full VM (Yes).
-\sd{why not ST, Plugin, VM?}
+\sd{why not ST, Plugin, VM?} \sk{Yes and No answers to the "recompilation needed" question, so maybe we should keep it that way}
 	\item \emph{Implementation language familiar to developers:} Smalltalk application developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code.

### 2018-05-16 paper.tex
fix after some comments

@@
 The Slang implementation is written inside the interpreter and has access to the full interpreter APIs.
-Since the same primitive index \sd{no idea what is it} is used for the primitive with two or three arguments, the primitive needs to deal with both cases in a way that the common case is efficient.
+This primitive takes with two or three arguments and needs to deal with both cases in a way that the common case is efficient.
 The primitive is now part of the core of the VM, it cannot be easily added or removed as plugin primitives, but it is simpler to read and write.
@@
 
-\subsection{Micro-benchmarks}
+\subsection{Micro-benchmarks} \sk{Results?}
 
@@
 \paragraph{Optimization potential on large strings comparison.}
-The plugins come with an overhead \sk{not an overhead when it comes to inlining.
-Just not optimized.
-need rephrasing} due to the cross-file inlining and the passage through an InterpreterProxy, like said in Section~\ref{sec:implem_cmp}.
-This is one of the factor explaining the performance gap between these versions and the Slang and Slang+JIT ones.
+The plugins come with an overhead due the use of an interface.
+Besides, cross-file inlining is not performed for plugins, like said in Section~\ref{sec:implem_cmp}.
+These are some of the factors explaining the performance gap between these versions and the Pure Slang and Slang+JIT ones.
 When it comes to comparing large strings (100 characters and more), the execution time is mainly spent in the comparison loop (Figure~\ref{fig:Asm}).

### 2018-05-16 paper.tex
remove old spec

@@
 \label{subsec:primSpec}
-
-\cba{Cette partie je faisais la queue dans la gare en mm temps, a relire}
-\sd{je ne suis pas sur que je garderais la specification originale car cela n'apporte pas grand-chose.}
- The primitive specification is a bit tedious to explain, since as we were trying to improve its performance, we had to change the specifications.
-\sk{rephrase bc subsections deleted} In this section we describe first the string implementation, then the original primitive specification and why they were problematic and lastly the new specifications.
+In this section we describe first the string representations then the string comparison primitive specification.
 
@@
 
-\paragraph{Original specification.}  \sk{delete}The original primitive took three operands.
-The two first operands were the two ByteStrings to compare and the third operand was the order table, a ByteArray encoding the order of characters.
-The default order is the ASCII order, hence each entry corresponds to its index in the byte array (The ByteArray has a size of 256, and accessing the element i, with 0 $\leq$ i $\leq$ 255, answered i).
-However, in some case, the program may do case insensitive ByteStrings comparison.
-In this case, the order ByteArray is different, since some entries answer the same value (for example, the entry for the 41 and 61, respectively characters A and a, both answer the same value).
-To compare two ByteStrings, the primitive conceptually iterates byte by byte over the two objects until it reaches the end of one them, comparing for each byte that the value present in the order ByteArray at the index is greater, equal or smaller than the value computed from the other ByteString.
-If a difference is met, the value returns based on the result, if not, at the end of the iteration the primitive compares the ByteString sizes and returns.
-Based on the three parameters, the primitive answered 3, 2, or 1 depending if the first ByteString was greater, equal or smaller than the second one.
-
-\paragraph{Specification issues.} \sk{delete} As we were trying to optimize string comparison but specifically string equality, we encountered two main issues:
-\begin{itemize}
-	\item \emph{Returned value:} Other programming languages implement the primitive to return a negative integer, 0 or a positive integer instead of 1, 2 or 3.
-It looks like a minor difference, but it is important for performance.
-With a negative, 0 or positive value, the primitive can just answer the difference between two different characters or the difference in size if all first characters are equal.
-With 1, 2, or 3, the primitive has to end with multiple branches to know what number to return.
-In native code, a subtraction is on common architecture quicker to perform than one or multiple branches.
-In addition, the primitive code is easier to write with the subtraction when writing directly the assembly instructions and we did exactly that when writing the primitive on top of the JIT's back-end.
-	\item \emph{Order:} Having the order as the third operands is convenient in some cases (case insensitive comparison), but is just overhead in the common case (checking for string equality).
-\end{itemize}
-
-\paragraph{Final specification.} \sk{rename and possibly rephrase}The final specification we productized last month takes two or three parameters.
-The third parameter, the order, is optional.
-If the third parameter is absent, the VM assumes the byte ordering (ASCII order) is the order to compare against.
-Therefore, the VM does not need to go through the indirection order array in the common case.
+\paragraph{Primitive specification.} The String comparison optional primitive takes two or three parameters.
+The two first operands are the two ByteStrings to compare and the third operand, optional, is the order table (a ByteArray of size 256 encoding the order of characters).
+If no order is specified, the VM assumes the byte ordering (ASCII order) is the order to compare against, and compares directly the bytes of the ByteStrings.
+If an order is specified, for example a case insensitive order (the entries for the 41 and 61 in the order byte array, respectively characters A and a in the ASCII table, both answer the same value), the primitive compare the bytes through the indirection order.
+Having the order optional allows to have quicker string comparison in the common case where the ASCII order is used.
 The primitive answers a negative integer, zero, or a positive integer as the first parameter is less than, equal to, or greater than the second parameter.
@@
 In the paper we will focus on the comparison of two strings with the default ASCII ordering and measure the performance only of this case.
-The goal of this work was to improve the performance of common string operations, typically string equality which internally use the primitive, and the cases where the order is non ASCII are uncommon.
+The goal of this work was to improve the performance of common string operations, typically string equality which internally uses the primitive, and the cases where the order is non ASCII are considered uncommon.
 
@@
  \subsection{Different implementations and executions}
-We distinguish several implementations: the first one \sk{delete "the first one"? sounds odd}
+
+We distinguish several implementations: the baseline in pure Smalltalk, the  \sk{delete "the first one"? sounds odd}
 
@@
 \def\legBase{Baseline}
-\def\legSmartSyntaxPlug{Smart syntax}
+\def\legSmartSyntaxPlug{Syntax}
 \def\legSlangPlug{Plugin}
@@
 \def\legRTL{Slang+JIT}
-\def\plotHeight{6.4cm}
+\def\plotHeight{6.7cm}
 

### 2018-05-17 paper.tex
Pass over all except evaluation

@@
 
+%%%
+% Legend definition (so I can change it once for all)
+%%%
+
+\def\legBase{Baseline}
+\def\legSmartSyntaxPlug{Syntax}
+\def\legSlangPlug{Plugin}
+\def\legSlang{Slang}
+\def\legRTL{Slang+RTL}
+\def\plotHeight{7cm}
+
+%%%
+% End Legend
+%%%
+
 %% Title information
@@
 In a hybrid runtime, where code is executed by an interpreter and a just-in-time compiler, the implementor can choose to implement optional primitives in the client language, in the virtual machine implementation language (typically C or C++), or on top of the just-in-time compiler back-end.
-This raises the question of the tradeoffs of the different alternatives.
+This raises the question of the maintenance and performance trade-offs of the different alternatives.
 As a case study, we implemented the String comparison optional primitive in each case.
-This paper describes the different implementations and compares the execution time in Cog, a Smalltalk virtual machine.
-
-Without surprise, the paper shows that the implementation with the most reliable and highest performance is also the most complex and difficult to implement.
-Nevertheless, since hundreds of optional primitives could be wished, the virtual machine implementor has to carefully choose how to implement each of them to balance between maintenance cost and high performance.
+This paper describes the different implementations, discusses the maintenance cost of each of them and evaluates for different string sizes the execution time in Cog, a Smalltalk virtual machine.
 \end{abstract}
@@
 	\item High performance relies on a JIT with speculative optimizations, which is hard to implement, maintain and evolve.
-	\item The performance of the code not yet optimized is much worse than the peak performance code \sd{I did not really get it. indeed non optimized code is slower than optimized code. So what?}.
-	\item Even mature JITs fail to optimize some narrow cases, where the performance is drastically slower. \sd{do you have a ref?}\sd{clement said: Hotspot paper}
+	\item There is a big performance gap between the baseline performance and the peak performance.
+	\item Even mature JITs fail to optimize some narrow cases, where the performance is drastically slower~\cite{VMHotCold}.
 \end{itemize}
+
 Overall, this approach requires high engineering time to get a mature optimizing JIT with speculative optimizations and even then, it leads in practice to unreliable performance.
@@
 
-\sd{the link is missing: In this article we want to measure the complexities and gain of this different implementations}
-
-In this paper, we take the example of a string comparison optional primitive.
+\paragraph{Measurements.}
+In this paper, we want to measure the complexities and gain of different implementations.
+For this purpose, we take the example of a string comparison optional primitive.
 The primitive compares two strings and answers if one string is greater, equal or lower than the other string.
-The exact specifications are detailed in Section \ref{subsec:primSpec}.
-We implemented a first version in Pharo/Smalltalk.
-We then implemented multiple version compiling through C compilers to native code, using different part of our VM infrastructure.
+The exact specification is detailed in Section \ref{subsec:primSpec}.
+The primitive is convenient for measurements since it can be executed with strings of small sizes or large sizes, leading the primitive to execute many processor instructions or only a few of them.
+
+We implemented a first version of the optional primitive in Pharo/Smalltalk.
+We then implemented two versions compiling through C compilers to native code, using different parts of our VM infrastructure.
 Lastly, we implemented a version in the back-end of the JIT.
 Each time, we tried to write the code in the most efficient way possible.
-Then, we compared the execution time of all versions, showing that overall the most complex implementation is the fastest to execute.
+Then, we compared the execution time of all versions on strings of different sizes, showing that overall the most complex implementation is the fastest to execute.
 
@@
 \label{sec:VMCompil}
+
+\begin{figure}[bth!]
+		\centering
+		\includegraphics[width=0.65\linewidth]{figures/VMCompilation}
+		\caption{Cog VM compilation}
+		\label{fig:VMCompilation}
+\end{figure}
 
@@
 
-\begin{figure}[bth]
-		\centering
-		\includegraphics[width=0.6\linewidth]{figures/VMCompilation}
-		\caption{Cog VM compilation}
-		\label{fig:VMCompilation}
-\end{figure}
-
 The executable is generated in two steps as shown on Figure~\ref{fig:VMCompilation}, similarly to the RPython toolchain \cite{RPythonToolchain}.
@@
 
-Cog VM can be extended using plugins \cite{Guzd01b}.
-VM plugins enable the addition of new features to the VM. There are two types of plugins:  the external and the internal ones.
+\paragraph{Plugins.}The Cog VM can be extended using plugins \cite{Guzd01b}.
+VM plugins enable the addition of new features to the VM.
+The main benefit of writing a plugin over extending the VM is modularity, plugins are in separate code bases and can be easily added or removed from the VM.
+Plugins can be compiled as internal or external:
 \begin{itemize}
-\item \emph{External plugins:} Plugins that can be added or removed from the VM without the need to recompile the VM sources.
-\item \emph{Internal plugins:} Plugins that cannot be installed without recompiling the VM sources.
-However, these plugins can reach via an interface some parts of the VM internals such as data structures, routines,...
-that remains unreachable from an external plugin.
-This solution results in a substential overhead if called often.
+	\item \emph{External plugins} are compiled as a dynamic libraries distributed with the VM.
+They can be added or removed from a compiled VM: each dynamic library can be removed, or recompiled separately and modified.
+	\item \emph{Internal plugins} are compiled with the VM executable. They can be added or removed at VM compilation time.
 \end{itemize}
+
+There is a performance overhead for the plugins compared to normal VM code.
+For example, accesses to object headers are memory representation dependent and therefore need to be done indirectly through calls to main VM executable, since the plugin cannot know ahead of time which memory representation is used.
+Such calls cannot be inlined by the C compiler\footnote{In the case of specific internal plugins, C compiler linking time optimizations can be enabled to limit this overhead.}.
 
@@
 The interpreter uses a global look-up cache to speed-up virtual calls.
-On cache hits, it requests the JIT to compile that method and the native code version is used instead.
+On a cache hit, it requests the JIT to compile that method and the native code version is used instead.
 In most case, the JIT is used at the second activation of the method.
@@
 The VM code, including the interpreter code, is compiled by a C compiler and uses the C stack at runtime.
-During its execution, the interpreter modifies the execution stack of the client language, which is different from its own code stack as shown on Figure~\ref{fig:TwoStacks}.
+During its execution, the interpreter modifies the execution stack of the client language, which is different from its own C execution stack as shown on Figure~\ref{fig:TwoStacks}.
 This means for example that the frame pointer register in the processor refers to a stack frame on the C stack and that a processor push instruction pushes a value on the C stack.
@@
 Because of this design, switching the execution between the interpreter and the native code generated by the JIT has a cost.
-Effectively, the VM needs to switch the frame pointer, the stack pointer and depending on architecture the link register \sd{I do not get "the link register"} from one stack to the other one.
+Effectively, the VM needs to switch the frame pointer, the stack pointer and the link register (the latter only on some architectures) from one stack to the other one.
 It also needs to save those pointers to be able to resume execution from one stack to the other one.
 In addition, caller-saved registers on the C conventions or the JIT convention need to be saved since the VM cannot tell ahead what code is going to be used in the other execution model and what registers are required.
-We estimate the overhead to up to around a dozen of processor instruction, the exact number varies a lot so it is difficult to precise it (the number of register to edit depends on the processor and on the code executed).
+We estimate the overhead to up to around a dozen of processor instructions, the exact number of numbers varies a lot (the number of registers to edit depends on the processor used and on the code executed).
 
@@
 They benefit from the C compiler optimizations to be efficient.
-When called from the interpreter, activating a primitive is just a C call (the primitive function pointer is cached in the look-up cache next to the method to activate).
+When called from the interpreter, activating a primitive is just a C function call (the primitive function pointer is cached in the look-up cache next to the method to activate).
 When called from the native code generated by the JIT, activating a primitive requires to switch the execution from the client stack to the C stack, perform the primitive, and switch back to resume execution.
@@
 Primitives redefined in such way can be entirely or partially redefined.
-In the later case, only the common cases are re-implemented on top of the JIT back-end, and an unimplemented case happens at runtime, the generated native code calls the C primitive instead, adding overhead only in uncommon cases.
+In the later case, only the common cases are re-implemented on top of the JIT back-end.
+If an unimplemented case happens at runtime, the generated native code calls the C primitive instead, adding overhead only in uncommon cases.
 
@@
 
-The primitives can be associated either with a name, either with an index in the Cog VM. The named primitives can be added and modified without the need to recompile the VM.
+%The primitives can be associated either with a name, either with an index in the Cog VM. The named primitives can be added and modified without the need to recompile the VM.
 
@@
 
-\sk{So, do we agree that: 1- we bench all implem with new spec (i am rewriting the plugins right now) 2- we drop the sections "original spec", "spec issues", rename "final spec" and rephrase the implementations details on the sections below?}
- This section first describes the different ways of implementing a primitive. We then explain how each of these versions is executed by the VM and the pros and cons of each and every.
-
- \subsection{Different implementations and executions}
-
-We distinguish several implementations: the baseline in pure Smalltalk, the  \sk{delete "the first one"? sounds odd}
-
-\paragraph{Baseline: pure Smalltalk.}
+This section describes the different implementations of the string comparison optional primitive and then compare them.
+
+We then explain how each of these versions is executed by the VM and the pros and cons of each and every.
+
+ \subsection{Different implementations}
+
+We distinguish four implementations: the Baseline in pure Smalltalk, the Plugin version, the Slang version and the Slang+RTL version.
+
+\paragraph{\legBase.}
 To evaluate the primitive performance, we use as baseline an implementation in pure Smalltalk.
 There are multiple ways of writing the comparison of two strings in Smalltalk, following Smalltalk coding conventions, a standard developer would likely use high-level constructs which apply a closure on each element of the strings.
-We did not use high-level construct and chose to implement the Smalltalk version the most optimized way Smalltalk allows, \emph{i.e.,} we used the constructs to:do: and ifFalse:, which are both recognized by the Smalltalk to bytecode compiler and compiled respectively to loops and branches at the bytecode level.
+We did not use high-level construct and chose to implement the Smalltalk version the most optimized way Smalltalk allows, \emph{i.e.,} we used the constructs to:do: and ifFalse:, which are both recognized by the Smalltalk to bytecode compiler and compiled respectively to loops and branches at the bytecode level\footnote{The bytecode compiler moves the upper limit of the loop ahead of the loop, so (length1 min: length2) is not compute at each iteration.}.
 The implementation follows the new specification, hence two methods are available, baselineCompareWith: and baselineCompareWith:order:.
@@
 
-\paragraph{Existing: Smart Syntax.}
-
-The Smart Syntax Plugin is one of the version \sd{of what} existing until we started this work.
+%\paragraph{Existing: Smart Syntax.}
+%
+%The Smart Syntax Plugin is one of the version \sd{of what} existing until we started this work.
 The idea behind Smart Syntax is the following: the programmer writes code in Smalltalk using exclusively constructs that are recognized by the Smart Syntax compiler and that can be directly compiled to C (\emph{e.g.}, computing the size of a byte array, accessing its fields and doing integer arithmetic).
 With the addition of type annotations in the form of pragmas \cite{Duca16a} \footnote{Pragma are method annotation.}, the Smart Syntax plugin is able to generate C code from the Smalltalk code.
-\sd{A word on QuickTalk \cite{Ball86a}}
-With this approach, the programmer has to write the optional primitive code a single time in Smalltalk.
+%With this approach, the programmer has to write the optional primitive code a single time in Smalltalk.
 This code is used both as the fall-back code if the VM does not implement the primitive and at Slang to C compilation time to generate a C plugin from the code and the type annotations.
@@
 
-\paragraph{Existing: Plugin.}
-The Slang plugin is a recent complete rewrite of the Smart Syntax plugin.
-The main difference is that the Smalltalk version is used only as fall-back code, while at Slang-to-C compilation time the VM uses code written in Slang to generate the C code.
-One of the side-effect is that the plugin's code is now part of the VM code base and not the client code bases, which is convenient for maintenance since the Cog VM has multiple client languages.
-However, the implementor now needs to maintain and evolve two version of the primitive, the fall-back code in Smalltalk and the code in Slang.
-\sd{do we need the next sentence?}This primitive follows the old specification, the full code is available in Appendix~\ref{PluginAppendix}.
-
-\paragraph{Pure Slang.}
-
+\paragraph{\legSlangPlug.}
+The VM plugin implementation can be easily added and removed as explained in Section \ref{sec:VMCompil}.
+Written in Slang, it can be compiled to C independently to the VM and is then compiled to native code through the C compiler as internal or external plugin.
+Some accesses to the string objects (for example, checking if they are byte strings) are done indirectly through the main VM API and these calls generate some overhead.
+%\sd{do we need the next sentence?}This primitive follows the old specification, the full code is available in Appendix~\ref{PluginAppendix}.
+
+\paragraph{\legSlang.}
 The Slang implementation is written inside the interpreter and has access to the full interpreter APIs.
@@
 The primitive is now part of the core of the VM, it cannot be easily added or removed as plugin primitives, but it is simpler to read and write.
-This primitive follows the new specification, the full code is available in Appendix \ref{SlangAppendix}.
-
-\paragraph{Slang+JIT.} \sk{Id like to keep slang+JIT and not slang+RLT as JIT may ring a bell for the non-smalltalkers (while RLT may not).
-but as we are already using a lot of ST words (primitives, Slang...), does that actually make any sense ?}
-
+%This primitive follows the new specification, the full code is available in Appendix \ref{SlangAppendix}.
+
+\paragraph{\legRTL.} 
 In addition to the previous implementation, we've re-implemented here the primitive on top of Cog's JIT back-end.
@@
 For the three parameter primitive, the VM falls back to the Slang version.
-This primitive follows the new specification, the full code is available in Appendix \ref{RTLAppendix}.
+%This primitive follows the new specification, the full code is available in Appendix \ref{RTLAppendix}.
 
@@
   \hline
-  Baseline    & No & All  & 10  & 0 & No & - & - \\
+  \legBase    & No & All  & 10  & 0 & No & - & - \\
    \hline
-  SmartSyntax & Plugin & VM & 20  & 3+ & Yes  & Limited & Low \\
+  \legSlangPlug & Plugin & VM & 35  & 1  & Yes & Limited & Low \\
    \hline
-  SlangPlugin & Plugin & VM & 35  & 1  & Yes & Limited & Low \\
+  \legSlang   & Yes  & VM & 45  & 1  & Yes & High & Low \\
    \hline
-  PureSlang   & Yes  & VM & 45  & 1  & Yes & High & Low \\
-   \hline
-  Slang+JIT   & Yes  & VM-JIT & 85  & 2  & No & High & High \\
+  \legRTL   & Yes  & VM-JIT & 85  & 2  & No & High & High \\
   \hline
@@
 
- In this section we compare the five different implementations according to the following criteria:
+ In this section we compare the four different implementations according to the following criteria:
  \begin{itemize}
  	\item \emph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (Plugin) or requires to recompile the full VM (Yes).
-\sd{why not ST, Plugin, VM?} \sk{Yes and No answers to the "recompilation needed" question, so maybe we should keep it that way}
-	\item \emph{Implementation language familiar to developers:} Smalltalk application developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code.
+%\sd{why not ST, Plugin, VM?} %\sk{Yes and No answers to the "recompilation needed" question, so maybe we should keep it that way}
+	\item \emph{Language familiar to developers:} Smalltalk developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code.
 Only Cog's JIT implementors are familiar with its RTL, which is a subset of the VM developers.
 We use All if all communities are familiar with the language, VM if all VM developers are and VM-JIT if only a subset of VM developers are.
-	\item \emph{Readability:} To evaluate the approach readability, we use the number of lines of code even if this is arguable but used regularly in other papers \cite{xxx}.
-A better metric would be to use the cyclomatic complexity, however since the  Cog' s RTL generates branches at runtime, it is difficult to use it.
+	\item \emph{Number of lines of code:} We measured the number of lines of code of each implementation, which arguably can be used to evaluate the complexity level of each implementation.
 	\item \emph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to be executed if the primitive fails.
 This code is maintained by the client language implementors, and not by the VM implementors.
-Hence we do not count the fall-back code in the number of implementation to maintain, nor the pure Smalltalk implementation\sd{why?}.
-However, as a VM implementor, we need to maintain what generates the C code of the VM, hence, in the case of Smart Syntax, we need to maintain a different method in each client language.
-We also count all the Slang implementations as well as the ones on Cog's RTL.
+Hence we do not count the fall-back code in the number of implementation to maintain, nor the pure Smalltalk implementation.
+However, as a VM implementor, we need to maintain what generates the C code of the VM.
+We count all the Slang implementations as well as the ones on Cog's RTL.
 	\item \emph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
@@
 The generated code can therefore differ according the compilers and the optimizations they performed.
-On the contrary, the JIT will always generate the same native code.
+On the contrary, the JIT generates the same native code whichever C compiler is used.
 \end{itemize}
@@
 Also, a VM recompilation affects the overall complexity of the development process, but the resulting code might be faster to run.
-Likewise, the use of plugins brings modularity, but as you will see in Section~\ref{sec:eval}, the performance, although better than baseline version, could still be improved.
+Likewise, the use of plugins brings modularity at the cost of some performance loss.
 
@@
 
+\todo{Make a  pass cba}
+
 The SlangPlugin, PureSlang and Slang+JIT implementations of this string comparison primitive have been recently implemented in the Cog VM.
 In this section, we compare execution speed of these primitives against the other existing implementations.
-We measure their respective execution times for different string sizes.
-\sd{mention microbenchmarks here.}\sd{why there is only 4.3 and not something else?
+We measure their respective execution times for different string sizes.% \sd{mention microbenchmarks here.}\sd{why there is only 4.3 and not something else?
 Because this is strange from a shape structure.}
@@
 \subsection{Set-up}
+
+\cba{All benchs on same VM specifically compiled (normal VM has only one impl)}
 
@@
 Each benchmark was measured 30 times.
-
-%%%
-% Legend definition (so I can change it once for all)
-%%%
-
-\def\legBase{Baseline}
-\def\legSmartSyntaxPlug{Syntax}
-\def\legSlangPlug{Plugin}
-\def\legSlang{Pure slang}
-\def\legRTL{Slang+JIT}
-\def\plotHeight{6.7cm}
-
-%%%
-% End Legend
-%%%
 
@@
    \hline
-  \multirow{2}{*}{SmartSyntax} &GCC & 0.3411 +- 0.0007 & 0.6014 +- 0.0023 & 1.3891 +- 0.0023 &  8.6796 +- 0.0587 & 13.6051 +- 0.0589\\
- 							   &LLVM & 0.3147 +- 0.0023 & 0.5606 +- 0.0022 & 1.2505 +- 0.0046 & 10.3029 +- 0.0624 & 9.4065 +- 0.0367\\
-   \hline
   \multirow{2}{*}{SlangPlugin} &GCC& 0.3375 +- 0.0016 & 0.5654 +- 0.0078 & 1.3150 +- 0.0205 & 8.9585 +- 0.1170  & 14.7789 +- 0.1471 \\
@@
 The JIT overhead is negligible here for 1 second because we could not see any time spent in the JIT in the profiler + ref to your paper, i.e.
-less than 0.5ms in the JIT} \cite{CogProfiler} and allows the benchmarks to run at least 1 second in the slower implementations to avoid processor / OS noises.
+less than 0.5ms in the JIT} \cite{CogProfiler} and allows the benchmarks to run at least 1 second in the slower implementations to avoid processor and OS noises.
 The loop overhead is measured and removed from the measurements.
@@
 
-\subsection{Micro-benchmarks} \sk{Results?}
+\subsection{Results} %\sk{Results?}
 
@@
 \begin{itemize}
-\item The execution time of the SmartSyntax and Plugin implementations is either very close or greater than \sd{within which ranges? 10\% 200\%} the baseline's for the small strings.
+\item The execution time of the Plugin implementations is either very close or greater than \sd{within which ranges? 10\% 200\%} the baseline's for the small strings.
 \item The Slang+JIT implementation is more efficient than the baseline in every case.
@@
 As mentioned in Section~\ref{sec:VMexec}, all implementations written in Slang come with an overhead due to the switch between the C and Smalltalk stacks.
-This overhead is clearly visible in the bench results in Figure~\ref{fig:bench} for small strings (size 1 or less): the SmartSyntax and Plugin implementations are around 30 to 60\% slower than the baseline's, and the PureSlang performance is close to the baseline's.
+This overhead is clearly visible in the bench results in Figure~\ref{fig:bench} for small strings (size 1 or less): the Plugin implementation is around 30 to 60\% slower than the baseline's, and the PureSlang performance is close to the baseline's.
 As the string size goes larger, the overhead is less significant because it is absorbed by other performance gains.
@@
 
-\todo{add conclusion}
+\todo{add conclusion SK}
 
@@
 
-\cba{Maybe talk in those cases the number of representation of the primitives, this is a bit disconnected with the rest of the paper I feel}
-
-We identified two directly related works.
-In some languages (Strongtalk, Self, RSqueak), the implementors tried to decrease the overall number of optional primitives by improving the client language performance, reducing the need for such primitives \cite{Strongtalk, ursPHD, NoPrimTracing}.
-In addition, in the V8 project~\cite{V8}, the interpreter was entirely built on top of the V8 JIT's back-end using a specific DSL similar to Cog's RTL.
+In some programming languages (Strongtalk, Self, RSqueak), the implementors tried to decrease the overall number of optional primitives by improving the client language performance, reducing the need for such primitives \cite{Strongtalk, ursPHD, NoPrimTracing}.
+In other work \cite{Ball86a, Char13a}, optional primitives can be written directly in a subset of the client language and compiled at runtime to native code, hence they can be modified in the client language without recompiling the VM.
+Lastly, several VMs implemented their interpreter on top of an abstract assembly~\cite{V8,LLIntJSC}, allowing the interpreter to use the same register conventions as the code generated by the JIT and avoiding the stack switch cost.
 
@@
 Those performance drops were due to the optimizing JIT taking different optimization decisions.
-This problem was so significant that recently the V8 team \cite{V8} rewrote their primitives written in Javascript (named built-ins in their project) in a DSL compiled ahead of time through their JIT back-end to improve the performance of the primitives and therefore decrease the performance drop between baseline and peak performance.
+%This problem was so significant that recently the V8 team \cite{V8} rewrote their primitives written in Javascript (named built-ins in their project) in a DSL compiled ahead of time through their JIT back-end to improve the performance of the primitives and therefore decrease the performance drop between baseline and peak performance.
 Aside from the start-up performance and the performance unreliability problems described, these approaches also require a mature optimizing JIT, hard to implement and maintain.
 
+\subsection{Primitives written in the client language}
+
+In QuickTalk \cite{Ball86a} and Waterfall \cite{Char13a}, the implementors are able to write the primitives in a subset of the client language.
+The primitives are not part of the VM anymore, but are instead generated as part of the compiled code bytecode representation with a special encoding.
+They are executed by generating at runtime native code for the special primitive encoding.
+They can be modified without recompiling the VM and can be performed by the client language implementors instead of the VM implementors, lowering the maintenance cost for the VM implementor.
+
+However, this design has some issues that can lead to performance drops.
+Many VMs support multiple processor back-ends and multiple memory representation of objects.
+Primitives written in such DSL are required to be independent from the processor used and the memory representation of objects.
+If they were written in the VM, a different version could exist for each different processor and memory representation and a different version would be picked at compile-time.
+Having different versions allows to refine carefully the instructions generated for performance critical primitives (Array accesses for example) and can lead to noticeable speed-ups.
+
 \subsection{No stack switch overhead}
 
-To avoid paying the stack switch cost between the client Stack and the C stack, the V8 team recently wrote an interpreter named \emph{Ignition} in a DSL.
+To avoid paying the stack switch cost between the client Stack and the C stack, some VMs include an interpreter written on top of an abstract assembly.
+It allows the interpreter to have same register convention than the native code generated by the JIT, removing the stack switch overhead.
+For example, the V8 team recently wrote an interpreter named \emph{Ignition}.
 This interpreter is compiled ahead of time through their JIT back-end to native code.
-With their primitives and their interpreter written in the DSL, the V8 runtime rarely needs to switch between the client and the C stack (no need to switch to interpret code, no need to switch for primitives).
-In this context, the stack switch overhead is not a problem anymore.
+With their primitives and their interpreter written in such way, the V8 runtime rarely needs to switch between the client and the C stack (no need to switch to interpret code, no need to switch for primitives).
+In Javascript Core, a similar approach is taken with the interpreter named \emph{LLInt}~\cite{LLIntJSC}.
+LLInt is written in an abstract assembly code compiled ahead of time to native code compatible with the JIT's calling conventions.
 
@@
 Slang has some abstractions over native code, similarly to C.
-Compiling Slang ahead of time to native code efficiently requires us to re-implement optimizations implemented in C compilers and we don't want that complexity or to lower the abstractions level of Slang to express native code optimizations, which would increase the complexity of Slang, which we don't want either.
+Compiling Slang ahead of time to native code efficiently requires us to re-implement optimizations implemented in C compilers and we don't want that complexity.
+Alternatively, we could try to lower the abstractions level of Slang to express native code optimizations, but that would increase the complexity of Slang, which we don't want either.
 Rewriting the interpreter on top of Cog JIT's RTL is possible but requires a significant amount of work.
+It would be nice to do so as a future work.
 
@@
 
-All the primitives we compared are written in a processor independent way, due to the C compiler abstraction for the Slang versions, the abstract assembly abstraction for Cog's RTL version or the VM abstraction for the baseline implementation in Smalltalk.
-We could rewrite the primitive differently for each back-end currently supported (x86, x64, MIPSEL, ARMv6) and yet to support (ARMv8, ...).
+All the primitive implementations we compared are written in a processor independent way, due to the C compiler abstraction for the Slang versions, the abstract assembly abstraction for Cog's RTL version or the VM abstraction for the baseline implementation in Smalltalk.
+We could rewrite the primitive differently for each back-end we currently support in production (x86, x64, MIPSEL, ARMv6) and for the back-end yet to support (ARMv8, ...).
 It could be relevant in some cases performance-wise, for example, on Intel processors, the JVM implements string comparison using the SSE4.2 string comparison instructions when available~\cite{JVMSSE4StringComp}.
 We did not go in this direction because the maintenance cost is too high since we would have to maintain a different implementation per back-end.
-We already have four back-ends (x86, x64, MIPSEL, ARMv6) in production.
 
@@
 
-In this paper we compare five different implementations of the string primitives on top of the Cog VM.
+In this paper we compared four different implementations of the string primitives on top of the Cog VM.
+We showed the performance difference between each implementation for different string sizes.
 Choosing the right implementation for one's VM is a trade-off between maintenance cost and execution time, the quickest to execute the primitive is, the harder it is to maintain.
@@
 %% Appendix
-\null \vfill\null %switch column
-
-\appendix
-\section{Smart Syntax code}
-\label{SmartSyntaxAppendix}
-
-Source code of the smart syntax method, which is used to generate the C code for the plugin but also as Smalltalk fall-back code.
-
-\begin{code}
-compare: string1 with: string2 collated: order
-	"Return 1, 2 or 3, if string1 is <, =, or > string2, with the collating order of characters given by the order array."
-
-	| len1 len2 c1 c2 |
-	<primitive: 'primitiveCompareString' module: 'MiscPrimitivePlugin'>
-	<var: #string1 declareC: 'unsigned char *string1'>
-	<var: #string2 declareC: 'unsigned char *string2'>
-	<var: #order declareC: 'unsigned char *order'>
-
-	len1 := string1 size.
-	len2 := string2 size.
-	1 to: (len1 min: len2) do:
-		[:i |
-		c1 := order at: (string1 basicAt: i) + 1.
-		c2 := order at: (string2 basicAt: i) + 1.
-		c1 = c2 ifFalse: 
-			[c1 < c2 ifTrue: [^ 1] ifFalse: [^ 3]]].
-	len1 = len2 ifTrue: [^ 2].
-	len1 < len2 ifTrue: [^ 1] ifFalse: [^ 3].
-\end{code}
-
-\vfill\null %switch column
-\section{Slang Plugin code}
-\label{PluginAppendix}
-
-Source code of the plugin method generating the C code for the Slang plugin code. Most operations needs to go through the interpreter proxy.
-
-\begin{code}
-primitiveCompareString
-	"ByteString (class) compare: string1 with: string2 collated: order"
-	<export: true>
-	| len1 len2 order string1 string2 |
-
-	<var: 'order' type: #'unsigned char *'>
-	<var: 'string1' type: #'unsigned char *'>
-	<var: 'string2' type: #'unsigned char *'>
-	((interpreterProxy isBytes: (interpreterProxy stackValue: 0))
-	and: [(interpreterProxy isBytes: (interpreterProxy stackValue: 1))
-	and: [interpreterProxy isBytes: (interpreterProxy stackValue: 2)]]) ifFalse:
-		[^interpreterProxy primitiveFailFor: PrimErrBadArgument].
-	string1 := interpreterProxy firstIndexableField: (interpreterProxy stackValue: 2).
-	string2 := interpreterProxy firstIndexableField: (interpreterProxy stackValue: 1).
-	order := interpreterProxy firstIndexableField: (interpreterProxy stackValue: 0).
-	len1 := interpreterProxy sizeOfSTArrayFromCPrimitive: string1.
-	len2 := interpreterProxy sizeOfSTArrayFromCPrimitive: string2.
-	(interpreterProxy sizeOfSTArrayFromCPrimitive: order) < 256 ifTrue:
-		[^interpreterProxy primitiveFailFor: PrimErrBadArgument].
-	0 to: (len1 min: len2) - 1 do: 
-		[ :i | | c1 c2 |
-		c1 := order at: (string1 at: i).
-		c2 := order at: (string2 at: i).
-		c1 = c2 ifFalse:
-			[^interpreterProxy methodReturnInteger: (c1 < c2 ifTrue: [1] ifFalse: [3])]].
-	interpreterProxy methodReturnInteger:
-		(len1 = len2 ifTrue: [2] ifFalse: [len1 < len2 ifTrue: [1] ifFalse: [3]])
-\end{code}
-
-\vfill\null %switch column
-\section{Slang code}
-\label{SlangAppendix}
-
-Source code of the slang method generating the C code for the Slang string comparison primitive.
+%\null \vfill\null %switch column
+%
+%\appendix
+%\section{Smart Syntax code}
+%\label{SmartSyntaxAppendix}
+%
+%Source code of the smart syntax method, which is used to generate the C code for the plugin but also as Smalltalk fall-back code.
+%
+%\begin{code}
+%compare: string1 with: string2 collated: order
+%	"Return 1, 2 or 3, if string1 is <, =, or > string2, with the collating order of characters given by the order array."
+%
+%	| len1 len2 c1 c2 |
+%	<primitive: 'primitiveCompareString' module: 'MiscPrimitivePlugin'>
+%	<var: #string1 declareC: 'unsigned char *string1'>
+%	<var: #string2 declareC: 'unsigned char *string2'>
+%	<var: #order declareC: 'unsigned char *order'>
+%
+%	len1 := string1 size.
+%	len2 := string2 size.
+%	1 to: (len1 min: len2) do:
+%		[:i |
+%		c1 := order at: (string1 basicAt: i) + 1.
+%		c2 := order at: (string2 basicAt: i) + 1.
+%		c1 = c2 ifFalse: 
+%			[c1 < c2 ifTrue: [^ 1] ifFalse: [^ 3]]].
+%	len1 = len2 ifTrue: [^ 2].
+%	len1 < len2 ifTrue: [^ 1] ifFalse: [^ 3].
+%\end{code}
+%
+%\vfill\null %switch column
+%\section{Slang Plugin code}
+%\label{PluginAppendix}
+%
+%Source code of the plugin method generating the C code for the Slang plugin code. Most operations needs to go through the interpreter proxy. 
+%
+%\begin{code}
+%primitiveCompareString
+%	"ByteString (class) compare: string1 with: string2 collated: order"
+%	<export: true>
+%	| len1 len2 order string1 string2 |
+%
+%	<var: 'order' type: #'unsigned char *'>
+%	<var: 'string1' type: #'unsigned char *'>
+%	<var: 'string2' type: #'unsigned char *'>
+%	((interpreterProxy isBytes: (interpreterProxy stackValue: 0))
+%	and: [(interpreterProxy isBytes: (interpreterProxy stackValue: 1))
+%	and: [interpreterProxy isBytes: (interpreterProxy stackValue: 2)]]) ifFalse:
+%		[^interpreterProxy primitiveFailFor: PrimErrBadArgument].
+%	string1 := interpreterProxy firstIndexableField: (interpreterProxy stackValue: 2).
+%	string2 := interpreterProxy firstIndexableField: (interpreterProxy stackValue: 1).
+%	order := interpreterProxy firstIndexableField: (interpreterProxy stackValue: 0).
+%	len1 := interpreterProxy sizeOfSTArrayFromCPrimitive: string1.
+%	len2 := interpreterProxy sizeOfSTArrayFromCPrimitive: string2.
+%	(interpreterProxy sizeOfSTArrayFromCPrimitive: order) < 256 ifTrue:
+%		[^interpreterProxy primitiveFailFor: PrimErrBadArgument].
+%	0 to: (len1 min: len2) - 1 do: 
+%		[ :i | | c1 c2 |
+%		c1 := order at: (string1 at: i).
+%		c2 := order at: (string2 at: i).
+%		c1 = c2 ifFalse:
+%			[^interpreterProxy methodReturnInteger: (c1 < c2 ifTrue: [1] ifFalse: [3])]].
+%	interpreterProxy methodReturnInteger:
+%		(len1 = len2 ifTrue: [2] ifFalse: [len1 < len2 ifTrue: [1] ifFalse: [3]])
+%\end{code}
+%
+%\vfill\null %switch column
+%\section{Slang code}
+%\label{SlangAppendix}
+%
+%Source code of the slang method generating the C code for the Slang string comparison primitive.
 We split it in two methods, and forced inlining at Slang-to-C compilation time to remove the closures while avoiding code duplication.
-
-\begin{code}
-primitiveCompareWith
-	"<string1> primitiveCompareWith: string2 [collated: order] "
-	<export: true>	
-	| string1 string2 order strLength1 strLength2 result |
-	"1 - fetch the parameters from the stack"	
-	argumentCount = 1 ifFalse:
-		[argumentCount ~= 2 ifTrue:
-			[^self primitiveFailFor: PrimErrBadNumArgs].
-			 order := self stackTop.
-			 ((objectMemory isBytes: order)
-			 and: [(objectMemory numBytesOfBytes: order) = 256]) ifFalse:
-				[^self primitiveFailFor: PrimErrBadArgument]].
-	string1 := self stackValue: argumentCount.
-	string2 := self stackValue: argumentCount - 1. 		
-	"2 - check their types - all parameters are ByteObject"
-	((objectMemory isBytes: string1)
-	 and: [objectMemory isBytes: string2]) ifFalse: 
-		[^self primitiveFailFor: PrimErrBadArgument].
-	"3 - compare the strings"	
-	strLength1 := objectMemory numBytesOfBytes: string1.
-	strLength2 := objectMemory numBytesOfBytes: string2.
-	result := order 
-		ifNil: [self rawCompare: string1 length: strLength1 with: string2 length: strLength2 accessBlock:
-				[:str :index | objectMemory fetchByte: index ofObject: str ]]
-		ifNotNil: 
-			[self rawCompare: string1 length: strLength1 with: string2 length: strLength2 accessBlock:
-				[:str :index | objectMemory fetchByte: (objectMemory fetchByte: index ofObject: str) + 1 ofObject: order ]].
-
-	self methodReturnInteger: result
-
-rawCompare: string1 length: strLength1 with: string2 length: strLength2 accessBlock: accessBlock
-	| c1 c2 min |
-	<inline: true> "needs to be forced else slang does not inline it by default"
-	min := strLength1 min: strLength2.
-	0 to: min-1 do: 
-		[:i | c1 := accessBlock value: string1 value: i.
-			c2 := accessBlock value: string2 value: i.
-			c1 = c2 ifFalse: [^c1 - c2]].
-	^strLength1 - strLength2
-\end{code}
-
-\vfill\null %switch column
-\section{Cog's RTL code}
-\label{RTLAppendix}
-
-Source code of the method generating the native code of the string comparison primitive on top of Cog's RTL.
+%
+%\begin{code}
+%primitiveCompareWith
+%	"<string1> primitiveCompareWith: string2 [collated: order] "
+%	<export: true>	
+%	| string1 string2 order strLength1 strLength2 result |
+%	"1 - fetch the parameters from the stack"	
+%	argumentCount = 1 ifFalse:
+%		[argumentCount ~= 2 ifTrue:
+%			[^self primitiveFailFor: PrimErrBadNumArgs].
+%			 order := self stackTop.
+%			 ((objectMemory isBytes: order)
+%			 and: [(objectMemory numBytesOfBytes: order) = 256]) ifFalse:
+%				[^self primitiveFailFor: PrimErrBadArgument]].
+%	string1 := self stackValue: argumentCount.
+%	string2 := self stackValue: argumentCount - 1. 		
+%	"2 - check their types - all parameters are ByteObject"
+%	((objectMemory isBytes: string1)
+%	 and: [objectMemory isBytes: string2]) ifFalse: 
+%		[^self primitiveFailFor: PrimErrBadArgument].
+%	"3 - compare the strings"	
+%	strLength1 := objectMemory numBytesOfBytes: string1.
+%	strLength2 := objectMemory numBytesOfBytes: string2.
+%	result := order 
+%		ifNil: [self rawCompare: string1 length: strLength1 with: string2 length: strLength2 accessBlock:
+%				[:str :index | objectMemory fetchByte: index ofObject: str ]]
+%		ifNotNil: 
+%			[self rawCompare: string1 length: strLength1 with: string2 length: strLength2 accessBlock:
+%				[:str :index | objectMemory fetchByte: (objectMemory fetchByte: index ofObject: str) + 1 ofObject: order ]].
+%
+%	self methodReturnInteger: result
+%
+%
+%rawCompare: string1 length: strLength1 with: string2 length: strLength2 accessBlock: accessBlock
+%	| c1 c2 min |
+%	<inline: true> "needs to be forced else slang does not inline it by default"
+%	min := strLength1 min: strLength2.
+%	0 to: min-1 do: 
+%		[:i | c1 := accessBlock value: string1 value: i.
+%			c2 := accessBlock value: string2 value: i.
+%			c1 = c2 ifFalse: [^c1 - c2]].
+%	^strLength1 - strLength2
+%\end{code}
+%
+%\vfill\null %switch column
+%\section{Cog's RTL code}
+%\label{RTLAppendix}
+%
+%Source code of the method generating the native code of the string comparison primitive on top of Cog's RTL.
 Method names starting with uppercase (in blue) are directly generating an abstract assembly instruction.
@@
 We duplicated the code checking if each operand is a ByteString and extract its size to generate better native code than doing the two operations separatedly.
-
-\begin{code}
-genPrimitiveStringCompareWith
-	| instr jump jumpAbove jumpIncorrectFormat1 jumpIncorrectFormat2 jumpIncorrectFormat3 jumpIncorrectFormat4 jumpMidFailure jumpSuccess minSizeReg string1CharOrByteSizeReg string2CharOrByteSizeReg string1Reg string2Reg |
-	<var: #jumpIncorrectFormat1 type: #'AbstractInstruction *'>
-	<var: #jumpIncorrectFormat2 type: #'AbstractInstruction *'>
-	<var: #jumpIncorrectFormat3 type: #'AbstractInstruction *'>
-	<var: #jumpIncorrectFormat4 type: #'AbstractInstruction *'>
-	<var: #jumpAbove type: #'AbstractInstruction *'>
-	<var: #jumpSuccess type: #'AbstractInstruction *'>
-	<var: #jump type: #'AbstractInstruction *'>
-	<var: #jumpMidFailure type: #'AbstractInstruction *'>
-	"I redefine those name to ease program comprehension"
-	string1Reg := ReceiverResultReg.
-	string2Reg := Arg0Reg.
-	string1CharOrByteSizeReg := Arg1Reg.
-	string2CharOrByteSizeReg := ClassReg.
-	minSizeReg := SendNumArgsReg.
-	"Load arguments in reg"
-	cogit genLoadArgAtDepth: 0 into: string2Reg.
-	"checks if string1 is a byteobject and get its size in bytes"
-	self genGetFormatOf: string1Reg into: TempReg.
-	cogit CmpCq: objectMemory firstByteFormat R: TempReg.
-	jumpIncorrectFormat1 := cogit JumpLess: 0.
-	cogit CmpCq: objectMemory firstCompiledMethodFormat R: TempReg.
-	jumpIncorrectFormat2 := cogit JumpAboveOrEqual: 0.	
-	self genGetNumSlotsOf: string1Reg into: string1CharOrByteSizeReg.
-	(cogit LogicalShiftLeftCq: objectMemory shiftForWord R: string1CharOrByteSizeReg). 
-	cogit AndCq: objectMemory wordSize - 1 R: TempReg R: TempReg. 
-	cogit SubR: TempReg R: string1CharOrByteSizeReg. 
-	"checks if string2 is a byteobject and get its size in bytes"
-	self genGetFormatOf: string2Reg into: TempReg.
-	cogit CmpCq: objectMemory firstByteFormat R: TempReg.
-	jumpIncorrectFormat3 := cogit JumpLess: 0.
-	cogit CmpCq: objectMemory firstCompiledMethodFormat R: TempReg.
-	jumpIncorrectFormat4 := cogit JumpAboveOrEqual: 0.	
-	self genGetNumSlotsOf: string2Reg into: string2CharOrByteSizeReg.
-	(cogit LogicalShiftLeftCq: objectMemory shiftForWord R: string2CharOrByteSizeReg).
-	cogit AndCq: objectMemory wordSize - 1 R: TempReg R: TempReg.
-	cogit SubR: TempReg R: string2CharOrByteSizeReg.
-	\end{code}
-	\vfill\null %switch column
-	\begin{code}
-	"Type and number of arguments are correct"
-	"Compute the min"	 
-	cogit CmpR: string1CharOrByteSizeReg R: string2CharOrByteSizeReg.
-	jumpAbove := cogit JumpBelow: 0. 
-	cogit MoveR: string1CharOrByteSizeReg R: minSizeReg. 
-	jump := cogit Jump: 0. 
-	jumpAbove jmpTarget: (cogit MoveR: string2CharOrByteSizeReg R: minSizeReg). 
-	jump jmpTarget: (cogit CmpCq: 0 R: minSizeReg). 
-	jumpSuccess := cogit JumpZero: 0. "if one of the string is empty, no need to go through the comparing loop"
-	"Compare the bytes"
-	cogit MoveCq: objectMemory baseHeaderSize  R: TempReg.
-	cogit AddCq: objectMemory baseHeaderSize R: minSizeReg.
-	instr := cogit MoveXbr: TempReg R: string1Reg R: string1CharOrByteSizeReg.
-	cogit MoveXbr: TempReg R: string2Reg R: string2CharOrByteSizeReg.
-	cogit SubR: string2CharOrByteSizeReg R: string1CharOrByteSizeReg. 
-	jumpMidFailure := cogit JumpNonZero: 0. "the 2 compared characters are different, exit the loop"
-	cogit AddCq: 1 R: TempReg.
-	cogit CmpR: TempReg R: minSizeReg. 
-	cogit JumpNonZero: instr.
-	"all bytes from 1 to minSize are equal"
-	self genGetNumBytesOf: string1Reg into: string1CharOrByteSizeReg.
-	self genGetNumBytesOf: string2Reg into: string2CharOrByteSizeReg.
-	jumpSuccess jmpTarget: (cogit SubR: string2CharOrByteSizeReg R: string1CharOrByteSizeReg).
-	jumpMidFailure  jmpTarget: (cogit MoveR: string1CharOrByteSizeReg R: ReceiverResultReg).	
-	self genConvertIntegerToSmallIntegerInReg: ReceiverResultReg.
-	cogit genPrimReturn.
-	jumpIncorrectFormat4 
-		jmpTarget: (jumpIncorrectFormat3 
-			jmpTarget: (jumpIncorrectFormat2 
-				jmpTarget: (jumpIncorrectFormat1 jmpTarget: cogit Label))).
-	^ CompletePrimitive}
-\end{code}
+%
+%\begin{code}
+%genPrimitiveStringCompareWith
+%	| instr jump jumpAbove jumpIncorrectFormat1 jumpIncorrectFormat2 jumpIncorrectFormat3 jumpIncorrectFormat4 jumpMidFailure jumpSuccess minSizeReg string1CharOrByteSizeReg string2CharOrByteSizeReg string1Reg string2Reg |
+%	<var: #jumpIncorrectFormat1 type: #'AbstractInstruction *'>
+%	<var: #jumpIncorrectFormat2 type: #'AbstractInstruction *'>
+%	<var: #jumpIncorrectFormat3 type: #'AbstractInstruction *'>
+%	<var: #jumpIncorrectFormat4 type: #'AbstractInstruction *'>
+%	<var: #jumpAbove type: #'AbstractInstruction *'>
+%	<var: #jumpSuccess type: #'AbstractInstruction *'>
+%	<var: #jump type: #'AbstractInstruction *'>
+%	<var: #jumpMidFailure type: #'AbstractInstruction *'>
+%	"I redefine those name to ease program comprehension"
+%	string1Reg := ReceiverResultReg.
+%	string2Reg := Arg0Reg.
+%	string1CharOrByteSizeReg := Arg1Reg.
+%	string2CharOrByteSizeReg := ClassReg.
+%	minSizeReg := SendNumArgsReg.
+%	"Load arguments in reg"
+%	cogit genLoadArgAtDepth: 0 into: string2Reg.
+%	"checks if string1 is a byteobject and get its size in bytes"
+%	self genGetFormatOf: string1Reg into: TempReg.
+%	cogit CmpCq: objectMemory firstByteFormat R: TempReg.
+%	jumpIncorrectFormat1 := cogit JumpLess: 0.
+%	cogit CmpCq: objectMemory firstCompiledMethodFormat R: TempReg.
+%	jumpIncorrectFormat2 := cogit JumpAboveOrEqual: 0.	
+%	self genGetNumSlotsOf: string1Reg into: string1CharOrByteSizeReg.
+%	(cogit LogicalShiftLeftCq: objectMemory shiftForWord R: string1CharOrByteSizeReg). 
+%	cogit AndCq: objectMemory wordSize - 1 R: TempReg R: TempReg. 
+%	cogit SubR: TempReg R: string1CharOrByteSizeReg. 
+%	"checks if string2 is a byteobject and get its size in bytes"
+%	self genGetFormatOf: string2Reg into: TempReg.
+%	cogit CmpCq: objectMemory firstByteFormat R: TempReg.
+%	jumpIncorrectFormat3 := cogit JumpLess: 0.
+%	cogit CmpCq: objectMemory firstCompiledMethodFormat R: TempReg.
+%	jumpIncorrectFormat4 := cogit JumpAboveOrEqual: 0.	
+%	self genGetNumSlotsOf: string2Reg into: string2CharOrByteSizeReg.
+%	(cogit LogicalShiftLeftCq: objectMemory shiftForWord R: string2CharOrByteSizeReg).
+%	cogit AndCq: objectMemory wordSize - 1 R: TempReg R: TempReg.
+%	cogit SubR: TempReg R: string2CharOrByteSizeReg.
+%	\end{code}
+%	\vfill\null %switch column
+%	\begin{code}
+%	"Type and number of arguments are correct"
+%	"Compute the min"	 
+%	cogit CmpR: string1CharOrByteSizeReg R: string2CharOrByteSizeReg.
+%	jumpAbove := cogit JumpBelow: 0. 
+%	cogit MoveR: string1CharOrByteSizeReg R: minSizeReg. 
+%	jump := cogit Jump: 0. 
+%	jumpAbove jmpTarget: (cogit MoveR: string2CharOrByteSizeReg R: minSizeReg). 
+%	jump jmpTarget: (cogit CmpCq: 0 R: minSizeReg). 
+%	jumpSuccess := cogit JumpZero: 0. "if one of the string is empty, no need to go through the comparing loop"
+%	"Compare the bytes"
+%	cogit MoveCq: objectMemory baseHeaderSize  R: TempReg.
+%	cogit AddCq: objectMemory baseHeaderSize R: minSizeReg.
+%	instr := cogit MoveXbr: TempReg R: string1Reg R: string1CharOrByteSizeReg.
+%	cogit MoveXbr: TempReg R: string2Reg R: string2CharOrByteSizeReg.
+%	cogit SubR: string2CharOrByteSizeReg R: string1CharOrByteSizeReg. 
+%	jumpMidFailure := cogit JumpNonZero: 0. "the 2 compared characters are different, exit the loop"
+%	cogit AddCq: 1 R: TempReg.
+%	cogit CmpR: TempReg R: minSizeReg. 
+%	cogit JumpNonZero: instr.
+%	"all bytes from 1 to minSize are equal"
+%	self genGetNumBytesOf: string1Reg into: string1CharOrByteSizeReg.
+%	self genGetNumBytesOf: string2Reg into: string2CharOrByteSizeReg.
+%	jumpSuccess jmpTarget: (cogit SubR: string2CharOrByteSizeReg R: string1CharOrByteSizeReg).
+%	jumpMidFailure  jmpTarget: (cogit MoveR: string1CharOrByteSizeReg R: ReceiverResultReg).	
+%	self genConvertIntegerToSmallIntegerInReg: ReceiverResultReg.
+%	cogit genPrimReturn.
+%	jumpIncorrectFormat4 
+%		jmpTarget: (jumpIncorrectFormat3 
+%			jmpTarget: (jumpIncorrectFormat2 
+%				jmpTarget: (jumpIncorrectFormat1 jmpTarget: cogit Label))).
+%	^ CompletePrimitive}
+%\end{code}
 

### 2018-05-17 paper.tex
pass on evaluation

@@
 
+\vfill\null %Title not at bottom of page please
+
 \section{Evaluation}
@@
 
-\cba{All benchs on same VM specifically compiled (normal VM has only one impl)}
+To evaluate the performance, we had to compile a special VM with all the different primitive implementations included.
+The production VM normally only features one string comparison primitive implementation, not three of them.
+We compiled one VM for Linux with GCC, and one for Mac with LLVM (exact version numbers below).
+We ran all benchmarks on the same run of the same VM on both platforms.
 
@@
 The number of iterations of the loop depends on the string size: the comparison is repeated between 1414 and 10000 times for a short string (5 characters or less), and between 100 and 316 times for a long string (100 characters or more).
-This addresses the warm-up of the JIT \cba{???
-This needs to be explained differently.
-We get into jitted code after a small number of loop iteration (I think 20 currently) or second call.
-The JIT overhead is negligible here for 1 second because we could not see any time spent in the JIT in the profiler + ref to your paper, i.e.
-less than 0.5ms in the JIT} \cite{CogProfiler} and allows the benchmarks to run at least 1 second in the slower implementations to avoid processor and OS noises.
+The warm-up time of the JIT is in this context not noticeable on the performance results: the primitive is compiled to native code usually at the second call as explained in Section \ref{sec:VMexec} and the VM sampling profiler \cite{CogProfiler} confirms no significant time is spent in the interpreter loop (Between 0 and 1 sample over dozens of thousands are shown in the interpreter loop).
+
+The high number of iterations allows the benchmarks to run at least 1 second in the slower implementations to avoid processor and OS noises.
 The loop overhead is measured and removed from the measurements.
 
-The two compared strings are equal as this situation corresponds to the worst scenario for most implementations. There is one less branch in the two plugin implementations in this case.
+The two compared strings are equal as this situation corresponds to the worst scenario for most implementations. %There is one less branch in the two plugin implementations in this case.
 

### 2018-05-17 paper.tex
comment
update plot for linux
update plot for mac

@@
         x=0.5cm,
-        xtick={1,...,5},
+        xtick={1,...,4},
         xticklabel style={rotate=90},
@@
         enlarge x limits={abs=0.45cm},
-        xticklabels={\legBase, \legSmartSyntaxPlug, \legSlangPlug, \legSlang, \legRTL},
+        xticklabels={\legBase, \legSlangPlug, \legSlang, \legRTL},
         font={\footnotesize},
@@
 1   1           0
-2   0.3411212   0.0006746299
-3   0.3375418   0.001577878
-4   1.077762   0.005363974
-5   2.490423    0.004304882
+2   0.4222030   0.0008047936
+3   1.028828    0.004597225
+4   2.429007    0.002912642
 };
@@
 1   1           0
-2   0.6013716   0.0023206431
-3   0.5653467   0.007829291
-4   1.816869    0.019114984
-5   3.537800    0.012357500
+2   0.7673822   0.0011423885
+3   1.843189    0.011583771
+4   3.517044    0.019675605
 };
@@
 1   1           0
-2   1.3891212   0.0029489731
-3   1.3149840   0.020471694
-4   3.818961   0.054523074
-5   6.740474    0.010805042
+2   1.8905825   0.0119153480
+3   4.041207   0.061090881
+4   7.121638   0.010505603
 };
@@
 1   1           0
-2   8.6795487   0.0586762222
-3   8.9585208   0.116958804
-4   15.809515   0.300447885
-5   16.752667    0.106255682
+2   13.1217728   0.0170678541
+3   17.708440   0.105891879
+4   18.553670   0.023765562
 };
@@
 1   1           0
-2   13.6051267   0.0589164968
-3   14.7789145   0.147140189
-4   24.417751   0.186326223
-5   24.480071    0.095038374
+2   25.9913138   0.0320593050
+3   24.868577   0.062229078
+4   27.454826   0.044223466
 };
@@
 1   1           0
-2   0.3147064   0.002288488
-3   0.3035965   0.001786185
-4   0.8545431   0.006132950
-5   2.250732    0.04133638
+2   0.3841129   0.001065645
+3   0.2692461   0.0006797441
+4   2.477776   0.009248639
 };
@@
 1   1           0
-2   0.5605682   0.002198180
-3   0.5566675   0.001023225
-4   1.5605275   0.006372279
-5   3.754532    0.01109086
+2   0.7062775   0.001896893
+3   0.4939137   0.0013709049
+4   3.744060   0.012604598
 };
@@
 1   1           0
-2   1.2505193   0.004611354
-3   1.2295801   0.001810171
-4   2.8205415   0.008035024
-5   7.089806    0.01403785
+2   1.7187980   0.005024289
+3   1.1879773   0.0033351563
+4   7.245105   0.046617323
 };
@@
 1   1           0
-2   6.6951738   0.062391122
-3   6.6032572   0.024774416
-4   10.3029394  0.061481532
-5   17.979611   0.10127936
+2   12.6151983   0.080605928
+3   7.5970353   0.0396763134
+4   18.935631  0.124523834
 };
@@
 1   1           0
-2   9.4064534   0.036698187
-3   9.2138415   0.013982843
-4   13.4384830  0.040202717
-5   26.404792   0.09298324
+2   25.7537381   0.169939301
+3   11.6939697   0.0497562584
+4   27.829197  0.142897191
 };
@@
 
-\subsection{Results} %\sk{Results?}
-
-%\cba{Je pense string de taille 0-1 c'est cool pour montrer que le code C va moins vite que pur smalltalk dans ce cas (si c'est vraiment le cas)}
+\subsection{Results} 
+\cba{Please precise results. x\% for this size showing the problem. y\% for other size so non-existent problem. }
+
 We performed the benchmarks upon five different string sizes: 0, 1, 5, 100 and 1000 characters, using two different compilers: GCC on one hand and LLVM on the other hand.

### 2018-05-17 paper.tex
fix size 1000 on ubuntu

@@
 1   1           0
-2   25.9913138   0.0320593050
-3   24.868577   0.062229078
-4   27.454826   0.044223466
+2   25.35945   0.1742098
+3   25.42575   0.1621165
+4   26.76261   0.186075
 };
@@
 \item The Slang+JIT implementation is more efficient than the baseline in every case.
-Nevertheless, the performance gap between this version and the PureSlang one tends to decrease as the string size increases.
-\item The compiler used (either GCC or LLVM) does not lead to significant differences in terms of performance, except for the 100+-characters strings. 
+Nevertheless, the performance gap between this version and the PureSlang one \sk{all the Slang versions, in the end} tends to decrease as the string size increases.
+\item \sk{if we keep these results, then it actually leads to huge differences...} The compiler used (either GCC or LLVM) does not lead to significant differences in terms of performance, except for the 100+-characters strings.
 \end{itemize}

### 2018-05-17 paper.tex
update bench table + llvm results

@@
 1   1           0
-2   0.3841129   0.001065645
-3   0.2692461   0.0006797441
-4   2.477776   0.009248639
+2   0.3856066   0.0003752957
+3   0.2087539   0.0002019377
+4   2.571790   0.003453254
 };
@@
 1   1           0
-2   0.7062775   0.001896893
-3   0.4939137   0.0013709049
-4   3.744060   0.012604598
+2   0.6684630   0.0010425882
+3   0.3648824   0.0006104112
+4   3.591927   0.007633769
 };
@@
 1   1           0
-2   1.7187980   0.005024289
-3   1.1879773   0.0033351563
-4   7.245105   0.046617323
+2   1.6386585   0.0100987883
+3   0.8953226   0.0055539752
+4   6.956865   0.041610797
 };
@@
 1   1           0
-2   12.6151983   0.080605928
-3   7.5970353   0.0396763134
-4   18.935631  0.124523834
+2   11.6376234   0.0575507774
+3   6.1978094   0.0180202851
+4   17.274749  0.107117757
 };
@@
 1   1           0
-2   25.7537381   0.169939301
-3   11.6939697   0.0497562584
-4   27.829197  0.142897191
+2   23.9578625   0.1389707435
+3   10.6698346   0.0065863712
+4   25.399095  0.181054897
 };
@@
    \hline
-  \multirow{2}{*}{SlangPlugin} &GCC& 0.3375 +- 0.0016 & 0.5654 +- 0.0078 & 1.3150 +- 0.0205 & 8.9585 +- 0.1170  & 14.7789 +- 0.1471 \\
-  							   &LLVM& 0.3036 +- 0.0018 & 0.5567 +- 0.0010 & 1.2296 +- 0.0018 & 6.6033 +- 0.0248  & 9.2138 +- 0.0140 \\
+  \multirow{2}{*}{SlangPlugin} &GCC& 0.4222 +- 0.0008 & 0.7674 +- 0.0011 & 1.8906 +- 0.0119 & 13.1218 +- 0.0171  & 25.3595 +- 0.1742 \\
+  							   &LLVM& 0.3856 +- 0.0004 & 0.6685 +- 0.0010 & 1.6387 +- 0.0101 & 11.6376 +- 0.0576  & 23.9579 +- 0.1390 \\
    \hline
-  \multirow{2}{*}{PureSlang} &GCC& 1.0778 +- 0.0054 & 1.8169 +- 0.0191 & 3.8190 +- 0.0545 & 15.8095 +- 0.3005 & 24.4178 +- 0.1863\\
- 							 &LLVM& 0.8545 +- 0.0061 & 1.5605 +- 0.0064 & 2.8205 +- 0.0080 & 10.3029 +- 0.0615  & 13.4385 +- 0.0402 \\
+  \multirow{2}{*}{PureSlang} &GCC& 1.0288 +- 0.0046 & 1.8431 +- 0.0116 & 4.0412 +- 0.0611 & 17.7084 +- 0.1059 & 25.4258 +- 0.1621\\
+ 							 &LLVM& 0.2088 +- 0.0002 & 0.3649 +- 0.0064 & 0.0006 +- 0.0056 & 6.1978 +- 0.0180  & 10.6698 +- 0.0066 \\
    \hline
-  \multirow{2}{*}{Slang+JIT} &GCC& 2.4904 +- 0.0043 & 3.5378 +- 0.0124  & 6.7405 +- 0.0108 & 16.7527 +- 0.1063 & 24.4801 +- 0.0950\\
- 							&LLVM & 2.2507 +- 0.0413 & 3.7545 +- 0.0111 & 7.0898 +- 0.0140 & 17.9796 +- 0.1013  & 26.4048 +- 0.0930 \\
+  \multirow{2}{*}{Slang+JIT} &GCC& 2.4290 +- 0.0029 & 3.5170 +- 0.0197  & 7.1216 +- 0.0105 & 18.5537 +- 0.0238 & 26.7626 +- 0.1861\\
+ 							&LLVM & 2.5718 +- 0.0035 & 3.5919 +- 0.0076 & 6.9569 +- 0.0416 & 17.2747 +- 0.1071  & 25.3991 +- 0.1811 \\
   \hline

### 2018-05-18 paper.tex
update bench explanation. the LoC was not different from before

@@
 \subsection{Results} 
-\cba{Please precise results. x\% for this size showing the problem. y\% for other size so non-existent problem. }
-
 We performed the benchmarks upon five different string sizes: 0, 1, 5, 100 and 1000 characters, using two different compilers: GCC on one hand and LLVM on the other hand.
@@
 \begin{itemize}
-\item The execution time of the Plugin implementations is either very close or greater than \sd{within which ranges? 10\% 200\%} the baseline's for the small strings.
+\item The execution time of the Plugin implementations is greater than the baseline's for empty strings (0,58 times slower) or 1-character strings (0,23 times slower).
 \item The Slang+JIT implementation is more efficient than the baseline in every case.
-Nevertheless, the performance gap between this version and the PureSlang one \sk{all the Slang versions, in the end} tends to decrease as the string size increases.
-\item \sk{if we keep these results, then it actually leads to huge differences...} The compiler used (either GCC or LLVM) does not lead to significant differences in terms of performance, except for the 100+-characters strings.
+Nevertheless, the performance gap between this version and the Plugin and Slang ones tends to decrease as the string size increases, especially for GCC.
+For 1000-characters strings, the Slang+JIT version goes 0,25 times faster than the Plugin and the Slang versions.
+\item The compiler used (either GCC or LLVM) does lead to significant differences in terms of performance for the Slang implementation. 
 \end{itemize}
@@
 As mentioned in Section~\ref{sec:VMexec}, all implementations written in Slang come with an overhead due to the switch between the C and Smalltalk stacks.
-This overhead is clearly visible in the bench results in Figure~\ref{fig:bench} for small strings (size 1 or less): the Plugin implementation is around 30 to 60\% slower than the baseline's, and the PureSlang performance is close to the baseline's.
+This overhead is clearly visible in the bench results in Figure~\ref{fig:bench} for small strings (size 1 or less): the Plugin implementation is between 57,8 (gcc) to 61,4\% (llvm) slower than the baseline's for empty strings.
 As the string size goes larger, the overhead is less significant because it is absorbed by other performance gains.
@@
 \paragraph{Optimization potential on large strings comparison.}
-The plugins come with an overhead due the use of an interface.
+The Plugin come with an overhead due the use of an interface.
 Besides, cross-file inlining is not performed for plugins, like said in Section~\ref{sec:implem_cmp}.
-These are some of the factors explaining the performance gap between these versions and the Pure Slang and Slang+JIT ones.
+These are some of the factors explaining the performance gap between these versions and the Slang and Slang+JIT ones.
 When it comes to comparing large strings (100 characters and more), the execution time is mainly spent in the comparison loop (Figure~\ref{fig:Asm}).
-However, the switch overhead is still noticeable in the PureSlang implementation when strings of 100 characters are compared: PureSlang is 15.8 times faster than Baseline while Slang+JIT is 16.8 times faster than Baseline.
-This is not the case anymore on 1000-characters strings.
+However, the switch overhead is still noticeable in the Slang implementation when strings of 100 characters are compared: Slang is 17.7 times faster than Baseline while Slang+JIT is 18,6 times faster than Baseline.
 This last observation only applies to the GCC results: the code generated by the LLVM compiler induces a longer execution time, as explained in the next paragraph.
 
-One could notice a rather striking performance difference between the 2 compilers: when compiling with GCC, the performance between the PureSlang and Slang+JIT implementations is really close for long strings (100 characters and longer), with respectively 24,42 and 24,48 relative speed-up factors for a 1000-characters string.
-Yet performance greatly differs when LLVM is used, with respectively 13,44 and 26,40 relative speed-up factors.
+\paragraph{Control over generated code.} One could notice a rather striking performance difference between the 2 compilers: when compiling with GCC, the performance between the Slang and Slang+JIT implementations is really close for long strings (100 characters and longer), with respectively 25,42 and 26,76 relative speed-up factors for a 1000-characters string.
+Yet performance greatly differs when LLVM is used, with respectively 10,67 and 25,4 relative speed-up factors.
 
@@
 
-\todo{add conclusion SK}
+One can also notice the gap between the Plugin and Slang implementations in LLVM, which does not appear in GCC. \sk{one last comment: WHY?!}
+
+In this section, we compared the performance of the different implementations against a baseline (\ie{pure Smalltalk code}) implementation.
+The results showed the Slang+JIT implementation was always the fastest.
+Otherwise, most of the performance gaps between the primitive versions could be explained by 3 criteria: stack swithing, optimization for large strings, and the control over generated code.
 

### 2018-05-18 paper.tex
update table2 implem names
fix multiple things

@@
 
-\def\legBase{Baseline}
-\def\legSmartSyntaxPlug{Syntax}
-\def\legSlangPlug{Plugin}
-\def\legSlang{Slang}
-\def\legRTL{Slang+RTL}
+\def\legBase{Baseline\xspace}
+\def\legSmartSyntaxPlug{Syntax\xspace}
+\def\legSlangPlug{Plugin\xspace}
+\def\legSlang{Slang\xspace}
+\def\legRTL{Slang+RTL\xspace}
 \def\plotHeight{7cm}
@@
 
+Conceptually, the primitive with two parameters first checks if the operands are ByteStrings and fails if they are not.
+It then extract the ByteString sizes from their object header and computes the minimum size.
+Lastly, it iterates over the two ByteStrings until the minimum size is reached.
+If a difference is found, if answers the difference between the two different characters.
+If no difference is found, it answers the difference between the two string sizes.
+
 \section{Different primitive implementations}
@@
 
-We distinguish four implementations: the Baseline in pure Smalltalk, the Plugin version, the Slang version and the Slang+RTL version.
+We distinguish four implementations: the \legBase in pure Smalltalk, the \legSlangPlug version, the \legSlang version and the \legRTL version.
 
@@
 \paragraph{\legSlangPlug.}
-The VM plugin implementation can be easily added and removed as explained in Section \ref{sec:VMCompil}.
-Written in Slang, it can be compiled to C independently to the VM and is then compiled to native code through the C compiler as internal or external plugin.
-Some accesses to the string objects (for example, checking if they are byte strings) are done indirectly through the main VM API and these calls generate some overhead.
-%\sd{do we need the next sentence?}This primitive follows the old specification, the full code is available in Appendix~\ref{PluginAppendix}.
+Written in Slang, the VM plugin implementation is compiled to C independently to the VM and is then compiled to native code through the C compiler as an internal plugin, without enabling linking time optimizations.
+Accesses to the ByteString object header (to check if they are actually ByteString and not other objects as well as extracting their size), and accesses to the stack (to read the operands) are done indirectly through the main VM API since they're dependent on the memory representation of objects and stack layout used.
+These calls generate some overhead since they cannot be inlined.
+To avoid such calls in the main comparison loop, the primitive requests the main VM to provide a pointer to the first byte of each string and then assumes all bytes are then contiguous.
 
@@
 The Slang implementation is written inside the interpreter and has access to the full interpreter APIs.
-This primitive takes with two or three arguments and needs to deal with both cases in a way that the common case is efficient.
-The primitive is now part of the core of the VM, it cannot be easily added or removed as plugin primitives, but it is simpler to read and write.
-%This primitive follows the new specification, the full code is available in Appendix \ref{SlangAppendix}.
+It is compiled to C and then to native code as part of the main VM.
+There is no overhead in accessing the object memory representation or the stack as in the \legSlangPlug version.
 
@@
 The primitive is written in Cog's RTL, an abstract assembly whose instructions compile almost one-to-one to native instructions.
-This primitive requires a more important amount of work to be written: the number of line of code is larger, the implementor has to be careful to generate efficient machine code (there is no C compiler to optimize the code) and debugging such code requires the use of a processor simulator such as Bochs.
+This primitive requires a more important amount of work to be written: we had to be careful to generate efficient machine code (there is no C compiler to optimize the code) and debugging such code requires the use of a processor simulator such as Bochs.
 Since only performance critical implementations matter in this context, we've only re-implemented the case where there are two parameters.
 For the three parameter primitive, the VM falls back to the Slang version.
-%This primitive follows the new specification, the full code is available in Appendix \ref{RTLAppendix}.
 
@@
 
-\vfill\null %Title not at bottom of page please
+%\vfill\null %Title not at bottom of page please
 
@@
 
-\todo{Make a  pass cba}
-
-The SlangPlugin, PureSlang and Slang+JIT implementations of this string comparison primitive have been recently implemented in the Cog VM.
+The \legSlangPlug, \legSlang and \legRTL implementations of this string comparison primitive have been implemented in the Cog VM.
 In this section, we compare execution speed of these primitives against the other existing implementations.
@@
 \paragraph{Processor.}The following evaluation was performed on 32 bits Intel VMs (x86).
-
-\subsection{Methodology}
-
-We assessed the execution time of the different implementations by running the following benchmark:
-
-\begin{code}
-#(0 1 5 100 1000) collect: [ :size |
-	| iterations time overhead |
-	collection := ByteString new: size.
-	collection2 := ByteString new: size.
-	size < 100 
-		ifTrue: [iterations := (100000000 // (size*10 max: 1) sqrt) floor.]
-		ifFalse: [iterations := (10000000 // size sqrt) floor.].
-	overhead := [ 1 to: iterations do: [ :i | ] ] timeToRun.
-	time := [ 1 to: iterations do: [ :i |
-		collection compareWithSlang: collection2 ] ] timeToRun.
-	stream nextPutAll: ((time - overhead) asString)].
-\end{code}\\
-
-Each benchmark was measured 30 times.
 
@@
 \hline
-  \multirow{2}{*}{SlangPlugin} &GCC& 0.4222 +- 0.0008 & 0.7674 +- 0.0011 & 1.8906 +- 0.0119 & 13.1218 +- 0.0171  & 25.3595 +- 0.1742 \\
-  							   &LLVM& 0.3856 +- 0.0004 & 0.6685 +- 0.0010 & 1.6387 +- 0.0101 & 11.6376 +- 0.0576  & 23.9579 +- 0.1390 \\
+  \multirow{2}{*}{\legSlangPlug} &GCC-Linux& 0.4222 $\pm$ 0.0008 & 0.7674 $\pm$ 0.0011 & 1.8906 $\pm$ 0.0119 & 13.1218 $\pm$ 0.0171  & 25.3595 $\pm$ 0.1742 \\
+  						&LLVM-Mac& 0.3856 $\pm$ 0.0004 & 0.6685 $\pm$ 0.0010 & 1.6387 $\pm$ 0.0101 & 11.6376 $\pm$ 0.0576  & 23.9579 $\pm$ 0.1390 \\
    \hline
-  \multirow{2}{*}{PureSlang} &GCC& 1.0288 +- 0.0046 & 1.8431 +- 0.0116 & 4.0412 +- 0.0611 & 17.7084 +- 0.1059 & 25.4258 +- 0.1621\\
- 							 &LLVM& 0.2088 +- 0.0002 & 0.3649 +- 0.0064 & 0.0006 +- 0.0056 & 6.1978 +- 0.0180  & 10.6698 +- 0.0066 \\
+  \multirow{2}{*}{\legSlang} &GCC-Linux& 1.0288 $\pm$ 0.0046 & 1.8431 $\pm$ 0.0116 & 4.0412 $\pm$ 0.0611 & 17.7084 $\pm$ 0.1059 & 25.4258 $\pm$ 0.1621\\
+ 					 &LLVM-Mac& 0.2088 $\pm$ 0.0002 & 0.3649 $\pm$ 0.0064 & 0.8953 $\pm$ 0.0056 & 6.1978 $\pm$ 0.0180  & 10.6698 $\pm$ 0.0066 \\
    \hline
-  \multirow{2}{*}{Slang+JIT} &GCC& 2.4290 +- 0.0029 & 3.5170 +- 0.0197  & 7.1216 +- 0.0105 & 18.5537 +- 0.0238 & 26.7626 +- 0.1861\\
- 							&LLVM & 2.5718 +- 0.0035 & 3.5919 +- 0.0076 & 6.9569 +- 0.0416 & 17.2747 +- 0.1071  & 25.3991 +- 0.1811 \\
+  \multirow{2}{*}{\legRTL}  &GCC-Linux& 2.4290 $\pm$ 0.0029 & 3.5170 $\pm$ 0.0197  & 7.1216 $\pm$ 0.0105 & 18.5537 $\pm$ 0.0238 & 26.7626 $\pm$ 0.1861\\
+ 				 	&LLVM-Mac & 2.5718 $\pm$ 0.0035 & 3.5919 $\pm$ 0.0076 & 6.9569 $\pm$ 0.0416 & 17.2747 $\pm$ 0.1071  & 25.3991 $\pm$ 0.1811 \\
   \hline
@@
 
-In each benchmark, we run a string comparison in a loop and measure the total execution time.
+\subsection{Methodology}
+
+We assessed the execution time of the different implementations by running the following benchmark:
+
+\begin{code}
+#(0 1 5 100 1000) collect: [ :size |
+	| iterations time overhead |
+	collection := ByteString new: size.
+	collection2 := ByteString new: size.
+	size < 100 
+		ifTrue: [iterations := (100000000 // (size*10 max: 1) sqrt) floor.]
+		ifFalse: [iterations := (10000000 // size sqrt) floor.].
+	overhead := [ 1 to: iterations do: [ :i | ] ] timeToRun.
+	time := [ 1 to: iterations do: [ :i |
+		collection compareWithSlang: collection2 ] ] timeToRun.
+	stream nextPutAll: ((time - overhead) asString)].
+\end{code}\\
+
+Each benchmark was measured 30 times. Each time, we run a string comparison in a loop and measure the total execution time.
+
 The number of iterations of the loop depends on the string size: the comparison is repeated between 1414 and 10000 times for a short string (5 characters or less), and between 100 and 316 times for a long string (100 characters or more).
 The warm-up time of the JIT is in this context not noticeable on the performance results: the primitive is compiled to native code usually at the second call as explained in Section \ref{sec:VMexec} and the VM sampling profiler \cite{CogProfiler} confirms no significant time is spent in the interpreter loop (Between 0 and 1 sample over dozens of thousands are shown in the interpreter loop).
-
-The high number of iterations allows the benchmarks to run at least 1 second in the slower implementations to avoid processor and OS noises.
+The number of iterations allows the benchmarks to run for at least 1 second in the slower implementations to avoid processor and OS noises.
+
 The loop overhead is measured and removed from the measurements.
-
-The two compared strings are equal as this situation corresponds to the worst scenario for most implementations. %There is one less branch in the two plugin implementations in this case.
+The two compared strings are equal as this situation corresponds to the worst scenario for most implementations.
+%There is one less branch in the two plugin implementations in this case.
 
 \subsection{Results} 
-We performed the benchmarks upon five different string sizes: 0, 1, 5, 100 and 1000 characters, using two different compilers: GCC on one hand and LLVM on the other hand.
-Figure~\ref{fig:bench} shows the results in terms of relative speed-up towards the baseline implementation, respectively for GCC and LLVM.
+We performed the benchmarks upon five different string sizes: 0, 1, 5, 100 and 1000 characters.
+In the case of small strings, most of the time is spent activating the primitive and in the code before the main comparison loop, the primitive is performed with a little overall number of native instructions.
+In the case of large strings, the time spent executing the primitive is dominated by the main comparison loop.
+Figure~\ref{fig:bench} shows the results in terms of relative speed-up towards the baseline implementation, respectively for GCC-Linux and LLVM-Mac.
 
@@
 \begin{itemize}
-\item The execution time of the Plugin implementations is greater than the baseline's for empty strings (0,58 times slower) or 1-character strings (0,23 times slower).
-\item The Slang+JIT implementation is more efficient than the baseline in every case.
-Nevertheless, the performance gap between this version and the Plugin and Slang ones tends to decrease as the string size increases, especially for GCC.
-For 1000-characters strings, the Slang+JIT version goes 0,25 times faster than the Plugin and the Slang versions.
+\item The \legSlangPlug implementation is slower than the baseline version for small strings (0,58 times slower for empty strings, 0,23 times slower for 1-character strings).
+This is due to the overhead of the calls to the main VM to access the object headers and the stack.
+On large strings, the performance is comparable to
+\item The \legRTL implementation is more efficient than the baseline in every case.
+Nevertheless, the performance gap between this version and the \legSlangPlug and \legSlang ones tends to decrease as the string size increases, especially for GCC.
+For 1000-characters strings, the \legRTL version goes 0,25 times faster than the \legSlangPlug and the \legSlang versions.
 \item The compiler used (either GCC or LLVM) does lead to significant differences in terms of performance for the Slang implementation. 
@@
 As the string size goes larger, the overhead is less significant because it is absorbed by other performance gains.
-Regarding this criteria, the Slang+JIT version is always the fastest as it is not impacted by the switch overhead.
+Regarding this criteria, \legRTL version is always the fastest as it is not impacted by the switch overhead.
 
@@
 The Plugin come with an overhead due the use of an interface.
-Besides, cross-file inlining is not performed for plugins, like said in Section~\ref{sec:implem_cmp}.
-These are some of the factors explaining the performance gap between these versions and the Slang and Slang+JIT ones.
+Besides, cross-file inlining is not performed for plugin, like said in Section~\ref{sec:implem_cmp}.
+These are some of the factors explaining the performance gap between these versions and the \legSlang and \legRTL ones.
 When it comes to comparing large strings (100 characters and more), the execution time is mainly spent in the comparison loop (Figure~\ref{fig:Asm}).
@@
 
-In this section, we compared the performance of the different implementations against a baseline (\ie{pure Smalltalk code}) implementation.
+In this section, we compared the performance of the different implementations against a baseline (\emph{i.e.,}{pure Smalltalk code}) implementation.
 The results showed the Slang+JIT implementation was always the fastest.

### 2018-05-18 paper.tex
make things consistent

@@
 \def\legRTL{Slang+RTL\xspace}
+\def\LLVMMac{LLVM-Mac\xspace}
+\def\GCCLinux{GCC-Linux\xspace}
 \def\plotHeight{7cm}
@@
 
-\paragraph{GCC-Linux.}The GCC-Linux evaluation was performed on an Asus Zenbook with Ubuntu 16.04 LTS, a 2.2 Ghz processor Intel Core 5, 8 Gb 1600MHz DDR3 of RAM.
+\paragraph{\GCCLinux.}The \GCCLinux evaluation was performed on an Asus Zenbook with Ubuntu 16.04 LTS, a 2.2 Ghz processor Intel Core 5, 8 Gb 1600MHz DDR3 of RAM.
 The Linux VM was compiled using GCC version 5.4.0.
 
-\paragraph{LLVM-Mac.}The LLVM-Mac evaluation was performed on a MacBook pro with Mac OS 10.11.6, a 2.9 Ghz processor Intel Core 5, 8 Gb 1867MHz DDR3 of RAM.
+\paragraph{\LLVMMac.}The \LLVMMac evaluation was performed on a MacBook pro with Mac OS 10.11.6, a 2.9 Ghz processor Intel Core 5, 8 Gb 1867MHz DDR3 of RAM.
 The Mac VM was compiled using Apple LLVM version 8.0.0 (clang-800.0.38).
@@
    % \label{fig:gcc}
-   \subcaption*{\textbf{GCC-Linux}\vspace{0.25cm}}
+   \subcaption*{\textbf{\GCCLinux}\vspace{0.25cm}}
    \label{fig:gcc}
@@
    \end{subfigure}
-   \subcaption*{\textbf{LLVM-Mac}\vspace{-0.2cm}} %change vspace based on fig position
+   \subcaption*{\textbf{\LLVMMac}\vspace{-0.2cm}} %change vspace based on fig position
     \end{subfigure}
@@
 \hline
-  \multirow{2}{*}{\legSlangPlug} &GCC-Linux& 0.4222 $\pm$ 0.0008 & 0.7674 $\pm$ 0.0011 & 1.8906 $\pm$ 0.0119 & 13.1218 $\pm$ 0.0171  & 25.3595 $\pm$ 0.1742 \\
-  						&LLVM-Mac& 0.3856 $\pm$ 0.0004 & 0.6685 $\pm$ 0.0010 & 1.6387 $\pm$ 0.0101 & 11.6376 $\pm$ 0.0576  & 23.9579 $\pm$ 0.1390 \\
+  \multirow{2}{*}{\legSlangPlug} &\GCCLinux& 0.422 $\pm$ 0.0008 & 0.767 $\pm$ 0.0011 & 1.891 $\pm$ 0.0119 & 13.12 $\pm$ 0.0171  & 25.36 $\pm$ 0.1742 \\
+  						&\LLVMMac& 0.386 $\pm$ 0.0004 & 0.669 $\pm$ 0.0010 & 1.639 $\pm$ 0.0101 & 11.64 $\pm$ 0.0576  & 23.96 $\pm$ 0.1390 \\
    \hline
-  \multirow{2}{*}{\legSlang} &GCC-Linux& 1.0288 $\pm$ 0.0046 & 1.8431 $\pm$ 0.0116 & 4.0412 $\pm$ 0.0611 & 17.7084 $\pm$ 0.1059 & 25.4258 $\pm$ 0.1621\\
- 					 &LLVM-Mac& 0.2088 $\pm$ 0.0002 & 0.3649 $\pm$ 0.0064 & 0.8953 $\pm$ 0.0056 & 6.1978 $\pm$ 0.0180  & 10.6698 $\pm$ 0.0066 \\
+  \multirow{2}{*}{\legSlang} &\GCCLinux& 1.029 $\pm$ 0.0046 & 1.843 $\pm$ 0.0116 & 4.041 $\pm$ 0.0611 & 17.71 $\pm$ 0.1059 & 25.43 $\pm$ 0.1621\\
+ 					 &\LLVMMac& 0.209 $\pm$ 0.0002 & 0.365 $\pm$ 0.0064 & 0.895 $\pm$ 0.0056 & 6.198 $\pm$ 0.0180  & 10.67 $\pm$ 0.0066 \\
    \hline
-  \multirow{2}{*}{\legRTL}  &GCC-Linux& 2.4290 $\pm$ 0.0029 & 3.5170 $\pm$ 0.0197  & 7.1216 $\pm$ 0.0105 & 18.5537 $\pm$ 0.0238 & 26.7626 $\pm$ 0.1861\\
- 				 	&LLVM-Mac & 2.5718 $\pm$ 0.0035 & 3.5919 $\pm$ 0.0076 & 6.9569 $\pm$ 0.0416 & 17.2747 $\pm$ 0.1071  & 25.3991 $\pm$ 0.1811 \\
+  \multirow{2}{*}{\legRTL}  &\GCCLinux& 2.429 $\pm$ 0.0029 & 3.517 $\pm$ 0.0197  & 7.122 $\pm$ 0.0105 & 18.55 $\pm$ 0.0238 & 26.76 $\pm$ 0.1861\\
+ 				 	&\LLVMMac & 2.572 $\pm$ 0.0035 & 3.592 $\pm$ 0.0076 & 6.957 $\pm$ 0.0416 & 17.28 $\pm$ 0.1071  & 25.40 $\pm$ 0.1811 \\
   \hline
@@
 		\includegraphics[width=1\linewidth]{figures/asm_cmp.pdf}
-		\caption{GCC and LLVM: different optimizations on the comparison loop.}
+		\caption{\GCCLinux and \LLVMMac: different optimizations on the comparison loop.}
 		\label{fig:Asm}
@@
 	time := [ 1 to: iterations do: [ :i |
-		collection compareWithSlang: collection2 ] ] timeToRun.
+		collection primitiveStringCompare: collection2 ] ] timeToRun.
 	stream nextPutAll: ((time - overhead) asString)].
@@
 
-The loop overhead is measured and removed from the measurements.
+The loop overhead is measured and removed from the measurements (We made sure the JIT did not remove the dead loop to measure it).
 The two compared strings are equal as this situation corresponds to the worst scenario for most implementations.
@@
 In the case of large strings, the time spent executing the primitive is dominated by the main comparison loop.
-Figure~\ref{fig:bench} shows the results in terms of relative speed-up towards the baseline implementation, respectively for GCC-Linux and LLVM-Mac.
-
+Figure~\ref{fig:bench} shows the results in terms of relative speed-up towards the baseline implementation, respectively for \GCCLinux and \LLVMMac.
+
+\cba{Is next paragraph good ?}
 We can draw the following patterns from these results:
@@
 This is due to the overhead of the calls to the main VM to access the object headers and the stack.
-On large strings, the performance is comparable to
+On large strings, \legSlangPlug is one of the fastest implementation with
 \item The \legRTL implementation is more efficient than the baseline in every case.
-Nevertheless, the performance gap between this version and the \legSlangPlug and \legSlang ones tends to decrease as the string size increases, especially for GCC.
+Nevertheless, the performance gap between this version and the \legSlangPlug and \legSlang ones tends to decrease as the string size increases, especially for \GCCLinux.
 For 1000-characters strings, the \legRTL version goes 0,25 times faster than the \legSlangPlug and the \legSlang versions.
-\item The compiler used (either GCC or LLVM) does lead to significant differences in terms of performance for the Slang implementation. 
+\item The set-up used (either \GCCLinux or \LLVMMac) does lead to significant differences in terms of performance for the Slang implementation. 
 \end{itemize}
@@
 As mentioned in Section~\ref{sec:VMexec}, all implementations written in Slang come with an overhead due to the switch between the C and Smalltalk stacks.
-This overhead is clearly visible in the bench results in Figure~\ref{fig:bench} for small strings (size 1 or less): the Plugin implementation is between 57,8 (gcc) to 61,4\% (llvm) slower than the baseline's for empty strings.
+This overhead is clearly visible in the bench results in Figure~\ref{fig:bench} for small strings (size 1 or less): the Plugin implementation is between 57,8 (\GCCLinux) to 61,4\% (\LLVMMac) slower than the baseline's for empty strings.
 As the string size goes larger, the overhead is less significant because it is absorbed by other performance gains.
@@
 However, the switch overhead is still noticeable in the Slang implementation when strings of 100 characters are compared: Slang is 17.7 times faster than Baseline while Slang+JIT is 18,6 times faster than Baseline.
-This last observation only applies to the GCC results: the code generated by the LLVM compiler induces a longer execution time, as explained in the next paragraph.
-
-\paragraph{Control over generated code.} One could notice a rather striking performance difference between the 2 compilers: when compiling with GCC, the performance between the Slang and Slang+JIT implementations is really close for long strings (100 characters and longer), with respectively 25,42 and 26,76 relative speed-up factors for a 1000-characters string.
+This last observation only applies to the \GCCLinux results: the code generated by the LLVM compiler induces a longer execution time, as explained in the next paragraph.
+
+\paragraph{Control over generated code.} One could notice a rather striking performance difference between the 2 compilers: when compiling with GCC, the performance between the \legSlang and \legRTL implementations is really close for long strings (100 characters and longer), with respectively 25,42 and 26,76 relative speed-up factors for a 1000-characters string.
 Yet performance greatly differs when LLVM is used, with respectively 10,67 and 25,4 relative speed-up factors.
@@
 In this section, we compared the performance of the different implementations against a baseline (\emph{i.e.,}{pure Smalltalk code}) implementation.
-The results showed the Slang+JIT implementation was always the fastest.
+The results showed the \legRTL implementation was always the fastest.
 Otherwise, most of the performance gaps between the primitive versions could be explained by 3 criteria: stack swithing, optimization for large strings, and the control over generated code.

### 2018-05-18 paper.tex
update mac results

@@
 1   1           0
-2   0.3856066   0.0003752957
-3   0.2087539   0.0002019377
-4   2.571790   0.003453254
+2   0.3391346   0.0006874275
+3   0.9324989   0.001469945
+4   3.254642   0.007910739
 };
@@
 1   1           0
-2   0.6684630   0.0010425882
-3   0.3648824   0.0006104112
-4   3.591927   0.007633769
+2   0.6237533   0.0012023524
+3   1.6900368   0.003040773
+4   5.215819   0.015275079
 };
@@
 1   1           0
-2   1.6386585   0.0100987883
-3   0.8953226   0.0055539752
-4   6.956865   0.041610797
+2   1.5807038   0.0092529942
+3   3.2413871   0.006737660
+4   8.970703   0.035921295
 };
@@
 1   1           0
-2   11.6376234   0.0575507774
-3   6.1978094   0.0180202851
-4   17.274749  0.107117757
+2   11.4096285   0.0473858056
+3   10.1485203   0.032077214
+4   17.642998  0.059953679
 };
@@
 1   1           0
-2   23.9578625   0.1389707435
-3   10.6698346   0.0065863712
-4   25.399095  0.181054897
+2   23.2087956   0.1788766172
+3   12.9392674   0.017734225
+4   24.880755  0.198267254
 };
@@
   \multirow{2}{*}{\legSlangPlug} &\GCCLinux& 0.422 $\pm$ 0.0008 & 0.767 $\pm$ 0.0011 & 1.891 $\pm$ 0.0119 & 13.12 $\pm$ 0.0171  & 25.36 $\pm$ 0.1742 \\
-  						&\LLVMMac& 0.386 $\pm$ 0.0004 & 0.669 $\pm$ 0.0010 & 1.639 $\pm$ 0.0101 & 11.64 $\pm$ 0.0576  & 23.96 $\pm$ 0.1390 \\
+  						&\LLVMMac& 0.339 $\pm$ 0.0007 & 0.6238 $\pm$ 0.0012 & 1.5807 $\pm$ 0.0092 & 11.4096 $\pm$ 0.0474  & 23.21 $\pm$ 0.1789 \\
    \hline
   \multirow{2}{*}{\legSlang} &\GCCLinux& 1.029 $\pm$ 0.0046 & 1.843 $\pm$ 0.0116 & 4.041 $\pm$ 0.0611 & 17.71 $\pm$ 0.1059 & 25.43 $\pm$ 0.1621\\
- 					 &\LLVMMac& 0.209 $\pm$ 0.0002 & 0.365 $\pm$ 0.0064 & 0.895 $\pm$ 0.0056 & 6.198 $\pm$ 0.0180  & 10.67 $\pm$ 0.0066 \\
+ 					 &\LLVMMac& 0.932 $\pm$ 0.0015 & 1.6900 $\pm$ 0.0030 & 3.241 $\pm$ 0.0067 & 10.149 $\pm$ 0.0321  & 12.939 $\pm$ 0.0177 \\
    \hline
   \multirow{2}{*}{\legRTL}  &\GCCLinux& 2.429 $\pm$ 0.0029 & 3.517 $\pm$ 0.0197  & 7.122 $\pm$ 0.0105 & 18.55 $\pm$ 0.0238 & 26.76 $\pm$ 0.1861\\
- 				 	&\LLVMMac & 2.572 $\pm$ 0.0035 & 3.592 $\pm$ 0.0076 & 6.957 $\pm$ 0.0416 & 17.28 $\pm$ 0.1071  & 25.40 $\pm$ 0.1811 \\
+ 				 	&\LLVMMac & 3.255 $\pm$ 0.0079 & 5.216 $\pm$ 0.0153 & 8.971 $\pm$ 0.0359 & 17.643 $\pm$ 0.06  & 24.881 $\pm$ 0.1983 \\
   \hline
@@
 This is due to the overhead of the calls to the main VM to access the object headers and the stack.
-On large strings, \legSlangPlug is one of the fastest implementation with
+On large strings, \legSlangPlug is one of the fastest implementation.
 \item The \legRTL implementation is more efficient than the baseline in every case.

### 2018-05-18 paper.tex
update linux results
update linux results - fix the 117 nonsense
update the asm comparison loop and figure 3 title
fix mistale

@@
 1   1           0
-2   0.4222030   0.0008047936
-3   1.028828    0.004597225
-4   2.429007    0.002912642
+2   0.413386   0.0004942463
+3   1.286553    0.004837948
+4   3.343083    0.02105667
 };
@@
 1   1           0
-2   0.7673822   0.0011423885
-3   1.843189    0.011583771
-4   3.517044    0.019675605
+2   0.745517   0.0008472452
+3   2.256624    0.011554253
+4   5.171014    0.02650453
 };
@@
 1   1           0
-2   1.8905825   0.0119153480
-3   4.041207   0.061090881
-4   7.121638   0.010505603
+2   1.804201   0.0004602098
+3   4.958408   0.011106844
+4   8.714720   0.02207683
 };
@@
 1   1           0
-2   13.1217728   0.0170678541
-3   17.708440   0.105891879
-4   18.553670   0.023765562
+2   12.047182   0.0088056909
+3   17.240628   0.051498584
+4   17.656653   0.04813202
 };
@@
 1   1           0
-2   25.35945   0.1742098
-3   25.42575   0.1621165
-4   26.76261   0.186075
+2   23.492410   0.0134896734
+3   22.541872   0.077485955
+4   24.927379   0.03087542
 };
@@
     \end{subfigure}
-   \caption{String comparison benchmarks for strings of different sizes on different architectures.}
+   \caption{String comparison benchmarks for strings of different sizes on different architectures. Average relative speed-up regarding baseline. Higher than 1 is better.}
    \label{fig:bench}
@@
 \hline
-  \multirow{2}{*}{\legSlangPlug} &\GCCLinux& 0.422 $\pm$ 0.0008 & 0.767 $\pm$ 0.0011 & 1.891 $\pm$ 0.0119 & 13.12 $\pm$ 0.0171  & 25.36 $\pm$ 0.1742 \\
+  \multirow{2}{*}{\legSlangPlug} &\GCCLinux& 0.413 $\pm$ 0.0005 & 0.746 $\pm$ 0.0008 & 1.804 $\pm$ 0.0005 & 12.047 $\pm$ 0.0088 & 23.492 $\pm$ 0.0135 \\
   						&\LLVMMac& 0.339 $\pm$ 0.0007 & 0.6238 $\pm$ 0.0012 & 1.5807 $\pm$ 0.0092 & 11.4096 $\pm$ 0.0474  & 23.21 $\pm$ 0.1789 \\
    \hline
-  \multirow{2}{*}{\legSlang} &\GCCLinux& 1.029 $\pm$ 0.0046 & 1.843 $\pm$ 0.0116 & 4.041 $\pm$ 0.0611 & 17.71 $\pm$ 0.1059 & 25.43 $\pm$ 0.1621\\
+  \multirow{2}{*}{\legSlang} &\GCCLinux& 1.287 $\pm$ 0.0048 & 2.257 $\pm$ 0.0116 & 4.958 $\pm$ 0.0111 & 17.241 $\pm$ 0.0515 & 22.542 $\pm$ 0.0775\\
  					 &\LLVMMac& 0.932 $\pm$ 0.0015 & 1.6900 $\pm$ 0.0030 & 3.241 $\pm$ 0.0067 & 10.149 $\pm$ 0.0321  & 12.939 $\pm$ 0.0177 \\
    \hline
-  \multirow{2}{*}{\legRTL}  &\GCCLinux& 2.429 $\pm$ 0.0029 & 3.517 $\pm$ 0.0197  & 7.122 $\pm$ 0.0105 & 18.55 $\pm$ 0.0238 & 26.76 $\pm$ 0.1861\\
+  \multirow{2}{*}{\legRTL}  &\GCCLinux& 3.343 $\pm$ 0.0211 & 5.171 $\pm$ 0.0265  & 8.715 $\pm$ 0.0221 & 17.657 $\pm$ 0.0481 & 24.927 $\pm$ 0.0309\\
  				 	&\LLVMMac & 3.255 $\pm$ 0.0079 & 5.216 $\pm$ 0.0153 & 8.971 $\pm$ 0.0359 & 17.643 $\pm$ 0.06  & 24.881 $\pm$ 0.1983 \\
@@
 		\centering
-		\includegraphics[width=1\linewidth]{figures/asm_cmp.pdf}
+		\makebox[0pt]{
+		\includegraphics[width=1.09\linewidth]{figures/asm_cmp.pdf}}
 		\caption{\GCCLinux and \LLVMMac: different optimizations on the comparison loop.}
@@
 
-\cba{Is next paragraph good ?}
-We can draw the following patterns from these results:
-\begin{itemize}
-\item The \legSlangPlug implementation is slower than the baseline version for small strings (0,58 times slower for empty strings, 0,23 times slower for 1-character strings).
+%\cba{Is next paragraph good ?}
+%We can draw the following patterns from these results:
+%\begin{itemize}
+%\item The \legSlangPlug implementation is slower than the baseline version for small strings (0,58 times slower for empty strings, 0,23 times slower for 1-character strings).
 This is due to the overhead of the calls to the main VM to access the object headers and the stack.
 On large strings, \legSlangPlug is one of the fastest implementation.
-\item The \legRTL implementation is more efficient than the baseline in every case.
+%\item The \legRTL implementation is more efficient than the baseline in every case.
 Nevertheless, the performance gap between this version and the \legSlangPlug and \legSlang ones tends to decrease as the string size increases, especially for \GCCLinux.
 For 1000-characters strings, the \legRTL version goes 0,25 times faster than the \legSlangPlug and the \legSlang versions.
-\item The set-up used (either \GCCLinux or \LLVMMac) does lead to significant differences in terms of performance for the Slang implementation. 
-\end{itemize}
-
-We can explain these results based on the last 3 criteria detailed in Section~\ref{sec:implem_cmp}:
+%\item The set-up used (either \GCCLinux or \LLVMMac) does lead to significant differences in terms of performance for the Slang implementation. 
+%\end{itemize}
+
+These results can be explained based on the last 3 criteria detailed in Section~\ref{sec:implem_cmp}:
 
@@
 \paragraph{Optimization potential on large strings comparison.}
-The Plugin come with an overhead due the use of an interface.
+The Plugin comes with an overhead due the use of an interface.
 Besides, cross-file inlining is not performed for plugin, like said in Section~\ref{sec:implem_cmp}.

### 2018-05-18 paper.tex
consistency for llvm gcc names, add a sentence about the performance gap within llvm

@@
 The Plugin comes with an overhead due the use of an interface.
-Besides, cross-file inlining is not performed for plugin, like said in Section~\ref{sec:implem_cmp}.
+Besides, cross-file inlining is not performed for plugins, like said in Section~\ref{sec:implem_cmp}.
 These are some of the factors explaining the performance gap between these versions and the \legSlang and \legRTL ones.
 When it comes to comparing large strings (100 characters and more), the execution time is mainly spent in the comparison loop (Figure~\ref{fig:Asm}).
-However, the switch overhead is still noticeable in the Slang implementation when strings of 100 characters are compared: Slang is 17.7 times faster than Baseline while Slang+JIT is 18,6 times faster than Baseline.
+However, the switch overhead is still noticeable in the Slang implementation when strings of 100 characters are compared: Slang is 17.2 times faster than Baseline while Slang+JIT is 18,6 times faster than Baseline.
 This last observation only applies to the \GCCLinux results: the code generated by the LLVM compiler induces a longer execution time, as explained in the next paragraph.
 
-\paragraph{Control over generated code.} One could notice a rather striking performance difference between the 2 compilers: when compiling with GCC, the performance between the \legSlang and \legRTL implementations is really close for long strings (100 characters and longer), with respectively 25,42 and 26,76 relative speed-up factors for a 1000-characters string.
-Yet performance greatly differs when LLVM is used, with respectively 10,67 and 25,4 relative speed-up factors.
-
-This difference is mainly due to the optimizations performed by both the compilers: as depicted in Figure~\ref{fig:Asm}, GCC generates instructions that are very close to the ones generated by the JIT, hence the similar speed-up factors.
-On the contrary, LLVM generates more instructions: it especially fails at preserving registers holding the strings (edi and esi), leading to two extra instructions (bolded in the figure) that  load the strings at each iteration.
-
-One can also notice the gap between the Plugin and Slang implementations in LLVM, which does not appear in GCC. \sk{one last comment: WHY?!}
-
-In this section, we compared the performance of the different implementations against a baseline (\emph{i.e.,}{pure Smalltalk code}) implementation.
+\paragraph{Control over generated code.} One could notice a rather striking performance difference between the 2 compilers: when compiling with GCC, the performance between the \legSlang and \legRTL implementations is really close for long strings (100 characters and longer), with respectively 22.54 and 24.93 relative speed-up factors for a 1000-characters string.
+Yet performance greatly differs when LLVM is used, with respectively 12,94 and 24,88 relative speed-up factors.
+
+This difference is mainly due to the optimizations performed by both the compilers: as depicted in Figure~\ref{fig:Asm}, \GCCLinux generates instructions that are very close to the ones generated by the JIT, hence the similar speed-up factors.
+On the contrary, \LLVMMac generates more instructions: it especially fails at preserving registers holding the strings (edi and esi), leading to two extra instructions (bolded in the figure) that  load the strings at each iteration.
+
+One can also notive a significant gap between the Plugin and Slang implementations in \LLVMMac.
+This can also be explained by a difference in optimizations.
+As seen in Figure \ref{fig:Asm}, the comparison loop is indeed optimized in the Plugin version and is similar to the JIT implementation.
+
+In this section, we compared the performance of the different implementations against a baseline (\emph{i.e., }{pure Smalltalk code}) implementation.
 The results showed the \legRTL implementation was always the fastest.

### 2018-05-18 paper.tex
update plot and table
update figures in paragraphs

@@
 1   1           0
-2   23.492410   0.0134896734
-3   22.541872   0.077485955
-4   24.927379   0.03087542
+2   25.35945   0.1742098
+3   25.42575   0.1621165
+4   26.76261   0.186075
 };
@@
 \hline
-  \multirow{2}{*}{\legSlangPlug} &\GCCLinux& 0.413 $\pm$ 0.0005 & 0.746 $\pm$ 0.0008 & 1.804 $\pm$ 0.0005 & 12.047 $\pm$ 0.0088 & 23.492 $\pm$ 0.0135 \\
+  \multirow{2}{*}{\legSlangPlug} &\GCCLinux& 0.413 $\pm$ 0.0005 & 0.746 $\pm$ 0.0008 & 1.804 $\pm$ 0.0005 & 12.047 $\pm$ 0.0088 & 25.359 $\pm$ 0.1742 \\
   						&\LLVMMac& 0.339 $\pm$ 0.0007 & 0.6238 $\pm$ 0.0012 & 1.5807 $\pm$ 0.0092 & 11.4096 $\pm$ 0.0474  & 23.21 $\pm$ 0.1789 \\
    \hline
-  \multirow{2}{*}{\legSlang} &\GCCLinux& 1.287 $\pm$ 0.0048 & 2.257 $\pm$ 0.0116 & 4.958 $\pm$ 0.0111 & 17.241 $\pm$ 0.0515 & 22.542 $\pm$ 0.0775\\
+  \multirow{2}{*}{\legSlang} &\GCCLinux& 1.287 $\pm$ 0.0048 & 2.257 $\pm$ 0.0116 & 4.958 $\pm$ 0.0111 & 17.241 $\pm$ 0.0515 & 25.426 $\pm$ 0.1621\\
  					 &\LLVMMac& 0.932 $\pm$ 0.0015 & 1.6900 $\pm$ 0.0030 & 3.241 $\pm$ 0.0067 & 10.149 $\pm$ 0.0321  & 12.939 $\pm$ 0.0177 \\
    \hline
-  \multirow{2}{*}{\legRTL}  &\GCCLinux& 3.343 $\pm$ 0.0211 & 5.171 $\pm$ 0.0265  & 8.715 $\pm$ 0.0221 & 17.657 $\pm$ 0.0481 & 24.927 $\pm$ 0.0309\\
+  \multirow{2}{*}{\legRTL}  &\GCCLinux& 3.343 $\pm$ 0.0211 & 5.171 $\pm$ 0.0265  & 8.715 $\pm$ 0.0221 & 17.657 $\pm$ 0.0481 & 26.763 $\pm$ 0.1861\\
  				 	&\LLVMMac & 3.255 $\pm$ 0.0079 & 5.216 $\pm$ 0.0153 & 8.971 $\pm$ 0.0359 & 17.643 $\pm$ 0.06  & 24.881 $\pm$ 0.1983 \\
@@
 As mentioned in Section~\ref{sec:VMexec}, all implementations written in Slang come with an overhead due to the switch between the C and Smalltalk stacks.
-This overhead is clearly visible in the bench results in Figure~\ref{fig:bench} for small strings (size 1 or less): the Plugin implementation is between 57,8 (\GCCLinux) to 61,4\% (\LLVMMac) slower than the baseline's for empty strings.
+This overhead is clearly visible in the bench results in Figure~\ref{fig:bench} for small strings (size 1 or less): the Plugin implementation is between 58,7 (\GCCLinux) to 66,1\% (\LLVMMac) slower than the baseline's for empty strings.
 As the string size goes larger, the overhead is less significant because it is absorbed by other performance gains.
@@
 
-\paragraph{Control over generated code.} One could notice a rather striking performance difference between the 2 compilers: when compiling with GCC, the performance between the \legSlang and \legRTL implementations is really close for long strings (100 characters and longer), with respectively 22.54 and 24.93 relative speed-up factors for a 1000-characters string.
+\paragraph{Control over generated code.} One could notice a rather striking performance difference between the 2 compilers: when compiling with GCC, the performance between the \legSlang and \legRTL implementations is really close for long strings (100 characters and longer), with respectively 25.24 and 26.76 relative speed-up factors for a 1000-characters string.
 Yet performance greatly differs when LLVM is used, with respectively 12,94 and 24,88 relative speed-up factors.

### 2018-05-18 paper.tex
fix bib ref in between collumns
foo

--- 2018-05-18 paper.tex
@@
 \def\GCCLinux{GCC-Linux\xspace}
-\def\plotHeight{7cm}
+\def\plotHeight{6.8cm}
 
@@
     \end{subfigure}
-   \caption{String comparison benchmarks for strings of different sizes on different architectures. Average relative speed-up regarding baseline. Higher than 1 is better.}
+   \caption{String comparison benchmarks for strings of different sizes on different architectures. Average relative speed-up regarding baseline. The higher the better.}
    \label{fig:bench}
@@
 \hline
-  \multirow{2}{*}{\legSlangPlug} &\GCCLinux& 0.413 $\pm$ 0.0005 & 0.746 $\pm$ 0.0008 & 1.804 $\pm$ 0.0005 & 12.047 $\pm$ 0.0088 & 23.492 $\pm$ 0.0135 \\
-  						&\LLVMMac& 0.339 $\pm$ 0.0007 & 0.6238 $\pm$ 0.0012 & 1.5807 $\pm$ 0.0092 & 11.4096 $\pm$ 0.0474  & 23.21 $\pm$ 0.1789 \\
+  \multirow{2}{*}{\legSlangPlug} &\GCCLinux& 0.413 $\pm$ 0.0005 & 0.746 $\pm$ 0.0008 & 1.804 $\pm$ 0.0005 & 12.05 $\pm$ 0.0088 & 23.49 $\pm$ 0.0135 \\
+  						&\LLVMMac& 0.339 $\pm$ 0.0007 & 0.624 $\pm$ 0.0012 & 1.581 $\pm$ 0.0092 & 11.41 $\pm$ 0.0474  & 23.21 $\pm$ 0.1789 \\
    \hline
-  \multirow{2}{*}{\legSlang} &\GCCLinux& 1.287 $\pm$ 0.0048 & 2.257 $\pm$ 0.0116 & 4.958 $\pm$ 0.0111 & 17.241 $\pm$ 0.0515 & 22.542 $\pm$ 0.0775\\
- 					 &\LLVMMac& 0.932 $\pm$ 0.0015 & 1.6900 $\pm$ 0.0030 & 3.241 $\pm$ 0.0067 & 10.149 $\pm$ 0.0321  & 12.939 $\pm$ 0.0177 \\
+  \multirow{2}{*}{\legSlang} &\GCCLinux& 1.287 $\pm$ 0.0048 & 2.257 $\pm$ 0.0116 & 4.958 $\pm$ 0.0111 & 17.24 $\pm$ 0.0515 & 22.54 $\pm$ 0.0775\\
+ 					 &\LLVMMac& 0.932 $\pm$ 0.0015 & 1.690 $\pm$ 0.0030 & 3.241 $\pm$ 0.0067 & 10.15 $\pm$ 0.0321  & 12.94 $\pm$ 0.0177 \\
    \hline
-  \multirow{2}{*}{\legRTL}  &\GCCLinux& 3.343 $\pm$ 0.0211 & 5.171 $\pm$ 0.0265  & 8.715 $\pm$ 0.0221 & 17.657 $\pm$ 0.0481 & 24.927 $\pm$ 0.0309\\
- 				 	&\LLVMMac & 3.255 $\pm$ 0.0079 & 5.216 $\pm$ 0.0153 & 8.971 $\pm$ 0.0359 & 17.643 $\pm$ 0.06  & 24.881 $\pm$ 0.1983 \\
+  \multirow{2}{*}{\legRTL}  &\GCCLinux& 3.343 $\pm$ 0.0211 & 5.171 $\pm$ 0.0265  & 8.715 $\pm$ 0.0221 & 17.66 $\pm$ 0.0481 & 24.93 $\pm$ 0.0309\\
+ 				 	&\LLVMMac & 3.255 $\pm$ 0.0079 & 5.216 $\pm$ 0.0153 & 8.971 $\pm$ 0.0359 & 17.64 $\pm$ 0.0600  & 24.88 $\pm$ 0.1983 \\
   \hline
@@
 		\makebox[0pt]{
-		\includegraphics[width=1.09\linewidth]{figures/asm_cmp.pdf}}
+		\includegraphics[width=\linewidth]{figures/asm_cmp.pdf}}
 		\caption{\GCCLinux and \LLVMMac: different optimizations on the comparison loop.}
@@
 This overhead is clearly visible in the bench results in Figure~\ref{fig:bench} for small strings (size 1 or less): the Plugin implementation is between 57,8 (\GCCLinux) to 61,4\% (\LLVMMac) slower than the baseline's for empty strings.
-As the string size goes larger, the overhead is less significant because it is absorbed by other performance gains.
+The problem is even more important for the plugin version, since the beginning of the primitive code, before the main comparison loop, is slower due to the calls to the main VM code.
+As the string size goes larger, the overhead is less significant because the execution time becomes dominated by the time spent in the main comparison loop.
 Regarding this criteria, \legRTL version is always the fastest as it is not impacted by the switch overhead.
@@
 
-\paragraph{Control over generated code.} One could notice a rather striking performance difference between the 2 compilers: when compiling with GCC, the performance between the \legSlang and \legRTL implementations is really close for long strings (100 characters and longer), with respectively 22.54 and 24.93 relative speed-up factors for a 1000-characters string.
-Yet performance greatly differs when LLVM is used, with respectively 12,94 and 24,88 relative speed-up factors.
-
-This difference is mainly due to the optimizations performed by both the compilers: as depicted in Figure~\ref{fig:Asm}, \GCCLinux generates instructions that are very close to the ones generated by the JIT, hence the similar speed-up factors.
-On the contrary, \LLVMMac generates more instructions: it especially fails at preserving registers holding the strings (edi and esi), leading to two extra instructions (bolded in the figure) that  load the strings at each iteration.
+\paragraph{Control over generated code.} One can notice a rather striking performance difference between the 2 set-ups we have.In \GCCLinux, the performance between the \legSlang and \legRTL implementations is really close for long strings (1000 characters wide), with respectively 22.54 and 24.93 relative speed-up factors for a 1000-characters string.
+Yet performance greatly differs when the \LLVMMac set-up is used, with respectively 12,94 and 24,88 relative speed-up factors.
+
+This difference is mainly due to the optimizations performed or not by both the compilers for the \legSlang version.
+As depicted in Figure~\ref{fig:Asm}, \GCCLinux generates instructions that are very close to the ones generated by the JIT, hence the execution time is similar.
+On the contrary, \LLVMMac generates more instructions: it fails at keeping all the operands live across the comparison loops, leading to two extra memory reads per iteration (bolded in the figure).
+
+This is an example of one native function in the whole VM and we do not conclude anything about the C compiler capabilites from it.
+VMs compiled with LLVM and GCC are around equally fast except in narrow cases like this one and LLVM is able to optimize correctly the \legSlangPlug version.
+LLVM likely fails at optimizing the \legSlang version likely due to the old C extensions we use in the interpreter code present in the same file to force variables into specific registers, which are better supported by GCC, historically the only compiler we supported.
+However, we can see that the \legRTL version has the most reliable performance since it does not depend on what optimization the C compiler is able to perform or not.
+This is even more relevant in less common back-ends not shown in this evaluation, such as the MIPSEL back-end we support, where C compilers are usually less clever at performing back-end specific optimizations since less engineering time has been invested into such optimizations.
 
@@
 The results showed the \legRTL implementation was always the fastest.
-Otherwise, most of the performance gaps between the primitive versions could be explained by 3 criteria: stack swithing, optimization for large strings, and the control over generated code.
+Otherwise, most of the performance gaps between the primitive versions could be explained by 3 criteria: stack switching, optimization for large strings, and the control over generated code.
 

### 2018-05-18 paper.tex

@@
 	\item There is a big performance gap between the baseline performance and the peak performance.
-	\item Even mature JITs fail to optimize some narrow cases, where the performance is drastically slower~\cite{VMHotCold}.
+	\item Even mature JITs fail to optimize some narrow cases, where the performance is drastically slower \\~\cite{VMHotCold}.
 \end{itemize}
@@
 		\centering
-		\includegraphics[width=0.65\linewidth]{figures/VMCompilation}
+		\includegraphics[width=0.62\linewidth]{figures/VMCompilation}
 		\caption{Cog VM compilation}
@@
 		\centering
-		\includegraphics[width=0.99\linewidth]{figures/TwoStacks}
+		\includegraphics[width=0.96\linewidth]{figures/TwoStacks}
 		\caption{Client and C stacks.}
@@
 	\item \emph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
-	\item \emph{Optimization potential on large string comparison:}
-	C cross-file inlining is not performed by default for internal plugins. External plugins don't have access to the whole VM API and have to go through an interface to do so, leading to an overhead.
+	\item \emph{Optimization potential:}
+	C cross-file inlining is not performed by default for internal plugins.
+External plugins don't have access to the whole VM API and have to go through an interface to do so, leading to an overhead, limiting the performance unless the string size is really high.
 	\item \emph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code.
@@
 \hline
-  \multirow{2}{*}{\legSlangPlug} &\GCCLinux& 0.413 $\pm$ 0.0005 & 0.746 $\pm$ 0.0008 & 1.804 $\pm$ 0.0005 & 12.05 $\pm$ 0.0088 & 23.49 $\pm$ 0.0135 \\
+
+  \multirow{2}{*}{\legSlangPlug} &\GCCLinux& 0.413 $\pm$ 0.0005 & 0.746 $\pm$ 0.0008 & 1.804 $\pm$ 0.0005 & 12.05 $\pm$ 0.0088 & 25.36 $\pm$ 0.1742 \\
   						&\LLVMMac& 0.339 $\pm$ 0.0007 & 0.624 $\pm$ 0.0012 & 1.581 $\pm$ 0.0092 & 11.41 $\pm$ 0.0474  & 23.21 $\pm$ 0.1789 \\
    \hline
-  \multirow{2}{*}{\legSlang} &\GCCLinux& 1.287 $\pm$ 0.0048 & 2.257 $\pm$ 0.0116 & 4.958 $\pm$ 0.0111 & 17.24 $\pm$ 0.0515 & 22.54 $\pm$ 0.0775\\
+  \multirow{2}{*}{\legSlang} &\GCCLinux& 1.287 $\pm$ 0.0048 & 2.257 $\pm$ 0.0116 & 4.958 $\pm$ 0.0111 & 17.24 $\pm$ 0.0515 & 25.43 $\pm$ 0.1621\\
  					 &\LLVMMac& 0.932 $\pm$ 0.0015 & 1.690 $\pm$ 0.0030 & 3.241 $\pm$ 0.0067 & 10.15 $\pm$ 0.0321  & 12.94 $\pm$ 0.0177 \\
    \hline
-  \multirow{2}{*}{\legRTL}  &\GCCLinux& 3.343 $\pm$ 0.0211 & 5.171 $\pm$ 0.0265  & 8.715 $\pm$ 0.0221 & 17.66 $\pm$ 0.0481 & 24.93 $\pm$ 0.0309\\
- 				 	&\LLVMMac & 3.255 $\pm$ 0.0079 & 5.216 $\pm$ 0.0153 & 8.971 $\pm$ 0.0359 & 17.64 $\pm$ 0.0600  & 24.88 $\pm$ 0.1983 \\
+  \multirow{2}{*}{\legRTL}  &\GCCLinux& 3.343 $\pm$ 0.0211 & 5.171 $\pm$ 0.0265  & 8.715 $\pm$ 0.0221 & 17.66 $\pm$ 0.0481 & 26.76 $\pm$ 0.1861\\
+ 				 	&\LLVMMac & 3.255 $\pm$ 0.0079 & 5.216 $\pm$ 0.0153 & 8.971 $\pm$ 0.0359 & 17.64 $\pm$ 0.06  & 24.88 $\pm$ 0.1983 \\
   \hline
@@
 
-These results can be explained based on the last 3 criteria detailed in Section~\ref{sec:implem_cmp}:
+These results can be explained based on the last 3 criteria detailed in Section~\ref{sec:implem_cmp}.
 
@@
 As mentioned in Section~\ref{sec:VMexec}, all implementations written in Slang come with an overhead due to the switch between the C and Smalltalk stacks.
-This overhead is clearly visible in the bench results in Figure~\ref{fig:bench} for small strings (size 1 or less): the Plugin implementation is between 57,8 (\GCCLinux) to 61,4\% (\LLVMMac) slower than the baseline's for empty strings.
-The problem is even more important for the plugin version, since the beginning of the primitive code, before the main comparison loop, is slower due to the calls to the main VM code.
-As the string size goes larger, the overhead is less significant because the execution time becomes dominated by the time spent in the main comparison loop.
+This overhead is clearly visible in the bench results in Figure~\ref{fig:bench} for small strings (size 1 or less): the Plugin implementation is between 58,7 (\GCCLinux) to 66,1\% (\LLVMMac) slower than the baseline's for empty strings.The problem is even more important for the \legSlangPlug version, since the beginning of its code, before the main comparison loop, is slower due to the calls to the main VM code.
+As the string size goes larger, the overhead is less significant because it is absorbed by other performance gains.
 Regarding this criteria, \legRTL version is always the fastest as it is not impacted by the switch overhead.
 
-\paragraph{Optimization potential on large strings comparison.}
-The Plugin comes with an overhead due the use of an interface.
-Besides, cross-file inlining is not performed for plugins, like said in Section~\ref{sec:implem_cmp}.
-These are some of the factors explaining the performance gap between these versions and the \legSlang and \legRTL ones.
+\paragraph{Optimization potential.}
 When it comes to comparing large strings (100 characters and more), the execution time is mainly spent in the comparison loop (Figure~\ref{fig:Asm}).
-However, the switch overhead is still noticeable in the Slang implementation when strings of 100 characters are compared: Slang is 17.2 times faster than Baseline while Slang+JIT is 18,6 times faster than Baseline.
-This last observation only applies to the \GCCLinux results: the code generated by the LLVM compiler induces a longer execution time, as explained in the next paragraph.
-
-\paragraph{Control over generated code.} One can notice a rather striking performance difference between the 2 set-ups we have.In \GCCLinux, the performance between the \legSlang and \legRTL implementations is really close for long strings (1000 characters wide), with respectively 22.54 and 24.93 relative speed-up factors for a 1000-characters string.
+The stack switch overhead is still noticeable in the \legSlang implementation when strings of 100 characters are compared: in \GCCLinux, Slang is 17.2 times faster than Baseline while \legRTL is 18,6 times faster than Baseline.
+Due to the beginning of its code calling the VM code, the \legSlangPlug version is still around 0.68 times slower than the \legSlang and \legRTL version for a string of size 100 (Still in \GCCLinux).
+For strings of size 1000, the performance of all primitives is very similar, since they have almost the same comparison loop, except for the \legSlang version on \LLVMMac, as we are about to discuss in the next paragraph.
+
+\paragraph{Control over generated code.} One can notice a rather striking performance difference between the 2 set-ups we have.In \GCCLinux, the performance between the \legSlang and \legRTL implementations is really close for long strings (1000 characters wide), with respectively 25.24 and 26.76 relative speed-up factors for a 1000-characters string.
 Yet performance greatly differs when the \LLVMMac set-up is used, with respectively 12,94 and 24,88 relative speed-up factors.
@@
 VMs compiled with LLVM and GCC are around equally fast except in narrow cases like this one and LLVM is able to optimize correctly the \legSlangPlug version.
-LLVM likely fails at optimizing the \legSlang version likely due to the old C extensions we use in the interpreter code present in the same file to force variables into specific registers, which are better supported by GCC, historically the only compiler we supported.
+LLVM likely fails at optimizing the \legSlang version due to the old GNU C extensions we use in the interpreter code present in the same file to force variables into specific registers, which are better supported by GCC, the only compiler we supported back in the days.
+
 However, we can see that the \legRTL version has the most reliable performance since it does not depend on what optimization the C compiler is able to perform or not.
-This is even more relevant in less common back-ends not shown in this evaluation, such as the MIPSEL back-end we support, where C compilers are usually less clever at performing back-end specific optimizations since less engineering time has been invested into such optimizations.
-
-One can also notive a significant gap between the Plugin and Slang implementations in \LLVMMac.
-This can also be explained by a difference in optimizations.
-As seen in Figure \ref{fig:Asm}, the comparison loop is indeed optimized in the Plugin version and is similar to the JIT implementation.
-
-In this section, we compared the performance of the different implementations against a baseline (\emph{i.e., }{pure Smalltalk code}) implementation.
+This is even more relevant in less common back-ends not shown in this evaluation, such as the MIPSEL back-end we support.
+On such uncommon processors, C compilers are usually less clever at performing back-end specific optimizations since less engineering time has been invested into it.
+Cog's RTL is closer to the native instructions and can therefore often generate better code in this case.
+
+\subsection*{Conclusion}In this section, we compared the performance of the different implementations against a baseline (\emph{i.e., }{pure Smalltalk code}) implementation.
 The results showed the \legRTL implementation was always the fastest.

### 2018-05-18 paper.tex
reducing a bit the paper length

@@
  In this section we compare the four different implementations according to the following criteria:
- \begin{itemize}
- 	\item \emph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (Plugin) or requires to recompile the full VM (Yes).
-%\sd{why not ST, Plugin, VM?} %\sk{Yes and No answers to the "recompilation needed" question, so maybe we should keep it that way}
-	\item \emph{Language familiar to developers:} Smalltalk developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code.
+% \begin{itemize}
+% 	\item \emph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (Plugin) or requires to recompile the full VM (Yes).
+%	\item \emph{Language familiar to developers:} Smalltalk developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code.
 Only Cog's JIT implementors are familiar with its RTL, which is a subset of the VM developers.
 We use All if all communities are familiar with the language, VM if all VM developers are and VM-JIT if only a subset of VM developers are.
-	\item \emph{Number of lines of code:} We measured the number of lines of code of each implementation, which arguably can be used to evaluate the complexity level of each implementation.
-	\item \emph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to be executed if the primitive fails.
+%	\item \emph{Number of lines of code:} We measured the number of lines of code of each implementation, which arguably can be used to evaluate the complexity level of each implementation.
+%	\item \emph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to be executed if the primitive fails.
 This code is maintained by the client language implementors, and not by the VM implementors.
@@
 We count all the Slang implementations as well as the ones on Cog's RTL.
-	\item \emph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
-	\item \emph{Optimization potential:}
+%	\item \emph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
+%	\item \emph{Optimization potential:}
+%	C cross-file inlining is not performed by default for internal plugins.
+External plugins don't have access to the whole VM API and have to go through an interface to do so, leading to an overhead, limiting the performance unless the string size is really high.
+%	\item \emph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code.
+The generated code can therefore differ according the compilers and the optimizations they performed.
+On the contrary, the JIT generates the same native code whichever C compiler is used.
+%\end{itemize}
+
+%Vspce to fit all on page.
+\vspace{-0.1cm}
+	\paragraph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (Plugin) or requires to recompile the full VM (Yes).
+\vspace{-0.1cm}
+	\paragraph{Language familiar to developers:} Smalltalk developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code.
+Only Cog's JIT implementors are familiar with its RTL, which is a subset of the VM developers.
+We use All if all communities are familiar with the language, VM if all VM developers are and VM-JIT if only a subset of VM developers are.
+\vspace{-0.1cm}
+	\paragraph{Number of lines of code:} We measured the number of lines of code of each implementation, which arguably can be used to evaluate the complexity level of each implementation.
+\vspace{-0.1cm}
+	\paragraph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to be executed if the primitive fails.
+This code is maintained by the client language implementors, and not by the VM implementors.
+Hence we do not count the fall-back code in the number of implementation to maintain, nor the pure Smalltalk implementation.
+However, as a VM implementor, we count all the Slang implementations as well as the Cog's RTL one.
+\vspace{-0.1cm}
+	\paragraph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
+\vspace{-0.1cm}
+	\paragraph{Optimization potential:}
 	C cross-file inlining is not performed by default for internal plugins.
 External plugins don't have access to the whole VM API and have to go through an interface to do so, leading to an overhead, limiting the performance unless the string size is really high.
-	\item \emph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code.
+\vspace{-0.1cm}
+	\paragraph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code.
 The generated code can therefore differ according the compilers and the optimizations they performed.
 On the contrary, the JIT generates the same native code whichever C compiler is used.
-\end{itemize}
-
-One or another implementation could be preferred depending on the intended aim: the main trade-off is between  engineering time and  performance.
-Table \ref{tab:cmp_implem} summarizes the main pros and cons related to each implementation: the engineering time tends to increase when more low-level languages are used, such as Cog's RTL for instance.
-Also, a VM recompilation affects the overall complexity of the development process, but the resulting code might be faster to run.
+\vspace{0.2cm}
+
+Table \ref{tab:cmp_implem} summarizes the main pros and cons related to each implementation: the engineering time tends to increase when more low-level languages are used.
+VM recompilation slows slightly the development process, but the resulting code is usually faster to run.
 Likewise, the use of plugins brings modularity at the cost of some performance loss.

### 2018-05-18 paper.tex
update asm comp figures, nr of loc, some typos

@@
    \hline
-  \legSlang   & Yes  & VM & 45  & 1  & Yes & High & Low \\
+  \legSlang   & Yes  & VM & 35  & 1  & Yes & High & Low \\
    \hline
@@
 
-\paragraph{Control over generated code.} One can notice a rather striking performance difference between the 2 set-ups we have.In \GCCLinux, the performance between the \legSlang and \legRTL implementations is really close for long strings (1000 characters wide), with respectively 25.24 and 26.76 relative speed-up factors for a 1000-characters string.
+\paragraph{Control over generated code.} One can notice a rather striking performance difference between the 2 set-ups we have.
+In \GCCLinux, the performance between the \legSlang and \legRTL implementations is really close for long strings (1000 characters wide), with respectively 25.24 and 26.76 relative speed-up factors for a 1000-characters string.
 Yet performance greatly differs when the \LLVMMac set-up is used, with respectively 12,94 and 24,88 relative speed-up factors.

### 2018-05-18 paper.tex
fix typos

@@
 Slang is compiled to C and then to native code through standard C compilers.
-The execution engine (the memory manager, the interpreter and the baseline JIT) are entirely written in Slang.
+The execution engine (the memory manager, the interpreter and the baseline JIT) is entirely written in Slang.
 The two main purposes of using Slang code over plain C are:
@@
 Primitives redefined in such way can be entirely or partially redefined.
-In the later case, only the common cases are re-implemented on top of the JIT back-end.
+In the latter case, only the common cases are re-implemented on top of the JIT back-end.
 If an unimplemented case happens at runtime, the generated native code calls the C primitive instead, adding overhead only in uncommon cases.
@@
 If no order is specified, the VM assumes the byte ordering (ASCII order) is the order to compare against, and compares directly the bytes of the ByteStrings.
-If an order is specified, for example a case insensitive order (the entries for the 41 and 61 in the order byte array, respectively characters A and a in the ASCII table, both answer the same value), the primitive compare the bytes through the indirection order.
+If an order is specified, for example a case insensitive order (the entries for the 41 and 61 in the order byte array, respectively characters A and a in the ASCII table, both answer the same value), the primitive compares the bytes through the indirection order.
 Having the order optional allows to have quicker string comparison in the common case where the ASCII order is used.

### 2018-06-26 paper.tex
fix typos from review 1

@@
 High-level object-oriented programming languages are often implemented on top of a virtual machine (VM).
-As VMs has become more popular, different techniques have been set up using Just-In-Time (JIT) compilation to improve the runtime overall performance: method JITs \cite{ursPHD} which usually compile methods to native code and tracing JITs \cite{Dynamo, PyPyTracing} which usually compile linear traces of execution into native code.
+As VMs have become more popular, different techniques have been set up using Just-In-Time (JIT) compilation to improve the overall runtime performance: method JITs \cite{ursPHD} which usually compile methods to native code and tracing JITs \cite{Dynamo, PyPyTracing} which usually compile linear traces of execution into native code.
 
@@
 
-We implemented a first version of the optional primitive in Pharo/Smalltalk.
+We implemented a first version of the optional primitive in Pharo.
 We then implemented two versions compiling through C compilers to native code, using different parts of our VM infrastructure.
@@
 There is a performance overhead for the plugins compared to normal VM code.
-For example, accesses to object headers are memory representation dependent and therefore need to be done indirectly through calls to main VM executable, since the plugin cannot know ahead of time which memory representation is used.
+For example, accesses to object headers are dependent on the memory representation and therefore need to be done indirectly through calls to main VM executable, since the plugin cannot know ahead of time which memory representation is used.
 Such calls cannot be inlined by the C compiler\footnote{In the case of specific internal plugins, C compiler linking time optimizations can be enabled to limit this overhead.}.
@@
 As depicted in Figure~\ref{fig:Asm}, \GCCLinux generates instructions that are very close to the ones generated by the JIT, hence the execution time is similar.
-On the contrary, \LLVMMac generates more instructions: it fails at keeping all the operands live across the comparison loops, leading to two extra memory reads per iteration (bolded in the figure).
+On the contrary, \LLVMMac generates more instructions: it fails at keeping all the operands live across the comparison loops, leading to two extra memory reads per iteration (bolded in Figure \ref{fig:Asm}).
 
@@
 The philosophy of Self and Strongtalk, both high performance VMs, were to improve the performance of the client language with a mature optimizing JIT featuring advanced optimizations such as speculative optimizations \cite{Strongtalk, ursPHD}.
-In this context, the need of primitive is reduced, since the performance gain from using a primitive over the client language is low or inexistant.
+In this context, the need of primitives is reduced, since the performance gain from using a primitive over the client language is low or inexistant.
 More recently, Felgentreff \emph{et al.} \cite{NoPrimTracing} analysed the speed difference between primitives implemented in C, RPython, and Smalltalk.
@@
 To avoid paying the stack switch cost between the client Stack and the C stack, some VMs include an interpreter written on top of an abstract assembly.
-It allows the interpreter to have same register convention than the native code generated by the JIT, removing the stack switch overhead.
+It allows the interpreter to have the same register convention as the native code generated by the JIT, removing the stack switch overhead.
 For example, the V8 team recently wrote an interpreter named \emph{Ignition}.
@@
 
-In this paper we compared four different implementations of the string primitives on top of the Cog VM.
+In this paper we compared four different implementations of the string comparison primitive on top of the Cog VM.
 We showed the performance difference between each implementation for different string sizes.

### 2018-06-27 paper.tex
add a bunch of todos (mainly finding refs)

@@
 %% Title information
-\title[Assessing primitives performance on multi-stage execution]{Assessing primitives performance\\ on multi-stage execution}
+\title[Assessing primitives performance on multi-stage execution]{Assessing primitives performance\\ on multi-stage execution \todo{modify maybe}}
 
@@
 Some other programming languages, such as Javascript, prefer to use the term built-in instead of primitive.}, which are performed directly by the interpreter rather than by evaluating expressions in a method.
-We distinguish two kinds of primitives:
+We distinguish two kinds of primitives: \todo{motivate the distinction.
+See the algorithmic primitives paper.
+Is there a reference stating the different types of primitives ?
+}
 \begin{itemize}
@@
 \begin{itemize}
-	\item High performance relies on a JIT with speculative optimizations, which is hard to implement, maintain and evolve.
-	\item There is a big performance gap between the baseline performance and the peak performance.
+	\item High performance relies on a JIT with speculative optimizations, which is hard to implement, maintain and evolve. \todo{a ref would be nice}
+	\item There is a big performance gap between the baseline performance and the peak performance. \todo{a ref would be nice}
 	\item Even mature JITs fail to optimize some narrow cases, where the performance is drastically slower \\~\cite{VMHotCold}.
@@
 The encoding usually follows the extended ASCII standard.
-WideStrings encode each character in 32 bits.
+\todo{we should not use this term} WideStrings encode each character in 32 bits.
 The encoding usually follows Unicode specifications.
@@
 
+\todo{mention baseline here}
 The \legSlangPlug, \legSlang and \legRTL implementations of this string comparison primitive have been implemented in the Cog VM.
@@
 The two compared strings are equal as this situation corresponds to the worst scenario for most implementations.
-%There is one less branch in the two plugin implementations in this case.
+\todo{explain how the relative speed up is computed} %There is one less branch in the two plugin implementations in this case.
 
@@
 In the case of large strings, the time spent executing the primitive is dominated by the main comparison loop.
-Figure~\ref{fig:bench} shows the results in terms of relative speed-up towards the baseline implementation, respectively for \GCCLinux and \LLVMMac.
+Figure~\ref{fig:bench} and Table \ref{tab:detail_bench} show the results in terms of relative speed-up towards the baseline implementation, respectively for \GCCLinux and \LLVMMac.
 
@@
 
-This difference is mainly due to the optimizations performed or not by both the compilers for the \legSlang version.
+This difference is mainly due to the optimizations performed or not by the GCC and Clang compilers for the \legSlang version.
 As depicted in Figure~\ref{fig:Asm}, \GCCLinux generates instructions that are very close to the ones generated by the JIT, hence the execution time is similar.
@@
 \section{Discussion and related work}
+\todo{primitives in other system. (generalize more after Rpython. }
 
@@
 They showed that with the RPython toolchain framework, the speed difference between all three implementations is not that big, hence they could choose to implement some optional primitives in the client language over C or RPython without performance drop to reduce the number of primitives to maintain.
+\todo{emphasize the difference more}
 

### 2018-06-28 paper.tex
comply wih template workshop. Will have to modify reference as well

@@
-%%%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
-\documentclass[10pt, sigplan]{acmart}
-%%\settopmatter{printfolios=true,printccs=false,printacmref=false}
-%% For double-blind review submission, w/ CCS and ACM Reference
-%\documentclass[sigplan,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true}
-%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
-%\documentclass[sigplan,10pt,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
-%% For single-blind review submission, w/ CCS and ACM Reference
-%\documentclass[sigplan,10pt,review]{acmart}\settopmatter{printfolios=true}
-%% For final camera-ready submission, w/ required CCS and ACM Reference
-%\documentclass[sigplan,10pt]{acmart}\settopmatter{}
-
-%% Conference information
-%% Supplied to authors by publisher for camera-ready submission;
-%% use defaults for review submission.
-%\acmConference[PL'17]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
-%\acmYear{2017}
-%\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
-%\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
-%\startPage{1}
-
-%% Copyright information
-%% Supplied to authors (based on authors' rights management selection;
-%% see authors.acm.org) by publisher for camera-ready submission;
-%% use 'none' for review submission.
-\setcopyright{none}
-%\setcopyright{acmcopyright}
-%\setcopyright{acmlicensed}
-%\setcopyright{rightsretained}
-%\copyrightyear{2017}           %% If different from \acmYear
-
-%% Bibliography style
-
-%\bibliographystyle{ACM-Reference-Format}
-
-%% Citation style
-%\citestyle{acmauthoryear}  %% For author/year citations
-\citestyle{acmnumeric}     %% For numeric citations
-%\setcitestyle{nosort}      %% With 'acmnumeric', to disable automatic
-                            %% sorting of references within a single citation;
-                            %% e.g., \cite{Smith99,Carpenter05,Baker12}
-                            %% rendered as [14,5,2] rather than [2,5,14].
-%\setcitesyle{nocompress}   %% With 'acmnumeric', to disable automatic
-                            %% compression of sequential references within a
-                            %% single citation;
-                            %% e.g., \cite{Baker12,Baker14,Baker16}
-                            %% rendered as [2,3,4] rather than [2-4].
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Note: Authors migrating a paper from traditional SIGPLAN
-%% proceedings format to PACMPL format must update the
-%% '\documentclass' and topmatter commands above; see
-%% 'acmart-pacmpl-template.tex'.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%-----------------------------------------------------------------------------
+%
+%               Template for sigplanconf LaTeX Class
+%
+% Name:         sigplanconf-template.tex
+%
+% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
+%               file for SIGPLAN conference proceedings.
+%
+% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
+%               sigplanconf-guide.pdf
+%
+% Author:       Paul C. Anagnostopoulos
+%               Windfall Software
+%               978 371-2316
+%               paul@windfall.com
+%
+% Created:      15 February 2005
+%
+%-----------------------------------------------------------------------------
+
+\documentclass[preprint]{sigplanconf}
+
+% The following \documentclass options may be useful:
+
+% preprint       Remove this option only once the paper is in final form.
+%  9pt           Set paper in  9-point type (instead of default 10-point)
+% 11pt           Set paper in 11-point type (instead of default 10-point).
+% numbers        Produce numeric citations with natbib (instead of default author/year).
+% authorversion  Prepare an author version, with appropriate copyright-space text.
+
+\usepackage{amsmath}
 
@@
 
-\begin{document}
-
 %%%
@@
 
-%% Title information
-\title[Assessing primitives performance on multi-stage execution]{Assessing primitives performance\\ on multi-stage execution \todo{modify maybe}}
-
-%% Author with single affiliation.
-\author{Sophie Kaleba}
-                                        %% can be repeated if necessary
-\affiliation{
-  %\position{Position1}
-  \department{RMoD}              %% \department is recommended
-  \institution{Inria}            %% \institution is required
- % \streetaddress{Street1 Address1}
-  \city{Lille}
-  %\state{France}
-  %\postcode{Post-Code1}
-  \country{France}                    %% \country is recommended
-}
-\email{sophie.kaleba@etudiant.univ-lille1.fr}          %% \email is recommended
-
-%% Author with two affiliations and emails.
-\author{Cl\'ement B\'era}
-\affiliation{
-  % \position{}
-	\department{Software Languages Lab}              %% \department is recommended
-	\institution{Vrije Universiteit Brussel}            %% \institution is required
-	\city{Brussel}
-  % \state{}
-  % \postcode{}
-	\country{Belgium}                    %% \country is recommended
-}
-\email{clement.bera@vub.ac.be}          %% \email is recommended
-
-%% Author with two affiliations and emails.
-\author{St\'ephane Ducasse}
-\affiliation{
- % \position{Position2b}
-  \department{RMoD}             %% \department is recommended
-  \institution{Inria}           %% \institution is required
-  %\streetaddress{Street3b Address2b}
-  \city{Lille}
-  %\state{State2b}
-  %\postcode{Post-Code2b}
-  \country{France}                   %% \country is recommended
-}
-\email{stephane.ducasse@inria.fr}         %% \email is recommended
-
-%% Abstract
-%% Note: \begin{abstract}...\end{abstract} environment must come
-%% before \maketitle command
-\begin{abstract}
+\newcommand{\cL}{{\cal L}}
+
+\begin{document}
+
+\special{papersize=8.5in,11in}
+\setlength{\pdfpageheight}{\paperheight}
+\setlength{\pdfpagewidth}{\paperwidth}
+
+\conferenceinfo{CONF'yy}{Month d--d, 20yy, City, ST, Country}
+\copyrightyear{20yy}
+\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}\reprintprice{\$15.00}
+\copyrightdoi{nnnnnnn.nnnnnnn}
+
+% For compatibility with auto-generated ACM eRights management
+% instructions, the following alternate commands are also supported.
+%\CopyrightYear{2016}
+%\conferenceinfo{CONF'yy,}{Month d--d, 20yy, City, ST, Country}
+%\isbn{978-1-nnnn-nnnn-n/yy/mm}\acmPrice{\$15.00}
+%\doi{http://dx.doi.org/10.1145/nnnnnnn.nnnnnnn}
+
+% Uncomment the publication rights used.
+%\setcopyright{acmcopyright}
+\setcopyright{acmlicensed}  % default
+%\setcopyright{rightsretained}
+
+%\titlebanner{banner above paper title}        % These are ignored unless
+\preprintfooter{Assessing primitives performance on multi-stage execution}   % 'preprint' option specified.
+
+\title{Assessing primitives performance\\ on multi-stage execution \todo{modify maybe}}
+%\subtitle{Subtitle Text, if any\titlenote{with optional subtitle note}}
+
+\authorinfo{Sophie Kaleba}
+           {RMoD - Inria Lille}
+           %{Lille, France}
+           {sophie.kaleba@etudiant.univ-lille1.fr}
+
+\authorinfo{Cl\'ement B\'era}
+           {Software Languages Lab - Vrije Universiteit Brussel}
+           {clement.bera@vub.ac.be}
+
+\authorinfo{{St\'ephane Ducasse}}
+           {RMoD - Inria Lille}
+           {stepahne.ducasse@inria.fr}
+
+\maketitle
 
@@
 
-%% 2012 ACM Computing Classification System (CSS) concepts
-%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
+% 2012 ACM Computing Classification System (CSS) concepts
+% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
 %\begin{CCSXML}
@@
 %<concept>
-%<concept_id>10003456.10003457.10003521.10003525</concept_id>
-%<concept_desc>Social and professional topics~History of programming languages</concept_desc>
+%<concept_id>10003752.10010124.10010138.10010143</concept_id>
+%<concept_desc>Theory of computation~Program analysis</concept_desc>
 %<concept_significance>300</concept_significance>
@@
 %\end{CCSXML}
-
+%
 %\ccsdesc[500]{Software and its engineering~General programming languages}
-%\ccsdesc[300]{Social and professional topics~History of programming languages}
-%% End of generated code
-
-%% Keywords
-%% comma separated list
-\keywords{Just-in-Time compiler, Primitive, Virtual machine, Managed runtime}  %% \keywords are mandatory in final camera-ready submission
-
-%% \maketitle
-%% Note: \maketitle command must come after title commands, author
-%% commands, abstract environment, Computing Classification System
-%% environment and commands, and keywords command.
-\maketitle
+%\ccsdesc[300]{Theory of computation~Program analysis}
+% end generated code
+
+% Legacy 1998 ACM Computing Classification System categories are also
+% supported, but not recommended.
+%\category{CR-number}{subcategory}{third-level}[fourth-level]
+%\category{D.3.0}{Programming Languages}{General}
+%\category{F.3.2}{Logics and Meanings of Programs}{Semantics of Programming Languages}[Program analysis]
+
+\keywords
+Just-in-Time compiler, Primitive, Virtual machine, Managed runtime
 
@@
 
-%% Acknowledgments
-%%\begin{acks}                            %% acks environment is optional
-                                        %% contents suppressed with 'anonymous'
-  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
-  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
-  %% acknowledge financial support and will be used by metadata
-  %% extraction tools.
-%  This material is based upon work supported by the
-%  \grantsponsor{GS100000001}{National Science
-%    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
-%  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
-%  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
-%  conclusions or recommendations expressed in this material are those
-%  of the author and do not necessarily reflect the views of the
-%  National Science Foundation.
-%\end{acks}
-
-%% Bibliography
+%\appendix
+%\section{Appendix Title}
+%
+%This is the text of the appendix, if you need one.
+
+\acks
+\todo{add acknowledgments}
+Acknowledgments, if needed.
+
+% The 'abbrvnat' bibliography style is recommended.
 \bibliographystyle{alpha}
 \bibliography{sista,rmod,others}
-
-%\vspace{10cm} %Adapt to make vfill work
-
-%% Appendix
-%\null \vfill\null %switch column
-%
-%\appendix
-%\section{Smart Syntax code}
-%\label{SmartSyntaxAppendix}
-%
-%Source code of the smart syntax method, which is used to generate the C code for the plugin but also as Smalltalk fall-back code.
-%
-%\begin{code}
-%compare: string1 with: string2 collated: order
-%	"Return 1, 2 or 3, if string1 is <, =, or > string2, with the collating order of characters given by the order array."
-%
-%	| len1 len2 c1 c2 |
-%	<primitive: 'primitiveCompareString' module: 'MiscPrimitivePlugin'>
-%	<var: #string1 declareC: 'unsigned char *string1'>
-%	<var: #string2 declareC: 'unsigned char *string2'>
-%	<var: #order declareC: 'unsigned char *order'>
-%
-%	len1 := string1 size.
-%	len2 := string2 size.
-%	1 to: (len1 min: len2) do:
-%		[:i |
-%		c1 := order at: (string1 basicAt: i) + 1.
-%		c2 := order at: (string2 basicAt: i) + 1.
-%		c1 = c2 ifFalse: 
-%			[c1 < c2 ifTrue: [^ 1] ifFalse: [^ 3]]].
-%	len1 = len2 ifTrue: [^ 2].
-%	len1 < len2 ifTrue: [^ 1] ifFalse: [^ 3].
-%\end{code}
-%
-%\vfill\null %switch column
-%\section{Slang Plugin code}
-%\label{PluginAppendix}
-%
-%Source code of the plugin method generating the C code for the Slang plugin code. Most operations needs to go through the interpreter proxy. 
-%
-%\begin{code}
-%primitiveCompareString
-%	"ByteString (class) compare: string1 with: string2 collated: order"
-%	<export: true>
-%	| len1 len2 order string1 string2 |
-%
-%	<var: 'order' type: #'unsigned char *'>
-%	<var: 'string1' type: #'unsigned char *'>
-%	<var: 'string2' type: #'unsigned char *'>
-%	((interpreterProxy isBytes: (interpreterProxy stackValue: 0))
-%	and: [(interpreterProxy isBytes: (interpreterProxy stackValue: 1))
-%	and: [interpreterProxy isBytes: (interpreterProxy stackValue: 2)]]) ifFalse:
-%		[^interpreterProxy primitiveFailFor: PrimErrBadArgument].
-%	string1 := interpreterProxy firstIndexableField: (interpreterProxy stackValue: 2).
-%	string2 := interpreterProxy firstIndexableField: (interpreterProxy stackValue: 1).
-%	order := interpreterProxy firstIndexableField: (interpreterProxy stackValue: 0).
-%	len1 := interpreterProxy sizeOfSTArrayFromCPrimitive: string1.
-%	len2 := interpreterProxy sizeOfSTArrayFromCPrimitive: string2.
-%	(interpreterProxy sizeOfSTArrayFromCPrimitive: order) < 256 ifTrue:
-%		[^interpreterProxy primitiveFailFor: PrimErrBadArgument].
-%	0 to: (len1 min: len2) - 1 do: 
-%		[ :i | | c1 c2 |
-%		c1 := order at: (string1 at: i).
-%		c2 := order at: (string2 at: i).
-%		c1 = c2 ifFalse:
-%			[^interpreterProxy methodReturnInteger: (c1 < c2 ifTrue: [1] ifFalse: [3])]].
-%	interpreterProxy methodReturnInteger:
-%		(len1 = len2 ifTrue: [2] ifFalse: [len1 < len2 ifTrue: [1] ifFalse: [3]])
-%\end{code}
-%
-%\vfill\null %switch column
-%\section{Slang code}
-%\label{SlangAppendix}
-%
-%Source code of the slang method generating the C code for the Slang string comparison primitive.
-We split it in two methods, and forced inlining at Slang-to-C compilation time to remove the closures while avoiding code duplication.
-%
-%\begin{code}
-%primitiveCompareWith
-%	"<string1> primitiveCompareWith: string2 [collated: order] "
-%	<export: true>	
-%	| string1 string2 order strLength1 strLength2 result |
-%	"1 - fetch the parameters from the stack"	
-%	argumentCount = 1 ifFalse:
-%		[argumentCount ~= 2 ifTrue:
-%			[^self primitiveFailFor: PrimErrBadNumArgs].
-%			 order := self stackTop.
-%			 ((objectMemory isBytes: order)
-%			 and: [(objectMemory numBytesOfBytes: order) = 256]) ifFalse:
-%				[^self primitiveFailFor: PrimErrBadArgument]].
-%	string1 := self stackValue: argumentCount.
-%	string2 := self stackValue: argumentCount - 1. 		
-%	"2 - check their types - all parameters are ByteObject"
-%	((objectMemory isBytes: string1)
-%	 and: [objectMemory isBytes: string2]) ifFalse: 
-%		[^self primitiveFailFor: PrimErrBadArgument].
-%	"3 - compare the strings"	
-%	strLength1 := objectMemory numBytesOfBytes: string1.
-%	strLength2 := objectMemory numBytesOfBytes: string2.
-%	result := order 
-%		ifNil: [self rawCompare: string1 length: strLength1 with: string2 length: strLength2 accessBlock:
-%				[:str :index | objectMemory fetchByte: index ofObject: str ]]
-%		ifNotNil: 
-%			[self rawCompare: string1 length: strLength1 with: string2 length: strLength2 accessBlock:
-%				[:str :index | objectMemory fetchByte: (objectMemory fetchByte: index ofObject: str) + 1 ofObject: order ]].
-%
-%	self methodReturnInteger: result
-%
-%
-%rawCompare: string1 length: strLength1 with: string2 length: strLength2 accessBlock: accessBlock
-%	| c1 c2 min |
-%	<inline: true> "needs to be forced else slang does not inline it by default"
-%	min := strLength1 min: strLength2.
-%	0 to: min-1 do: 
-%		[:i | c1 := accessBlock value: string1 value: i.
-%			c2 := accessBlock value: string2 value: i.
-%			c1 = c2 ifFalse: [^c1 - c2]].
-%	^strLength1 - strLength2
-%\end{code}
-%
-%\vfill\null %switch column
-%\section{Cog's RTL code}
-%\label{RTLAppendix}
-%
-%Source code of the method generating the native code of the string comparison primitive on top of Cog's RTL.
-Method names starting with uppercase (in blue) are directly generating an abstract assembly instruction.
-Other methods called, starting with gen, generate a sequence of abstract assembly instructions.
-We duplicated the code checking if each operand is a ByteString and extract its size to generate better native code than doing the two operations separatedly.
-%
-%\begin{code}
-%genPrimitiveStringCompareWith
-%	| instr jump jumpAbove jumpIncorrectFormat1 jumpIncorrectFormat2 jumpIncorrectFormat3 jumpIncorrectFormat4 jumpMidFailure jumpSuccess minSizeReg string1CharOrByteSizeReg string2CharOrByteSizeReg string1Reg string2Reg |
-%	<var: #jumpIncorrectFormat1 type: #'AbstractInstruction *'>
-%	<var: #jumpIncorrectFormat2 type: #'AbstractInstruction *'>
-%	<var: #jumpIncorrectFormat3 type: #'AbstractInstruction *'>
-%	<var: #jumpIncorrectFormat4 type: #'AbstractInstruction *'>
-%	<var: #jumpAbove type: #'AbstractInstruction *'>
-%	<var: #jumpSuccess type: #'AbstractInstruction *'>
-%	<var: #jump type: #'AbstractInstruction *'>
-%	<var: #jumpMidFailure type: #'AbstractInstruction *'>
-%	"I redefine those name to ease program comprehension"
-%	string1Reg := ReceiverResultReg.
-%	string2Reg := Arg0Reg.
-%	string1CharOrByteSizeReg := Arg1Reg.
-%	string2CharOrByteSizeReg := ClassReg.
-%	minSizeReg := SendNumArgsReg.
-%	"Load arguments in reg"
-%	cogit genLoadArgAtDepth: 0 into: string2Reg.
-%	"checks if string1 is a byteobject and get its size in bytes"
-%	self genGetFormatOf: string1Reg into: TempReg.
-%	cogit CmpCq: objectMemory firstByteFormat R: TempReg.
-%	jumpIncorrectFormat1 := cogit JumpLess: 0.
-%	cogit CmpCq: objectMemory firstCompiledMethodFormat R: TempReg.
-%	jumpIncorrectFormat2 := cogit JumpAboveOrEqual: 0.	
-%	self genGetNumSlotsOf: string1Reg into: string1CharOrByteSizeReg.
-%	(cogit LogicalShiftLeftCq: objectMemory shiftForWord R: string1CharOrByteSizeReg). 
-%	cogit AndCq: objectMemory wordSize - 1 R: TempReg R: TempReg. 
-%	cogit SubR: TempReg R: string1CharOrByteSizeReg. 
-%	"checks if string2 is a byteobject and get its size in bytes"
-%	self genGetFormatOf: string2Reg into: TempReg.
-%	cogit CmpCq: objectMemory firstByteFormat R: TempReg.
-%	jumpIncorrectFormat3 := cogit JumpLess: 0.
-%	cogit CmpCq: objectMemory firstCompiledMethodFormat R: TempReg.
-%	jumpIncorrectFormat4 := cogit JumpAboveOrEqual: 0.	
-%	self genGetNumSlotsOf: string2Reg into: string2CharOrByteSizeReg.
-%	(cogit LogicalShiftLeftCq: objectMemory shiftForWord R: string2CharOrByteSizeReg).
-%	cogit AndCq: objectMemory wordSize - 1 R: TempReg R: TempReg.
-%	cogit SubR: TempReg R: string2CharOrByteSizeReg.
-%	\end{code}
-%	\vfill\null %switch column
-%	\begin{code}
-%	"Type and number of arguments are correct"
-%	"Compute the min"	 
-%	cogit CmpR: string1CharOrByteSizeReg R: string2CharOrByteSizeReg.
-%	jumpAbove := cogit JumpBelow: 0. 
-%	cogit MoveR: string1CharOrByteSizeReg R: minSizeReg. 
-%	jump := cogit Jump: 0. 
-%	jumpAbove jmpTarget: (cogit MoveR: string2CharOrByteSizeReg R: minSizeReg). 
-%	jump jmpTarget: (cogit CmpCq: 0 R: minSizeReg). 
-%	jumpSuccess := cogit JumpZero: 0. "if one of the string is empty, no need to go through the comparing loop"
-%	"Compare the bytes"
-%	cogit MoveCq: objectMemory baseHeaderSize  R: TempReg.
-%	cogit AddCq: objectMemory baseHeaderSize R: minSizeReg.
-%	instr := cogit MoveXbr: TempReg R: string1Reg R: string1CharOrByteSizeReg.
-%	cogit MoveXbr: TempReg R: string2Reg R: string2CharOrByteSizeReg.
-%	cogit SubR: string2CharOrByteSizeReg R: string1CharOrByteSizeReg. 
-%	jumpMidFailure := cogit JumpNonZero: 0. "the 2 compared characters are different, exit the loop"
-%	cogit AddCq: 1 R: TempReg.
-%	cogit CmpR: TempReg R: minSizeReg. 
-%	cogit JumpNonZero: instr.
-%	"all bytes from 1 to minSize are equal"
-%	self genGetNumBytesOf: string1Reg into: string1CharOrByteSizeReg.
-%	self genGetNumBytesOf: string2Reg into: string2CharOrByteSizeReg.
-%	jumpSuccess jmpTarget: (cogit SubR: string2CharOrByteSizeReg R: string1CharOrByteSizeReg).
-%	jumpMidFailure  jmpTarget: (cogit MoveR: string1CharOrByteSizeReg R: ReceiverResultReg).	
-%	self genConvertIntegerToSmallIntegerInReg: ReceiverResultReg.
-%	cogit genPrimReturn.
-%	jumpIncorrectFormat4 
-%		jmpTarget: (jumpIncorrectFormat3 
-%			jmpTarget: (jumpIncorrectFormat2 
-%				jmpTarget: (jumpIncorrectFormat1 jmpTarget: cogit Label))).
-%	^ CompletePrimitive}
-%\end{code}
+%\bibliographystyle{abbrvnat}
+%
+%% The bibliography should be embedded for final submission.
+%
+%\begin{thebibliography}{}
+%\softraggedright
+%
+%\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
+%P. Q. Smith, and X. Y. Jones. ...reference text...
+%
+%\end{thebibliography}
 

### 2018-06-28 paper.tex
comply with reference style
fix margin overrun

@@
 
-\paragraph{\GCCLinux.}The \GCCLinux evaluation was performed on an Asus Zenbook with Ubuntu 16.04 LTS, a 2.2 Ghz processor Intel Core 5, 8 Gb 1600MHz DDR3 of RAM.
+\paragraph{\GCCLinux.}The \GCCLinux evaluation was performed on an Asus Zenbook with Ubuntu 16.04.4 LTS, a 2.2 Ghz processor Intel Core 5, 8 Gb 1600MHz DDR3 of RAM.
 The Linux VM was compiled using GCC version 5.4.0.
@@
 
-\paragraph{Processor.}The following evaluation was performed on 32 bits Intel VMs (x86).
+\paragraph{}The following evaluation was performed on 32 bits Intel VMs (x86).
 
@@
 
-In some programming languages (Strongtalk, Self, RSqueak), the implementors tried to decrease the overall number of optional primitives by improving the client language performance, reducing the need for such primitives \cite{Strongtalk, ursPHD, NoPrimTracing}.
+In some programming languages (Strongtalk, Self, RSque-ak), the implementors tried to decrease the overall number of optional primitives by improving the client language performance, reducing the need for such primitives \cite{Strongtalk, ursPHD, NoPrimTracing}.
 In other work \cite{Ball86a, Char13a}, optional primitives can be written directly in a subset of the client language and compiled at runtime to native code, hence they can be modified in the client language without recompiling the VM.
@@
 % The 'abbrvnat' bibliography style is recommended.
-\bibliographystyle{alpha}
+%\bibliographystyle{alpha}
+\bibliographystyle{abbrvnat}
 \bibliography{sista,rmod,others}
-%\bibliographystyle{abbrvnat}
+
 %

### 2018-06-28 paper.tex
add another bunch of todo's and fix other stuff

@@
 \usepackage{multirow}
+\usepackage{soul}
 \usepgfplotslibrary{statistics}
@@
 The encoding usually follows the extended ASCII standard.
-\todo{we should not use this term} WideStrings encode each character in 32 bits.
+\todo{we should not use this term.
+Maybre prefer Latin-1 standard} WideStrings encode each character in 32 bits.
 The encoding usually follows Unicode specifications.
@@
 To sort arrays of strings, strings can be compared.
-They can be compared in the default order (ASCII or Unicode) or using different orders, for example, the case insensitive order.
+They can be compared in the default order (\st{ASCII or }Unicode) or using different orders, for example, the case insensitive order.
 
@@
 The two first operands are the two ByteStrings to compare and the third operand, optional, is the order table (a ByteArray of size 256 encoding the order of characters).
-If no order is specified, the VM assumes the byte ordering (ASCII order) is the order to compare against, and compares directly the bytes of the ByteStrings.
+If no order is specified, the VM assumes the byte ordering (\st{ASCII} \sk{Unicode} order) is the order to compare against, and compares directly the bytes of the ByteStrings.
 If an order is specified, for example a case insensitive order (the entries for the 41 and 61 in the order byte array, respectively characters A and a in the ASCII table, both answer the same value), the primitive compares the bytes through the indirection order.
-Having the order optional allows to have quicker string comparison in the common case where the ASCII order is used.
+Having the order optional allows to have quicker string comparison in the common case where the \st{ASCII} \sk{Unicode} order is used.
 The primitive answers a negative integer, zero, or a positive integer as the first parameter is less than, equal to, or greater than the second parameter.
 
-In the paper we will focus on the comparison of two strings with the default ASCII ordering and measure the performance only of this case.
-The goal of this work was to improve the performance of common string operations, typically string equality which internally uses the primitive, and the cases where the order is non ASCII are considered uncommon.
+In the paper we will focus on the comparison of two strings with the default \st{ASCII}\sk{Unicode} ordering and measure the performance only of this case.
+The goal of this work was to improve the performance of common string operations, typically string equality which internally uses the primitive, and the cases where the order is non- \sk{Unicode }ASCII are considered uncommon.
 
@@
 The implementation follows the new specification, hence two methods are available, baselineCompareWith: and baselineCompareWith:order:.
-Since the focus is only on ASCII ordered comparison, we show the code only of baselineCompareWith:.
+Since the focus is only on ASCII \sk{Unicode} ordered comparison, we show the code only of baselineCompareWith:.
 
@@
 \vspace{-0.1cm}
-	\paragraph{Number of lines of code:} We measured the number of lines of code of each implementation, which arguably can be used to evaluate the complexity level of each implementation.
+	\paragraph{Number of lines of code:} We measured the number of lines of code of each implementation, \todo {} \st{which arguably can be used to evaluate the complexity level of each implementation.}
 \vspace{-0.1cm}
@@
 \subsection{Methodology}
+\todo{so now I have to figure out how to actually bootstrap errors computed that way}
 

### 2018-07-01 paper.tex
fix some todos

@@
 
-\title{Assessing primitives performance\\ on multi-stage execution \todo{modify maybe}}
+\title{Assessing primitives performance\\ on multi-stage \st{execution} implementations}
 %\subtitle{Subtitle Text, if any\titlenote{with optional subtitle note}}
@@
 
+\begin{abstract}
 Virtual machines, besides the interpreter and just-in-time compiler optimization facilities, also include a set of primitive operations that the client language can use.
@@
 \subsection{Methodology}
-\todo{so now I have to figure out how to actually bootstrap errors computed that way}
 
@@
 Figure~\ref{fig:bench} and Table \ref{tab:detail_bench} show the results in terms of relative speed-up towards the baseline implementation, respectively for \GCCLinux and \LLVMMac.
+The relative speed-up is obtained by dividing the baseline results by the execution times. Both the figure and the table shows bootstrapped standard errors. \todo{add ref for bootstrap ?}
 
@@
 \todo{add acknowledgments}
+Eliot, Levente
 Acknowledgments, if needed.

### 2018-07-01 paper.tex
latest version before the bootstrap thingie

@@
 
-\title{Assessing primitives performance\\ on multi-stage \st{execution} implementations}
+\title{Assessing primitives performance\\ on multi-stage execution}
 %\subtitle{Subtitle Text, if any\titlenote{with optional subtitle note}}
@@
 Some other programming languages, such as Javascript, prefer to use the term built-in instead of primitive.}, which are performed directly by the interpreter rather than by evaluating expressions in a method.
-We distinguish two kinds of primitives: \todo{motivate the distinction.
-See the algorithmic primitives paper.
-Is there a reference stating the different types of primitives ?
-}
+We distinguish two kinds of primitives:
+%\todo{motivate the distinction. See the algorithmic primitives paper. Is there a reference stating the different types of primitives ? }
 \begin{itemize}
@@
 \begin{itemize}
-	\item High performance relies on a JIT with speculative optimizations, which is hard to implement, maintain and evolve. \todo{a ref would be nice}
-	\item There is a big performance gap between the baseline performance and the peak performance. \todo{a ref would be nice}
+	\item High performance relies on a JIT with speculative optimizations, which is hard to implement, maintain and evolve. %\todo{a ref would be nice}
+	\item There is a big performance gap between the baseline performance and the peak performance. %\todo{a ref would be nice}
 	\item Even mature JITs fail to optimize some narrow cases, where the performance is drastically slower \\~\cite{VMHotCold}.
@@
 ByteStrings encode each character of the string in a single byte.
-The encoding usually follows the extended ASCII standard.
-\todo{we should not use this term.
-Maybre prefer Latin-1 standard} WideStrings encode each character in 32 bits.
+The encoding usually follows the Latin-1 (ISO 8859-1) standard.
+WideStrings encode each character in 32 bits.
 The encoding usually follows Unicode specifications.
@@
 To sort arrays of strings, strings can be compared.
-They can be compared in the default order (\st{ASCII or }Unicode) or using different orders, for example, the case insensitive order.
+They can be compared in the default order (Latin-1) or using different orders, for example, the case insensitive order.
 
@@
 The two first operands are the two ByteStrings to compare and the third operand, optional, is the order table (a ByteArray of size 256 encoding the order of characters).
-If no order is specified, the VM assumes the byte ordering (\st{ASCII} \sk{Unicode} order) is the order to compare against, and compares directly the bytes of the ByteStrings.
+If no order is specified, the VM assumes the byte ordering (Latin-1 order) is the order to compare against, and compares directly the bytes of the ByteStrings.
 If an order is specified, for example a case insensitive order (the entries for the 41 and 61 in the order byte array, respectively characters A and a in the ASCII table, both answer the same value), the primitive compares the bytes through the indirection order.
-Having the order optional allows to have quicker string comparison in the common case where the \st{ASCII} \sk{Unicode} order is used.
+Having the order optional allows to have quicker string comparison in the common case where the Latin-1 order is used.
 The primitive answers a negative integer, zero, or a positive integer as the first parameter is less than, equal to, or greater than the second parameter.
 
-In the paper we will focus on the comparison of two strings with the default \st{ASCII}\sk{Unicode} ordering and measure the performance only of this case.
-The goal of this work was to improve the performance of common string operations, typically string equality which internally uses the primitive, and the cases where the order is non- \sk{Unicode }ASCII are considered uncommon.
+In the paper we will focus on the comparison of two strings with the default Latin-1 ordering and measure the performance only of this case.
+The goal of this work was to improve the performance of common string operations, typically string equality which internally uses the primitive, and the cases where the order is non-Latin-1 are considered uncommon.
 
@@
 The implementation follows the new specification, hence two methods are available, baselineCompareWith: and baselineCompareWith:order:.
-Since the focus is only on ASCII \sk{Unicode} ordered comparison, we show the code only of baselineCompareWith:.
+Since the focus is only on Latin-1 ordered comparison, we show the code only of baselineCompareWith:.
 
@@
 \vspace{-0.1cm}
-	\paragraph{Number of lines of code:} We measured the number of lines of code of each implementation, \todo {} \st{which arguably can be used to evaluate the complexity level of each implementation.}
+	\paragraph{Number of lines of code:} We measured the number of lines of code of each implementation.
 \vspace{-0.1cm}
@@
 
-\todo{mention baseline here}
+%\todo{mention baseline here}
 The \legSlangPlug, \legSlang and \legRTL implementations of this string comparison primitive have been implemented in the Cog VM.
@@
 The two compared strings are equal as this situation corresponds to the worst scenario for most implementations.
-\todo{explain how the relative speed up is computed} %There is one less branch in the two plugin implementations in this case.
+%\todo{explain how the relative speed up is computed} %There is one less branch in the two plugin implementations in this case.
 
@@
 Figure~\ref{fig:bench} and Table \ref{tab:detail_bench} show the results in terms of relative speed-up towards the baseline implementation, respectively for \GCCLinux and \LLVMMac.
-The relative speed-up is obtained by dividing the baseline results by the execution times. Both the figure and the table shows bootstrapped standard errors. \todo{add ref for bootstrap ?}
+%The relative speed-up is obtained by dividing the baseline results by the execution times. Both the figure and the table shows bootstrapped standard errors. \todo{add ref for bootstrap ?}
 
@@
 \section{Discussion and related work}
-\todo{primitives in other system. (generalize more after Rpython. }
+%\todo{primitives in other system. (generalize more after Rpython. }
 
@@
 They showed that with the RPython toolchain framework, the speed difference between all three implementations is not that big, hence they could choose to implement some optional primitives in the client language over C or RPython without performance drop to reduce the number of primitives to maintain.
-\todo{emphasize the difference more}
+%\todo{emphasize the difference more}
 
@@
 \acks
-\todo{add acknowledgments}
-Eliot, Levente
-Acknowledgments, if needed.
+We thank Levente Uzonyi for providing the benchmark template used in this article and Eliot Miranda for his advice on the implementation.
 
